---
title: "Benchmarking DMRsegal Against Other Methods"
author: "DMRsegal Package"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Benchmarking DMRsegal Against Other Methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    warning = TRUE,
    message = TRUE,
    fig.width = 8,
    fig.height = 5,
    dpi = 150
)
```

## Introduction

This vignette demonstrates how DMRsegal compares to other popular DMR detection methods for WGBS, 
specifically **dmrseq** and **DSS**. DMRsegal uses a seed-guided assembly approach 
starting from differentially methylated positions (seeds), expanding regions based on 
correlation and statistical significance. We compare the methods in terms of:

- Number of DMRs detected
- DMR size distribution  
- Overlap between methods
- Selectivity and sensitivity metrics
- Computational efficiency

## Setup

First, we'll set up the necessary directories and load required libraries:

```{r load_libraries}
library(progressr)
try(progressr::handlers(global = TRUE), silent = TRUE)
benchmark_output_dir <- "benchmark_data_wgbs"
dir.create(benchmark_output_dir, showWarnings = FALSE)
```

## Loading and Preparing Data

We'll create a synthetic WGBS dataset for benchmarking purposes:

```{r load_data}
if (!requireNamespace("MethylSeqData", quietly = TRUE)) {
    if (!requireNamespace("BiocManager", quietly = TRUE)) {
        install.packages("BiocManager")
    }
    BiocManager::install("MethylSeqData", update = FALSE, ask = FALSE)
}
library(MethylSeqData)
brain <- RizzardiHickeyBrain()
brain
```

## Finding Differentially Methylated Positions (DMPs)

Before running DMR detection, we need to identify individual DMPs.

```{r get_dmps}
# Find DMPs using minfi's dmpFinder
dmps <- suppressWarnings(dmpFinder(mvalues,
    pheno = pheno$casecontrol,
    type = "categorical",
    shrinkVar = TRUE
))
dmps <- dmps[!is.na(dmps$pval), ]
# Add adjusted p-value to dmps results
dmps$pval_adj <- p.adjust(dmps$pval, method = "BH")

# Filter significant DMPs
sig_dmps <- dmps[dmps$pval_adj < 0.05, ]
cat(sprintf("Found %d significant DMPs (FDR < 0.05)\n", nrow(sig_dmps)))
```

## Running DMRsegal

DMRsegal uses a seed-based approach, starting from significant DMPs and extending them into coherent regions:

```{r DMRsegal}
DMRsegal_file <- file.path(benchmark_output_dir, "dmrs.DMRsegal.rds") # nolint
if (!file.exists(DMRsegal_file)) {
    # Write DMPs to temp file
    options("DMRsegal.verbose" = 1)
    options("DMRsegal.njobs" = 8)
    progressr::handlers("cli")
    dmrs_segal <- progressr::with_progress(DMRsegal::findDMRsFromSeeds(
        beta = beta_handler,
        seeds = sig_dmps,
        pheno = pheno,
        sample_group_col = "Sample_Group",
        casecontrol_col = "casecontrol",
        min_seeds = 2,
        min_cpgs = 3,
        max_lookup_dist = 1000,
        max_pval = 0.05,
        pval_mode = "parametric",
        group_concordance_strategy = "relaxed"
    ))
    saveRDS(dmrs_segal, DMRsegal_file)
    rm(beta_handler)
}
```

## Running DMRcate

DMRcate is a popular method that uses Gaussian kernel smoothing to identify DMRs:

```{r dmrcate}
dmrcate_file <- file.path(benchmark_output_dir, "dmrs.dmrcate.rds")
if (!file.exists(dmrcate_file)) {
    if (!requireNamespace("DMRcate", quietly = TRUE)) {
        if (!requireNamespace("BiocManager", quietly = TRUE)) {
            install.packages("BiocManager")
        }
        BiocManager::install("DMRcate", update = FALSE, ask = FALSE)
    }
    library(DMRcate)
    design <- model.matrix(~Sample_Group, data = pheno)
    colnames(design) <- c("(Intercept)", "cancer_vs_normal")

    myannotation <- cpg.annotate("array", mvalues,
        what = "M", arraytype = "450K",
        analysis.type = "differential", design = design,
        coef = 2, fdr = 0.05
    )

    # Run DMRcate with the same p-value cutoff we used for DMPs
    dmrcate_results <- dmrcate(myannotation, C = 2, lambda = 1000)
    dmrs_dmrcate <- extractRanges(dmrcate_results, genome = "hg19")
    saveRDS(dmrs_dmrcate, dmrcate_file)
    detach("package:DMRcate", unload = TRUE)
}
```

## Running bumphunter

bumphunter uses permutation-based testing to identify regions with consistent methylation differences:

```{r bumphunter}
bumphunter_file <- file.path(benchmark_output_dir, "dmrs.bumphunter.rds")
if (!file.exists(bumphunter_file)) {
    if (!requireNamespace("bumphunter", quietly = TRUE)) {
        if (!requireNamespace("BiocManager", quietly = TRUE)) {
            install.packages("BiocManager")
        }
        BiocManager::install("bumphunter", update = FALSE, ask = FALSE)
    }
    library(bumphunter)
    # bumphunter uses the same mset and design matrix as other methods
    # Run bumphunter with the same design matrix
    doParallel::registerDoParallel(cores = 8)
    design <- model.matrix(~Sample_Group, data = pheno)

    bumphunter_results <- bumphunter(
        beta_mat,
        chr = locs$chr,
        pos = locs$start,
        design = design,
        pickCutoff = TRUE,
        B = 200, # Number of permutations
        type = "Beta"
    )
    dmrs_bumphunter <- makeGRangesFromDataFrame(bumphunter_results$table)
    saveRDS(dmrs_bumphunter, bumphunter_file)
}
```

## Comparing Results

Now let's load all results and compare them:

```{r compare_results}
# Function to get DMR stats
get_dmr_stats <- function(dmrs, method) {
    data.frame(
        Method = method,
        Number_of_DMRs = length(dmrs),
        Mean_DMR_Width = mean(IRanges::width(dmrs)),
        Median_DMR_Width = median(IRanges::width(dmrs)),
        Total_Coverage_bp = sum(IRanges::width(dmrs)),
        Min_Width = min(IRanges::width(dmrs)),
        Max_Width = max(IRanges::width(dmrs))
    )
}
dmrs_segal <- readRDS(DMRsegal_file)
dmrs_dmrcate <- readRDS(dmrcate_file)
dmrs_bumphunter <- readRDS(bumphunter_file)

# Collect stats
stats_list <- list(
    get_dmr_stats(dmrs_segal, "DMRsegal"),
    get_dmr_stats(dmrs_dmrcate, "DMRcate"),
    get_dmr_stats(dmrs_bumphunter, "bumphunter")
)

# Combine stats
results_comparison <- do.call(rbind, stats_list)
knitr::kable(results_comparison, digits = 2, caption = "Basic DMR Statistics")
```

## Overlap Analysis

Let's examine how the DMRs from different methods overlap:

```{r overlap}
# Convert all results to GRanges
gr_segal <- dmrs_segal
gr_dmrcate <- dmrs_dmrcate
gr_bumphunter <- dmrs_bumphunter

intersection <- function(gr1, gr2) {
    sum(IRanges::width(GenomicRanges::intersect(gr1, gr2)))
}

# IoU metric
iou <- function(gr1, gr2) {
    intersection <- sum(IRanges::width(GenomicRanges::intersect(gr1, gr2)))
    union <- sum(IRanges::width(GenomicRanges::union(gr1, gr2)))
    if (union == 0) {
        return(0)
    }
    intersection / union
}

# Jaccard index for regions (count-based)
jaccard_index <- function(gr1, gr2) {
    overlap_count <- length(IRanges::subsetByOverlaps(gr1, gr2))
    total_unique <- length(unique(c(gr1, gr2)))
    if (total_unique == 0) {
        return(0)
    }
    overlap_count / total_unique
}

# Calculate overlap metrics
# How many DMRsegal DMRs overlap with each method
overlap_segal_dmrcate <- length(subsetByOverlaps(gr_segal, gr_dmrcate))
overlap_segal_bumphunter <- length(subsetByOverlaps(gr_segal, gr_bumphunter))
overlap_dmrcate_bumphunter <- length(subsetByOverlaps(gr_dmrcate, gr_bumphunter))

# Reciprocal: how many DMRcate/Bumphunter DMRs overlap with DMRsegal
overlap_dmrcate_segal <- length(subsetByOverlaps(gr_dmrcate, gr_segal))
overlap_bumphunter_segal <- length(subsetByOverlaps(gr_bumphunter, gr_segal))
overlap_bumphunter_dmrcate <- length(subsetByOverlaps(gr_bumphunter, gr_dmrcate))

# Calculate IoU (Intersection over Union) for genomic coverage
iou_segal_dmrcate <- iou(gr_segal, gr_dmrcate)
iou_segal_bumphunter <- iou(gr_segal, gr_bumphunter)
iou_dmrcate_bumphunter <- iou(gr_dmrcate, gr_bumphunter)

# Calculate Jaccard indices
jaccard_segal_dmrcate <- jaccard_index(gr_segal, gr_dmrcate)
jaccard_segal_bumphunter <- jaccard_index(gr_segal, gr_bumphunter)
jaccard_dmrcate_bumphunter <- jaccard_index(gr_dmrcate, gr_bumphunter)

# Calculate percentage overlaps
pct_segal_in_dmrcate <- (overlap_segal_dmrcate / length(gr_segal)) * 100
pct_segal_in_bumphunter <- (overlap_segal_bumphunter / length(gr_segal)) * 100
pct_dmrcate_in_segal <- (overlap_dmrcate_segal / length(gr_dmrcate)) * 100
pct_bumphunter_in_segal <- (overlap_bumphunter_segal / length(gr_bumphunter)) * 100

# Create detailed overlap comparison
overlap_comparison <- data.frame(
    Comparison = c(
        "DMRsegal → DMRcate",
        "DMRsegal → Bumphunter",
        "DMRcate → DMRsegal",
        "Bumphunter → DMRsegal",
        "DMRcate → Bumphunter",
        "Bumphunter → DMRcate"
    ),
    Overlapping_DMRs = c(
        overlap_segal_dmrcate,
        overlap_segal_bumphunter,
        overlap_dmrcate_segal,
        overlap_bumphunter_segal,
        overlap_dmrcate_bumphunter,
        overlap_bumphunter_dmrcate
    ),
    Percentage_Overlap = c(
        pct_segal_in_dmrcate,
        pct_segal_in_bumphunter,
        pct_dmrcate_in_segal,
        pct_bumphunter_in_segal,
        (overlap_dmrcate_bumphunter / length(gr_dmrcate)) * 100,
        (overlap_bumphunter_dmrcate / length(gr_bumphunter)) * 100
    )
)

knitr::kable(overlap_comparison, digits = 2, caption = "Detailed Overlap Analysis")
```

## Similarity Metrics

We can quantify the similarity between methods using IoU (Intersection over Union) and Jaccard index:

```{r similarity_metrics}
# Create IoU and Jaccard comparison
similarity_metrics <- data.frame(
    Method_Pair = c(
        "DMRsegal vs DMRcate",
        "DMRsegal vs Bumphunter",
        "DMRcate vs Bumphunter"
    ),
    IoU_Genomic_Coverage = c(
        iou_segal_dmrcate,
        iou_segal_bumphunter,
        iou_dmrcate_bumphunter
    ),
    Jaccard_Index = c(
        jaccard_segal_dmrcate,
        jaccard_segal_bumphunter,
        jaccard_dmrcate_bumphunter
    )
)

knitr::kable(similarity_metrics, digits = 4, caption = "Similarity Metrics")
```

## Specificity Analysis

Let's analyze how specific DMRsegal is compared to other methods:

```{r specificity}
# Analyze specificity of DMRsegal
specificity_analysis <- data.frame(
    Metric = c(
        "DMRs found by DMRsegal",
        "DMRs found by DMRcate",
        "DMRs found by Bumphunter",
        "DMRsegal unique (not in DMRcate or Bumphunter)",
        "DMRsegal in both DMRcate and Bumphunter",
        "DMRsegal in DMRcate only",
        "DMRsegal in Bumphunter only"
    ),
    Count = c(
        length(gr_segal),
        length(gr_dmrcate),
        length(gr_bumphunter),
        length(gr_segal) - length(subsetByOverlaps(gr_segal, c(gr_dmrcate, gr_bumphunter))),
        length(subsetByOverlaps(subsetByOverlaps(gr_segal, gr_dmrcate), gr_bumphunter)),
        length(subsetByOverlaps(gr_segal, gr_dmrcate)) - length(subsetByOverlaps(subsetByOverlaps(gr_segal, gr_dmrcate), gr_bumphunter)),
        length(subsetByOverlaps(gr_segal, gr_bumphunter)) - length(subsetByOverlaps(subsetByOverlaps(gr_segal, gr_bumphunter), gr_dmrcate))
    )
)

knitr::kable(specificity_analysis, caption = "DMRsegal Specificity Analysis")
```

## Quality and Selectivity Metrics

Finally, let's examine quality metrics that highlight DMRsegal's selective approach:

```{r quality_metrics}
# Additional quality metrics for DMRsegal's selective approach

# Calculate the "capture rate" - what fraction of other methods' DMRs does DMRsegal capture
capture_metrics <- data.frame(
    Metric = c(
        "DMRsegal captures % of DMRcate genomic space",
        "DMRsegal captures % of Bumphunter genomic space",
        "Average DMR width - DMRsegal",
        "Average DMR width - DMRcate",
        "Average DMR width - Bumphunter",
        "DMRsegal selectivity ratio vs DMRcate",
        "DMRsegal selectivity ratio vs Bumphunter"
    ),
    Value = c(
        (sum(width(GenomicRanges::intersect(gr_segal, gr_dmrcate))) / sum(width(gr_dmrcate))) * 100,
        (sum(width(GenomicRanges::intersect(gr_segal, gr_bumphunter))) / sum(width(gr_bumphunter))) * 100,
        mean(width(gr_segal)),
        mean(width(gr_dmrcate)),
        mean(width(gr_bumphunter)),
        length(gr_dmrcate) / length(gr_segal),
        length(gr_bumphunter) / length(gr_segal)
    )
)

knitr::kable(capture_metrics, digits = 2, caption = "Quality and Selectivity Metrics")
```

## Visualizations

Let's create some visualizations to better understand the differences between methods:

```{r visualizations, message=FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) == FALSE) {
    install.packages("ggplot2", quiet = TRUE, update = FALSE)
}
library(ggplot2)

# Create visualization comparing method characteristics
method_chars <- data.frame(
    Method = rep(c("DMRsegal", "DMRcate", "Bumphunter"), each = 2),
    Characteristic = rep(c("Number of DMRs", "Total Coverage (kb)"), 3),
    Value = c(
        length(gr_segal), sum(width(gr_segal)) / 1000,
        length(gr_dmrcate), sum(width(gr_dmrcate)) / 1000,
        length(gr_bumphunter), sum(width(gr_bumphunter)) / 1000
    )
)

p1 <- ggplot(method_chars, aes(x = Method, y = Value, fill = Method)) +
    geom_bar(stat = "identity") +
    facet_wrap(~Characteristic, scales = "free_y") +
    theme_minimal() +
    labs(
        title = "DMR Detection Methods Comparison",
        y = "Value"
    ) +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none"
    )

print(p1)
```

```{r dmr_width_distribution}
# Create box plot comparing DMR widths
width_data <- data.frame(
    Method = c(
        rep("DMRsegal", length(gr_segal)),
        rep("DMRcate", length(gr_dmrcate)),
        rep("Bumphunter", length(gr_bumphunter))
    ),
    Width = c(width(gr_segal), width(gr_dmrcate), width(gr_bumphunter))
)

p2 <- ggplot(width_data, aes(x = Method, y = Width, fill = Method)) +
    geom_boxplot() +
    scale_y_log10() +
    theme_minimal() +
    labs(
        title = "Distribution of DMR Widths",
        y = "Width (bp, log scale)"
    ) +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none"
    )

print(p2)
```

## Summary

```{r summary, echo=FALSE, results='asis'}
cat(sprintf(
    "- DMRsegal found **%d DMRs** compared to **%d (DMRcate)** and **%d (Bumphunter)**\n",
    length(gr_segal), length(gr_dmrcate), length(gr_bumphunter)
))
cat(sprintf("- **%.1f%%** of DMRsegal DMRs overlap with DMRcate\n", pct_segal_in_dmrcate))
cat(sprintf("- **%.1f%%** of DMRsegal DMRs overlap with Bumphunter\n", pct_segal_in_bumphunter))
cat(sprintf("- DMRsegal is **%.1fx more selective** than DMRcate\n", length(gr_dmrcate) / length(gr_segal)))
cat(sprintf("- DMRsegal is **%.1fx more selective** than Bumphunter\n", length(gr_bumphunter) / length(gr_segal)))
```

## Session Info

```{r session_info}
sessionInfo()
```