---
title: "DMRsegal vs comb-p, probeLasso, DMRcate, and Bumphunter"
author: "DMRsegal Package"
date: "`r Sys.Date()`"
output: 
    BiocStyle::html_document:
        code_folding: hide
vignette: >
  %\VignetteIndexEntry{DMRsegal vs comb-p, probeLasso, DMRcate, and Bumphunter}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
    results = "asis",
    echo = TRUE,
    warning = FALSE,
    message = FALSE,
    fig.width = 12,
    fig.height = 8,
    dpi = 300
)
```

## Introduction

This vignette provides an extended benchmark comparing DMRsegal against multiple popular DMR detection methods:

- **DMRsegal**: Correlation-based seed-guided assembly approach
- **comb-p**: Combines adjacent p-values using Stouffer-Liptak-Kechris correction
- **probeLasso**: Lasso-based feature selection for DMR detection
- **DMRcate**: Gaussian kernel smoothing approach
- **Bumphunter**: Permutation-based regional testing

We evaluate these methods on:

- Number and size distribution of DMRs
- Overlap and concordance between methods
- Biological relevance of detected regions

The dataset used is DMRsegaldata, which contains methylation profiles for 5 cancer vs 5 normal samples.

## Setup and Dependencies

```{r, message=FALSE}
devtools::load_all()
library(minfi)
library(data.table)
library(ggplot2)
library(dplyr)
library(GenomicRanges)
library(progressr)

progressr::handlers("cli")
output_dir <- "dmrsegaldata"
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)
```

### Install External Dependencies

```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

if (!requireNamespace("DMRcate", quietly = TRUE)) {
    BiocManager::install("DMRcate", update = FALSE, ask = FALSE)
}

if (!requireNamespace("bumphunter", quietly = TRUE)) {
    BiocManager::install("bumphunter", update = FALSE, ask = FALSE)
}

if (!requireNamespace("doParallel", quietly = TRUE)) {
    install.packages("doParallel")
}

if (!requireNamespace("reticulate", quietly = TRUE)) {
    install.packages("reticulate")
}

if (!requireNamespace("ChAMP", quietly = TRUE)) {
    BiocManager::install("ChAMP", update = FALSE, ask = FALSE)
}
```

### Setup comb-p (conda required)

**Note:** comb-p requires both the tool itself and bedtools. Install via conda:

```bash
# Install comb-p and bedtools
conda install -yc bioconda combined-pvalues bedtools

# Or create a dedicated environment
conda create -n combp python=3.9
conda activate combp
conda install -yc bioconda combined-pvalues bedtools
```

```{r, eval=FALSE}
# Check if comb-p and bedtools are available
if (Sys.which("comb-p") == "") {
    warning("comb-p not found. Install with: conda install -yc bioconda combined-pvalues")
}

if (Sys.which("bedtools") == "") {
    warning("bedtools not found. Install with: conda install -yc bioconda bedtools")
}
```

## Load Example Data

```{r}
beta <- loadExampleInputData("beta")
pheno <- loadExampleInputData("pheno")
array_type <- loadExampleInputData("array_type")
genome <- "hg19"

beta_handler <- DMRsegal::getBetaHandler(beta, array = array_type, genome = genome)
beta_mat <- as.matrix(beta_handler$getBeta())
locs <- beta_handler$getBetaLocs()
mvalues <- log2(beta_mat / (1 - beta_mat + 1e-6) + 1e-6)
pheno$casecontrol <- pheno$Sample_Group == "cancer"
```

## Find Differentially Methylated Positions

```{r}
dmps <- suppressWarnings(dmpFinder(mvalues,
    pheno = pheno$casecontrol,
    type = "categorical",
    shrinkVar = TRUE
))
dmps <- dmps[!is.na(dmps$pval), ]
dmps$pval_adj <- p.adjust(dmps$pval, method = "BH")

case_idx <- which(pheno$casecontrol)
ctrl_idx <- which(!pheno$casecontrol)
dmps$delta_beta <- rowMeans(beta_mat[rownames(dmps), case_idx, drop = FALSE]) -
    rowMeans(beta_mat[rownames(dmps), ctrl_idx, drop = FALSE])

sig_dmps <- dmps[dmps$pval_adj < 0.01 & abs(dmps$delta_beta) >= 0.05, ]
cat(sprintf("Found %d significant DMPs (FDR < 0.01, |delta beta| >= 0.05)\n", nrow(sig_dmps)))

dmps_gr <- makeGRangesFromDataFrame(
    data.frame(
        chr = locs$chr,
        start = locs$start,
        end = locs$end,
        pval = dmps$pval,
        qval = dmps$pval_adj,
        stat = dmps$qval
    ),
    keep.extra.columns = TRUE
)
names(dmps_gr) <- rownames(dmps)
```

## Method 1: DMRsegal

```{r}
DMRsegal_file <- file.path(output_dir, "dmrs.DMRsegal.rds")
if (!file.exists(DMRsegal_file)) {
    options("DMRsegal.verbose" = 1)
    options("DMRsegal.njobs" = 8)

    dmrs_segal <- progressr::with_progress(DMRsegal::findDMRsFromSeeds(
        beta = beta_handler,
        seeds = sig_dmps,
        pheno = pheno,
        sample_group_col = "Sample_Group",
        casecontrol_col = "casecontrol",
        covariates = c("Age", "Gender"),
        min_seeds = 2,
        min_cpgs = 3,
        max_lookup_dist = 1000,
        max_pval = 0.05,
        pval_mode = "parametric",
        entanglement = "weak",
        rank_dmrs = TRUE
    ))
    saveRDS(dmrs_segal, DMRsegal_file)
    rm(beta_handler)
} else {
    dmrs_segal <- readRDS(DMRsegal_file)
}

cat(sprintf("DMRsegal: Found %d DMRs\n", length(dmrs_segal)))
```

## Method 2: comb-p

```{r}
combp_file <- file.path(output_dir, "dmrs.combp.rds")

run_combp <- function(dmps_gr, output_dir, seed = 500, dist = 500) {
    if (Sys.which("bedtools") == "") {
        warning("bedtools not found in PATH. comb-p requires bedtools to be installed.")
        warning("Install with: conda install -yc bioconda bedtools")
        return(GRanges())
    }

    if (Sys.which("comb-p") == "") {
        warning("comb-p not found in PATH.")
        warning("Install with: conda install -yc bioconda combined-pvalues")
        return(GRanges())
    }

    bed_file <- file.path(output_dir, "dmps_for_combp.bed")
    bed_df <- data.frame(
        chrom = as.character(seqnames(dmps_gr)),
        start = start(dmps_gr) - 1,
        end = end(dmps_gr),
        pval = mcols(dmps_gr)$pval
    )
    bed_df <- bed_df[order(bed_df$chrom, bed_df$start), ]

    write.table(bed_df, bed_file,
        quote = FALSE, sep = "\t",
        row.names = FALSE, col.names = TRUE
    )

    output_prefix <- file.path(output_dir, "combp_output")

    cmd <- sprintf(
        "comb-p pipeline -c 4 --seed %d --dist %d -p %s --region-filter-p 0.01 --region-filter-n 3 %s",
        seed, dist, output_prefix, bed_file
    )

    tryCatch(
        {
            exit_code <- system(cmd, ignore.stdout = FALSE, ignore.stderr = FALSE)

            if (exit_code != 0) {
                warning(sprintf("comb-p exited with code %d", exit_code))
                return(GRanges())
            }

            regions_file <- paste0(output_prefix, ".regions-p.bed.gz")
            if (file.exists(regions_file)) {
                regions <- fread(cmd = sprintf("zcat %s", regions_file))
                colnames(regions) <- c("chr", "start", "end", "min_p", "n_probes", "z_p", "z_sidak_p")

                regions <- regions[regions$z_sidak_p < 0.05 & regions$n_probes >= 3, ]

                dmrs_combp <- makeGRangesFromDataFrame(regions,
                    seqnames.field = "chr",
                    start.field = "start",
                    end.field = "end",
                    keep.extra.columns = TRUE
                )
                return(dmrs_combp)
            } else {
                warning("comb-p did not produce output file")
                return(GRanges())
            }
        },
        error = function(e) {
            warning(sprintf("comb-p failed: %s", e$message))
            return(GRanges())
        }
    )
}

if (!file.exists(combp_file)) {
    dmrs_combp <- run_combp(dmps_gr, output_dir)
    saveRDS(dmrs_combp, combp_file)
} else {
    dmrs_combp <- readRDS(combp_file)
}

cat(sprintf("comb-p: Found %d DMRs\n", length(dmrs_combp)))
```

## Method 3: probeLasso
```{r}
probeLasso_file <- file.path(output_dir, "dmrs.probeLasso.rds")

run_probeLasso_champ <- function(beta_mat, pheno, array_type = "450K") {
    if (!requireNamespace("ChAMP", quietly = TRUE)) {
        warning("ChAMP not installed. probeLasso requires ChAMP.")
        warning("Install with: BiocManager::install('ChAMP')")
        return(GRanges())
    }

    library(ChAMP)

    tryCatch(
        {
            result <- champ.DMR(
                beta = beta_mat,
                pheno = pheno$Sample_Group,
                arraytype = array_type,
                method = "ProbeLasso",
                minProbes = 3,
                adjPvalDmr = 0.05,
                adjPvalProbe = 0.1,
                meanLassoRadius = 500,
                minDmrSep = 500,
                minDmrSize = 50,
                PDFplot = FALSE,
                Rplot = FALSE,
                resultsDir = "/tmp/champ_probeLasso"
            )$ProbeLassoDMR

            return(makeGRangesFromDataFrame(result, keep.extra.columns = TRUE))
        },
        error = function(e) {
            warning(sprintf("probeLasso (ChAMP) failed: %s", e$message))
            return(GRanges())
        }
    )
}

if (!file.exists(probeLasso_file)) {
    dmrs_probeLasso <- run_probeLasso_champ(beta_mat, pheno, array_type = "450K")
    saveRDS(dmrs_probeLasso, probeLasso_file)
} else {
    dmrs_probeLasso <- readRDS(probeLasso_file)
}

cat(sprintf("probeLasso: Found %d DMRs\n", length(dmrs_probeLasso)))
```

## Method 4: DMRcate

```{r}
dmrcate_file <- file.path(output_dir, "dmrs.dmrcate.rds")
if (!file.exists(dmrcate_file)) {
    library(DMRcate)

    design <- model.matrix(~Sample_Group, data = pheno)
    colnames(design) <- c("(Intercept)", "cancer_vs_normal")

    myannotation <- cpg.annotate("array", mvalues,
        what = "M", arraytype = "450K",
        analysis.type = "differential", design = design,
        coef = 2, fdr = 0.05
    )

    dmrcate_results <- dmrcate(myannotation, C = 2, lambda = 1000, min.cpgs = 3, betacutoff = 0.05)
    dmrs_dmrcate <- extractRanges(dmrcate_results, genome = "hg19")
    saveRDS(dmrs_dmrcate, dmrcate_file)
} else {
    dmrs_dmrcate <- readRDS(dmrcate_file)
}

cat(sprintf("DMRcate: Found %d DMRs\n", length(dmrs_dmrcate)))
```

## Method 5: Bumphunter

```{r}
bumphunter_file <- file.path(output_dir, "dmrs.bumphunter.rds")
if (!file.exists(bumphunter_file)) {
    library(bumphunter)
    library(doParallel)

    # Safer defaults for local/WSL rendering; override via options().
    bh_cores <- as.integer(getOption("DMRsegal.bumphunter_cores", 1L))
    bh_B <- as.integer(getOption("DMRsegal.bumphunter_B", 100L))
    bh_max_dmrs <- as.integer(getOption("DMRsegal.bumphunter_max_dmrs_for_cutoff", 12000L))
    bh_cores <- ifelse(is.na(bh_cores) || bh_cores < 1L, 1L, bh_cores)
    bh_B <- ifelse(is.na(bh_B) || bh_B < 0L, 100L, bh_B)
    bh_max_dmrs <- ifelse(is.na(bh_max_dmrs) || bh_max_dmrs < 1L, 12000L, bh_max_dmrs)

    doParallel::registerDoParallel(cores = bh_cores)
    on.exit({
        try(doParallel::stopImplicitCluster(), silent = TRUE)
        gc()
    }, add = TRUE)

    design <- model.matrix(~Sample_Group, data = pheno)
    max_gap <- 1000
    dmrs_cutoff <- 0.1
    dmrs_cnt <- Inf
    iter <- 0L
    max_iter <- 10L
    while (dmrs_cnt > bh_max_dmrs && iter < max_iter) {
        iter <- iter + 1L
        dmrs <- bumphunter(beta_mat,
            design = design,
            cutoff = dmrs_cutoff,
            chr = as.character(locs$chr),
            pos = locs$start,
            maxGap = max_gap,
            nullMethod = "bootstrap",
            B = 0,
            type = "Beta",
            coef = 2
        )
        dmrs_cnt <- nrow(dmrs$table)
        dmrs_cutoff <- dmrs_cutoff * 1.2
        rm(dmrs)
        gc()
    }
    final_cutoff <- dmrs_cutoff / 1.2
    cat(sprintf(
        "Bumphunter: cutoff=%.4f, candidate_bumps=%d, cores=%d, B=%d\n",
        final_cutoff, dmrs_cnt, bh_cores, bh_B
    ))
    bumphunter_results <- bumphunter(beta_mat,
        design = design,
        chr = as.character(locs$chr),
        pos = locs$start,
        cutoff = final_cutoff,
        maxGap = max_gap,
        nullMethod = "bootstrap",
        B = bh_B,
        type = "Beta",
        coef = 2
    )

    dmr_table <- bumphunter_results$table
    dmr_table <- dmr_table[dmr_table$L >= 3 & dmr_table$fwer <= 0.05, ]

    dmrs_bumphunter <- makeGRangesFromDataFrame(dmr_table,
        seqnames.field = "chr",
        start.field = "start",
        end.field = "end",
        keep.extra.columns = TRUE
    )

    saveRDS(dmrs_bumphunter, bumphunter_file)
} else {
    dmrs_bumphunter <- readRDS(bumphunter_file)
}

cat(sprintf("Bumphunter: Found %d DMRs\n", length(dmrs_bumphunter)))
```

## Comparative Analysis

### Basic Statistics

```{r}
get_dmr_stats <- function(dmrs, method) {
    if (length(dmrs) == 0) {
        return(data.frame(
            Method = method,
            Number_of_DMRs = 0,
            Mean_DMR_Width = NA,
            Median_DMR_Width = NA,
            Total_Coverage_bp = 0,
            Min_Width = NA,
            Max_Width = NA
        ))
    }

    data.frame(
        Method = method,
        Number_of_DMRs = length(dmrs),
        Mean_DMR_Width = mean(width(dmrs)),
        Median_DMR_Width = median(width(dmrs)),
        Total_Coverage_bp = sum(width(dmrs)),
        Min_Width = min(width(dmrs)),
        Max_Width = max(width(dmrs))
    )
}

stats_list <- list(
    get_dmr_stats(dmrs_segal, "DMRsegal"),
    get_dmr_stats(dmrs_combp, "comb-p"),
    get_dmr_stats(dmrs_probeLasso, "probeLasso"),
    get_dmr_stats(dmrs_dmrcate, "DMRcate"),
    get_dmr_stats(dmrs_bumphunter, "Bumphunter")
)

results_comparison <- do.call(rbind, stats_list)
knitr::kable(results_comparison, digits = 2, caption = "Basic DMR Statistics Across Methods")
```

### Pairwise Overlap Analysis

```{r, results='asis'}
methods_list <- list(
    DMRsegal = dmrs_segal,
    `comb-p` = dmrs_combp,
    probeLasso = dmrs_probeLasso,
    DMRcate = dmrs_dmrcate,
    Bumphunter = dmrs_bumphunter
)

methods_list <- methods_list[sapply(methods_list, length) > 0]

calculate_overlap <- function(gr1, gr2) {
    if (length(gr1) == 0 || length(gr2) == 0) {
        return(list(count = 0, percentage = 0))
    }
    overlap_count <- length(subsetByOverlaps(gr1, gr2))
    pct <- (overlap_count / length(gr1)) * 100
    list(count = overlap_count, percentage = pct)
}

overlap_matrix <- matrix(0, nrow = length(methods_list), ncol = length(methods_list))
pct_matrix <- matrix(0, nrow = length(methods_list), ncol = length(methods_list))
rownames(overlap_matrix) <- colnames(overlap_matrix) <- names(methods_list)
rownames(pct_matrix) <- colnames(pct_matrix) <- names(methods_list)

for (i in seq_along(methods_list)) {
    for (j in seq_along(methods_list)) {
        if (i != j) {
            overlap_result <- calculate_overlap(methods_list[[i]], methods_list[[j]])
            overlap_matrix[i, j] <- overlap_result$count
            pct_matrix[i, j] <- overlap_result$percentage
        }
    }
}

cat("\n#### Overlap Counts (rows overlap with columns)\n")
knitr::kable(overlap_matrix, digits = 0, caption = "Number of DMRs from Method A (rows) overlapping with Method B (columns)")

cat("\n#### Overlap Percentages\n")
knitr::kable(pct_matrix, digits = 2, caption = "Percentage of DMRs from Method A (rows) overlapping with Method B (columns)")
```

### Jaccard Index and IoU

```{r}
jaccard_index <- function(gr1, gr2) {
    if (length(gr1) == 0 || length(gr2) == 0) {
        return(0)
    }
    overlap_count <- length(subsetByOverlaps(gr1, gr2))
    total_unique <- length(unique(c(gr1, gr2)))
    if (total_unique == 0) {
        return(0)
    }
    overlap_count / total_unique
}

iou <- function(gr1, gr2) {
    if (length(gr1) == 0 || length(gr2) == 0) {
        return(0)
    }
    intersection <- sum(width(GenomicRanges::intersect(gr1, gr2)))
    union <- sum(width(GenomicRanges::union(gr1, gr2)))
    if (union == 0) {
        return(0)
    }
    intersection / union
}

jaccard_matrix <- matrix(0, nrow = length(methods_list), ncol = length(methods_list))
iou_matrix <- matrix(0, nrow = length(methods_list), ncol = length(methods_list))
rownames(jaccard_matrix) <- colnames(jaccard_matrix) <- names(methods_list)
rownames(iou_matrix) <- colnames(iou_matrix) <- names(methods_list)

for (i in seq_along(methods_list)) {
    for (j in seq_along(methods_list)) {
        if (i < j) {
            jaccard_matrix[i, j] <- jaccard_matrix[j, i] <- jaccard_index(methods_list[[i]], methods_list[[j]])
            iou_matrix[i, j] <- iou_matrix[j, i] <- iou(methods_list[[i]], methods_list[[j]])
        } else if (i == j) {
            jaccard_matrix[i, j] <- 1
            iou_matrix[i, j] <- 1
        }
    }
}

cat("\n#### Jaccard Index\n")
knitr::kable(jaccard_matrix, digits = 4, caption = "Jaccard Index (region count-based similarity)")

cat("\n#### Intersection over Union (IoU)\n")
knitr::kable(iou_matrix, digits = 4, caption = "IoU (genomic coverage-based similarity)")
```

## Visualizations

### DMR Count and Coverage Comparison

```{r, fig.width=10, fig.height=5}
viz_data <- results_comparison %>%
    dplyr::select(Method, Number_of_DMRs, Total_Coverage_bp) %>%
    tidyr::pivot_longer(
        cols = c(Number_of_DMRs, Total_Coverage_bp),
        names_to = "Metric", values_to = "Value"
    )

viz_data$Metric <- ifelse(viz_data$Metric == "Number_of_DMRs",
    "Number of DMRs",
    "Total Coverage (bp)"
)

p1 <- ggplot(viz_data, aes(x = Method, y = Value, fill = Method)) +
    geom_bar(stat = "identity") +
    facet_wrap(~Metric, scales = "free_y") +
    theme_minimal() +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none",
        panel.grid.major.x = element_blank()
    ) +
    labs(
        title = "DMR Detection: Count and Coverage Comparison",
        y = "Value", x = ""
    )

print(p1)
```

### DMR Width Distribution

```{r, fig.width=10, fig.height=6}
width_data <- do.call(rbind, lapply(names(methods_list), function(method) {
    if (length(methods_list[[method]]) > 0) {
        data.frame(
            Method = method,
            Width = width(methods_list[[method]])
        )
    }
}))

p2 <- ggplot(width_data, aes(x = Method, y = Width, fill = Method)) +
    geom_violin(alpha = 0.7) +
    geom_boxplot(width = 0.2, alpha = 0.5, outlier.size = 0.5) +
    scale_y_log10(labels = scales::comma) +
    theme_minimal() +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none"
    ) +
    labs(
        title = "Distribution of DMR Widths",
        y = "Width (bp, log scale)", x = ""
    )

print(p2)
```

### Overlap Heatmap

```{r, fig.width=8, fig.height=6}
library(reshape2)

overlap_df <- melt(pct_matrix)
colnames(overlap_df) <- c("Method1", "Method2", "Percentage")

p3 <- ggplot(overlap_df, aes(x = Method2, y = Method1, fill = Percentage)) +
    geom_tile() +
    geom_text(aes(label = sprintf("%.1f%%", Percentage)), size = 3) +
    scale_fill_gradient2(
        low = "white", mid = "lightblue", high = "darkblue",
        midpoint = 50, limits = c(0, 100)
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(
        title = "DMR Overlap Matrix",
        subtitle = "% of row method's DMRs overlapping with column method",
        x = "", y = "", fill = "Overlap %"
    )

print(p3)
```

### Similarity Matrix (IoU)

```{r, fig.width=8, fig.height=6}
iou_df <- melt(iou_matrix)
colnames(iou_df) <- c("Method1", "Method2", "IoU")

p4 <- ggplot(iou_df, aes(x = Method2, y = Method1, fill = IoU)) +
    geom_tile() +
    geom_text(aes(label = sprintf("%.3f", IoU)), size = 3) +
    scale_fill_gradient2(
        low = "white", mid = "yellow", high = "red",
        midpoint = 0.5, limits = c(0, 1)
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(
        title = "Method Similarity Matrix (IoU)",
        subtitle = "Intersection over Union of genomic coverage",
        x = "", y = "", fill = "IoU"
    )

print(p4)
```

## Biological Validation

### Effect Size Analysis (Delta Beta)

```{r, results='asis'}
calculate_mean_delta_beta <- function(dmrs, beta_mat, pheno, locs, group_col = "Sample_Group", max_dmrs = 5000) {
    if (length(dmrs) == 0) {
        return(NA)
    }

    if (length(dmrs) > max_dmrs) {
        set.seed(42)
        dmrs <- dmrs[sample(length(dmrs), max_dmrs)]
    }

    case_samples <- colnames(beta_mat)[pheno[[group_col]] == "cancer"]
    control_samples <- colnames(beta_mat)[pheno[[group_col]] == "normal"]

    cpg_gr <- makeGRangesFromDataFrame(locs)
    names(cpg_gr) <- rownames(beta_mat)

    hits <- findOverlaps(cpg_gr, dmrs)
    if (length(hits) == 0) {
        return(NA)
    }

    cpg_idx <- queryHits(hits)
    dmr_idx <- subjectHits(hits)

    case_means <- rowMeans(beta_mat[cpg_idx, case_samples, drop = FALSE], na.rm = TRUE)
    control_means <- rowMeans(beta_mat[cpg_idx, control_samples, drop = FALSE], na.rm = TRUE)
    delta_per_cpg <- abs(case_means - control_means)

    dmr_deltas <- tapply(delta_per_cpg, dmr_idx, mean, na.rm = TRUE)
    mean(dmr_deltas, na.rm = TRUE)
}

effect_size_list <- lapply(names(methods_list), function(method) {
    data.frame(
        Method = method,
        Mean_Delta_Beta = calculate_mean_delta_beta(methods_list[[method]], beta_mat, pheno, locs)
    )
})

effect_size_df <- do.call(rbind, effect_size_list)
knitr::kable(effect_size_df, digits = 4, caption = "Mean Effect Size (Delta Beta) per Method")

ggplot(effect_size_df, aes(x = reorder(Method, -Mean_Delta_Beta), y = Mean_Delta_Beta, fill = Method)) +
    geom_bar(stat = "identity") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
    labs(title = "Mean Effect Size by Method", x = "", y = "Mean |Delta Beta|")
```

### Gene Annotation Analysis

```{r}
if (requireNamespace("TxDb.Hsapiens.UCSC.hg19.knownGene", quietly = TRUE)) {
    library(TxDb.Hsapiens.UCSC.hg19.knownGene)
    txdb <- TxDb.Hsapiens.UCSC.hg19.knownGene

    promoters_gr <- promoters(txdb, upstream = 2000, downstream = 500)
    genes_gr <- genes(txdb)

    annotate_dmrs <- function(dmrs, method_name) {
        if (length(dmrs) == 0) {
            return(data.frame(
                Method = method_name,
                Total_DMRs = 0,
                In_Promoters = 0,
                In_Gene_Bodies = 0,
                Intergenic = 0,
                Pct_Promoters = 0,
                Pct_Gene_Bodies = 0,
                Pct_Intergenic = 0
            ))
        }

        in_promoters <- sum(overlapsAny(dmrs, promoters_gr))
        in_genes <- sum(overlapsAny(dmrs, genes_gr))
        intergenic <- length(dmrs) - sum(overlapsAny(dmrs, c(promoters_gr, genes_gr)))

        data.frame(
            Method = method_name,
            Total_DMRs = length(dmrs),
            In_Promoters = in_promoters,
            In_Gene_Bodies = in_genes,
            Intergenic = intergenic,
            Pct_Promoters = round(100 * in_promoters / length(dmrs), 1),
            Pct_Gene_Bodies = round(100 * in_genes / length(dmrs), 1),
            Pct_Intergenic = round(100 * intergenic / length(dmrs), 1)
        )
    }

    gene_annotation_list <- lapply(names(methods_list), function(method) {
        annotate_dmrs(methods_list[[method]], method)
    })

    gene_annotation_df <- do.call(rbind, gene_annotation_list)
    knitr::kable(gene_annotation_df, caption = "DMR Gene Context Annotation")

    gene_plot_data <- gene_annotation_df %>%
        dplyr::select(Method, Pct_Promoters, Pct_Gene_Bodies, Pct_Intergenic) %>%
        tidyr::pivot_longer(cols = -Method, names_to = "Context", values_to = "Percentage")

    gene_plot_data$Context <- gsub("Pct_", "", gene_plot_data$Context)
    gene_plot_data$Context <- gsub("_", " ", gene_plot_data$Context)

    ggplot(gene_plot_data, aes(x = Method, y = Percentage, fill = Context)) +
        geom_bar(stat = "identity", position = "dodge") +
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        labs(title = "DMR Distribution by Gene Context", x = "", y = "Percentage of DMRs")
}
```

### CpG Context Distribution

```{r}
if (requireNamespace("IlluminaHumanMethylation450kanno.ilmn12.hg19", quietly = TRUE)) {
    library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
    anno <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)

    cpg_context_gr <- makeGRangesFromDataFrame(
        data.frame(
            chr = anno$chr,
            start = anno$pos,
            end = anno$pos,
            context = anno$Relation_to_Island
        ),
        keep.extra.columns = TRUE
    )
    names(cpg_context_gr) <- rownames(anno)

    get_cpg_context_distribution <- function(dmrs, method_name) {
        if (length(dmrs) == 0) {
            return(data.frame(Method = method_name, Context = NA, Count = 0, Percentage = 0))
        }

        overlapping <- subsetByOverlaps(cpg_context_gr, dmrs)
        context_counts <- table(mcols(overlapping)$context)

        data.frame(
            Method = method_name,
            Context = names(context_counts),
            Count = as.vector(context_counts),
            Percentage = round(100 * as.vector(context_counts) / sum(context_counts), 1)
        )
    }

    cpg_context_list <- lapply(names(methods_list), function(method) {
        get_cpg_context_distribution(methods_list[[method]], method)
    })

    cpg_context_df <- do.call(rbind, cpg_context_list)
    cpg_context_df <- cpg_context_df[!is.na(cpg_context_df$Context), ]

    ggplot(cpg_context_df, aes(x = Method, y = Percentage, fill = Context)) +
        geom_bar(stat = "identity", position = "stack") +
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        scale_fill_brewer(palette = "Set2") +
        labs(
            title = "CpG Context Distribution (Island/Shore/Shelf/OpenSea)",
            x = "", y = "Percentage of CpGs in DMRs"
        )
}
```

## Consensus Analysis

### UpSet Plot of Method Intersections

```{r, fig.width=12, fig.height=7}
if (requireNamespace("UpSetR", quietly = TRUE) && length(methods_list) >= 2) {
    library(UpSetR)

    all_dmrs_reduced <- GenomicRanges::reduce(do.call(c, unname(methods_list)))

    membership_matrix <- sapply(methods_list, function(method_dmrs) {
        as.integer(overlapsAny(all_dmrs_reduced, method_dmrs))
    })

    membership_df <- as.data.frame(membership_matrix)
    colnames(membership_df) <- names(methods_list)

    upset(membership_df,
        sets = names(methods_list),
        order.by = "freq",
        main.bar.color = "steelblue",
        sets.bar.color = "darkgreen",
        text.scale = 1.3,
        point.size = 3,
        line.size = 1
    )
}
```

### Consensus DMR Analysis

```{r}
find_consensus <- function(methods_list, min_methods = 2) {
    all_dmrs <- do.call(c, unname(methods_list))
    all_dmrs <- GenomicRanges::reduce(all_dmrs)

    overlap_matrix <- sapply(methods_list, function(method_dmrs) {
        overlapsAny(all_dmrs, method_dmrs)
    })
    consensus_counts <- rowSums(overlap_matrix)

    consensus_dmrs <- all_dmrs[consensus_counts >= min_methods]
    mcols(consensus_dmrs)$n_methods <- consensus_counts[consensus_counts >= min_methods]

    return(consensus_dmrs)
}

consensus_2plus <- find_consensus(methods_list, min_methods = 2)
consensus_3plus <- find_consensus(methods_list, min_methods = 3)
consensus_4plus <- find_consensus(methods_list, min_methods = 4)
cat(sprintf("- Consensus DMRs (2+ methods): %d\n", length(consensus_2plus)))
cat(sprintf("- Consensus DMRs (3+ methods): %d\n", length(consensus_3plus)))
cat(sprintf("- Consensus DMRs (4+ methods): %d\n", length(consensus_4plus)))

capture_rate <- sapply(names(methods_list), function(method) {
    if (length(consensus_3plus) == 0) {
        return(0)
    }
    sum(overlapsAny(consensus_3plus, methods_list[[method]])) / length(consensus_3plus) * 100
})

capture_df <- data.frame(
    Method = names(capture_rate),
    Capture_Rate_Pct = capture_rate
)
capture_df <- capture_df[order(-capture_df$Capture_Rate_Pct), ]

knitr::kable(capture_df,
    digits = 1, row.names = FALSE,
    caption = "% of High-Confidence (3+ methods) DMRs Captured by Each Method"
)

ggplot(capture_df, aes(x = reorder(Method, Capture_Rate_Pct), y = Capture_Rate_Pct, fill = Method)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    theme_minimal() +
    theme(legend.position = "none") +
    geom_hline(yintercept = 100, linetype = "dashed", color = "red") +
    labs(
        title = "Capture Rate of High-Confidence DMRs",
        subtitle = "% of consensus (3+ methods) DMRs detected by each method",
        x = "", y = "Capture Rate (%)"
    )
```

### Method-Specific vs Consensus DMRs

```{r}
find_specific_dmrs <- function(target_method, methods_list) {
    target_dmrs <- methods_list[[target_method]]
    if (length(target_dmrs) == 0) {
        return(GRanges())
    }

    other_methods <- methods_list[names(methods_list) != target_method]
    all_others <- do.call(c, unname(other_methods))

    if (length(all_others) == 0) {
        return(target_dmrs)
    }

    specific <- target_dmrs[countOverlaps(target_dmrs, all_others) == 0]
    return(specific)
}

specificity_df <- do.call(rbind, lapply(names(methods_list), function(method) {
    specific <- find_specific_dmrs(method, methods_list)
    shared_2plus <- methods_list[[method]][overlapsAny(methods_list[[method]], consensus_2plus)]
    shared_3plus <- methods_list[[method]][overlapsAny(methods_list[[method]], consensus_3plus)]

    data.frame(
        Method = method,
        Total_DMRs = length(methods_list[[method]]),
        Method_Specific = length(specific),
        Shared_2plus = length(shared_2plus),
        Shared_3plus = length(shared_3plus),
        Pct_Specific = round(100 * length(specific) / max(1, length(methods_list[[method]])), 1),
        Pct_Shared_2plus = round(100 * length(shared_2plus) / max(1, length(methods_list[[method]])), 1),
        Pct_Shared_3plus = round(100 * length(shared_3plus) / max(1, length(methods_list[[method]])), 1)
    )
}))

knitr::kable(specificity_df, caption = "Method-Specific vs Shared DMRs")

specificity_plot_data <- specificity_df %>%
    dplyr::select(Method, Pct_Specific, Pct_Shared_2plus) %>%
    tidyr::pivot_longer(cols = -Method, names_to = "Type", values_to = "Percentage")

specificity_plot_data$Type <- ifelse(specificity_plot_data$Type == "Pct_Specific",
    "Method-Specific", "Shared (2+ methods)"
)

ggplot(specificity_plot_data, aes(x = Method, y = Percentage, fill = Type)) +
    geom_bar(stat = "identity", position = "stack") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_fill_manual(values = c("Method-Specific" = "#E41A1C", "Shared (2+ methods)" = "#377EB8")) +
    labs(title = "Method-Specific vs Shared DMRs", x = "", y = "Percentage of DMRs", fill = "")
```

## Example DMR Locus Visualization

```{r, fig.width=12, fig.height=8}
if (length(consensus_3plus) > 0) {
    top_consensus <- consensus_3plus[order(-mcols(consensus_3plus)$n_methods)][1]

    extended_region <- GRanges(
        seqnames = seqnames(top_consensus),
        ranges = IRanges(
            start = max(1, start(top_consensus) - 2000),
            end = end(top_consensus) + 2000
        )
    )

    cat(sprintf(
        "\nExample locus: %s:%d-%d (detected by %d methods)\n",
        as.character(seqnames(top_consensus)),
        start(top_consensus), end(top_consensus),
        mcols(top_consensus)$n_methods
    ))

    method_colors <- c(
        "DMRsegal" = "#E41A1C", "comb-p" = "#377EB8",
        "probeLasso" = "#4DAF4A", "DMRcate" = "#984EA3",
        "Bumphunter" = "#FF7F00"
    )

    plot_data <- do.call(rbind, lapply(names(methods_list), function(method) {
        overlapping <- subsetByOverlaps(methods_list[[method]], extended_region)
        if (length(overlapping) > 0) {
            data.frame(
                Method = method,
                start = start(overlapping),
                end = end(overlapping),
                y = which(names(methods_list) == method)
            )
        } else {
            NULL
        }
    }))

    if (!is.null(plot_data) && nrow(plot_data) > 0) {
        p_locus <- ggplot(plot_data) +
            geom_rect(aes(
                xmin = start, xmax = end,
                ymin = y - 0.4, ymax = y + 0.4,
                fill = Method
            ), alpha = 0.7) +
            geom_vline(
                xintercept = c(start(top_consensus), end(top_consensus)),
                linetype = "dashed", color = "gray40"
            ) +
            scale_y_continuous(
                breaks = seq_along(methods_list),
                labels = names(methods_list)
            ) +
            scale_fill_manual(values = method_colors) +
            theme_minimal() +
            theme(
                legend.position = "none",
                panel.grid.minor = element_blank()
            ) +
            labs(
                title = sprintf(
                    "Example Consensus DMR: %s:%d-%d",
                    as.character(seqnames(top_consensus)),
                    start(top_consensus), end(top_consensus)
                ),
                subtitle = sprintf("Detected by %d methods", mcols(top_consensus)$n_methods),
                x = "Genomic Position", y = ""
            )

        print(p_locus)
    }
}
```

## DMRsegal Ranking Quality Assessment

```{r}
if ("rank" %in% names(mcols(dmrs_segal)) && length(dmrs_segal) > 10) {
    dmrs_segal_ranked <- dmrs_segal[order(mcols(dmrs_segal)$rank)]

    n_bins <- min(10, floor(length(dmrs_segal_ranked) / 5))
    bin_size <- ceiling(length(dmrs_segal_ranked) / n_bins)

    ranking_quality <- do.call(rbind, lapply(seq_len(n_bins), function(i) {
        start_idx <- (i - 1) * bin_size + 1
        end_idx <- min(i * bin_size, length(dmrs_segal_ranked))
        bin_dmrs <- dmrs_segal_ranked[start_idx:end_idx]

        in_consensus <- sum(overlapsAny(bin_dmrs, consensus_3plus))

        data.frame(
            Rank_Bin = sprintf("Top %d%%", i * 10),
            Bin_Order = i,
            N_DMRs = length(bin_dmrs),
            In_Consensus = in_consensus,
            Pct_In_Consensus = round(100 * in_consensus / length(bin_dmrs), 1)
        )
    }))

    knitr::kable(ranking_quality, caption = "DMRsegal Ranking Quality: Consensus Overlap by Rank Decile")

    ggplot(ranking_quality, aes(x = reorder(Rank_Bin, Bin_Order), y = Pct_In_Consensus)) +
        geom_bar(stat = "identity", fill = "steelblue") +
        geom_line(aes(group = 1), color = "red", size = 1) +
        geom_point(color = "red", size = 3) +
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        labs(
            title = "DMRsegal Ranking Quality Assessment",
            subtitle = "Higher-ranked DMRs should have higher consensus overlap",
            x = "Rank Decile", y = "% in Consensus (3+ methods)"
        )
}
```

## Summary Statistics

```{r}
summary_table <- results_comparison %>%
    left_join(effect_size_df, by = "Method") %>%
    left_join(capture_df, by = "Method") %>%
    left_join(specificity_df %>% dplyr::select(Method, Pct_Specific, Pct_Shared_2plus, Pct_Shared_3plus), by = "Method")

knitr::kable(summary_table, digits = 2, caption = "Comprehensive Method Comparison Summary")
```

```{r}
cat("\n## Key Findings\n\n")

if (nrow(results_comparison) > 0) {
    max_dmrs <- results_comparison[which.max(results_comparison$Number_of_DMRs), ]
    min_dmrs <- results_comparison[which.min(results_comparison$Number_of_DMRs), ]

    cat(sprintf(
        "- **%s** detected the most DMRs (%d regions)\n",
        max_dmrs$Method, max_dmrs$Number_of_DMRs
    ))
    cat(sprintf(
        "- **%s** was most selective (%d regions)\n",
        min_dmrs$Method, min_dmrs$Number_of_DMRs
    ))

    if (exists("capture_df") && nrow(capture_df) > 0) {
        best_capture <- capture_df[which.max(capture_df$Capture_Rate_Pct), ]
        cat(sprintf(
            "- **%s** captured %.1f%% of high-confidence (3+ methods) DMRs\n",
            best_capture$Method, best_capture$Capture_Rate_Pct
        ))
    }

    if (exists("effect_size_df") && nrow(effect_size_df) > 0) {
        best_effect <- effect_size_df[which.max(effect_size_df$Mean_Delta_Beta), ]
        cat(sprintf(
            "- **%s** had the highest mean effect size (delta beta = %.4f)\n",
            best_effect$Method, best_effect$Mean_Delta_Beta
        ))
    }

    if (exists("specificity_df") && "DMRsegal" %in% specificity_df$Method) {
        segal_row <- specificity_df[specificity_df$Method == "DMRsegal", ]
        cat(sprintf(
            "- **DMRsegal**: %.1f%% method-specific, %.1f%% shared with at least 2 methods, %.1f%% shared with at least 3 methods\n",
            segal_row$Pct_Specific, segal_row$Pct_Shared_2plus, segal_row$Pct_Shared_3plus
        ))
    }
}
```

## Session Info

```{r}
sessionInfo()
```
