---
title: "DMRsegal vs comb-p, probeLasso, DMRcate, and Bumphunter"
author: "DMRsegal Package"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{DMRsegal vs comb-p, probeLasso, DMRcate, and Bumphunter}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    warning = TRUE,
    message = TRUE,
    fig.width = 10,
    fig.height = 6,
    dpi = 150
)
```

## Introduction

This vignette provides an extended benchmark comparing DMRsegal against multiple popular DMR detection methods:

- **DMRsegal**: Correlation-based seed-guided assembly approach
- **comb-p**: Combines adjacent p-values using Stouffer-Liptak-Kechris correction
- **probeLasso**: Lasso-based feature selection for DMR detection
- **DMRcate**: Gaussian kernel smoothing approach
- **Bumphunter**: Permutation-based regional testing

We evaluate these methods on:

- Number and size distribution of DMRs
- Overlap and concordance between methods
- Computational performance
- Biological relevance of detected regions

## Setup and Dependencies

```{r load_libraries, message=FALSE}
devtools::load_all()
library(minfi)
library(data.table)
library(ggplot2)
library(dplyr)
library(GenomicRanges)
library(progressr)

progressr::handlers("cli")
benchmark_output_dir <- "benchmark_data"
dir.create(benchmark_output_dir, showWarnings = FALSE, recursive = TRUE)
```

### Install External Dependencies

```{r install_external, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

if (!requireNamespace("DMRcate", quietly = TRUE)) {
    BiocManager::install("DMRcate", update = FALSE, ask = FALSE)
}

if (!requireNamespace("bumphunter", quietly = TRUE)) {
    BiocManager::install("bumphunter", update = FALSE, ask = FALSE)
}

if (!requireNamespace("doParallel", quietly = TRUE)) {
    install.packages("doParallel")
}

if (!requireNamespace("reticulate", quietly = TRUE)) {
    install.packages("reticulate")
}

if (!requireNamespace("ChAMP", quietly = TRUE)) {
    BiocManager::install("ChAMP", update = FALSE, ask = FALSE)
}
```

### Setup comb-p (conda required)

**Note:** comb-p requires both the tool itself and bedtools. Install via conda:

```bash
# Install comb-p and bedtools
conda install -yc bioconda combined-pvalues bedtools

# Or create a dedicated environment
conda create -n combp python=3.9
conda activate combp
conda install -yc bioconda combined-pvalues bedtools
```

```{r check_combp, eval=FALSE}
# Check if comb-p and bedtools are available
if (Sys.which("comb-p") == "") {
    warning("comb-p not found. Install with: conda install -yc bioconda combined-pvalues")
}

if (Sys.which("bedtools") == "") {
    warning("bedtools not found. Install with: conda install -yc bioconda bedtools")
}
```

## Load Example Data

```{r load_data}
beta <- loadExampleInputData("beta")
pheno <- loadExampleInputData("pheno")
array_type <- loadExampleInputData("array_type")
genome <- "hg19"

beta_handler <- DMRsegal::getBetaHandler(beta, array = array_type, genome = genome)
beta_mat <- as.matrix(beta_handler$getBeta())
locs <- beta_handler$getBetaLocs()
mvalues <- log2(beta_mat / (1 - beta_mat + 1e-6) + 1e-6)
pheno$casecontrol <- pheno$Sample_Group == "cancer"
```

## Find Differentially Methylated Positions

```{r get_dmps}
dmps <- suppressWarnings(dmpFinder(mvalues,
    pheno = pheno$casecontrol,
    type = "categorical",
    shrinkVar = TRUE
))
dmps <- dmps[!is.na(dmps$pval), ]
dmps$pval_adj <- p.adjust(dmps$pval, method = "BH")

sig_dmps <- dmps[dmps$pval_adj < 0.05, ]
cat(sprintf("Found %d significant DMPs (FDR < 0.05)\n", nrow(sig_dmps)))

dmps_gr <- makeGRangesFromDataFrame(
    data.frame(
        chr = locs$chr,
        start = locs$start,
        end = locs$end,
        pval = dmps$pval,
        qval = dmps$pval_adj,
        stat = dmps$qval
    ),
    keep.extra.columns = TRUE
)
names(dmps_gr) <- rownames(dmps)
```

## Method 1: DMRsegal

```{r DMRsegal}
DMRsegal_file <- file.path(benchmark_output_dir, "dmrs.DMRsegal.rds")
if (!file.exists(DMRsegal_file)) {
    options("DMRsegal.verbose" = 1)
    options("DMRsegal.njobs" = 8)

    dmrs_segal <- progressr::with_progress(DMRsegal::findDMRsFromSeeds(
        beta = beta_handler,
        seeds = sig_dmps,
        pheno = pheno,
        sample_group_col = "Sample_Group",
        casecontrol_col = "casecontrol",
        covariates = c("Age", "Gender"),
        min_seeds = 2,
        min_cpgs = 3,
        max_lookup_dist = 1000,
        max_pval = 0.05,
        pval_mode = "parametric",
        group_concordance_strategy = "relaxed",
        rank_dmrs = TRUE
    ))
    saveRDS(dmrs_segal, DMRsegal_file)
    rm(beta_handler)
} else {
    dmrs_segal <- readRDS(DMRsegal_file)
}

cat(sprintf("DMRsegal: Found %d DMRs\n", length(dmrs_segal)))
```

## Method 2: comb-p

comb-p combines adjacent p-values using autocorrelation correction.

```{r combp}
combp_file <- file.path(benchmark_output_dir, "dmrs.combp.rds")

run_combp <- function(dmps_gr, output_dir, seed = 1000, dist = 1000) {
    if (Sys.which("bedtools") == "") {
        warning("bedtools not found in PATH. comb-p requires bedtools to be installed.")
        warning("Install with: conda install -yc bioconda bedtools")
        return(GRanges())
    }
    
    if (Sys.which("comb-p") == "") {
        warning("comb-p not found in PATH.")
        warning("Install with: conda install -yc bioconda combined-pvalues")
        return(GRanges())
    }

    bed_file <- file.path(output_dir, "dmps_for_combp.bed")
    bed_df <- data.frame(
        chrom = as.character(seqnames(dmps_gr)),
        start = start(dmps_gr) - 1,
        end = end(dmps_gr),
        pval = mcols(dmps_gr)$pval
    )
    bed_df <- bed_df[order(bed_df$chrom, bed_df$start), ]

    write.table(bed_df, bed_file,
        quote = FALSE, sep = "\t",
        row.names = FALSE, col.names = TRUE
    )

    output_prefix <- file.path(output_dir, "combp_output")

    cmd <- sprintf(
        "comb-p pipeline -c 4 --seed %d --dist %d -p %s --region-filter-p 0.05 --region-filter-n 2 %s",
        seed, dist, output_prefix, bed_file
    )

    tryCatch(
        {
            exit_code <- system(cmd, ignore.stdout = FALSE, ignore.stderr = FALSE)
            
            if (exit_code != 0) {
                warning(sprintf("comb-p exited with code %d", exit_code))
                return(GRanges())
            }

            regions_file <- paste0(output_prefix, ".regions-p.bed.gz")
            if (file.exists(regions_file)) {
                regions <- fread(cmd = sprintf("zcat %s", regions_file))
                colnames(regions) <- c("chr", "start", "end", "min_p", "n_probes", "z_p", "z_sidak_p")

                dmrs_combp <- makeGRangesFromDataFrame(regions,
                    seqnames.field = "chr",
                    start.field = "start",
                    end.field = "end",
                    keep.extra.columns = TRUE
                )
                return(dmrs_combp)
            } else {
                warning("comb-p did not produce output file")
                return(GRanges())
            }
        },
        error = function(e) {
            warning(sprintf("comb-p failed: %s", e$message))
            return(GRanges())
        }
    )
}

if (!file.exists(combp_file)) {
    dmrs_combp <- run_combp(dmps_gr, benchmark_output_dir)
    saveRDS(dmrs_combp, combp_file)
} else {
    dmrs_combp <- readRDS(combp_file)
}

cat(sprintf("comb-p: Found %d DMRs\n", length(dmrs_combp)))
```

## Method 3: probeLasso (via ChAMP)

probeLasso uses lasso regression for feature selection to identify DMRs. It must be run through the ChAMP package.

```{r probeLasso}
probeLasso_file <- file.path(benchmark_output_dir, "dmrs.probeLasso.rds")

run_probeLasso_champ <- function(beta_mat, pheno, array_type = "450K") {
    if (!requireNamespace("ChAMP", quietly = TRUE)) {
        warning("ChAMP not installed. probeLasso requires ChAMP.")
        warning("Install with: BiocManager::install('ChAMP')")
        return(GRanges())
    }

    library(ChAMP)

    tryCatch(
        {
            result <- champ.DMR(
                beta = beta_mat,
                pheno = pheno$Sample_Group,
                arraytype = array_type,
                method = "ProbeLasso",
                resultsDir = "/tmp/champ_probeLasso",
            )$ProbeLassoDMR

            return(makeGRangesFromDataFrame(result, keep.extra.columns = TRUE))
        },
        error = function(e) {
            warning(sprintf("probeLasso (ChAMP) failed: %s", e$message))
            return(GRanges())
        }
    )
}

if (!file.exists(probeLasso_file)) {
    dmrs_probeLasso <- run_probeLasso_champ(beta_mat, pheno, array_type = "450K")
    saveRDS(dmrs_probeLasso, probeLasso_file)
} else {
    dmrs_probeLasso <- readRDS(probeLasso_file)
}

cat(sprintf("probeLasso: Found %d DMRs\n", length(dmrs_probeLasso)))
```

## Method 4: DMRcate

```{r dmrcate}
dmrcate_file <- file.path(benchmark_output_dir, "dmrs.dmrcate.rds")
if (!file.exists(dmrcate_file)) {
    library(DMRcate)

    design <- model.matrix(~Sample_Group, data = pheno)
    colnames(design) <- c("(Intercept)", "cancer_vs_normal")

    myannotation <- cpg.annotate("array", mvalues,
        what = "M", arraytype = "450K",
        analysis.type = "differential", design = design,
        coef = 2, fdr = 0.05
    )

    dmrcate_results <- dmrcate(myannotation, C = 2, lambda = 1000)
    dmrs_dmrcate <- extractRanges(dmrcate_results, genome = "hg19")
    saveRDS(dmrs_dmrcate, dmrcate_file)
    detach("package:DMRcate", unload = TRUE)
} else {
    dmrs_dmrcate <- readRDS(dmrcate_file)
}

cat(sprintf("DMRcate: Found %d DMRs\n", length(dmrs_dmrcate)))
```

## Method 5: Bumphunter

```{r bumphunter}
bumphunter_file <- file.path(benchmark_output_dir, "dmrs.bumphunter.rds")
if (!file.exists(bumphunter_file)) {
    library(bumphunter)
    library(doParallel)

    doParallel::registerDoParallel(cores = 4)
    design <- model.matrix(~Sample_Group, data = pheno)

    dmrs_cutoff <- 0.2
    dmrs_cnt <- Inf
    while (dmrs_cnt > 30000) {
        dmrs <- bumphunter(beta_mat,
            design = design,
            cutoff = dmrs_cutoff,
            chr = as.character(locs$chr),
            pos = locs$start,
            B = 0,
            type = "Beta",
            coef = 2
        )
        dmrs_cnt <- nrow(dmrs$table)
        dmrs_cutoff <- dmrs_cutoff * 1.3
    }
    bumphunter_results <- bumphunter(beta_mat,
        design = design,
        chr = as.character(locs$chr),
        pos = locs$start,
        cutoff = dmrs_cutoff / 1.3,
        B = 100,
        type = "Beta",
        coef = 2
    )

    dmrs_bumphunter <- makeGRangesFromDataFrame(bumphunter_results$table,
        seqnames.field = "chr",
        start.field = "start",
        end.field = "end",
        keep.extra.columns = TRUE
    )

    saveRDS(dmrs_bumphunter, bumphunter_file)
} else {
    dmrs_bumphunter <- readRDS(bumphunter_file)
}

cat(sprintf("Bumphunter: Found %d DMRs\n", length(dmrs_bumphunter)))
```

## Comparative Analysis

### Basic Statistics

```{r basic_stats}
get_dmr_stats <- function(dmrs, method) {
    if (length(dmrs) == 0) {
        return(data.frame(
            Method = method,
            Number_of_DMRs = 0,
            Mean_DMR_Width = NA,
            Median_DMR_Width = NA,
            Total_Coverage_bp = 0,
            Min_Width = NA,
            Max_Width = NA
        ))
    }

    data.frame(
        Method = method,
        Number_of_DMRs = length(dmrs),
        Mean_DMR_Width = mean(width(dmrs)),
        Median_DMR_Width = median(width(dmrs)),
        Total_Coverage_bp = sum(width(dmrs)),
        Min_Width = min(width(dmrs)),
        Max_Width = max(width(dmrs))
    )
}

stats_list <- list(
    get_dmr_stats(dmrs_segal, "DMRsegal"),
    get_dmr_stats(dmrs_combp, "comb-p"),
    get_dmr_stats(dmrs_probeLasso, "probeLasso"),
    get_dmr_stats(dmrs_dmrcate, "DMRcate"),
    get_dmr_stats(dmrs_bumphunter, "Bumphunter")
)

results_comparison <- do.call(rbind, stats_list)
knitr::kable(results_comparison, digits = 2, caption = "Basic DMR Statistics Across Methods")
```

### Pairwise Overlap Analysis

```{r overlap_analysis}
methods_list <- list(
    DMRsegal = dmrs_segal,
    comb_p = dmrs_combp,
    probeLasso = dmrs_probeLasso,
    DMRcate = dmrs_dmrcate,
    Bumphunter = dmrs_bumphunter
)

methods_list <- methods_list[sapply(methods_list, length) > 0]

calculate_overlap <- function(gr1, gr2) {
    if (length(gr1) == 0 || length(gr2) == 0) {
        return(list(count = 0, percentage = 0))
    }
    overlap_count <- length(subsetByOverlaps(gr1, gr2))
    pct <- (overlap_count / length(gr1)) * 100
    list(count = overlap_count, percentage = pct)
}

overlap_matrix <- matrix(0, nrow = length(methods_list), ncol = length(methods_list))
pct_matrix <- matrix(0, nrow = length(methods_list), ncol = length(methods_list))
rownames(overlap_matrix) <- colnames(overlap_matrix) <- names(methods_list)
rownames(pct_matrix) <- colnames(pct_matrix) <- names(methods_list)

for (i in seq_along(methods_list)) {
    for (j in seq_along(methods_list)) {
        if (i != j) {
            overlap_result <- calculate_overlap(methods_list[[i]], methods_list[[j]])
            overlap_matrix[i, j] <- overlap_result$count
            pct_matrix[i, j] <- overlap_result$percentage
        }
    }
}

cat("\n### Overlap Counts (rows overlap with columns)\n")
knitr::kable(overlap_matrix, digits = 0, caption = "Number of DMRs from Method A (rows) overlapping with Method B (columns)")

cat("\n### Overlap Percentages\n")
knitr::kable(pct_matrix, digits = 2, caption = "Percentage of DMRs from Method A (rows) overlapping with Method B (columns)")
```

### Jaccard Index and IoU

```{r similarity_metrics}
jaccard_index <- function(gr1, gr2) {
    if (length(gr1) == 0 || length(gr2) == 0) {
        return(0)
    }
    overlap_count <- length(subsetByOverlaps(gr1, gr2))
    total_unique <- length(unique(c(gr1, gr2)))
    if (total_unique == 0) {
        return(0)
    }
    overlap_count / total_unique
}

iou <- function(gr1, gr2) {
    if (length(gr1) == 0 || length(gr2) == 0) {
        return(0)
    }
    intersection <- sum(width(GenomicRanges::intersect(gr1, gr2)))
    union <- sum(width(GenomicRanges::union(gr1, gr2)))
    if (union == 0) {
        return(0)
    }
    intersection / union
}

jaccard_matrix <- matrix(0, nrow = length(methods_list), ncol = length(methods_list))
iou_matrix <- matrix(0, nrow = length(methods_list), ncol = length(methods_list))
rownames(jaccard_matrix) <- colnames(jaccard_matrix) <- names(methods_list)
rownames(iou_matrix) <- colnames(iou_matrix) <- names(methods_list)

for (i in seq_along(methods_list)) {
    for (j in seq_along(methods_list)) {
        if (i < j) {
            jaccard_matrix[i, j] <- jaccard_matrix[j, i] <- jaccard_index(methods_list[[i]], methods_list[[j]])
            iou_matrix[i, j] <- iou_matrix[j, i] <- iou(methods_list[[i]], methods_list[[j]])
        } else if (i == j) {
            jaccard_matrix[i, j] <- 1
            iou_matrix[i, j] <- 1
        }
    }
}

cat("\n### Jaccard Index\n")
knitr::kable(jaccard_matrix, digits = 4, caption = "Jaccard Index (region count-based similarity)")

cat("\n### Intersection over Union (IoU)\n")
knitr::kable(iou_matrix, digits = 4, caption = "IoU (genomic coverage-based similarity)")
```

## Visualizations

### DMR Count and Coverage Comparison

```{r viz_basic, fig.width=10, fig.height=5}
viz_data <- results_comparison %>%
    select(Method, Number_of_DMRs, Total_Coverage_bp) %>%
    tidyr::pivot_longer(
        cols = c(Number_of_DMRs, Total_Coverage_bp),
        names_to = "Metric", values_to = "Value"
    )

viz_data$Metric <- ifelse(viz_data$Metric == "Number_of_DMRs",
    "Number of DMRs",
    "Total Coverage (bp)"
)

p1 <- ggplot(viz_data, aes(x = Method, y = Value, fill = Method)) +
    geom_bar(stat = "identity") +
    facet_wrap(~Metric, scales = "free_y") +
    theme_minimal() +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none",
        panel.grid.major.x = element_blank()
    ) +
    labs(
        title = "DMR Detection: Count and Coverage Comparison",
        y = "Value", x = ""
    )

print(p1)
```

### DMR Width Distribution

```{r viz_width, fig.width=10, fig.height=6}
width_data <- do.call(rbind, lapply(names(methods_list), function(method) {
    if (length(methods_list[[method]]) > 0) {
        data.frame(
            Method = method,
            Width = width(methods_list[[method]])
        )
    }
}))

p2 <- ggplot(width_data, aes(x = Method, y = Width, fill = Method)) +
    geom_violin(alpha = 0.7) +
    geom_boxplot(width = 0.2, alpha = 0.5, outlier.size = 0.5) +
    scale_y_log10(labels = scales::comma) +
    theme_minimal() +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none"
    ) +
    labs(
        title = "Distribution of DMR Widths",
        y = "Width (bp, log scale)", x = ""
    )

print(p2)
```

### Overlap Heatmap

```{r viz_overlap, fig.width=8, fig.height=6}
library(reshape2)

overlap_df <- melt(pct_matrix)
colnames(overlap_df) <- c("Method1", "Method2", "Percentage")

p3 <- ggplot(overlap_df, aes(x = Method2, y = Method1, fill = Percentage)) +
    geom_tile() +
    geom_text(aes(label = sprintf("%.1f%%", Percentage)), size = 3) +
    scale_fill_gradient2(
        low = "white", mid = "lightblue", high = "darkblue",
        midpoint = 50, limits = c(0, 100)
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(
        title = "DMR Overlap Matrix",
        subtitle = "% of row method's DMRs overlapping with column method",
        x = "", y = "", fill = "Overlap %"
    )

print(p3)
```

### Similarity Matrix (IoU)

```{r viz_iou, fig.width=8, fig.height=6}
iou_df <- melt(iou_matrix)
colnames(iou_df) <- c("Method1", "Method2", "IoU")

p4 <- ggplot(iou_df, aes(x = Method2, y = Method1, fill = IoU)) +
    geom_tile() +
    geom_text(aes(label = sprintf("%.3f", IoU)), size = 3) +
    scale_fill_gradient2(
        low = "white", mid = "yellow", high = "red",
        midpoint = 0.5, limits = c(0, 1)
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(
        title = "Method Similarity Matrix (IoU)",
        subtitle = "Intersection over Union of genomic coverage",
        x = "", y = "", fill = "IoU"
    )

print(p4)
```

## Consensus DMRs

Let's identify consensus regions detected by multiple methods:

```{r consensus}
find_consensus <- function(methods_list, min_methods = 2) {
    all_dmrs <- do.call(c, unname(methods_list))
    all_dmrs <- GenomicRanges::reduce(all_dmrs)

    consensus_counts <- vapply(seq_along(all_dmrs), function(i) {
        region <- all_dmrs[i]
        sum(sapply(methods_list, function(method_dmrs) {
            length(subsetByOverlaps(region, method_dmrs)) > 0
        }))
    }, integer(1))

    consensus_dmrs <- all_dmrs[consensus_counts >= min_methods]
    mcols(consensus_dmrs)$n_methods <- consensus_counts[consensus_counts >= min_methods]

    return(consensus_dmrs)
}

if (length(methods_list) > 1) {
    consensus_2plus <- find_consensus(methods_list, min_methods = 2)
    consensus_3plus <- find_consensus(methods_list, min_methods = 3)

    cat(sprintf("\nConsensus DMRs (detected by 2+ methods): %d\n", length(consensus_2plus)))
    cat(sprintf("Consensus DMRs (detected by 3+ methods): %d\n", length(consensus_3plus)))

    if (length(consensus_3plus) > 0) {
        consensus_table <- table(mcols(consensus_3plus)$n_methods)
        cat("\nDistribution of consensus DMRs by number of methods:\n")
        print(consensus_table)
    }
}
```

## Method-Specific DMRs

```{r specific_dmrs}
find_specific_dmrs <- function(target_method, methods_list) {
    target_dmrs <- methods_list[[target_method]]
    if (length(target_dmrs) == 0) {
        return(GRanges())
    }

    other_methods <- methods_list[names(methods_list) != target_method]
    all_others <- do.call(c, unname(other_methods))

    if (length(all_others) == 0) {
        return(target_dmrs)
    }

    specific <- target_dmrs[countOverlaps(target_dmrs, all_others) == 0]
    return(specific)
}

specific_dmrs_list <- lapply(names(methods_list), function(method) {
    specific <- find_specific_dmrs(method, methods_list)
    data.frame(
        Method = method,
        Total_DMRs = length(methods_list[[method]]),
        Specific_DMRs = length(specific),
        Percentage_Specific = ifelse(length(methods_list[[method]]) > 0,
            (length(specific) / length(methods_list[[method]])) * 100,
            0
        )
    )
})

specific_summary <- do.call(rbind, specific_dmrs_list)
knitr::kable(specific_summary, digits = 2, caption = "Method-Specific DMRs (not overlapping with any other method)")
```

## Performance Summary

```{r summary_table}
summary_df <- results_comparison %>%
    left_join(specific_summary, by = "Method") %>%
    select(
        Method, Number_of_DMRs, Mean_DMR_Width, Median_DMR_Width,
        Total_Coverage_bp, Specific_DMRs, Percentage_Specific
    )

knitr::kable(summary_df,
    digits = 2,
    caption = "Comprehensive Performance Summary"
)
```

## Key Findings

```{r findings, echo=FALSE, results='asis'}
if (nrow(results_comparison) > 0) {
    max_dmrs <- results_comparison[which.max(results_comparison$Number_of_DMRs), ]
    min_dmrs <- results_comparison[which.min(results_comparison$Number_of_DMRs), ]

    cat(sprintf(
        "- **%s** detected the most DMRs (%d regions)\n",
        max_dmrs$Method, max_dmrs$Number_of_DMRs
    ))
    cat(sprintf(
        "- **%s** was most selective (%d regions)\n",
        min_dmrs$Method, min_dmrs$Number_of_DMRs
    ))

    if (length(consensus_3plus) > 0) {
        cat(sprintf(
            "- **%d DMRs** were detected by 3 or more methods (high confidence)\n",
            length(consensus_3plus)
        ))
    }

    segal_idx <- which(results_comparison$Method == "DMRsegal")
    if (length(segal_idx) > 0) {
        segal_specific_pct <- specific_summary$Percentage_Specific[
            specific_summary$Method == "DMRsegal"
        ]
        cat(sprintf(
            "- **DMRsegal** found %.1f%% method-specific DMRs\n",
            segal_specific_pct
        ))
    }
}
```


## Session Info

```{r session_info}
sessionInfo()
```
