---
title: "Registry (SQLite) vs data.frame vs DelayedDataFrame: Large-Input Benchmarks"
author: "DMRsegal Package"
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
        code_folding: hide
vignette: >
  %\VignetteIndexEntry{Registry (SQLite) vs data.frame vs DelayedDataFrame: Large-Input Benchmarks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    warning = FALSE,
    message = FALSE,
    fig.width = 11,
    fig.height = 6,
    dpi = 180
)
```

## Goal

This vignette benchmarks SQL-backed `Registry` operations against equivalent
`data.frame` and `DelayedDataFrame` operations on large synthetic inputs.

Benchmarked operation families:

- input loading from gzipped TSV
- row/column subsetting
- row subsetting via `master_index` (character selector in `[` row position)
- indexed lookup via `queryOnIndex` (single and composite index)
- column extraction (`$`) and two-column `paste(...)`
- materialization (`as.data.frame`) of active views

## Setup

```{r}
library(data.table)
library(ggplot2)
library(R.utils)

if (requireNamespace("DMRsegal", quietly = TRUE)) {
    suppressPackageStartupMessages(library(DMRsegal))
}
if (!exists("getRegistry", mode = "function")) {
    if (!requireNamespace("devtools", quietly = TRUE)) {
        stop("Could not load getRegistry(). Install DMRsegal or install 'devtools'.")
    }
    devtools::load_all(".", quiet = TRUE)
}
stopifnot(exists("getRegistry", mode = "function"))

if (!requireNamespace("DelayedArray", quietly = TRUE)) {
    BiocManager::install(c("DelayedArray", "S4Vectors"), ask = FALSE, update = FALSE)
}
set.seed(42)

bench_n_rows <- as.integer(Sys.getenv("DMRSEGAL_REGISTRY_BENCH_N_ROWS", "250000"))
bench_n_beta_cols <- as.integer(Sys.getenv("DMRSEGAL_REGISTRY_BENCH_N_BETA_COLS", "8"))
bench_iters <- as.integer(Sys.getenv("DMRSEGAL_REGISTRY_BENCH_ITERS", "3"))
bench_ingest_iters <- as.integer(Sys.getenv("DMRSEGAL_REGISTRY_BENCH_INGEST_ITERS", "3"))

stopifnot(bench_n_rows > 0L, bench_n_beta_cols > 0L, bench_iters > 0L, bench_ingest_iters > 0L)

cat("DelayedArray/S4Vectors not available; DelayedDataFrame benchmarks will be skipped.\n")
```

## Generate Large Synthetic Input

```{r}
n <- bench_n_rows
p <- bench_n_beta_cols

ids <- sprintf("cg%08d", seq_len(n))
chr_levels <- paste0("chr", c(1:22, "X", "Y"))

df <- data.frame(
    id = ids,
    chr = sample(chr_levels, n, replace = TRUE),
    start = sample.int(200000000, n, replace = TRUE),
    width = sample.int(600, n, replace = TRUE),
    strand = sample(c("+", "-"), n, replace = TRUE),
    group = sample(sprintf("g%02d", 1:40), n, replace = TRUE),
    stringsAsFactors = FALSE,
    check.names = FALSE
)
df$end <- df$start + df$width

beta_mat <- matrix(runif(n * p), nrow = n, ncol = p)
colnames(beta_mat) <- sprintf("beta_%02d", seq_len(p))
df <- cbind(df, as.data.frame(beta_mat, stringsAsFactors = FALSE, check.names = FALSE))

rownames(df) <- df$id
ncol(df)
nrow(df)
```

## Benchmark Utilities

```{r}
run_benchmark <- function(label, fun, iterations = 3L) {
    elapsed <- numeric(iterations)
    for (k in seq_len(iterations)) {
        gc(verbose = FALSE)
        elapsed[[k]] <- system.time({
            out <- fun()
            invisible(out)
        })[["elapsed"]]
    }
    data.frame(
        operation = label,
        iteration = seq_len(iterations),
        elapsed_sec = elapsed,
        stringsAsFactors = FALSE,
        check.names = FALSE
    )
}

summarize_benchmark <- function(res) {
    ops <- unique(res$operation)
    out <- lapply(ops, function(op) {
        vals <- res$elapsed_sec[res$operation == op]
        data.frame(
            operation = op,
            median_sec = median(vals),
            min_sec = min(vals),
            max_sec = max(vals),
            stringsAsFactors = FALSE,
            check.names = FALSE
        )
    })
    out <- do.call(rbind, out)
    out[order(out$median_sec), , drop = FALSE]
}

make_key <- function(dat, cols) {
    do.call(paste, c(dat[, cols, drop = FALSE], sep = "\r"))
}

make_delayed_df <- function(input_df) {
    delayed_cols <- lapply(input_df, function(col) {
        DelayedArray::DelayedArray(array(col, dim = c(length(col))))
    })
    out <- S4Vectors::DataFrame(delayed_cols, check.names = FALSE)
    rownames(out) <- rownames(input_df)
    out
}
```

## Prepare Equivalent Query Objects

```{r}
subset_rows <- seq_len(min(30000L, n))
subset_cols <- c("id", "chr", "start", "group", "beta_01", "beta_02")
subset_cols <- subset_cols[subset_cols %in% colnames(df)]

query_ids <- sample(df$id, size = min(80000L, n), replace = TRUE)
query_ids <- c(query_ids, sprintf("missing_%06d", seq_len(2000L)))

query_chr_start <- df[
    sample.int(n, size = min(50000L, n), replace = TRUE),
    c("chr", "start"),
    drop = FALSE
]
query_chr_start <- rbind(
    query_chr_start,
    data.frame(chr = rep("chrNA", 1000L), start = rep(-1L, 1000L), stringsAsFactors = FALSE)
)

id_to_row <- setNames(seq_len(nrow(df)), df$id)
df_chr_start_key <- make_key(df, c("chr", "start"))
chr_start_to_rows <- split(seq_len(nrow(df)), df_chr_start_key)

    delayed_df <- make_delayed_df(df)


df_subset_master <- function(ids_in, cols_out = NULL) {
    idx <- unname(id_to_row[ids_in])
    idx <- idx[!is.na(idx)]
    if (length(idx) == 0L) {
        return(df[0, cols_out, drop = FALSE])
    }
    if (is.null(cols_out)) {
        return(df[idx, , drop = FALSE])
    }
    df[idx, cols_out, drop = FALSE]
}

df_query_chr_start <- function(lookup_df, cols_out) {
    keys <- make_key(lookup_df, c("chr", "start"))
    idx <- unlist(chr_start_to_rows[keys], use.names = FALSE)
    if (length(idx) == 0L) {
        return(df[0, cols_out, drop = FALSE])
    }
    df[idx, cols_out, drop = FALSE]
}

delayed_subset_master <- function(ids_in, cols_out = NULL) {
    idx <- unname(id_to_row[ids_in])
    idx <- idx[!is.na(idx)]
    if (length(idx) == 0L) {
        if (is.null(cols_out)) {
            return(delayed_df[0, , drop = FALSE])
        }
        return(delayed_df[0, cols_out, drop = FALSE])
    }
    if (is.null(cols_out)) {
        return(delayed_df[idx, , drop = FALSE])
    }
    delayed_df[idx, cols_out, drop = FALSE]
}

delayed_query_chr_start <- function(lookup_df, cols_out) {
    keys <- make_key(lookup_df, c("chr", "start"))
    idx <- unlist(chr_start_to_rows[keys], use.names = FALSE)
    if (length(idx) == 0L) {
        return(delayed_df[0, cols_out, drop = FALSE])
    }
    delayed_df[idx, cols_out, drop = FALSE]
}
```

## Create SQL-Backed Registry

```{r}
sqlite_main <- tempfile(pattern = "registry_bench_main_", fileext = ".sqlite")
idx_defs <- list(
    id = "id",
    chr_start = c("chr", "start")
)

registry_sql <- getRegistry(
    data = df,
    indices = idx_defs,
    master_index = "id",
    memory_threshold_mb = 0,
    sqlite_path = sqlite_main,
    chunk_size = 50000L,
    query_chunk_size = 50000L
)

stopifnot(registry_sql$backend() == "sqlite")
registry_sql
```

## Correctness Spot Checks

```{r}
reg_q1 <- registry_sql$queryOnIndex("id", query_ids, columns = subset_cols)
df_q1 <- df_subset_master(query_ids, cols_out = subset_cols)
stopifnot(identical(reg_q1$id, df_q1$id))

comp_cols <- c("id", "chr", "start")
reg_q2 <- registry_sql$queryOnIndex("chr_start", query_chr_start, columns = comp_cols)
df_q2 <- df_query_chr_start(query_chr_start, cols_out = comp_cols)
stopifnot(identical(reg_q2$id, df_q2$id))

    delayed_q1 <- as.data.frame(delayed_subset_master(query_ids, cols_out = subset_cols))
    delayed_q2 <- as.data.frame(delayed_query_chr_start(query_chr_start, cols_out = comp_cols))
    stopifnot(identical(delayed_q1$id, df_q1$id))
    stopifnot(identical(delayed_q2$id, df_q2$id))

```

## Ingestion Benchmark (gzipped TSV)

```{r}
tsv_plain <- tempfile(pattern = "registry_bench_", fileext = ".tsv")
tsv_gz <- paste0(tsv_plain, ".gz")
utils::write.table(
    df,
    file = tsv_plain,
    sep = "\t",
    quote = FALSE,
    row.names = FALSE,
    col.names = TRUE
)
R.utils::gzip(tsv_plain, destname = tsv_gz, overwrite = TRUE, remove = FALSE)

ingest_cols <- c("id", "chr", "start", "group", "beta_01", "beta_02")
ingest_cols <- ingest_cols[ingest_cols %in% colnames(df)]

bench_ingest_list <- list(
    run_benchmark(
        "ingest_gz:data.frame_fread",
        function() data.table::fread(tsv_gz, select = ingest_cols, data.table = FALSE),
        iterations = bench_ingest_iters
    ),
    run_benchmark(
        "ingest_gz:registry_sql",
        function() {
            tmp_sql <- tempfile(pattern = "registry_bench_ingest_", fileext = ".sqlite")
            tmp_reg <- getRegistry(
                data = tsv_gz,
                select = ingest_cols,
                indices = list(id = "id", chr_start = c("chr", "start")),
                master_index = "id",
                memory_threshold_mb = 0,
                sqlite_path = tmp_sql,
                chunk_size = 50000L,
                query_chunk_size = 50000L
            )
            tmp_reg$close()
            unlink(tmp_sql, force = TRUE)
            invisible(TRUE)
        },
        iterations = bench_ingest_iters
    )
)

    bench_ingest_list <- c(
        bench_ingest_list,
        list(
            run_benchmark(
                "ingest_gz:delayed_df",
                function() {
                    tmp_df <- data.table::fread(tsv_gz, select = ingest_cols, data.table = FALSE)
                    make_delayed_df(tmp_df)
                },
                iterations = bench_ingest_iters
            )
        )
    )


bench_ingest <- do.call(rbind, bench_ingest_list)

summarize_benchmark(bench_ingest)
```

## Query and Subsetting Benchmarks

```{r}
bench_results_list <- list(
    run_benchmark(
        "subset_range:data.frame",
        function() df[subset_rows, subset_cols, drop = FALSE],
        iterations = bench_iters
    ),
    run_benchmark(
        "subset_range:registry_sql",
        function() as.data.frame(registry_sql[subset_rows, subset_cols]),
        iterations = bench_iters
    ),
    run_benchmark(
        "subset_master_index:data.frame",
        function() df_subset_master(query_ids, cols_out = subset_cols),
        iterations = bench_iters
    ),
    run_benchmark(
        "subset_master_index:registry_sql",
        function() as.data.frame(registry_sql[query_ids, subset_cols]),
        iterations = bench_iters
    ),
    run_benchmark(
        "queryOnIndex_id:data.frame",
        function() df_subset_master(query_ids, cols_out = subset_cols),
        iterations = bench_iters
    ),
    run_benchmark(
        "queryOnIndex_id:registry_sql",
        function() registry_sql$queryOnIndex("id", query_ids, columns = subset_cols),
        iterations = bench_iters
    ),
    run_benchmark(
        "queryOnIndex_chr_start:data.frame",
        function() df_query_chr_start(query_chr_start, cols_out = comp_cols),
        iterations = bench_iters
    ),
    run_benchmark(
        "queryOnIndex_chr_start:registry_sql",
        function() registry_sql$queryOnIndex("chr_start", query_chr_start, columns = comp_cols),
        iterations = bench_iters
    ),
    run_benchmark(
        "column_extract:data.frame",
        function() df$start,
        iterations = bench_iters
    ),
    run_benchmark(
        "column_extract:registry_sql",
        function() registry_sql$start,
        iterations = bench_iters
    ),
    run_benchmark(
        "paste_two_cols:data.frame",
        function() paste(df[subset_rows, "chr"], ":", df[subset_rows, "start"]),
        iterations = bench_iters
    ),
    run_benchmark(
        "paste_two_cols:registry_sql",
        function() paste(registry_sql[subset_rows, "chr"], ":", registry_sql[subset_rows, "start"]),
        iterations = bench_iters
    ),
    run_benchmark(
        "materialize_active_view:data.frame",
        function() {
            idx <- sample.int(n, min(40000L, n))
            df[idx, subset_cols, drop = FALSE]
        },
        iterations = bench_iters
    ),
    run_benchmark(
        "materialize_active_view:registry_sql",
        function() {
            idx <- sample.int(n, min(40000L, n))
            as.data.frame(registry_sql[idx, subset_cols])
        },
        iterations = bench_iters
    ),
    run_benchmark(
        "nrow:data.frame",
        function() nrow(df),
        iterations = bench_iters
    ),
    run_benchmark(
        "nrow:registry_sql",
        function() nrow(registry_sql),
        iterations = bench_iters
    )
)

    bench_results_list <- c(
        bench_results_list,
        list(
            run_benchmark(
                "subset_range:delayed_df",
                function() as.data.frame(delayed_df[subset_rows, subset_cols, drop = FALSE]),
                iterations = bench_iters
            ),
            run_benchmark(
                "subset_master_index:delayed_df",
                function() as.data.frame(delayed_subset_master(query_ids, cols_out = subset_cols)),
                iterations = bench_iters
            ),
            run_benchmark(
                "queryOnIndex_id:delayed_df",
                function() as.data.frame(delayed_subset_master(query_ids, cols_out = subset_cols)),
                iterations = bench_iters
            ),
            run_benchmark(
                "queryOnIndex_chr_start:delayed_df",
                function() as.data.frame(delayed_query_chr_start(query_chr_start, cols_out = comp_cols)),
                iterations = bench_iters
            ),
            run_benchmark(
                "column_extract:delayed_df",
                function() as.vector(delayed_df$start),
                iterations = bench_iters
            ),
            run_benchmark(
                "paste_two_cols:delayed_df",
                function() paste(delayed_df[subset_rows, "chr"], ":", delayed_df[subset_rows, "start"]),
                iterations = bench_iters
            ),
            run_benchmark(
                "materialize_active_view:delayed_df",
                function() {
                    idx <- sample.int(n, min(40000L, n))
                    as.data.frame(delayed_df[idx, subset_cols, drop = FALSE])
                },
                iterations = bench_iters
            ),
            run_benchmark(
                "nrow:delayed_df",
                function() nrow(delayed_df),
                iterations = bench_iters
            )
        )
    )


bench_results <- do.call(rbind, bench_results_list)

bench_summary <- summarize_benchmark(bench_results)
bench_summary
```

## Visual Summary

```{r}
plot_df <- bench_summary
backend_code <- sub("^.*:", "", plot_df$operation)
backend_map <- c(
    "registry_sql" = "Registry (SQLite)",
    "data.frame" = "data.frame",
    "delayed_df" = "DelayedDataFrame"
)
plot_df$backend <- unname(backend_map[backend_code])
plot_df$backend[is.na(plot_df$backend)] <- backend_code[is.na(plot_df$backend)]
plot_df$operation <- sub(":(registry_sql|data.frame|delayed_df)$", "", plot_df$operation)

ggplot(plot_df, aes(x = reorder(operation, median_sec), y = median_sec, fill = backend)) +
    geom_col(position = "dodge") +
    coord_flip() +
    labs(
        title = "Median elapsed time per operation",
        x = "Operation",
        y = "Elapsed time (seconds)"
    ) +
    theme_minimal(base_size = 12)
```

## Notes

- `Registry` timings here are explicitly for the SQL backend (`memory_threshold_mb = 0`).
- `DelayedDataFrame` benchmarks are included when both `DelayedArray` and `S4Vectors`
  are available; otherwise they are skipped.
- In this vignette, `DelayedDataFrame` refers to an `S4Vectors::DataFrame`
  whose columns are 1D `DelayedArray` objects.
- `data.frame` is in-memory and will usually be faster for pure vectorized extraction.
- `Registry` is designed to stay scalable for indexed retrieval and very large lookup sets.
- To increase workload, set environment variables before rendering:
  - `DMRSEGAL_REGISTRY_BENCH_N_ROWS`
  - `DMRSEGAL_REGISTRY_BENCH_N_BETA_COLS`
  - `DMRSEGAL_REGISTRY_BENCH_ITERS`
  - `DMRSEGAL_REGISTRY_BENCH_INGEST_ITERS`

## Cleanup

```{r}
registry_sql$close()
unlink(c(sqlite_main, tsv_plain, tsv_gz), force = TRUE)
```
