# Benchmarking DMRSegal against other DMR detection tools
# 
# This notebook compares DMRSegal with other popular DMR detection tools using 
# real-world methylation data. We'll use a small subset of the colon cancer dataset
# from the minfiData package, which contains tumor vs normal samples.

## Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = TRUE)
```

## Load Required Libraries

```{r load_libraries}
library(minfi)
library(DMRcate)
library(bumphunter)
library(DMRSegal)
library(minfiData)  # Contains example 450k array data
library(microbenchmark)
library(ggplot2)
library(progressr)
# Configure progressr for RStudio
# progressr::handlers(global = TRUE)
handlers("rstudio")
```

## Load and Prepare Data

We'll use a subset of the colon cancer dataset from minfiData MsetEx to keep the example manageable.

```{r load_data}
# Load the example data from minfiData
data(MsetEx)

# Get sample type information from MsetEx
sample_types <- pData(MsetEx)$Sample_Group
sample_types[sample_types == "GroupA"] <- 'normal'
sample_types[sample_types == "GroupB"] <- 'cancer'
sample_groups <- factor(sample_types, levels = c("normal", "cancer"))

# Subset to a smaller set for the example (3 cancer, 3 normal)
set.seed(123)
samples_per_group <- 3
cancer_idx <- sample(which(sample_groups == "cancer"), samples_per_group)
normal_idx <- sample(which(sample_groups == "normal"), samples_per_group)
selected_samples <- c(cancer_idx, normal_idx)

# Create the final subset - MsetEx is already preprocessed
mset <- MsetEx[, selected_samples]
pheno <- data.frame(
    status = sample_groups[selected_samples],
    row.names = colnames(mset)
)
pheno$group <- pheno$status
pheno$casecontrol <- pheno$status == "cancer"
```

## Get DMPs using minfi

```{r get_dmps}
# Create design matrix for all methods to use consistently
design <- model.matrix(~pheno$status)
colnames(design) <- c("Intercept", "StatusCancer")

# Get probe locations from annotation
ann450k <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)

# Find DMPs using limma
dmps <- dmpFinder(getBeta(mset), 
                  pheno = pheno$status,
                  type = "categorical")
dmps <- dmps[!is.na(dmps$pval),]
# Add delta beta to dmps results
dmps$pval_adj <- p.adjust(dmps$pval, method = "BH")

# Filter significant DMPs
sig_dmps <- dmps[dmps$pval_adj < 0.1, ]
```

## Run DMR Detection Methods

### DMRSegal

```{r dmrsegal}
# The phenotype data is already created above in the data loading section
# But let's rename the column to match our expected format


# Write beta values to temp file
beta_file <- tempfile(fileext = ".txt")
beta_mat <- getBeta(mset)


write.table(
    cbind(ID = rownames(beta_mat), beta_mat),
    file = beta_file,
    sep = "\t",
    quote = FALSE,
    row.names = FALSE
)

beta_file <- sortBetaFileByCoordinates(beta_file, overwrite=TRUE)

# Write DMPs to temp file
dmps_file <- tempfile(fileext = ".txt")
write.table(
    sig_dmps,
    file = dmps_file,
    sep = "\t",
    quote = FALSE,
    row.names = TRUE
)

# Find DMRs using DMRSegal
dmrs_segal <- findDMRsFromDMPs(
    beta_file = beta_file,
    dmps_tsv_file = dmps_file,
    pheno = pheno,
    sample_group_col = "group",
    min_dmps = 2,
    min_cpgs = 3,
    njobs = 4,
    verbose = TRUE
)
```

### DMRcate

```{r dmrcate}
# DMRcate uses the same mset and design matrix as other methods
M <- getM(mset)

# Fit the same design matrix we created above
fit <- limma::lmFit(M, design)
fit <- limma::eBayes(fit)

# Run DMRcate with the same p-value cutoff we used for DMPs
dmrcate_results <- dmrcate(fit, C = as.matrix(design[,2]), pcutoff = 0.1)
dmrs_dmrcate <- extractRanges(dmrcate_results)
```

### bumphunter

```{r bumphunter}
# bumphunter uses the same mset and design matrix as other methods
# Convert to a matrix format that bumphunter expects
beta_mat <- getBeta(mset)

# Run bumphunter with the same design matrix
bumphunter_results <- bumphunter(
    beta_mat,
    design = design,
    coef = 2,  # Use the cancer vs normal coefficient
    cutoff = 0.1,  # Use same significance threshold
    B = 1000,  # Number of permutations
    type = "Beta"
)
```

## Performance Comparison

```{r prepare_benchmark_data}
# Prepare data matrices for benchmarking (reuse the same data across methods)
M <- getM(mset)  # M-values for DMRcate
beta_mat <- getBeta(mset)  # Beta values for bumphunter
```

```{r benchmark}
# Function wrappers for benchmarking
run_dmrsegal <- function() {
    findDMRsFromDMPs(
        beta_file = beta_file,
        dmps_tsv_file = dmps_file,
        pheno = pheno,
        sample_group_col = "group",
        min_dmps = 2,
        min_cpgs = 3,
        njobs = 1,
        verbose = FALSE
    )
}

run_dmrcate <- function() {
    # Use the same M and design matrices as in the main analysis
    fit <- limma::lmFit(M, design)
    fit <- limma::eBayes(fit)
    dmrcate_results <- dmrcate(fit, C = as.matrix(design[,2]), pcutoff = 0.1)
    extractRanges(dmrcate_results)
}

run_bumphunter <- function() {
    # Use the same beta_mat and design as in the main analysis
    bumphunter(
        beta_mat,
        design = design,
        coef = 2,
        cutoff = 0.1,
        B = 100,  # Reduced for benchmarking
        type = "Beta"
    )
}

# Run benchmark
bench <- microbenchmark(
    DMRSegal = run_dmrsegal(),
    DMRcate = run_dmrcate(),
    bumphunter = run_bumphunter(),
    times = 5
)

# Plot results
autoplot(bench) +
    ggtitle("DMR Detection Method Performance") +
    theme_minimal()
```

## Results Comparison

All three methods now use the same underlying dataset (mset), same phenotype data (pheno), same design matrix (design), and comparable significance thresholds (p < 0.1) for fair comparison.

```{r compare_results}
# Function to get DMR stats
get_dmr_stats <- function(dmrs, method) {
    data.frame(
        Method = method,
        Number_of_DMRs = length(dmrs),
        Mean_DMR_Width = mean(width(dmrs)),
        Median_DMR_Width = median(width(dmrs))
    )
}

# Collect stats
stats_list <- list(
    get_dmr_stats(dmrs_segal, "DMRSegal"),
    get_dmr_stats(dmrs_dmrcate, "DMRcate"),
    get_dmr_stats(bumphunter_results$table, "bumphunter")
)

# Combine stats
results_comparison <- do.call(rbind, stats_list)
knitr::kable(results_comparison)
```

## DMR Overlap Analysis

```{r overlap}
# Convert all results to GRanges
gr_segal <- dmrs_segal
gr_dmrcate <- dmrs_dmrcate
gr_bumphunter <- GRanges(
    seqnames = bumphunter_results$table$chr,
    ranges = IRanges(
        start = bumphunter_results$table$start,
        end = bumphunter_results$table$end
    )
)

# Find overlaps
overlap_segal_dmrcate <- length(subsetByOverlaps(gr_segal, gr_dmrcate))
overlap_segal_bumphunter <- length(subsetByOverlaps(gr_segal, gr_bumphunter))
overlap_dmrcate_bumphunter <- length(subsetByOverlaps(gr_dmrcate, gr_bumphunter))

# Create overlap matrix
overlap_matrix <- matrix(
    c(
        length(gr_segal), overlap_segal_dmrcate, overlap_segal_bumphunter,
        overlap_segal_dmrcate, length(gr_dmrcate), overlap_dmrcate_bumphunter,
        overlap_segal_bumphunter, overlap_dmrcate_bumphunter, length(gr_bumphunter)
    ),
    nrow = 3,
    dimnames = list(
        c("DMRSegal", "DMRcate", "bumphunter"),
        c("DMRSegal", "DMRcate", "bumphunter")
    )
)

# Plot heatmap
pheatmap::pheatmap(
    overlap_matrix,
    main = "DMR Overlap Between Methods",
    display_numbers = TRUE
)
```

## Clean Up

```{r cleanup}
# Remove temporary files
unlink(c(beta_file, dmps_file))
```

## Conclusions

This benchmark analysis compares DMRSegal with two popular DMR detection methods (DMRcate and bumphunter) using the same underlying methylation dataset for fair comparison. All methods use:

- Same methylation data (MsetEx subset)
- Same phenotype information (cancer vs normal)
- Same design matrix (model.matrix(~pheno$status))
- Comparable significance thresholds (p < 0.1)

Key findings:

1. **Performance:**
   - Runtime comparison between methods
   - Memory usage patterns
   
2. **Results:**
   - Number of DMRs detected by each method
   - DMR width distributions
   - Overlap between methods
   
3. **Methodological Differences:**
   - DMRSegal: DMP-based region expansion approach
   - DMRcate: Gaussian kernel smoothing approach  
   - bumphunter: Bootstrap-based bump finding
   
4. **Practical Considerations:**
   - DMRSegal's advantage: can work with pre-computed DMPs
   - Trade-offs between sensitivity and specificity
   - Use case recommendations based on data availability