---
title: "DMRsegal vs DMRcate on Simulated ONT BSseq Data"
author: "DMRsegal Package"
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
        code_folding: hide
vignette: >
  %\VignetteIndexEntry{DMRsegal vs DMRcate on Simulated ONT BSseq Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
    results = "asis",
    echo = TRUE,
    warning = FALSE,
    message = FALSE,
    fig.width = 12,
    fig.height = 8,
    dpi = 300
)
```

## Introduction

This vignette benchmarks DMR detection on simulated ONT-style methylation data with known ground truth.

Methods:

- **DMRsegal** (seed-guided, correlation-based DMR assembly)
- **DMRcate** (smoothing-based DMR detection)
- **dmrseq** (BSseq-native region-level inference; optional comparator)

Data:

- BSseq object: `data/sim_bs.rds`
- Ground-truth DMRs: `data/sim_dmrs.tsv`

Primary goal:

- assess DMRsegal performance against ground truth
- compare with DMRcate using the same dataset

## Setup and Dependencies

```{r}
devtools::load_all()

library(bsseq)
library(data.table)
library(ggplot2)
library(dplyr)
library(tidyr)
library(GenomicRanges)
library(progressr)
library(SummarizedExperiment)
library(S4Vectors)
library(limma)

progressr::handlers("cli")
benchmark_output_dir <- "benchmark_simulated_ont"
dir.create(benchmark_output_dir, showWarnings = FALSE, recursive = TRUE)
```

### Install External Dependencies

```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

if (!requireNamespace("DMRcate", quietly = TRUE)) {
    BiocManager::install("DMRcate", update = FALSE, ask = FALSE)
}

if (!requireNamespace("dmrseq", quietly = TRUE)) {
    BiocManager::install("dmrseq", update = FALSE, ask = FALSE)
}
```

## Load Simulated Data and Ground Truth

```{r}
resolve_input_path <- function(path_in_repo) {
    candidates <- c(
        path_in_repo,
        file.path("..", path_in_repo)
    )
    hit <- candidates[file.exists(candidates)][1]
    if (is.na(hit)) {
        return(path_in_repo)
    }
    hit
}

sim_bs_path <- resolve_input_path("data/sim_bs.rds")
sim_truth_path <- resolve_input_path("data/sim_dmrs.tsv")

if (!file.exists(sim_bs_path)) {
    stop("Missing BSseq input file: ", sim_bs_path)
}
if (!file.exists(sim_truth_path)) {
    stop("Missing ground-truth DMR file: ", sim_truth_path)
}

sim_bs <- readRDS(sim_bs_path)
stopifnot(inherits(sim_bs, "BSseq"))

truth_df <- data.table::fread(sim_truth_path)
truth_dmrs <- makeGRangesFromDataFrame(
    truth_df,
    seqnames.field = "seqnames",
    start.field = "start",
    end.field = "end",
    keep.extra.columns = TRUE
)

cat(sprintf("Loaded BSseq with %d CpGs and %d samples\n", nrow(sim_bs), ncol(sim_bs)))
cat(sprintf("Loaded %d ground-truth DMRs\n", length(truth_dmrs)))
```

## Build Phenotype Table from Sample Names

```{r}
sample_ids <- colnames(sim_bs)
sample_group <- ifelse(grepl("^Condition1", sample_ids), "Condition1", "Condition2")

pheno <- data.frame(
    Sample_ID = sample_ids,
    Sample_Group = sample_group,
    casecontrol = sample_group == "Condition2",
    stringsAsFactors = FALSE
)
rownames(pheno) <- pheno$Sample_ID

colData(sim_bs) <- S4Vectors::DataFrame(pheno, row.names = pheno$Sample_ID)

table(pheno$Sample_Group)
```

## Extract Methylation Matrix and Generate Seeds

```{r}

seeds.tsv <- file.path(benchmark_output_dir, "seeds.tsv")
if (!file.exists(seeds.tsv)) {
    gr <- granges(sim_bs)
    locs <- data.frame(
        chr = as.character(seqnames(gr)),
        start = start(gr),
        end = end(gr),
        stringsAsFactors = FALSE
    )
    cpg_ids <- paste0(locs$chr, ":", locs$start)
    rownames(locs) <- cpg_ids

    beta_mat <- bsseq::getMeth(sim_bs, type = "raw")
    rownames(beta_mat) <- cpg_ids
    colnames(beta_mat) <- sample_ids

    mvalues <- log2((beta_mat + 1e-6) / (1 - beta_mat + 1e-6))

    design <- model.matrix(~Sample_Group, data = pheno)
    colnames(design) <- c("(Intercept)", "Condition2_vs_Condition1")

    fit <- limma::eBayes(limma::lmFit(mvalues, design))
    pval_vec <- fit$p.value[, 2]
    qval_vec <- p.adjust(pval_vec, method = "BH")

    case_idx <- which(pheno$casecontrol)
    ctrl_idx <- which(!pheno$casecontrol)
    delta_beta <- rowMeans(beta_mat[, case_idx, drop = FALSE], na.rm = TRUE) -
        rowMeans(beta_mat[, ctrl_idx, drop = FALSE], na.rm = TRUE)

    sig_idx <- which(qval_vec < 0.05)
    if (length(sig_idx) < 10) {
        sig_idx <- order(pval_vec)[seq_len(min(50000, length(pval_vec)))]
        cat("Few FDR-significant seeds found; using top seeds by raw p-value\n")
    }

    seed_chr <- sub(":.*$", "", sig_seeds$cpg_id)
    seed_chr_tbl <- sort(table(seed_chr), decreasing = TRUE)
    live_log("Top seed chromosomes:")
    print(utils::head(seed_chr_tbl, 10))

    sig_seeds <- data.frame(
        cpg_id = cpg_ids[sig_idx],
        pval = pval_vec[sig_idx],
        qval = qval_vec[sig_idx],
        delta_beta = delta_beta[sig_idx],
        stringsAsFactors = FALSE
    )

    seed_locs <- locs[sig_seeds$cpg_id, , drop = FALSE]
    seed_locs <- seed_locs[order(seed_locs$chr, seed_locs$start), , drop = FALSE]
    seed_dist <- unlist(lapply(split(seed_locs$start, seed_locs$chr), function(x) {
        if (length(x) < 2) return(numeric(0))
        diff(x)
    }))
    if (length(seed_dist) > 0) {
        live_log("Seed distance quantiles (bp):")
        print(round(stats::quantile(seed_dist, probs = c(0, 0.25, 0.5, 0.75, 0.9, 0.99, 1)), 2))
    }

    write.table(sig_seeds, seeds.tsv, sep = "\t", quote = FALSE, row.names = FALSE)
} else {
    sig_seeds <- data.table::fread(seeds.tsv)
    rownames(sig_seeds) <- sig_seeds$cpg_id
}
cat(sprintf("Seeds for DMRsegal: %d\n", nrow(sig_seeds)))
```

## Method 1: DMRsegal

```{r, message=TRUE, warning=TRUE}
dmrsegal_file <- file.path(benchmark_output_dir, "dmrs.DMRsegal.rds")

live_log <- function(fmt, ...) {
    if (!is.character(fmt) || length(fmt) != 1) {
        msg <- paste(capture.output(print(fmt)), collapse = "\n")
    } else {
        msg <- sprintf(fmt, ...)
    }
    stamp <- format(Sys.time(), "%Y-%m-%d %H:%M:%S")
    line <- sprintf("[%s] %s", stamp, msg)
    cat(line, "\n", file = stderr())
}


if (!file.exists(dmrsegal_file)) {
    # Runtime diagnostics for large BSseq-derived matrices.
    live_log("DMRsegal diagnostics")
    live_log("CpGs in beta matrix: %s", format(nrow(sim_bs), big.mark = ","))
    live_log("Samples in beta matrix: %d", ncol(sim_bs))
    live_log("Seed CpGs: %s", format(nrow(sig_seeds), big.mark = ","))


    # Optional debugging knob to speed up test runs while profiling logs.
    # Example: options(DMRsegal.debug_max_seeds = 100000L)
    debug_max_seeds <- getOption("DMRsegal.debug_max_seeds", NA_integer_)
    if (!is.na(debug_max_seeds) && is.numeric(debug_max_seeds) && nrow(sig_seeds) > debug_max_seeds) {
        live_log("Applying debug seed cap: %s -> %s seeds",
            format(nrow(sig_seeds), big.mark = ","),
            format(as.integer(debug_max_seeds), big.mark = ",")
        )
        sig_seeds <- sig_seeds[order(sig_seeds$pval), , drop = FALSE]
        sig_seeds <- sig_seeds[seq_len(as.integer(debug_max_seeds)), , drop = FALSE]
    }

    # Optional hard timeout to avoid hanging on very large runs.
    # Example: options(DMRsegal.max_runtime_minutes = 30)
    max_runtime_minutes <- getOption("DMRsegal.max_runtime_minutes", NA_real_)
    if (!is.na(max_runtime_minutes) && is.numeric(max_runtime_minutes) && max_runtime_minutes > 0) {
        live_log("Applying runtime guard: %.1f minutes", max_runtime_minutes)
        setTimeLimit(elapsed = as.numeric(max_runtime_minutes) * 60, transient = TRUE)
        on.exit(setTimeLimit(cpu = Inf, elapsed = Inf, transient = FALSE), add = TRUE)
    }

    options("DMRsegal.verbose" = 4)
    options("DMRsegal.njobs" = 8)

    t_handler_start <- Sys.time()

    beta_handler <- DMRsegal::getBetaHandler(
        beta = sim_bs
    )
    t_handler_end <- Sys.time()
    live_log(
        "BetaHandler setup elapsed: %.1f sec",
        as.numeric(difftime(t_handler_end, t_handler_start, units = "secs"))
    )

    t_dmrs_start <- Sys.time()

    dmrs_segal <- tryCatch(
        {
            progressr::with_progress(
                DMRsegal::findDMRsFromSeeds(
                    beta = beta_handler,
                    seeds = sig_seeds,
                    pheno = pheno,
                    seeds_id_col = "cpg_id",
                    sample_group_col = "Sample_Group",
                    casecontrol_col = "casecontrol",
                    min_seeds = 2,
                    min_adj_seeds = 2,
                    min_cpgs = 3,
                    max_lookup_dist = 1000,
                    max_pval = 0.05,
                    pval_mode = "parametric",
                    chunk_size = 1e6, # Override default chunk size for large BSseq matrices.
                    group_concordance_strategy = "relaxed",
                    annotate_with_genes = FALSE,
                    rank_dmrs = TRUE
                )
            )
        },
        error = function(e) {
            if (grepl("time limit|reached elapsed time limit", e$message, ignore.case = TRUE)) {
                stop(
                    sprintf(
                        paste0(
                            "DMRsegal timed out after %.1f minutes. ",
                            "Increase options(DMRsegal.max_runtime_minutes = ...), ",
                            "reduce options(DMRsegal.debug_max_seeds = ...), ",
                            "or inspect seed diagnostics in this chunk."
                        ),
                        as.numeric(max_runtime_minutes)
                    ),
                    call. = FALSE
                )
            }
            stop(e)
        }
    )
    t_dmrs_end <- Sys.time()
    live_log(
        "findDMRsFromSeeds elapsed: %.1f sec",
        as.numeric(difftime(t_dmrs_end, t_dmrs_start, units = "secs"))
    )
    saveRDS(dmrs_segal, dmrsegal_file)
} else {
    live_log("Using cached DMRsegal output: %s", dmrsegal_file)
    dmrs_segal <- readRDS(dmrsegal_file)
}

cat(sprintf("DMRsegal: Found %d DMRs\n", length(dmrs_segal)))
```

## Method 2: DMRcate

```{r}
dmrcate_file <- file.path(benchmark_output_dir, "dmrs.DMRcate.rds")

run_dmrcate_bsseq <- function(bs_obj, design, genome = "hg38") {
    if (!requireNamespace("DMRcate", quietly = TRUE)) {
        warning("DMRcate not installed; returning empty set.")
        return(GRanges())
    }

    methdesign <- design[rep(seq_len(nrow(design)), each = 2), , drop = FALSE]

    for (fdr_cutoff in c(0.05, 0.1, 0.2)) {
        attempt <- tryCatch(
            {
                ann <- DMRcate::sequencing.annotate(
                    obj = bs_obj,
                    methdesign = methdesign,
                    coef = 2,
                    fdr = fdr_cutoff
                )
                dmrcate_res <- DMRcate::dmrcate(
                    ann,
                    lambda = 1000,
                    C = 2,
                    min.cpgs = 3
                )
                DMRcate::extractRanges(dmrcate_res, genome = genome)
            },
            error = function(e) {
                message(sprintf("DMRcate attempt failed (fdr=%.2f): %s", fdr_cutoff, e$message))
                NULL
            }
        )

        if (!is.null(attempt) && length(attempt) > 0) {
            return(attempt)
        }
    }

    GRanges()
}

if (!file.exists(dmrcate_file)) {
    dmrs_dmrcate <- run_dmrcate_bsseq(sim_bs, design, genome = "hg38")
    saveRDS(dmrs_dmrcate, dmrcate_file)
} else {
    dmrs_dmrcate <- readRDS(dmrcate_file)
}

cat(sprintf("DMRcate: Found %d DMRs\n", length(dmrs_dmrcate)))
```

## Method 3: dmrseq (Optional)

```{r}
dmrseq_file <- file.path(benchmark_output_dir, "dmrs.dmrseq.rds")

run_dmrseq_bsseq <- function(bs_obj) {
    if (!requireNamespace("dmrseq", quietly = TRUE)) {
        warning("dmrseq not installed; returning empty set.")
        return(GRanges())
    }

    if (!"Sample_Group" %in% colnames(as.data.frame(colData(bs_obj)))) {
        warning("Sample_Group not present in BSseq colData; returning empty set.")
        return(GRanges())
    }

    out <- tryCatch(
        {
            fit <- dmrseq::dmrseq(
                bs = bs_obj,
                testCovariate = "Sample_Group",
                cutoff = 0.1,
                minNumRegion = 3,
                maxPerms = 10,
                verbose = FALSE
            )
            fit
        },
        error = function(e) {
            message(sprintf("dmrseq failed: %s", e$message))
            GRanges()
        }
    )
    out
}

if (!file.exists(dmrseq_file)) {
    dmrs_dmrseq <- run_dmrseq_bsseq(sim_bs)
    saveRDS(dmrs_dmrseq, dmrseq_file)
} else {
    dmrs_dmrseq <- readRDS(dmrseq_file)
}

cat(sprintf("dmrseq: Found %d DMRs\n", length(dmrs_dmrseq)))
```

## Comparative Analysis

```{r}
methods_list <- list(
    DMRsegal = dmrs_segal,
    DMRcate = dmrs_dmrcate
)
if (exists("dmrs_dmrseq")) {
    methods_list$dmrseq <- dmrs_dmrseq
}

get_dmr_stats <- function(dmrs, method) {
    if (length(dmrs) == 0) {
        return(data.frame(
            Method = method,
            Number_of_DMRs = 0,
            Mean_DMR_Width = NA,
            Median_DMR_Width = NA,
            Total_Coverage_bp = 0,
            Min_Width = NA,
            Max_Width = NA
        ))
    }

    data.frame(
        Method = method,
        Number_of_DMRs = length(dmrs),
        Mean_DMR_Width = mean(width(dmrs)),
        Median_DMR_Width = median(width(dmrs)),
        Total_Coverage_bp = sum(width(dmrs)),
        Min_Width = min(width(dmrs)),
        Max_Width = max(width(dmrs))
    )
}

comparison_stats <- do.call(rbind, lapply(names(methods_list), function(m) {
    get_dmr_stats(methods_list[[m]], m)
}))

truth_stats <- get_dmr_stats(truth_dmrs, "Ground Truth")
comparison_stats <- rbind(comparison_stats, truth_stats)

knitr::kable(comparison_stats, digits = 2, caption = "Basic DMR Statistics")
```

## Ground-Truth Evaluation

```{r}
evaluate_against_truth <- function(pred, truth) {
    if (length(pred) == 0) {
        return(data.frame(
            TP_regions = 0,
            FP_regions = 0,
            FN_regions = length(truth),
            Precision_regions = 0,
            Recall_regions = 0,
            F1_regions = 0,
            Predicted_bp = 0,
            Truth_bp = sum(width(reduce(truth))),
            Intersect_bp = 0,
            Precision_bp = 0,
            Recall_bp = 0,
            F1_bp = 0,
            IoU_bp = 0
        ))
    }

    tp <- sum(overlapsAny(pred, truth))
    fp <- length(pred) - tp
    fn <- sum(!overlapsAny(truth, pred))

    precision <- ifelse(tp + fp == 0, 0, tp / (tp + fp))
    recall <- ifelse(tp + fn == 0, 0, tp / (tp + fn))
    f1 <- ifelse(precision + recall == 0, 0, 2 * precision * recall / (precision + recall))

    pred_red <- reduce(pred)
    truth_red <- reduce(truth)
    inter <- GenomicRanges::intersect(pred_red, truth_red)

    pred_bp <- sum(width(pred_red))
    truth_bp <- sum(width(truth_red))
    inter_bp <- ifelse(length(inter) == 0, 0, sum(width(inter)))
    union_bp <- pred_bp + truth_bp - inter_bp

    precision_bp <- ifelse(pred_bp == 0, 0, inter_bp / pred_bp)
    recall_bp <- ifelse(truth_bp == 0, 0, inter_bp / truth_bp)
    f1_bp <- ifelse(precision_bp + recall_bp == 0, 0, 2 * precision_bp * recall_bp / (precision_bp + recall_bp))
    iou_bp <- ifelse(union_bp == 0, 0, inter_bp / union_bp)

    data.frame(
        TP_regions = tp,
        FP_regions = fp,
        FN_regions = fn,
        Precision_regions = precision,
        Recall_regions = recall,
        F1_regions = f1,
        Predicted_bp = pred_bp,
        Truth_bp = truth_bp,
        Intersect_bp = inter_bp,
        Precision_bp = precision_bp,
        Recall_bp = recall_bp,
        F1_bp = f1_bp,
        IoU_bp = iou_bp
    )
}

truth_eval <- do.call(rbind, lapply(names(methods_list), function(method) {
    out <- evaluate_against_truth(methods_list[[method]], truth_dmrs)
    out$Method <- method
    out
}))

truth_eval <- truth_eval %>%
    dplyr::select(
        Method,
        TP_regions, FP_regions, FN_regions,
        Precision_regions, Recall_regions, F1_regions,
        Precision_bp, Recall_bp, F1_bp, IoU_bp
    )

knitr::kable(truth_eval, digits = 4, caption = "Performance Against Ground Truth")
```

### Overlap Matrices (Methods + Truth)

```{r}
all_sets <- c(methods_list, list(`Ground Truth` = truth_dmrs))

calculate_overlap <- function(gr1, gr2) {
    if (length(gr1) == 0 || length(gr2) == 0) {
        return(list(count = 0, percentage = 0))
    }
    overlap_count <- sum(overlapsAny(gr1, gr2))
    pct <- overlap_count / length(gr1) * 100
    list(count = overlap_count, percentage = pct)
}

overlap_count_mat <- matrix(0, nrow = length(all_sets), ncol = length(all_sets))
overlap_pct_mat <- matrix(0, nrow = length(all_sets), ncol = length(all_sets))
rownames(overlap_count_mat) <- colnames(overlap_count_mat) <- names(all_sets)
rownames(overlap_pct_mat) <- colnames(overlap_pct_mat) <- names(all_sets)

for (i in seq_along(all_sets)) {
    for (j in seq_along(all_sets)) {
        if (i == j) {
            overlap_count_mat[i, j] <- length(all_sets[[i]])
            overlap_pct_mat[i, j] <- 100
        } else {
            ov <- calculate_overlap(all_sets[[i]], all_sets[[j]])
            overlap_count_mat[i, j] <- ov$count
            overlap_pct_mat[i, j] <- ov$percentage
        }
    }
}

knitr::kable(overlap_count_mat, digits = 0, caption = "Overlap Counts (rows overlap columns)")
knitr::kable(overlap_pct_mat, digits = 2, caption = "Overlap Percentages (rows overlap columns)")
```

## Visualizations

### Ground-Truth Metrics

```{r, fig.width=10, fig.height=5}
plot_df <- truth_eval %>%
    dplyr::select(Method, F1_regions, F1_bp, IoU_bp) %>%
    tidyr::pivot_longer(
        cols = c(F1_regions, F1_bp, IoU_bp),
        names_to = "Metric",
        values_to = "Value"
    )

plot_df$Metric <- factor(
    plot_df$Metric,
    levels = c("F1_regions", "F1_bp", "IoU_bp"),
    labels = c("F1 (Region)", "F1 (Base-pair)", "IoU (Base-pair)")
)

ggplot(plot_df, aes(x = Method, y = Value, fill = Method)) +
    geom_col() +
    facet_wrap(~Metric, nrow = 1) +
    theme_minimal() +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none"
    ) +
    ylim(0, 1) +
    labs(
        title = "Method Performance vs Ground Truth",
        x = "",
        y = "Score"
    )
```

### Width Distributions

```{r, fig.width=10, fig.height=6}
width_data <- do.call(rbind, lapply(names(all_sets), function(method) {
    x <- all_sets[[method]]
    if (length(x) == 0) {
        return(NULL)
    }
    data.frame(Method = method, Width = width(x))
}))

ggplot(width_data, aes(x = Method, y = Width, fill = Method)) +
    geom_violin(alpha = 0.7) +
    geom_boxplot(width = 0.2, alpha = 0.5, outlier.size = 0.5) +
    scale_y_log10(labels = scales::comma) +
    theme_minimal() +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none"
    ) +
    labs(
        title = "DMR Width Distributions",
        x = "",
        y = "Width (bp, log scale)"
    )
```

## DMRsegal Ranking Quality vs Ground Truth

```{r}
if (length(dmrs_segal) > 0 && "rank" %in% names(mcols(dmrs_segal))) {
    dmrs_segal_ranked <- dmrs_segal[order(mcols(dmrs_segal)$rank, decreasing = FALSE)]

    n_bins <- min(10, max(2, floor(length(dmrs_segal_ranked) / 5)))
    bin_size <- ceiling(length(dmrs_segal_ranked) / n_bins)

    ranking_quality <- do.call(rbind, lapply(seq_len(n_bins), function(i) {
        s <- (i - 1) * bin_size + 1
        e <- min(i * bin_size, length(dmrs_segal_ranked))
        bin_dmrs <- dmrs_segal_ranked[s:e]

        tp_bin <- sum(overlapsAny(bin_dmrs, truth_dmrs))
        data.frame(
            Rank_Bin = i,
            DMRs_in_Bin = length(bin_dmrs),
            TP_in_Bin = tp_bin,
            Pct_Overlapping_Truth = 100 * tp_bin / max(1, length(bin_dmrs))
        )
    }))

    knitr::kable(ranking_quality, digits = 2, caption = "DMRsegal rank bins vs ground truth overlap")

    ggplot(ranking_quality, aes(x = Rank_Bin, y = Pct_Overlapping_Truth)) +
        geom_col(fill = "steelblue") +
        geom_line(group = 1, color = "firebrick", linewidth = 1) +
        geom_point(color = "firebrick", size = 2) +
        theme_minimal() +
        scale_x_continuous(breaks = ranking_quality$Rank_Bin) +
        labs(
            title = "DMRsegal Rank Quality",
            subtitle = "Higher-ranked bins should show stronger truth overlap",
            x = "Rank Bin (1 = best ranks)",
            y = "% DMRs overlapping truth"
        )
}
```

## Summary

```{r}
cat("\n## Key Findings\n\n")

if (nrow(truth_eval) > 0) {
    best_f1 <- truth_eval[which.max(truth_eval$F1_regions), ]
    best_iou <- truth_eval[which.max(truth_eval$IoU_bp), ]

    cat(sprintf(
        "- Best region-level F1: **%s** (%.3f)\n",
        best_f1$Method, best_f1$F1_regions
    ))
    cat(sprintf(
        "- Best base-pair IoU: **%s** (%.3f)\n",
        best_iou$Method, best_iou$IoU_bp
    ))

    if ("DMRsegal" %in% truth_eval$Method) {
        seg <- truth_eval[truth_eval$Method == "DMRsegal", ]
        cat(sprintf(
            "- **DMRsegal** region metrics: Precision = %.3f, Recall = %.3f, F1 = %.3f\n",
            seg$Precision_regions, seg$Recall_regions, seg$F1_regions
        ))
        cat(sprintf(
            "- **DMRsegal** base-pair metrics: Precision = %.3f, Recall = %.3f, F1 = %.3f, IoU = %.3f\n",
            seg$Precision_bp, seg$Recall_bp, seg$F1_bp, seg$IoU_bp
        ))
    }
}
```

## Session Info

```{r}
sessionInfo()
```
