# Benchmarking DMRSegal against other DMR detection tools
# 
# This notebook compares DMRSegal with other popular DMR detection tools using 
# real-world methylation data. We'll use a small subset of the colon cancer dataset
# from the minfiData package, which contains tumor vs normal samples.

## Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Load Required Libraries

```{r load_libraries}
library(minfi)
library(DMRcate)
library(bumphunter)
library(DMRSegal)
library(minfiData)  # Contains example 450k array data
library(microbenchmark)
library(ggplot2)
```

## Load and Prepare Data

We'll use a subset of the AML dataset from ExperimentHub to keep the example manageable.

```{r load_data}
# Load the example colon cancer dataset
data(MsetEx)
data(sheet)

# Select tumor and normal samples
sample_groups <- factor(sheet$status, levels = c("normal", "cancer"))
mset <- preprocessFunnorm(MsetEx)

# Subset to a smaller set for the example (3 cancer, 3 normal)
set.seed(123)
samples_per_group <- 3
cancer_idx <- sample(which(sheet$status == "cancer"), samples_per_group)
normal_idx <- sample(which(sheet$status == "normal"), samples_per_group)
selected_samples <- c(cancer_idx, normal_idx)

# Create the final subset
mset <- mset[, selected_samples]
pheno <- data.frame(
    status = sheet$status[selected_samples],
    row.names = colnames(mset)
)
```

## Get DMPs using minfi

```{r get_dmps}
# Create design matrix
design <- model.matrix(~pheno$status)
colnames(design) <- c("Intercept", "StatusCancer")

# Find DMPs using limma
dmps <- dmpFinder(getBeta(mset), 
                  pheno = pheno$status,
                  type = "categorical")

# Filter significant DMPs
sig_dmps <- dmps[dmps$pval < 0.05 & abs(dmps$deltaBeta) > 0.2, ]

# Create DMPs data frame for DMRSegal
dmrs_input <- data.frame(
    dmp = rownames(sig_dmps),
    chr = gsub("chr", "", seqnames(rowRanges(mset))[match(rownames(sig_dmps), rownames(mset))]),
    pos = start(rowRanges(mset))[match(rownames(sig_dmps), rownames(mset))],
    pval = sig_dmps$pval,
    pval_adj = p.adjust(sig_dmps$pval, method = "BH"),
    delta_beta = sig_dmps$deltaBeta,
    stringsAsFactors = FALSE
)
```

## Run DMR Detection Methods

### DMRSegal

```{r dmrsegal}
# The phenotype data is already created above in the data loading section
# But let's rename the column to match our expected format
pheno$group <- pheno$status
pheno$status <- NULL

# Write beta values to temp file
beta_file <- tempfile(fileext = ".txt")
beta_mat <- getBeta(mset)
write.table(
    cbind(ID = rownames(beta_mat), beta_mat),
    file = beta_file,
    sep = "\t",
    quote = FALSE,
    row.names = FALSE
)

# Write DMPs to temp file
dmps_file <- tempfile(fileext = ".txt")
write.table(
    dmrs_input,
    file = dmps_file,
    sep = "\t",
    quote = FALSE,
    row.names = FALSE
)

# Find DMRs using DMRSegal
dmrs_segal <- findDMRsFromDMPs(
    beta.file = beta_file,
    dmps.tsv.file = dmps_file,
    pheno = pheno,
    sample_group.col = "group",
    min.dmps = 2,
    min.cpgs = 3
)
```

### DMRcate

```{r dmrcate}
# Create model matrix for DMRcate
M <- getM(mset)
design <- model.matrix(~aml_samples$CellType)

# Run DMRcate
fit <- limma::lmFit(M, design)
fit <- limma::eBayes(fit)
dmrcate_results <- dmrcate(fit, C = as.matrix(design), pcutoff = 0.05)
dmrs_dmrcate <- extractRanges(dmrcate_results)
```

### bumphunter

```{r bumphunter}
# Run bumphunter
bumphunter_results <- bumphunter(
    mset,
    design = design,
    cutoff = 0.2,
    B = 1000,
    type = "Beta"
)
```

## Performance Comparison

```{r benchmark}
# Function wrappers for benchmarking
run_dmrsegal <- function() {
    findDMRsFromDMPs(
        beta.file = beta_file,
        dmps.tsv.file = dmps_file,
        pheno = pheno,
        sample_group.col = "group",
        min.dmps = 2,
        min.cpgs = 3
    )
}

run_dmrcate <- function() {
    fit <- limma::lmFit(M, design)
    fit <- limma::eBayes(fit)
    dmrcate_results <- dmrcate(fit, C = as.matrix(design), pcutoff = 0.05)
    extractRanges(dmrcate_results)
}

run_bumphunter <- function() {
    bumphunter(
        mset,
        design = design,
        cutoff = 0.2,
        B = 100,  # Reduced for benchmarking
        type = "Beta"
    )
}

# Run benchmark
bench <- microbenchmark(
    DMRSegal = run_dmrsegal(),
    DMRcate = run_dmrcate(),
    bumphunter = run_bumphunter(),
    times = 5
)

# Plot results
autoplot(bench) +
    ggtitle("DMR Detection Method Performance") +
    theme_minimal()
```

## Results Comparison

```{r compare_results}
# Function to get DMR stats
get_dmr_stats <- function(dmrs, method) {
    data.frame(
        Method = method,
        Number_of_DMRs = length(dmrs),
        Mean_DMR_Width = mean(width(dmrs)),
        Median_DMR_Width = median(width(dmrs))
    )
}

# Collect stats
stats_list <- list(
    get_dmr_stats(dmrs_segal, "DMRSegal"),
    get_dmr_stats(dmrs_dmrcate, "DMRcate"),
    get_dmr_stats(bumphunter_results$table, "bumphunter")
)

# Combine stats
results_comparison <- do.call(rbind, stats_list)
knitr::kable(results_comparison)
```

## DMR Overlap Analysis

```{r overlap}
# Convert all results to GRanges
gr_segal <- dmrs_segal
gr_dmrcate <- dmrs_dmrcate
gr_bumphunter <- GRanges(
    seqnames = bumphunter_results$table$chr,
    ranges = IRanges(
        start = bumphunter_results$table$start,
        end = bumphunter_results$table$end
    )
)

# Find overlaps
overlap_segal_dmrcate <- length(subsetByOverlaps(gr_segal, gr_dmrcate))
overlap_segal_bumphunter <- length(subsetByOverlaps(gr_segal, gr_bumphunter))
overlap_dmrcate_bumphunter <- length(subsetByOverlaps(gr_dmrcate, gr_bumphunter))

# Create overlap matrix
overlap_matrix <- matrix(
    c(
        length(gr_segal), overlap_segal_dmrcate, overlap_segal_bumphunter,
        overlap_segal_dmrcate, length(gr_dmrcate), overlap_dmrcate_bumphunter,
        overlap_segal_bumphunter, overlap_dmrcate_bumphunter, length(gr_bumphunter)
    ),
    nrow = 3,
    dimnames = list(
        c("DMRSegal", "DMRcate", "bumphunter"),
        c("DMRSegal", "DMRcate", "bumphunter")
    )
)

# Plot heatmap
pheatmap::pheatmap(
    overlap_matrix,
    main = "DMR Overlap Between Methods",
    display_numbers = TRUE
)
```

## Clean Up

```{r cleanup}
# Remove temporary files
unlink(c(beta_file, dmps_file))
```

## Conclusions

This benchmark analysis compares DMRSegal with two popular DMR detection methods (DMRcate and bumphunter) using a real-world AML dataset. Key findings:

1. Performance:
   - Runtime comparison between methods
   - Memory usage patterns
   
2. Results:
   - Number of DMRs detected by each method
   - DMR width distributions
   - Overlap between methods
   
3. Practical Considerations:
   - DMRSegal's DMP-based approach
   - Trade-offs between methods
   - Use case recommendations