---
title: "DMRsegal Configuration"
author: "DMRsegal Package"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{DMRsegal Configuration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
try(devtools::load_all(), silent = TRUE)
suppressPackageStartupMessages({
    try(library(DMRsegal), silent = TRUE)
})
knitr::opts_chunk$set(
    echo = TRUE,
    warning = FALSE,
    message = FALSE,
    eval = FALSE
)
```

# Overview

This vignette provides a comprehensive guide to the configuration available in DMRsegal. Understanding these parameters will help you optimize the package for your specific analysis needs.

# Function Parameters

## Core Input Parameters

### `beta`
**Type:** Character, matrix, BetaHandler object, or BED file  
**Required:** Yes  
**Description:** Input methylation data. Can be:

- Path to a beta value file (tab-separated). The beta file will be loaded into memory if its size is below `getOption("DMRsegal.beta_in_mem_threshold_mb")`Megabytes or genomically sorted and converted to tabix for faster access, if samtools tabix is installed.
- Path to a tabix-indexed file (.bed.gz with .tbi index)
- A beta matrix with CpG IDs as rownames and sample IDs as column names.
- A BetaHandler object (see `?BetaHandler`)
- A BED file with columns `bed_chrom_col` and `bed_start_col`, followed by sample columns, existing as row names in the provided pheno data

**Example:**
```{r}
beta <- loadExampleInputData("beta")
```

### `seeds`
**Type:** Character or data.frame  
**Required:** Yes  
**Description:** CpG sites to use as seeds for DMR detection. Can be:

- Path to a file with line separated CpG IDs
- A data.frame with DMP information

**Format requirements:**
- Row names or first column or the column given by `seeds_id_col` should contain CpG IDs
- The CpG IDs must match those in the beta data, either as Illumina IDs or genomic coordinates (chr:start). The latter is required when using BED files for beta.

**Example:**
```{r}
dmps <- loadExampleInputData("dmps")
```

### `pheno`
**Type:** Data.frame  
**Required:** Yes  
**Description:** Sample phenotype information.

**Format requirements:**
- Row names should match column names in beta data
- Must contain sample group information column (specified by `sample_group_col`)
- May contain case/control status column (specified by `casecontrol_col`)

**Example:**
```{r}
pheno <- loadExampleInputData("pheno")
```

## Sample Grouping Parameters

### `sample_group_col`
**Type:** Character  
**Default:** `"Sample_Group"`  
**Description:** Column name in `pheno` that specifies sample groups (e.g., "case" vs "control", "treated" vs "untreated"). More than two groups are supported.

**Example:**
```{r}
dmrs <- findDMRsFromSeeds(
    beta = beta,
    seeds = seeds,
    pheno = pheno,
    sample_group_col = "Sample_Group"
)
```

### `casecontrol_col`
**Type:** Character  
**Default:** `NULL`  
**Description:** Column name in `pheno` for case (TRUE/1) vs control (FALSE/0) status, for delta beta computations. If `NULL`, controls are assumed to be the first level found at `sample_group_col`.

**Example:**
```{r}
dmrs <- findDMRsFromSeeds(
    beta = beta,
    seeds = seeds,
    pheno = pheno,
    sample_group_col = "Sample_Group",
    casecontrol_col = "casecontrol"
)
```

### `ignored_sample_groups`
**Type:** Character vector  
**Default:** `NULL`  
**Description:** Sample groups to exclude while considering connection and expansion. Can also be "case" or "control".

**Example:**
```{r}
dmrs <- findDMRsFromSeeds(
    beta = beta,
    seeds = seeds,
    pheno = pheno,
    ignored_sample_groups = c("excluded_group1", "excluded_group2")
)
```

## Array and Genome Parameters

### `array`
**Type:** Character  
**Options:** `"450K"`, `"27K"`, `"EPIC"`, `"EPICv2"`  
**Default:** `"450K"`  
**Description:** Type of methylation array platform used. Ignored when using mouse genomes or when beta is provided as a BED file.

**Example:**
```{r}
dmrs <- findDMRsFromSeeds(
    beta = beta,
    seeds = seeds,
    pheno = pheno,
    array = "EPIC"
)
```

### `genome`
**Type:** Character
**Default:** `"hg38"`  
**Description:** Reference genome version.

**Example:**
```{r}
dmrs <- findDMRsFromSeeds(
    beta = beta,
    seeds = seeds,
    pheno = pheno,
    genome = "hg38"
)
```

## Filtering Parameters

### `min_cpg_delta_beta`
**Type:** Numeric  
**Default:** `0.1`  
**Range:** 0 to 1  
**Description:** Minimum absolute delta beta value for CpGs to be included in DMRs, during the **second stage** of DMR extension. Use this to filter out low-confidence DMRs.

**Recommendation:** Keep 0.1 for balanced precision/recall. Lower to 0.05 or 0 only when prioritizing sensitivity.

**Example:**
```{r}
dmrs <- findDMRsFromSeeds(
    beta = beta,
    seeds = seeds,
    pheno = pheno,
    min_cpg_delta_beta = 0.1
)
```

### `adaptive_min_cpg_delta_beta`
**Type:** Logical  
**Default:** `TRUE`  
**Description:** When enabled, DMRsegal adaptively increases `min_cpg_delta_beta` based on the seed-level delta-beta distribution (never below `min_cpg_delta_beta`).

**Example:**
```{r}
dmrs <- findDMRsFromSeeds(
    beta = beta,
    seeds = seeds,
    pheno = pheno,
    adaptive_min_cpg_delta_beta = TRUE
)
```

### `min_seeds`
**Type:** Integer  
**Default:** `1`  
**Description:** Minimum number of connected seeds required in a DMR.

**Recommendation:** Increase this value (e.g., to 3 or 4) for higher-confidence DMRs.

**Example:**
```{r}
dmrs <- findDMRsFromSeeds(
    beta = beta,
    seeds = seeds,
    pheno = pheno,
    min_seeds = 2
)
```

### `min_adj_seeds`
**Type:** Integer  
**Default:** `2`  
**Description:** Minimum number of seeds in a DMR after adjusting for CpG density. Minimum 2. This accounts for regions with different CpG coverage, and it is mostly applicable to array-based data. Only used when min_seeds < min_adj_seeds.

**Example:**
```{r}
dmrs <- findDMRsFromSeeds(
    beta = beta,
    seeds = seeds,
    pheno = pheno,
    min_adj_seeds = 2
)
```

### `min_cpgs`
**Type:** Integer  
**Default:** `50`  
**Description:** Minimum number of CpGs required in a DMR after region expansion. Minimum 2.

**Recommendation:** Lower this value (e.g., 3-5) for array-based data, keep higher (50+) for WGBS data.

**Example:**
```{r}
dmrs <- findDMRsFromSeeds(
    beta = beta,
    seeds = seeds,
    pheno = pheno,
    min_cpgs = 3
)
```

## Region Building Parameters

### `max_lookup_dist`
**Type:** Integer  
**Default:** `10000`  
**Unit:** Base pairs  
**Description:** Maximum genomic distance between adjacent seeds to be considered part of the same DMR.

**Recommendation:**
- 1,000-5,000 bp for tightly connected regions
- 10,000 bp (default) for moderate spacing
- 20,000+ bp for broader regions

**Example:**
```{r}
dmrs <- findDMRsFromSeeds(
    beta = beta,
    seeds = seeds,
    pheno = pheno,
    max_lookup_dist = 5000
)
```

### `expansion_window`
**Type:** Numeric  
**Default:** `1e6`  
**Description:** Stage 2 connectivity is computed only around seed-derived Stage 1 neighborhoods, using this total window width in base pairs.

Set to `<= 0` to compute connectivity genome-wide.

### `max_bridge_seeds_gaps`
**Type:** Integer  
**Default:** `1`  
**Description:** In Stage 1 seed connectivity, allows bridging up to this many consecutive p-value-driven failed edges when both flanking edges are connected.

### `max_bridge_extension_gaps`
**Type:** Integer
**Default:** `1`
**Description:** In Stage 2 DMR extension, allows bridging up to this many consecutive p-value-driven failed edges when both flanking edges are connected.

### `expansion_step`
**Type:** Integer  
**Default:** `500`  
**Unit:** Base pairs  
**Description:** Index-specific step size for expanding DMRs around seeds. Larger values increase memory usage but improve computation speed.

**Trade-off:**
- Smaller values (100-300): Lower memory, slower computation
- Larger values (500-1000): Higher memory, faster computation

**Example:**
```{r}
dmrs <- findDMRsFromSeeds(
    beta = beta,
    seeds = seeds,
    pheno = pheno,
    expansion_step = 1000
)
```

## Statistical Parameters

### `max_pval`
**Type:** Numeric  
**Default:** `0.05`  
**Range:** 0 to 1  
**Description:** Maximum p-value threshold for considering correlation between seeds as significant during the first stage of connectivity testing, and between proxial CpGs during the second stage DMR extension.

**Example:**
```{r}
dmrs <- findDMRsFromSeeds(
    beta = beta,
    seeds = seeds,
    pheno = pheno,
    max_pval = 0.01
)
```

### `entanglement`
**Type:** Character  
**Options:** `"strong"`, `"weak"`  
**Default:** `"strong"`  
**Description:** Strategy for determining connectivity between CpG sites across sample groups:

- `"strong"`: Requires all sample groups to show significant correlation for two sites to be considered connected. This is more conservative and ensures consistent methylation patterns across all groups.
- `"weak"`: Requires at least one sample group to show significant correlation. This is more permissive and may identify DMRs that are specific to certain groups.

**Recommendation:** Use `"strong"` (default) for most cases to ensure robust, reproducible DMRs. Use `"weak"` when you want to capture group-specific methylation patterns or when working with heterogeneous sample groups.

**Example:**
```{r}
dmrs <- findDMRsFromSeeds(
    beta = beta,
    seeds = seeds,
    pheno = pheno,
    entanglement = "weak"
)
```

### `pval_mode`
**Type:** Character  
**Options:** `"parametric"`, `"empirical"`, `"auto"`  
**Default:** `"auto"`  
**Description:** Method for calculating p-values during connectivity testing:

- `"parametric"`: Uses t-based correlation p-values (faster, assumes normal distribution)
- `"empirical"`: Uses permutation-based p-values (slower, no distribution assumptions)
- `"auto"`: Evaluates correlation test assumptions per sample group and chooses `"parametric"` only when diagnostics are acceptable; otherwise switches to `"empirical"`

**Recommendation:** Use `"auto"` when you want robust defaults across heterogeneous datasets. Use `"parametric"` when assumptions are known to hold and runtime is critical, and `"empirical"` when assumptions are clearly questionable.

**Example:**
```{r}
dmrs <- findDMRsFromSeeds(
    beta = beta,
    seeds = seeds,
    pheno = pheno,
    pval_mode = "empirical"
)
```

### `empirical_strategy`
**Type:** Character  
**Options:** `"auto"`, `"montecarlo"`, `"permutations"`  
**Default:** `"auto"`  
**Description:** Strategy for empirical p-value calculation (only applies when `pval_mode = "empirical"`):

- `"auto"`: Uses Monte Carlo for groups <6 samples, permutations for groups ≥6 samples
- `"montecarlo"`: Always uses Monte Carlo simulation
- `"permutations"`: Always uses exact permutations

**Example:**
```{r}
dmrs <- findDMRsFromSeeds(
    beta = beta,
    seeds = seeds,
    pheno = pheno,
    pval_mode = "empirical",
    empirical_strategy = "permutations"
)
```

### `ntries`
**Type:** Integer  
**Default:** `200`  
**Description:** Number of permutations/simulations when `pval_mode = "empirical"`.

**Recommendation:**
- 100-500: Faster, less precise
- 1,000-10,000: Slower, more precise

**Example:**
```{r}
dmrs <- findDMRsFromSeeds(
    beta = beta,
    seeds = seeds,
    pheno = pheno,
    pval_mode = "empirical",
    ntries = 1000
)
```

### `mid_p`
**Type:** Logical  
**Default:** `FALSE`  
**Description:** Whether to use mid-p correction in empirical p-value calculation.

**Example:**
```{r}
dmrs <- findDMRsFromSeeds(
    beta = beta,
    seeds = seeds,
    pheno = pheno,
    pval_mode = "empirical",
    mid_p = TRUE
)
```

### `aggfun`
**Type:** Character or function  
**Options:** `"median"`, `"mean"`, or a custom function  
**Default:** `"median"`  
**Description:** Aggregation function for calculating DMR-level statistics (delta beta, p-values).

**Recommendation:** Median is more robust to outliers; mean may be more sensitive.

**Example:**
```{r}
dmrs <- findDMRsFromSeeds(
    beta = beta,
    seeds = seeds,
    pheno = pheno,
    aggfun = "mean"
)
```

## Performance Parameters

### `njobs`
**Type:** Integer  
**Default:** `getOption("DMRsegal.njobs", min(8, future::availableCores() - 1))`  
**Description:** Number of parallel jobs to use for computation.

**Recommendation:**
- Use `-1` to automatically use all available cores minus 1
- Limit to avoid overwhelming system resources
- Consider memory requirements when increasing parallelization

**Example:**
```{r}
dmrs <- findDMRsFromSeeds(
    beta = beta,
    seeds = seeds,
    pheno = pheno,
    njobs = 4
)
```

### `chunk_size`
**Type:** Integer
**Default:** `getOption("DMRsegal.chunk_size", 1000)`
**Description:** Number of CpGs to process in each chunk. Adjusting this can help manage memory usage and computation time.

**Example:**
```{r}
dmrs <- findDMRsFromSeeds(
    beta = beta,
    seeds = seeds,
    pheno = pheno,
    chunk_size = 10000
)
```

### `verbose`
**Type:** Integer  
**Range:** 0 to 5  
**Default:** `getOption("DMRsegal.verbose", 1)`  
**Description:** Level of verbosity for logging messages:

- `0`: No messages
- `1`: Essential messages only
- `2`: Standard progress information
- `3`: Detailed progress information
- `4-5`: Very detailed debugging information

**Example:**
```{r}
dmrs <- findDMRsFromSeeds(
    beta = beta,
    seeds = seeds,
    pheno = pheno,
    verbose = 2
)
```

## Input/Output Parameters

### `seeds_id_col`
**Type:** Character or integer  
**Default:** `NULL`  
**Description:** Column name or index for seed identifiers in the seeds file. If `NULL`, uses row names if present, otherwise the first column.

**Example:**
```{r}
dmrs <- findDMRsFromSeeds(
    beta = beta,
    seeds = seeds_file,
    pheno = pheno,
    seeds_id_col = "CpG_ID"
)
```

### `output_prefix`
**Type:** Character  
**Default:** `NULL`  
**Description:** Prefix for output files. If provided, results will be saved to files with this prefix. If `NULL`, no files are saved.

**Example:**
```{r}
dmrs <- findDMRsFromSeeds(
    beta = beta,
    seeds = seeds,
    pheno = pheno,
    output_prefix = "my_analysis"
)
```

### `beta_row_names_file`
**Type:** Character  
**Default:** `NULL`  
**Description:** Path to a file containing row names for beta values. Useful for large beta files where reading row names separately is more efficient.

**Example:**
```{r}
dmrs <- findDMRsFromSeeds(
    beta = beta,
    pheno = pheno,
    seeds = seeds,
    beta_row_names_file = "beta_rownames.txt"
)
```

## BED File Parameters

### `bed_provided`
**Type:** Logical  
**Default:** `FALSE`  
**Description:** Whether the beta file is provided as a BED file. Automatically set to `TRUE` if the input file has a `.bed` extension.

**Example:**
```{r}
dmrs <- findDMRsFromSeeds(
    beta = "methylation.bed",
    seeds = seeds,
    pheno = pheno,
    bed_provided = TRUE
)
```

### `bed_chrom_col`
**Type:** Character  
**Default:** `"chrom"`  
**Description:** Column name for chromosome in BED files.

**Example:**
```{r}
dmrs <- findDMRsFromSeeds(
    beta = "methylation.bed",
    seeds = seeds,
    pheno = pheno,
    bed_chrom_col = "chr"
)
```

### `bed_start_col`
**Type:** Character  
**Default:** `"start"`  
**Description:** Column name for start position in BED files.

**Example:**
```{r}
dmrs <- findDMRsFromSeeds(
    beta = "methylation.bed",
    seeds = seeds,
    pheno = pheno,
    bed_start_col = "pos"
)
```

## Annotation Parameters

### `annotate_with_genes`
**Type:** Logical  
**Default:** `TRUE`  
**Description:** Whether to annotate DMRs with overlapping genes.

**Example:**
```{r}
dmrs <- findDMRsFromSeeds(
    beta = beta,
    seeds = seeds,
    pheno = pheno,
    annotate_with_genes = FALSE
)
```

### `.score_dmrs`
**Type:** Logical  
**Default:** `TRUE`  
**Description:** Whether to score DMRs based on their classification accuracy using cross-validated SVM. When enabled, each DMR is evaluated for its ability to discriminate between sample groups using stratified k-fold cross-validation with an RBF kernel SVM. The resulting `accuracy` value represents how well the DMR's methylation pattern can classify samples into their correct groups.

**Details:**
- Uses stratified k-fold cross-validation (default: 5-fold)
- Number of folds can be controlled with `options(DMRsegal.scoring_nfold = 5)`
- Random seed is controlled with `options(DMRsegal.random_seed = 42)`
- Accuracy values range from 0 to 1, where higher values indicate better discriminative power
- Requires the `e1071` package for SVM classification

**Example:**
```{r}
dmrs_scoring <- scoreDMRs(
    dmrs = dmrs,
    beta = beta,
    pheno = pheno,
    sample_group_col = "group"
)
```

## Advanced Parameters

### `.load_debug`
**Type:** Logical  
**Default:** `FALSE`  
**Description:** Enable debug mode for loading intermediate files, through short-circuiting. For internal development use only.

# Global Package Options

DMRsegal uses several global options that can be set using the `options()` function. These persist across function calls in your R session.

## Parallelism

### Option: `DMRsegal.njobs`
**Type:** Integer  
**Default:** `min(8, future::availableCores() - 1)`  
**Description:** Number of parallel jobs (defaults to the minimum of 8 and one less than the number of available CPU cores).

```{r}
options("DMRsegal.njobs" = 4)
```

## Verbosity

### Option: `DMRsegal.verbose`
**Type:** Integer  
**Default:** `1`  
**Description:** Default verbosity level.

```{r}
options("DMRsegal.verbose" = 2)
```

## Reproducibility

### Option: `DMRsegal.random_seed`
**Type:** Integer  
**Default:** `42`  
**Description:** Random seed for reproducibility.

```{r}
options("DMRsegal.random_seed" = 123)
```


## Memory Management

### Option: `DMRsegal.beta_in_mem_threshold_mb`
**Type:** Integer  
**Default:** `500`  
**Description:** Maximum size (in Megabytes) of beta files to load into memory. Files larger than this will be processed using disk-based methods.

```{r}
options("DMRsegal.beta_in_mem_threshold_mb" = 200)
```


### Option: `DMRsegal.chunk_size`
**Type:** Integer  
**Default:** `1000`  
**Description:** Number of CpGs to process in each chunk. Adjusting this can help manage memory usage and computation time.

```{r}
options("DMRsegal.chunk_size" = 10000)
```

## Caching

### Option: `DMRsegal.use_bed_cache`
**Type:** Logical  
**Default:** `FALSE`  
**Description:** Enable caching of BED file conversions. Useful for repeated analyses (e.g., with different parameters) of large datasets.

```{r}
options("DMRsegal.use_bed_cache" = TRUE)
```

### Option: `DMRsegal.bed_cache_dir`
**Type:** Character  
**Default:** `USER_CACHE_DIR/R/DMRsegal/tabix_cache`  
**Description:** Directory for BED file cache.

```{r}
options("DMRsegal.bed_cache_dir" = "/path/to/cache")
```

### Option: `DMRsegal.use_tabix_cache`
**Type:** Logical  
**Default:** `FALSE`  
**Description:** Enable caching of tabix-indexed files. Useful for repeated analyses (e.g., with different parameters) of large datasets.

```{r}
options("DMRsegal.use_tabix_cache" = TRUE)
```

### Option: `DMRsegal.tabix_cache_dir`
**Type:** Character  
**Default:** `USER_CACHE_DIR/R/DMRsegal/tabix_cache`  
**Description:** Directory for tabix file cache.

```{r}
options("DMRsegal.tabix_cache_dir" = "/path/to/cache")
```

### Option: `DMRsegal.use_annotation_cache`
**Type:** Logical  
**Default:** `TRUE`  
**Description:** Enable caching of gene annotations.

```{r}
options("DMRsegal.use_annotation_cache" = TRUE)
```

### Option: `DMRsegal.annotation_cache_dir`
**Type:** Character  
**Default:** `USER_CACHE_DIR/R/DMRsegal/annotation_cache`  
**Description:** Directory for annotation cache.

```{r}
options("DMRsegal.annotation_cache_dir" = "/path/to/cache")
```

### Option: `DMRsegal.jaspar_cache_dir`
**Type:** Character  
**Default:** `USER_CACHE_DIR/R/DMRsegal/jaspar_cache`  
**Description:** Directory for JASPAR motif database cache.

```{r}
options("DMRsegal.jaspar_cache_dir" = "/path/to/cache")
```

## Motif Analysis

### Option: `DMRsegal.jaspar_version`
**Type:** Integer  
**Default:** `2024`  
**Description:** JASPAR database version to use for motif analysis.

```{r}
options("DMRsegal.jaspar_version" = 2024)
```

### Option: `DMRsegal.jaspar_tax_group`
**Type:** Character  
**Default:** `"vertebrates"`  
**Description:** Taxonomic group for JASPAR motif filtering.

```{r}
options("DMRsegal.min_motif_similarity" = 0.75)
```

### Option: `DMRsegal.min_motif_similarity`
**Type:** Numeric  
**Default:** `0.8`  
**Description:** Minimum motif similarity threshold for DMR interaction analysis.

```{r}
options("DMRsegal.jaspar_tax_group" = "vertebrates")
```

### Option: `DMRsegal.jaspar_corr_threshold`
**Type:** Numeric  
**Default:** `0.9`  
**Description:** Correlation threshold for JASPAR motif similarity.

```{r}
options("DMRsegal.jaspar_corr_threshold" = 0.85)
```

### Option: `DMRsegal.make_debug_dir`
**Type:** Logical  
**Default:** `FALSE`  
**Description:** Create debug directory for troubleshooting.

```{r}
options("DMRsegal.make_debug_dir" = TRUE)
```

## DMR scoring

### Option: `DMRsegal.scoring_nfold`
**Type:** Integer  
**Default:** `5`  
**Description:** Number of folds for cross-validation when scoring DMRs.

```{r}
options("DMRsegal.scoring_nfold" = 3)
```

# Configuration Examples

## Example 1: High-Confidence DMRs with Strict Filtering

```{r}
dmrs <- findDMRsFromSeeds(
    beta = beta,
    seeds = seeds,
    pheno = pheno,
    sample_group_col = "Sample_Group",
    array = "EPIC",
    genome = "hg38",
    min_cpg_delta_beta = 0.1,
    min_seeds = 3,
    min_cpgs = 5,
    max_lookup_dist = 5000,
    max_pval = 0.01,
    njobs = 4
)
```

## Example 2: Broad Region Detection with Relaxed Parameters

```{r}
dmrs <- findDMRsFromSeeds(
    beta = beta,
    seeds = seeds,
    pheno = pheno,
    sample_group_col = "Sample_Group",
    min_seeds = 2,
    min_cpgs = 3,
    max_lookup_dist = 20000,
    max_pval = 0.05,
    njobs = 8
)
```

## Example 3: Empirical P-values for Small Sample Sizes

```{r}
dmrs <- findDMRsFromSeeds(
    beta = beta,
    seeds = seeds,
    pheno = pheno,
    sample_group_col = "Sample_Group",
    pval_mode = "empirical",
    empirical_strategy = "montecarlo",
    ntries = 5000,
    mid_p = TRUE,
    njobs = 4
)
```

## Example 4: Memory-Efficient Processing for Large Datasets

```{r}
options("DMRsegal.use_tabix_cache" = TRUE)

dmrs <- findDMRsFromSeeds(
    beta = "large_dataset.bed.gz",
    seeds = seeds,
    pheno = pheno,
    sample_group_col = "Sample_Group",
    expansion_step = 1000,
    njobs = 4
)
```

## Example 5: Custom Aggregation Function

```{r}
weighted_mean <- function(x) {
    weights <- abs(x - 0.5)
    sum(x * weights) / sum(weights)
}

dmrs <- findDMRsFromSeeds(
    beta = beta,
    seeds = seeds,
    pheno = pheno,
    sample_group_col = "Sample_Group",
    aggfun = weighted_mean
)
```

# Best Practices

1. **Start with default parameters** and adjust based on your specific needs.

2. **For array data** (450K, EPIC), use lower `min_cpgs` values (3-5) since CpG coverage is sparse.

3. **For WGBS data**, keep `min_cpgs` higher (50+) to ensure robust regions.

4. **Avoid heavy pre-filtering** of seeds based on effect size. Let DMRsegal handle filtering internally.

5. **Use empirical p-values** for small sample sizes (<10 per group) or when normality assumptions are questionable.

6. **Enable caching** for large datasets or when running multiple analyses.

8. **Use parallel processing** (`njobs > 1`) for faster computation, but be mindful of memory requirements.

9. **Save intermediate results** using `output_prefix` for large analyses.

10. **Document your configuration** by saving parameter settings for reproducibility.

# Troubleshooting

## Issue: Out of Memory Errors

**Solution:**
- Decrease `njobs`
- Decrease `getOption("DMRsegal.beta_in_mem_threshold_mb")` (default 500) to enable disk-based processing
- Use tabix-indexed files for very large datasets
- Enable caching options

## Issue: DMRs Too Small

**Solution:**
- Increase `max_lookup_dist`
- Increase `max_pval`
- Decrease `min_cpg_delta_beta`

## Issue: Too Many DMRs

**Solution:**
- Increase `min_seeds`
- Increase `min_cpgs`
- Increase `max_lookup_dist`
- Increase `min_cpg_delta_beta`
- Decrease `max_pval`

## Issue: Slow Performance

**Solution:**
- Increase `chunk_size` for larger processing chunks. Setting this to a larger value (e.g., 10e6) can reduce overhead and speed up processing, especially for genome-wide datasets.
- Increase `njobs` for parallel processing
- Increase `expansion_step`
- Use `pval_mode = "parametric"` instead of `"empirical"`
- Enable caching options
- Consider using tabix-indexed files

# Session Info

```{r sessionInfo}
sessionInfo()
```
