---
title: "DMRsegal End-to-End Example"
author: "DMRsegal Package"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    code_folding: hide
vignette: >
  %\VignetteIndexEntry{DMRsegal End-to-End Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    warning = FALSE,
    message = FALSE,
    fig.width = 12,
    fig.height = 8,
    dpi = 300,
    fig.retina = 2
)
```

```{r html_styles, echo=FALSE, results='asis'}
cat("
<style>
.main-container {
  max-width: 98vw !important;
  width: 98vw !important;
}
.container-fluid {
  max-width: 98vw !important;
}
</style>
")
```

## Goal

This notebook showcases all DMRsegal capabilities, including DMR scoring, motif extraction, motif-based DMR interactions, and visualizations. 

## Setup

```{r setup}
try(devtools::load_all(), silent = TRUE)
try(library(DMRsegal), silent = TRUE)
library(GenomicRanges)
library(dplyr)
library(ggplot2)

genome <- "hg19"
```

## Load Example Inputs and Benchmark DMRs

```{r load_inputs}
# loadExampleInputData is a helper function to load example data included in the package.
beta <- loadExampleInputData("beta")
pheno <- loadExampleInputData("pheno")
array_type <- loadExampleInputData("array_type")
# DMPs are produced from minfi, and have the form of a dataframe
dmps <- loadExampleInputData("dmps")
cat(sprintf("Beta matrix dimensions: %d CpGs x %d samples\n", nrow(beta), ncol(beta)))
cat(sprintf("Phenotype data dimensions: %d samples x %d columns\n", nrow(pheno), ncol(pheno)))
cat(sprintf("Array type: %s\n", array_type))
cat(sprintf("Number of DMPs (seeds): %d\n", nrow(dmps)))
cat(sprintf("Head of phenotype data:\n"))
print(head(pheno))
cat(sprintf("Head of DMPs:\n"))
print(head(dmps))


# The beta handler is used to retrieve beta values for the DMRs and their seeds, and to perform the SVM-based scoring of the DMRs.
# It is better to initialize it once and reuse it, as it performs some pre-processing steps that are common across the DMRs.
# The beta handler can be supplied as the `beta` argument for the DMR calling and scoring functions, to avoid redundant initialization.
beta_handler <- getBetaHandler(beta, array = array_type, genome = genome)

# For the sake of the example, we look only at chromosomes 16-20.
dmps_locs <- beta_handler$getGenomicLocs()[rownames(dmps), ]
dmps <- dmps[dmps_locs[, "chr"]  %in% paste0("chr", 16:20), ]
cat(sprintf("Number of DMPs (seeds) on chr16-20: %d\n", nrow(dmps)))

```

## DMR Calling from Seeds
The `findDMRsFromSeeds` function is used to call DMRs from the provided seeds, which in this case are the DMPs. The function performs a correlation-based expansion of the regions around the seeds, considering both statistical significance and biological relevance of methylation changes. The resulting DMRs are stored in a GRanges object, which includes metadata columns for the number of seeds and CpGs in each DMR, as well as the delta-beta between the sample groups. The number of DMRs found is reported.
```{r find_dmrs}
pheno[, "case_control"] <- pheno$Sample_Group == "cancer"
# Find DMRs using parallel processing with 4 cores, producing a GRanges object with the called DMRs and their metadata.
dmrs <- findDMRsFromSeeds(
    beta = beta_handler, # The beta can be an array-like object, tabix file path, or a BED file path. Here we use the beta handler, which is a wrapper around the beta matrix that allows for efficient retrieval of beta values for the DMRs and their seeds.
    seeds = dmps, # The seeds are the DMPs, which are pre-computed and provided as example data.
    seeds_id_col = NULL, # The column in the seeds data that contains the unique identifier for each seed, which will be used to link the seeds to the DMRs. If not provided, like here, the rownames of the seeds data will be used as identifiers. In case of a NGS experiment (of a BED, tabix or bsseq input), the names should be in the format "chr:position" to allow for correct linking to the genomic coordinates.
    pheno = pheno, # The phenotype data, which contains the sample information and covariates. It should have a column with the sample group labels, which will be used for the SVM-based scoring of the DMRs. It can be provided as a TSV file.
    min_cpgs = 3, # The minimum number of CpGs that a DMR should have to be considered valid. This is used to filter out small regions that may not be biologically relevant.
    max_bridge_seeds_gaps = 1, # The maximum number of non-significant seeds that can be bridged when expanding the DMRs. This allows for some flexibility in the region expansion, while still maintaining a focus on significant seeds.
    max_bridge_extension_gaps = 1, # The maximum number of non-significant CpGs that can be bridged when extending the DMRs beyond the seeds. This allows for some flexibility in the region extension, while still maintaining a focus on significant CpGs.
    casecontrol_col = "case_control", # The column in the phenotype data that contains the binary case-control labels for the samples, which will be used for the SVM-based scoring of the DMRs. It should be a logical or factor column with two levels, where one level is considered as "case" and the other as "control". If not provided, the first level of the sample group column will be considered as "control" and the second level as "case".
    sample_group_col = "Sample_Group", # The column in the phenotype data that contains the sample group labels for the SVM-based scoring.
    covariates = c("Age", "Gender"), # Optional covariates to include in the model, if available in the phenotype data.
    genome = genome, # The genome assembly to use for the genomic coordinates of the DMRs. It will be guessed if not provided, based on the array type or the input data. It is used for the genomic operations and annotations of the DMRs.
    array = array_type, # The array platform to use, available are 450K, EPIC, EPICv2.
    output_prefix = "tmp/DMRsegal_example" # The prefix for the output files. It is required to provide this if you want to use the DMRsegalViewer Shiny app, as it will look for the output files with this prefix. The output files include the DMRs in a GRanges object saved as an RDS file, and a summary table of the DMRs saved as a TSV file.  
)
cat(sprintf("DMRs found: %d\n", length(dmrs)))
cat(sprintf("DMRs metadata columns: %s\n", paste(colnames(mcols(dmrs)), collapse = ", ")))
cat(sprintf("DMRs genomic range example: %s\n", paste0(seqnames(dmrs)[1], ":", start(dmrs)[1], "-", end(dmrs)[1])))
```

## DMR Summary
The scoring of DMRs is based on a margin-sensitive score derived from the SVM CV prediction strategy, which reflects the predictive power of the DMR in distinguishing between sample groups. The CV accuracy is also included as a complementary metric. In addition, score trajectories are smoothed per chromosome and segmented, enabling assignment of localized block IDs that capture saddle-like peaks in Manhattan plots. The summary table includes the genomic coordinates, number of seeds and CpGs, delta-beta, score, CV accuracy, score, and block ID of each DMR.
```{r}
summary_tbl <- data.frame(
    chr = as.character(seqnames(dmrs)),
    start = start(dmrs),
    end = end(dmrs),
    width = width(dmrs),
    seeds_num = mcols(dmrs)$seeds_num,
    cpgs_num = mcols(dmrs)$cpgs_num,
    delta_beta = round(mcols(dmrs)$delta_beta, 4),
    score = round(mcols(dmrs)$score, 4),
    cv_accuracy = round(mcols(dmrs)$cv_accuracy, 4),
    block_id = mcols(dmrs)$block_id
)
summary_tbl <- summary_tbl[order(summary_tbl$score, decreasing = TRUE), ]
knitr::kable(head(summary_tbl, 5), caption = "Top 5 scoring DMRs (margin-sensitive score + raw CV accuracy)")
```

## Motif Discovery
Based on the seeds of the DMRs, motifs are extracted using the `extractDMRMotifs` function (called by default when `findDMRsFromSeeds` is used). The function retrieves the sequences around the seeds, applies a flank size of 5 bp, and identifies consensus sequences. The number of DMRs with extracted motifs and the number of unique motif consensus sequences are reported.
```{r}
n_motif_consensus <- sum(!is.na(mcols(dmrs)$consensus_seq))
n_unique_motif_consensus <- length(unique(stats::na.omit(mcols(dmrs)$consensus_seq)))
cat(sprintf("Motifs extracted for %d/%d DMRs.\n", n_motif_consensus, length(dmrs)))
cat(sprintf("Unique motif consensus sequences: %d\n", n_unique_motif_consensus))
```

## Motif-Based DMR Interactions
Once motifs are extracted, the `computeDMRsInteraction` function can be used to identify "interactions" between DMRs based on motif similarity. 
The rationale is that DMRs with similar motifs may be co-regulated or functionally related, even if they are not in close genomic proximity.
The function computes pairwise similarities between the PWMs of the extracted motifs,
excluding the center CpG positions, corrects for microarray design biases,
applies a minimum similarity threshold of 0.8, and identifies connected components of interacting DMRs. 
The resulting components are annotated with JASPAR motifs if available.
The number of interactions, components, and the size of the largest component are reported. Additionally, the number of components with JASPAR annotation is provided.
```{r}
interaction_results <- computeDMRsInteraction(
    dmrs = dmrs,
    genome = genome,
    array = array_type,
    min_similarity = 0.8,
    motif_cpg_flank_size = 5,
    min_component_size = 2,
    query_components_with_jaspar = TRUE
)

interaction_tbl <- interaction_results$interactions
component_tbl <- interaction_results$components
dmrs <- interaction_results$dmrs # The input DMRs are returned with additional metadata columns for the interactions and components.

cat(sprintf("Interactions found: %d\n", nrow(interaction_tbl)))
cat(sprintf("Components found: %d\n", nrow(component_tbl)))
if (nrow(component_tbl) > 0) {
    largest_component <- max(component_tbl$size, na.rm = TRUE)
    cat(
        sprintf(
            "Largest component size: %d / %d DMRs (%.2f%%)\n",
            largest_component, length(dmrs), 100 * largest_component / length(dmrs)
        )
    )
}
annotated_components_m <- !is.na(component_tbl$jaspar_names) & nzchar(component_tbl$jaspar_names)
annotated_components <- sum(annotated_components_m)
cat(sprintf("Components with JASPAR annotation: %d / %d\n", annotated_components, nrow(component_tbl)))

if (nrow(interaction_tbl) > 5) {
    knitr::kable(
        head(interaction_tbl[order(-interaction_tbl$sim), ], 5),
        caption = "Top 5 motif-similarity DMR interactions"
    )
}

if (annotated_components > 0) {
    components_tbl <- component_tbl[annotated_components_m, ] |>
        dplyr::select(component_id, size, consensus_seq, jaspar_names) |>
        dplyr::arrange(dplyr::desc(size))
    if (nrow(components_tbl) > 0) {
        knitr::kable(
            components_tbl,
            caption = "Motif interaction components with JASPAR annotation"
        )
    }
}
```

## Visualizations

### DMR Score Manhattan Plot

A manhattan plot of the DMRs is drawn, showing their scores, produced by the SVM CV prediction strategy,
their annotation as promoter or gene body DMRs, and translucent block squares for localized score peaks identified by the scoring block extension. The DMRs that belong to both promoter and gene body are omitted.
```{r manhattan_plot, fig.height=6, fig.width=14, dpi=150}
plotDMRsManhattan(
    dmrs = dmrs,
    promoter_col = "in_promoter_of",
    gene_body_col = "in_gene_body_of"
)
```

### Single DMR Plot
Each DMR can be plotted with the `plotDMR` function, which shows the beta values of the CpGs in the DMR, the seeds, and the motif-based interactions with other DMRs, 
along with JASPAR annotations if existing.
Here we select a DMR that has more than one merged DMR and has an extension beyond the seed region, in both sides.

```{r single_dmr_plot, fig.height=10}
merged_counts <- suppressWarnings(as.numeric(as.character(mcols(dmrs)$merged_dmrs_num)))
has_extension <- (mcols(dmrs)$start_seed != mcols(dmrs)$start_cpg) &
    (mcols(dmrs)$end_seed != mcols(dmrs)$end_cpg)

complex_candidates <- which(is.finite(merged_counts) & merged_counts > 1 & has_extension)
if (length(complex_candidates) == 0) {
    complex_candidates <- which(has_extension)
}
if (length(complex_candidates) == 0) {
    complex_candidates <- 1L
}

selected_dmr_idx <- complex_candidates[1]
cat(
    sprintf(
        "Selected DMR index %d (score=%s, seeds_num=%s, cpgs_num=%s)\n",
        selected_dmr_idx,
        as.character(mcols(dmrs)$score[selected_dmr_idx]),
        as.character(mcols(dmrs)$seeds_num[selected_dmr_idx]),
        as.character(mcols(dmrs)$cpgs_num[selected_dmr_idx])
    )
)

plotDMR(
    dmrs = dmrs,
    dmr_index = selected_dmr_idx,
    beta = beta_handler,
    pheno = pheno,
    sample_group_col = "Sample_Group",
    genome = genome,
    array = array_type,
    motif_cpg_flank_size = 5
)

```

### Summary Circos Plot
A circos plot consisting of the extracted information from DMRsegal is drawn along the genome or a specified region, here from chromosome 16 to 22. 
The information includes:

- The DMRs and their scores (at most 40 DMRs per chromosome, scoring by score, are shown)
- Selected beta values of the DMRs (at most 5 CpGs per DMR, scoring by absolute delta-beta, are shown)
- The motif-based interactions between DMRs, with links direction going from higher scoring to lower scoring (at most 120 interactions groups are shown)
- The JASPAR annotation of the motif-based groups, if available, on the legend.
```{r circos_plot, eval=TRUE, fig.height=10, fig.width=14, dpi=300, fig.retina=1.5}
circos_dmrs <- dmrs
circos_summary_plotted <- FALSE

tryCatch(
    {
        plotDMRsCircos(
            dmrs = circos_dmrs,
            beta = beta_handler,
            pheno = pheno,
            genome = genome,
            array = array_type,
            sample_group_col = "Sample_Group",
            min_similarity = 0.8,
            motif_cpg_flank_size = 5,
            max_num_samples_per_group = 5,
            max_dmrs_per_chr = 40,
            max_cpgs_per_dmr = 5,
            max_components = 120,
            chromosomes = paste0("chr", 16:22),
            degenerate_resolution = 1e6,
            verbose = 1
        )
        circos_summary_plotted <- TRUE
    },
    error = function(e) {
        message("Circos plot generation failed: ", conditionMessage(e))
        NULL
    }
)
if (!circos_summary_plotted) {
    plot.new()
    text(0.5, 0.5, "Circos summary plot unavailable for this dataset.", cex = 1.1)
}
```
It is also possible to plot the circos for one or multiple regions to assess connectivity, e.g, for chr19:14700kb-16100kb and for chr20:61600kb-63000kb, which are regions that in the earlier plot were shown to interact:
```{r circos_plot_region, eval=TRUE, fig.height=10, fig.width=14, dpi=300, fig.retina=1.5}
circos_region_plotted <- FALSE

tryCatch(
    {
        plotDMRsCircos(
            dmrs = circos_dmrs,
            beta = beta_handler,
            pheno = pheno,
            genome = genome,
            array = array_type,
            sample_group_col = "Sample_Group",
            min_similarity = 0.8,
            motif_cpg_flank_size = 5,
            max_num_samples_per_group = 5,
            max_dmrs_per_chr = 40,
            max_cpgs_per_dmr = 5,
            max_components = 120,
            region = "chr19:14700kb-16100kb, chr20:61600kb-63000kb",
            degenerate_resolution = 10e2, # If the width of the DMR is larger than that, the DMRs is shown as rectangles and ribbons are used instead of links to connect the DMRs.
            verbose = 1
        )
        circos_region_plotted <- TRUE
    },
    error = function(e) {
        message("Circos plot generation failed: ", conditionMessage(e))
        NULL
    }
)
if (!circos_region_plotted) {
    plot.new()
    text(0.5, 0.5, "Circos region plot unavailable for selected regions.", cex = 1.1)
}
```

### DMRsegal interactive Shiny App
The DMRsegal interactive Shiny app can be launched with the `launchDMRsegalViewer` function, 
which allows for interactive exploration of the DMRs and the aforementioned plots. In order to use it, the `output_prefix` argument should be provided in the `findDMRsFromSeeds` function, as the app looks for the output files with this prefix.
```{r launch_shiny_app, eval=FALSE}
launchDMRsegalViewer("tmp/DMRsegal_example")
```

## Session Info

```{r session_info}
sessionInfo()
```
