---
title: "Finding Differentially Methylated Regions with DMRSegal"
author: 
  - name: Vasileios Lemonidis
    affiliation: University of Antwerp
    email: vasileios.lemonidis@uantwerpen.be
package: DMRSegal
output: 
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Finding Differentially Methylated Regions with DMRSegal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    warning = FALSE,
    message = FALSE
)
```

# Introduction

The `DMRSegal` package provides methods for identifying Differentially Methylated 
Regions (DMRs) from pre-computed Differentially Methylated Positions (DMPs). The 
package implements a correlation-based approach to expand significant DMPs into 
regions, considering both the statistical significance and the biological 
relevance of methylation changes.

# Installation

You can install the released version of DMRSegal from Bioconductor with:

```{r eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("DMRSegal")
```

# Quick Start

Let's start with a simple example using the included example data:

```{r}
library(DMRSegal)
library(GenomicRanges)

# Load example data
data(example_beta)
data(example_dmps)
data(example_pheno)

# Look at the data structure
dim(example_beta)
head(example_dmps)
table(example_pheno$group)
```

The example data includes:
- A beta value matrix with 1000 CpGs and 10 samples
- Pre-computed DMPs with positions and p-values
- Sample phenotype information

# Basic Usage

## Using a Beta Matrix

First, let's save our beta values to a file (normally you would already have this):

```{r}
beta_file <- tempfile(fileext = ".txt")
write.table(
    cbind(ID = rownames(example_beta), example_beta),
    file = beta_file,
    sep = "\t",
    quote = FALSE,
    row.names = FALSE
)

# Save DMPs to a file
dmps_file <- tempfile(fileext = ".txt")
write.table(
    example_dmps,
    file = dmps_file,
    sep = "\t",
    quote = FALSE,
    row.names = TRUE
)
```

Now we can find DMRs:

```{r}
dmrs <- findDMRsFromDMPs(
    beta.file = beta_file,
    dmps.tsv.file = dmps_file,
    pheno = example_pheno,
    sample_group.col = "group",
    min.cpg.delta_beta = 0.1,
    expansion.step = 5,
    max.pval = 0.05,
    max.lookup.dist = 1000,
    min.dmps = 2,
    min.cpgs = 3
)

# Look at the results
dmrs

# Understanding DMR Output

Each DMR returned by `findDMRsFromDMPs` includes comprehensive metadata:

- `start_cpg`, `end_cpg`: First and last CpG IDs in the region
- `start`, `end`: Genomic coordinates of the region
- `n_cpgs`: Total number of CpGs in the region
- `mean_delta_beta`: Average methylation difference between groups
- `region_correlation`: Average correlation between CpGs in the region
- `min_correlation`: Minimum correlation between any pair of CpGs
- `max_pval_in_region`: Maximum p-value of correlation tests
- Expansion metadata:
  - `downstream_cpg_expansion_stop_reason`
  - `upstream_cpg_expansion_stop_reason`

```{r}
# Extract metadata as a data frame
dmr_info <- mcols(dmrs)
head(dmr_info)

# Examine correlation statistics
summary(dmr_info$region_correlation)
summary(dmr_info$min_correlation)

# Look at effect sizes
hist(abs(dmr_info$mean_delta_beta), 
     main="Distribution of Mean Delta Beta Values",
     xlab="Absolute Mean Delta Beta")
```

## Using a Tabix File

For large datasets, using a tabix-indexed file can be more efficient. Here's how
to create and use one:

```{r, eval=FALSE}
library(Rsamtools)

# Create sorted bed file (normally you would already have this)
bed_file <- tempfile(fileext = ".bed")
bed_data <- data.frame(
    chr = example_dmps$chr,
    start = example_dmps$pos - 1,  # BED format is 0-based
    end = example_dmps$pos,
    beta_values = t(example_beta)
)
write.table(bed_data, bed_file, sep="\t", quote=FALSE, row.names=FALSE)

# Sort and compress
sorted_bed <- tempfile(fileext = ".bed.gz")
system(paste("sort -k1,1 -k2,2n", bed_file, "| bgzip >", sorted_bed))

# Index with tabix
system(paste("tabix -p bed", sorted_bed))

# Find DMRs using tabix file
dmrs_tabix <- findDMRsFromDMPs(
    tabix.file = sorted_bed,
    dmps.tsv.file = dmps_file,
    pheno = example_pheno,
    sample_group.col = "group"
)
```

# Parameter Tuning

The DMR finding process can be tuned with several parameters:

## Parameter Selection

The DMR finding process is controlled by several key parameters that affect both the
statistical rigor and biological relevance of the results:

### Correlation and Effect Size Parameters

* `max.pval` (default: 0.05): Maximum p-value for correlation tests between adjacent CpGs
  - Controls statistical significance of methylation pattern similarity
  - Lower values (e.g., 0.01) create more stringent regions
  - Higher values allow more flexibility but may include noise
  
* `min.cpg.delta_beta` (default: 0): Minimum absolute difference in methylation
  - Ensures biological relevance of methylation changes
  - Typical values: 0.1-0.3 (10-30% methylation difference)
  - Set to 0 to focus only on correlation structure

```{r}
# Compare different correlation stringency
dmrs_strict <- findDMRsFromDMPs(
    beta.file = beta_file,
    dmps.tsv.file = dmps_file,
    pheno = example_pheno,
    sample_group.col = "group",
    max.pval = 0.01,           # More stringent correlation
    min.cpg.delta_beta = 0.2   # Require stronger effect
)

dmrs_loose <- findDMRsFromDMPs(
    beta.file = beta_file,
    dmps.tsv.file = dmps_file,
    pheno = example_pheno,
    sample_group.col = "group",
    max.pval = 0.1,            # More permissive correlation
    min.cpg.delta_beta = 0.05  # Allow smaller effects
)

# Compare results
data.frame(
    Setting = c("Default", "Strict", "Loose"),
    DMRs = c(length(dmrs), length(dmrs_strict), length(dmrs_loose)),
    Mean_Correlation = c(
        mean(mcols(dmrs)$region_correlation),
        mean(mcols(dmrs_strict)$region_correlation),
        mean(mcols(dmrs_loose)$region_correlation)
    )
)
```

### Region Building Parameters

* `expansion.step` (default: 500): Number of CpGs to examine in each expansion step
  - Larger values are more efficient but may miss small regions
  - Smaller values (5-50) provide finer control but are slower
  - Does not affect final results, only computation speed

* `max.lookup.dist` (default: 1000): Maximum base pairs between DMPs
  - Controls how far to look for nearby DMPs when building regions
  - Larger values allow merging more distant DMPs
  - Should reflect expected DMR sizes in your data

* `min.dmps`, `min.cpgs`: Minimum DMPs and CpGs per region
  - `min.dmps`: Required number of significant DMPs (default: 2)
  - `min.cpgs`: Required total CpGs including non-DMPs (default: 3)
  - Higher values create more robust but fewer regions

```{r}
# Example of different region building criteria

```{r}
# Require larger regions
dmrs_large <- findDMRsFromDMPs(
    beta.file = beta_file,
    dmps.tsv.file = dmps_file,
    pheno = example_pheno,
    sample_group.col = "group",
    min.dmps = 3,  # More supporting DMPs
    min.cpgs = 5,  # More CpGs per region
    max.lookup.dist = 500  # Smaller distance between DMPs
)

# Compare region sizes
summary(width(dmrs))
summary(width(dmrs_large))
```

# Advanced Analysis

## Filtering DMRs by Quality Metrics

You can use the metadata to filter DMRs based on various quality criteria:

```{r}
# Filter for high-confidence DMRs
high_conf_dmrs <- dmrs[
    mcols(dmrs)$region_correlation > 0.7 &  # High correlation
    mcols(dmrs)$min_correlation > 0.5 &     # No weak links
    abs(mcols(dmrs)$mean_delta_beta) > 0.2  # Strong effect size
]

# Look at the filtering results
length(dmrs)  # Before
length(high_conf_dmrs)  # After
```

## Understanding Region Expansion

The expansion metadata helps understand why regions stopped growing:

```{r}
# Examine expansion stop reasons
table(mcols(dmrs)$upstream_cpg_expansion_stop_reason)
table(mcols(dmrs)$downstream_cpg_expansion_stop_reason)

# Find regions that might be part of larger DMRs
potential_large_dmrs <- dmrs[
    mcols(dmrs)$upstream_cpg_expansion_stop_reason == "end-of-input" |
    mcols(dmrs)$downstream_cpg_expansion_stop_reason == "end-of-input"
]
```

## Analyzing DMR Properties

Let's look at relationships between different DMR properties:

```{r}
library(ggplot2)

# Create a data frame of DMR properties
dmr_props <- data.frame(
    width = width(dmrs),
    n_cpgs = mcols(dmrs)$n_cpgs,
    correlation = mcols(dmrs)$region_correlation,
    effect_size = abs(mcols(dmrs)$mean_delta_beta)
)

# Plot relationships
ggplot(dmr_props, aes(x=n_cpgs, y=correlation)) +
    geom_point() +
    geom_smooth(method="lm") +
    labs(title="Relationship between DMR Size and Correlation",
         x="Number of CpGs",
         y="Region Correlation")

# Look at effect sizes by region width
ggplot(dmr_props, aes(x=width, y=effect_size)) +
    geom_point() +
    geom_smooth() +
    labs(title="Effect Size vs Region Width",
         x="Region Width",
         y="Absolute Mean Delta Beta")
```

# Visualizing DMRs

The package provides several ways to visualize DMRs and their properties. Here are
some useful visualization approaches:

## Basic DMR Visualization

```{r}
library(ggbio)
library(GenomeInfoDb)

# Create a basic autoplot of DMR locations
dmrs_for_plot <- keepStandardChromosomes(dmrs, pruning.mode="coarse")
autoplot(dmrs_for_plot, layout="karyogram")
```

## Methylation Heatmaps

We can create heatmaps of methylation values within DMRs:

```{r}
library(ComplexHeatmap)
library(circlize)

# Function to extract beta values for a DMR
get_dmr_betas <- function(dmr, beta_matrix) {
    region_rows <- rownames(beta_matrix) >= mcols(dmr)$start_cpg &
                   rownames(beta_matrix) <= mcols(dmr)$end_cpg
    beta_matrix[region_rows, ]
}

# Plot methylation heatmap for top DMRs
top_dmrs <- head(sort(dmrs, by = ~ abs(mean_delta_beta), decreasing = TRUE))
# Load example data included with the package
data(example_beta)
data(example_pheno)

top_dmr_betas <- do.call(rbind, lapply(top_dmrs, get_dmr_betas, example_beta))

Heatmap(top_dmr_betas,
        name = "Beta Value",
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        column_split = example_pheno$group,
        column_title = "Sample Groups",
        row_title = "CpGs",
        col = colorRamp2(c(0, 0.5, 1), c("blue", "white", "red")))
```

## Quality Assessment Plots

Visualize DMR quality metrics to help parameter tuning:

```{r}
# Create multi-panel quality plot
par(mfrow=c(2,2))

# Region sizes
hist(width(dmrs), 
     main="DMR Width Distribution",
     xlab="Width (bp)")

# Correlation strength
hist(mcols(dmrs)$region_correlation,
     main="Region Correlation Distribution",
     xlab="Mean Correlation")

# Effect sizes
hist(abs(mcols(dmrs)$mean_delta_beta),
     main="Effect Size Distribution",
     xlab="Absolute Mean Delta Beta")

# CpG density
hist(mcols(dmrs)$n_cpgs,
     main="CpGs per DMR",
     xlab="Number of CpGs")

par(mfrow=c(1,1))
```

# Validation

We can compare our found DMRs with the true DMRs in our example data:

```{r}
data(example_true_dmrs)

# Convert true DMRs to GRanges
true_gr <- GRanges(
    seqnames = "chr1",
    ranges = IRanges(
        start = example_true_dmrs$start,
        end = example_true_dmrs$end
    )
)

# Find overlaps
overlaps <- findOverlaps(dmrs, true_gr)
length(overlaps)

# Calculate validation metrics
sensitivity <- length(unique(subjectHits(overlaps))) / length(true_gr)
precision <- length(unique(queryHits(overlaps))) / length(dmrs)

cat("Sensitivity:", round(sensitivity, 3), "\n")
cat("Precision:", round(precision, 3), "\n")
```

# Session Info

```{r}
sessionInfo()
```

# References

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

# Introduction

The `DMRSegal` package provides tools for identifying Differentially Methylated 
Regions (DMRs) from pre-computed Differentially Methylated Positions (DMPs). This
approach is particularly useful when you already have DMP data and want to identify
broader regions of differential methylation.

# Installation

You can install the released version of DMRSegal from Bioconductor with:

```{r eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("DMRSegal")
```

# Quick Start

Here's a basic example using the included example data:

```{r}
library(DMRSegal)

# Load example data
data("example_beta")
data("example_dmps")

# Look at the input data structure
head(example_dmps)
dim(example_beta)

# Find DMRs
dmrs <- findDMRsFromDMPs(
    betaMatrix = example_beta,
    dmps = example_dmps,
    minDMPs = 2,
    minCpGs = 5,
    maxDist = 1000
)

# View results
dmrs
```

# Detailed Usage

## Input Data Preparation

The package expects two main inputs:

1. A beta value matrix or SummarizedExperiment object:
   - Rows represent CpG sites
   - Columns represent samples
   - Values should be beta values (0-1 range)
   - Row names should match the CpG identifiers in your DMPs

2. A data frame containing DMP information:
   - Must contain columns: 'chr' (chromosome), 'pos' (position)
   - Must contain a p-value column (default name: 'pval_adj')
   - Row names should match the CpG identifiers in your beta matrix

Here's an example of creating a SummarizedExperiment object:

```{r}
library(SummarizedExperiment)

# Create a SummarizedExperiment object
se <- SummarizedExperiment(
    assays = list(beta = example_beta),
    rowData = example_dmps
)

# Use it with findDMRsFromDMPs
dmrs_se <- findDMRsFromDMPs(
    betaMatrix = se,
    dmps = example_dmps,
    minDMPs = 2,
    minCpGs = 5,
    maxDist = 1000
)
```

## Algorithm Overview

The DMR finding process involves several steps:

1. Input Validation:
   - Checks data format and consistency
   - Ensures all DMPs have corresponding beta values

2. DMP Clustering:
   - Converts DMPs to genomic ranges
   - Groups DMPs based on proximity (maxDist parameter)

3. Region Expansion:
   - Includes nearby CpGs in the region
   - Filters based on minimum CpG count

4. Result Generation:
   - Creates GRanges object with metadata
   - Includes statistics about each DMR

## Parameters

The main function `findDMRsFromDMPs()` has several parameters that can be tuned:

- `minDMPs`: Minimum number of DMPs required to form a DMR
  - Higher values create more stringent DMR calling
  - Default: 1

- `minCpGs`: Minimum number of CpGs in a region
  - Higher values ensure broader coverage
  - Default: 50

- `maxDist`: Maximum distance between DMPs
  - Controls how far apart DMPs can be to form a DMR
  - Default: 10000 (10kb)

## Output

The function returns a GRanges object containing the identified DMRs with 
metadata columns:

- `n_cpgs`: Number of CpG sites in the region
- `n_dmps`: Number of significant DMPs in the region
- `mean_pval`: Mean adjusted p-value of DMPs in the region

```{r}
# Example of accessing DMR metadata
GenomicRanges::mcols(dmrs)
```

# Session Info

```{r}
sessionInfo()
```

