---
title: "Visualizing DMRs with DMRSegal"
author: "DMRSegal Package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualizing DMRs with DMRSegal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 4,
  dpi = 150
)
```

## Introduction

This vignette demonstrates how to visualize Differentially Methylated Regions (DMRs)
identified by the `findDMRsFromSeeds()` function in the DMRSegal package.

## Overview of Visualization Approach

DMRs are complex genomic features built from:

1. **Seed DMPs**: Individual CpG sites with differential methylation
2. **Core DMR**: Connected DMPs forming a contiguous region
3. **Extended region**: Surrounding CpG sites included in the final DMR

The visualization functions in DMRSegal represent these hierarchical relationships
using a layered plotting approach.

## Loading Example Data

```{r load_data, eval=FALSE}
library(DMRSegal)
library(ggplot2)

# Load pre-computed DMR results
dmrs <- readRDS("benchmark_data/dmrs.dmrsegal.rds")
```

```{r load_real, echo=FALSE, eval=FALSE}
# For vignette building - adjust path as needed
if (file.exists("../benchmark_data/dmrs.dmrsegal.rds")) {
  dmrs <- readRDS("../benchmark_data/dmrs.dmrsegal.rds")
} else if (file.exists("benchmark_data/dmrs.dmrsegal.rds")) {
  dmrs <- readRDS("benchmark_data/dmrs.dmrsegal.rds")
}
```

## Basic DMR Visualization

### Plotting a Single DMR

The `plotDMR()` function creates a detailed visualization of a single DMR:

```{r plot_single, eval=FALSE}
# Plot the first DMR
p1 <- plotDMR(dmrs, dmr_index = 1)
print(p1)
```

The plot shows:

- **Gray vertical lines**: Extended CpG sites in the region
- **Blue arrows**: Seed DMPs (differentially methylated positions)
- **Red horizontal line**: Core DMR connecting the DMPs
- **Shaded regions**: DMR boundaries and extended region

### Understanding Plot Elements

The title provides key information about the DMR:

- Genomic coordinates (chromosome:start-end)
- Number of DMPs (seed sites)
- Number of CpGs (total sites in extended region)
- Delta beta (effect size)
- Adjusted p-value (significance)

## Plotting Multiple DMRs

To compare multiple DMRs, use `plotDMRs()`:

```{r plot_multiple, eval=FALSE}
# Plot first 6 DMRs in a grid
library(patchwork)  # or gridExtra
p_grid <- plotDMRs(dmrs, dmr_indices = 1:6, ncol = 3)
print(p_grid)
```

## Advanced Visualization with Beta Values

For detailed analysis, you can overlay beta values using `plotDMRWithBeta()`:

```{r plot_with_beta, eval=FALSE}
# Option 1: Using a BetaFileHandler object (recommended for multiple plots)
beta_handler <- BetaFileHandler$new(
  beta_file = "path/to/beta_values.txt",
  array = "450K",
  genome = "hg19"
)

p_detailed <- plotDMRWithBeta(
  dmrs = dmrs,
  dmr_index = 1,
  beta_handler = beta_handler,
  pheno = pheno_df,
  sample_group_col = "Sample_Group"
)
print(p_detailed)

# Option 2: Using a file path directly (handler created automatically)
p_detailed <- plotDMRWithBeta(
  dmrs = dmrs,
  dmr_index = 1,
  beta_handler = "path/to/beta_values.txt",
  pheno = pheno_df,
  sample_group_col = "Sample_Group"
)
print(p_detailed)
```

This creates a two-panel plot:
- **Top panel**: DMR structure (as in `plotDMR()`)
- **Bottom panel**: Heatmap of beta values across samples

## Customization

### Changing Plot Appearance

```{r customize, eval=FALSE}
# Create base plot
p <- plotDMR(dmrs, dmr_index = 1)

# Apply different theme
p <- p + theme_bw()

# Adjust text size for presentations
p <- p + theme(
  plot.title = element_text(size = 16, face = "bold"),
  axis.title = element_text(size = 14),
  axis.text = element_text(size = 12)
)

print(p)
```

### Selecting Specific DMRs

```{r select_dmrs, eval=FALSE}
# Extract DMR metadata
dmr_data <- as.data.frame(S4Vectors::mcols(dmrs))

# Find DMRs with strong signals
strong_dmrs <- which(
  dmr_data$dmps_num >= 3 &           # Multiple DMPs
  abs(dmr_data$delta_beta) > 0.3 &   # Strong effect
  dmr_data$pval_adj < 0.01            # Highly significant
)

# Plot selected DMRs
p_selected <- plotDMRs(dmrs, dmr_indices = strong_dmrs[1:4], ncol = 2)
print(p_selected)
```

## Saving Plots

### For Publications

```{r save_pub, eval=FALSE}
# High resolution PNG
ggsave("figure1_dmr.png", p, width = 10, height = 4, dpi = 600)

# Vector format (PDF)
ggsave("figure1_dmr.pdf", p, width = 10, height = 4)

# TIFF for some journals
ggsave("figure1_dmr.tiff", p, width = 10, height = 4, dpi = 600, compression = "lzw")
```

### For Presentations

```{r save_pres, eval=FALSE}
# Lower resolution, larger size
ggsave("presentation_dmr.png", p, width = 12, height = 5, dpi = 150)
```

## Complete Example Workflow

```{r workflow, eval=FALSE}
library(DMRSegal)
library(ggplot2)

# 1. Load DMR results
dmrs <- readRDS("my_dmrs.rds")

# 2. Quick overview
overview <- plotDMRs(dmrs, dmr_indices = 1:4, ncol = 2)
ggsave("dmr_overview.png", overview, width = 12, height = 8)

# 3. Identify interesting DMRs
dmr_data <- as.data.frame(S4Vectors::mcols(dmrs))
interesting_idx <- which(
  dmr_data$dmps_num >= 2 &
  abs(dmr_data$delta_beta) > 0.2 &
  dmr_data$pval_adj < 0.05
)

# 4. Create publication figure for top DMR
top_dmr <- interesting_idx[1]
p_pub <- plotDMR(dmrs, dmr_index = top_dmr)

# Enhance for publication
p_pub <- p_pub + 
  theme_classic() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(size = 12)
  )

# Save in multiple formats
ggsave("figure_main.png", p_pub, width = 10, height = 4, dpi = 600)
ggsave("figure_main.pdf", p_pub, width = 10, height = 4)
```

## Interpretation Guide

### CpG Density

Dense vertical gray lines indicate CpG-rich regions such as:
- CpG islands
- Gene promoters
- Regulatory elements

### DMR Span

The width of the red line shows:
- **Short DMRs**: Localized methylation changes
- **Long DMRs**: Extended regulatory regions or multiple functional elements

### Number of DMPs

More blue arrows (DMPs) indicate:
- Stronger, more consistent signal
- Multiple independent evidence points
- Higher confidence in the DMR

### Effect Size (Delta Beta)

Shown in the title, indicates:
- **|Δβ| < 0.2**: Small effect
- **|Δβ| 0.2-0.5**: Moderate effect
- **|Δβ| > 0.5**: Large effect

## Tips and Best Practices

1. **Start with overview plots** (`plotDMRs()`) to identify patterns
2. **Focus on significant DMRs** with adjusted p-value < 0.05
3. **Consider effect size** in addition to significance
4. **Look for clustered DMPs** as evidence of coordinated regulation
5. **Compare CpG density** to genomic context (islands, shores, etc.)

## See Also

- `findDMRsFromSeeds()`: Main function for DMR identification
- `extractCpgInfoFromResultDMRs()`: Extract detailed CpG information
- `annotateDMRsWithGenes()`: Annotate DMRs with gene information

## Session Info

```{r session_info}
sessionInfo()
```
