---
title: "Benchmarking CMEnt on Repeated Simulated Microarray Data"
subtitle: "CMEnt vs comb-p, probeLasso, DMRcate, and Bumphunter across repeated simulateDMRs runs"
author: "CMEnt Package"
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
        code_folding: hide
---

```{r, include=FALSE}
knitr::opts_chunk$set(
    results = "asis",
    cache = FALSE,
    echo = TRUE,
    dev = "png",
    warning = FALSE,
    message = FALSE,
    fig.width = 8,
    fig.height = 6,
    dpi = 300
)
```

## Introduction

This vignette repeats the simulated 450K benchmark over multiple independently
simulated datasets. Each dataset is produced from the same GSE40279 chr20/chr21
beta-value subset and 100-sample covariate subset used by the single-run
benchmark, with `CMEnt::simulateDMRs()` parameters drawn from fixed ranges under
a master seed of 42. Method outputs and simulation bundles are cached per
iteration, and summary plots show medians with observed interval ranges across
simulations.

Methods:

- **CMEnt**: correlation-based seed-guided assembly approach
- **comb-p**: adjacent p-value combination
- **probeLasso**: lasso-based feature selection for DMR detection
- **DMRcate**: Gaussian kernel smoothing approach
- **Bumphunter**: permutation-based regional testing

## Setup and Dependencies

```{r load_packages, message=FALSE, warning=FALSE}
.loadCMEnt <- function() {
    root <- if (file.exists("NAMESPACE")) "." else if (file.exists(file.path("..", "NAMESPACE"))) ".." else NULL
    if (!is.null(root) && requireNamespace("pkgload", quietly = TRUE)) {
        ok <- tryCatch({
            pkgload::load_all(root, quiet = TRUE)
            TRUE
        }, error = function(e) {
            message("Could not load local CMEnt source: ", conditionMessage(e))
            FALSE
        })
        if (ok) return(invisible(TRUE))
    }
    suppressPackageStartupMessages(library(CMEnt))
}
.loadCMEnt()
suppressPackageStartupMessages({
    library(data.table)
    library(dplyr)
    library(GenomicRanges)
    library(ggplot2)
})

getBenchmarkFile <- function(filename) {
    folder <- "simulated_microarray450k_multisim"
    benchmark_file <- if (getwd() == "notebooks") {
        file.path("cached_results", folder, filename)
    } else if ("NAMESPACE" %in% dir()) {
        file.path("notebooks", "cached_results", folder, filename)
    } else {
        file.path("cached_results", folder, filename)
    }
    dir.create(dirname(benchmark_file), showWarnings = FALSE, recursive = TRUE)
    benchmark_file
}

getSimulationFile <- function(sim_id, filename) {
    getBenchmarkFile(file.path(sprintf("sim_%02d", sim_id), filename))
}

read_cache <- function(path, params = NULL) {
    if (!file.exists(path)) return(NULL)
    out <- readRDS(path)
    if (is.null(params) || identical(out$params, params)) out else NULL
}

save_cache <- function(path, value, params = NULL) {
    saveRDS(list(value = value, params = params), path)
}

nsim <- suppressWarnings(as.integer(Sys.getenv("CMENT_BENCHMARK_NSIM", "10")))
nsim <- ifelse(is.na(nsim) || nsim < 1L, 10L, nsim)

set.seed(42)
simulation_plan <- data.frame(
    simulation_id = seq_len(nsim),
    seed = sample(0:1000, nsim),
    delta_beta_max = round(runif(nsim, 0.2, 0.5), 2),
    expected_corr = round(runif(nsim, 0.5, 0.8), 2)
)
NJOBS <- as.integer(
    Sys.getenv(
        "NJOBS",
        unset = as.character(max(1L, min(8L, parallel::detectCores(logical = TRUE) - 1L)))
    )
)
doParallel::registerDoParallel(cores = NJOBS)
plan_file <- getBenchmarkFile("simulation_parameters.tsv")
data.table::fwrite(simulation_plan, plan_file, sep = "\t")
knitr::kable(simulation_plan, caption = "Simulation parameters generated under master seed 42")
```

### Install External Dependencies

```{r setup_dependencies, message=FALSE, warning=FALSE}
r <- getOption("repos")
if (is.null(r) || is.na(r["CRAN"]) || r["CRAN"] %in% c("@CRAN@", "")) {
    options(repos = c(CRAN = "https://cloud.r-project.org"))
}
pkg_installed <- function(pkg) pkg %in% rownames(utils::installed.packages())
if (!pkg_installed("benchmarkme")) install.packages("benchmarkme")
if (!pkg_installed("BiocManager")) install.packages("BiocManager")
if (!pkg_installed("DMRcate")) BiocManager::install("DMRcate", update = FALSE, ask = FALSE)
if (!pkg_installed("bumphunter")) BiocManager::install("bumphunter", update = FALSE, ask = FALSE)
if (!pkg_installed("doParallel")) install.packages("doParallel")
if (!pkg_installed("tidyr")) install.packages("tidyr")
if (!pkg_installed("callr")) install.packages("callr")
if (!pkg_installed("ChAMP")) BiocManager::install("ChAMP", update = FALSE, ask = FALSE)
anno_pkg <- "IlluminaHumanMethylation450kanno.ilmn12.hg19"
if (!nzchar(system.file(package = anno_pkg))) BiocManager::install(anno_pkg, update = FALSE, ask = FALSE)
```

**Note:** comb-p requires the Python `combined-pvalues` command and bedtools:

```{bash}
touch .combp.sh
{ command -v bedtools >/dev/null 2>&1; } || {
    apt-get update && apt-get install -y bedtools
}
{ command -v comb-p >/dev/null 2>&1 && command -v bedtools >/dev/null 2>&1 && command -v comb-p > .combp.sh; } || { 
    echo >&2 "comb-p or bedtools not found. Installing with conda...";
    apt-get update && apt-get install -y curl git bedtools
    if ! command -v conda >/dev/null 2>&1; then
        curl -L https://repo.anaconda.com/miniconda/Miniconda3-latest-Linux-x86_64.sh -o miniconda.sh
        rm -rf $HOME/miniconda
        bash miniconda.sh -b -p $HOME/miniconda
        rm miniconda.sh
        MINICONDA_BIN="$HOME/miniconda/bin/conda"
        $MINICONDA_BIN init bash
        export PATH="$HOME/miniconda/bin:$PATH"
    else 
        MINICONDA_BIN=$(command -v conda)
    fi
    $MINICONDA_BIN tos accept --override-channels --channel https://repo.anaconda.com/pkgs/main
    $MINICONDA_BIN tos accept --override-channels --channel https://repo.anaconda.com/pkgs/r
    $MINICONDA_BIN create -n combp -y python=3.7
    $MINICONDA_BIN run -n combp -- pip install git+https://github.com/brentp/combined-pvalues.git numpy==1.21.6
    echo "$MINICONDA_BIN run -n combp comb-p" > .combp.sh
}
chmod +x .combp.sh
```

## Load Input Data and Simulate

```{r simulate_data}
array_type <- "450K"
genome <- "hg19"
input_beta_path <- getBenchmarkFile("GSE40279_beta_subset.tsv.gz")
input_samplesheet_path <- getBenchmarkFile("GSE40279_sim_samplesheet.tsv")
zenodo_url <- "https://zenodo.org/records/20592944/files/GSE40279_beta_subset.tsv.gz?download=1"
zenodo_samplesheet_url <- "https://zenodo.org/records/20592944/files/GSE40279_sim_samplesheet.tsv?download=1"
if (!file.exists(input_beta_path)) {
    curl::curl_download(zenodo_url, input_beta_path)
}
if (!file.exists(input_samplesheet_path)) {
    curl::curl_download(zenodo_samplesheet_url, input_samplesheet_path)
}

simulation_samplesheet <- data.table::fread(input_samplesheet_path, data.table = FALSE, check.names = FALSE)
rownames(simulation_samplesheet) <- simulation_samplesheet[[1L]]
simulation_samplesheet <- simulation_samplesheet[, -1L, drop = FALSE]
simulation_samplesheet$Sample_Group <- NULL

base_simulation <- list(
    delta_jitter = 0.1,
    max_gap = 500L,
    min_dmr_size = 50L,
    max_dmr_size = 2000L,
    min_sites = 5L,
    max_sites = 500L,
    num_dmrs = 500L,
    genome = genome,
    sample_ids = rownames(simulation_samplesheet),
    covariates = c("age..y..ch1", "predictedSex")
)

simulate_one <- function(plan_row) {
    sim_file <- getSimulationFile(plan_row$simulation_id, "simulation.rds")
    params <- c(base_simulation, list(
        seed = plan_row$seed,
        delta_beta_max = plan_row$delta_beta_max,
        expected_corr = plan_row$expected_corr
    ))
    cached <- read_cache(sim_file, params)
    if (!is.null(cached)) return(cached$value)

    set.seed(plan_row$seed)
    sim <- CMEnt::simulateDMRs(
        beta = input_beta_path,
        array = array_type,
        genome = genome,
        samplesheet = simulation_samplesheet,
        sample_group_col = "Sample_Group",
        covariates = base_simulation$covariates,
        num_dmrs = base_simulation$num_dmrs,
        delta_max0 = plan_row$delta_beta_max,
        delta_jitter = base_simulation$delta_jitter,
        expected_correlation = plan_row$expected_corr,
        max_gap = base_simulation$max_gap,
        min_dmr_size = base_simulation$min_dmr_size,
        max_dmr_size = base_simulation$max_dmr_size,
        min_sites = base_simulation$min_sites,
        max_sites = base_simulation$max_sites,
        verbose = 1
    )
    sim$pheno$Sample_ID <- rownames(sim$pheno)
    sim$pheno$casecontrol <- sim$pheno$Sample_Group == "case"
    data.table::fwrite(sim$truth, getSimulationFile(plan_row$simulation_id, "truth.tsv"), sep = "\t")
    data.table::fwrite(sim$pheno, getSimulationFile(plan_row$simulation_id, "samplesheet.tsv"), sep = "\t")
    save_cache(sim_file, sim, params)
    sim
}

benchmark_specs_file <- getBenchmarkFile("benchmark_specs.rds")
if (!file.exists(benchmark_specs_file)) {
    benchmark_specs <- list(
        cpu = benchmarkme::get_cpu(),
        ram = benchmarkme::get_ram(),
        platform = benchmarkme::get_platform_info(),
        rversion = R.version.string,
        njobs = NJOBS
    )
    saveRDS(benchmark_specs, benchmark_specs_file)
} else {
    benchmark_specs <- readRDS(benchmark_specs_file)
}
cpu <- benchmark_specs$cpu
ram <- benchmark_specs$ram
platform <- benchmark_specs$platform
rversion <- benchmark_specs$rversion
njobs <- benchmark_specs$njobs
cat("CPU:", cpu$vendor, cpu$model, "(", cpu$no_of_cores, "cores )\n")
cat("RAM:", capture.output(print(ram)), "\n")
cat("OS:", platform$OS.type, "\n")
cat("R Version:", rversion, "\n")
cat("Using", njobs, "parallel jobs\n")
```

## Method Runners

```{r method_helpers}
as_truth_gr <- function(sim) {
    force_genome(makeGRangesFromDataFrame(
        sim$truth,
        seqnames.field = "seqnames",
        start.field = "start",
        end.field = "end",
        keep.extra.columns = TRUE,
        seqinfo = GenomeInfoDb::Seqinfo(genome = genome)
    ))
}

force_genome <- function(gr) {
    GenomeInfoDb::genome(gr) <- genome
    gr
}

logit2 <- function(x) log2(pmax(pmin(x, 1 - 1e-6), 1e-6)) - log2(1 - pmax(pmin(x, 1 - 1e-6), 1e-6))

prepare_array_data <- function(sim) {
    beta_handler <- CMEnt::getBetaHandler(sim$simulated, sorted_locs = sim$beta_locs, njobs = njobs)
    beta_mat <- as.matrix(beta_handler$getBeta())
    locs <- beta_handler$getBetaLocs()
    pheno <- sim$pheno[colnames(beta_mat), , drop = FALSE]
    pheno$Sample_Group <- factor(pheno$Sample_Group, levels = c("control", "case"))
    pheno$casecontrol <- pheno$Sample_Group == "case"

    dmps <- CMEnt::findDMPsArray(
        beta_handler,
        pheno,
        sample_group_col = "Sample_Group",
        id_col = "Sample_ID",
        array = array_type,
        sorted_locs = sim$beta_locs,
        case_group = "case",
        njobs = njobs
    )
    rownames(dmps) <- dmps$site_id
    dmps <- dmps[!is.na(dmps$pval), ]
    dmp_locs <- locs[rownames(locs) %in% rownames(dmps), , drop = FALSE]
    dmps <- dmps[rownames(dmp_locs), , drop = FALSE]

    dmps_gr <- makeGRangesFromDataFrame(
        data.frame(
            chr = dmp_locs$chr,
            start = dmp_locs$start,
            end = dmp_locs$end,
            id = rownames(dmps),
            pval = dmps$pval,
            qval = dmps$qval,
            delta_beta = dmps$delta_beta
        ),
        keep.extra.columns = TRUE
    )
    dmps_gr <- sort(dmps_gr)
    names(dmps_gr) <- mcols(dmps_gr)$id

    list(
        beta_handler = beta_handler,
        beta_mat = beta_mat,
        locs = locs,
        pheno = pheno,
        dmps = dmps,
        dmps_gr = dmps_gr,
        seeds = mcols(dmps_gr[mcols(dmps_gr)$qval < 0.05])$id
    )
}

run_combp <- function(dmps_gr, params) {
    if (Sys.which("bedtools") == "" || Sys.which("comb-p") == "") {
        warning("comb-p or bedtools not found in PATH; returning empty set.")
        return(GRanges())
    }
    tmpdir <- tempdir()
    bed_file <- file.path(tmpdir, "dmps_for_combp.bed")
    bed_df <- data.frame(
        chrom = as.character(seqnames(dmps_gr)),
        start = start(dmps_gr) - 1L,
        end = end(dmps_gr),
        pval = mcols(dmps_gr)$pval
    )
    write.table(bed_df[order(bed_df$chrom, bed_df$start), ], bed_file, quote = FALSE, sep = "\t", row.names = FALSE)
    out_prefix <- file.path(tmpdir, "output.")
    cmd <- sprintf(
        "./.combp.sh pipeline -c 4 --seed %d --dist %d -p %s --region-filter-p %s --region-filter-n %d %s",
        params$seed, params$dist, out_prefix, params$region_filter_p, params$region_filter_n, bed_file
    )
    if (system(cmd, ignore.stdout = FALSE, ignore.stderr = FALSE) != 0L) return(GRanges())
    regions_file <- paste0(out_prefix, "regions-p.bed.gz")
    if (!file.exists(regions_file)) return(GRanges())
    regions <- data.table::fread(cmd = sprintf("zcat %s", shQuote(regions_file)))
    colnames(regions) <- c("chr", "start", "end", "min_p", "n_probes", "z_p", "z_sidak_p")
    regions <- regions[regions$z_sidak_p < params$region_sidak_p & regions$n_probes >= params$region_filter_n, ]
    if (nrow(regions) == 0L) return(GRanges())
    makeGRangesFromDataFrame(regions, seqnames.field = "chr", start.field = "start", end.field = "end", keep.extra.columns = TRUE)
}

run_probelasso_champ <- function(beta_mat, pheno, params) {
    if (!requireNamespace("callr", quietly = TRUE)) return(GRanges(seqinfo = GenomeInfoDb::Seqinfo(genome = params$genome)))
    pl_dmr <- tryCatch({
        callr::r(
            function(beta_mat, sample_group, params) {
                if (!requireNamespace("ChAMP", quietly = TRUE)) stop("ChAMP is not installed.")
                suppressPackageStartupMessages(library(ChAMP))
                result <- ChAMP::champ.DMR(
                    beta = beta_mat,
                    pheno = sample_group,
                    arraytype = params$array_type,
                    method = "ProbeLasso",
                    minProbes = params$minProbes,
                    adjPvalDmr = params$adjPvalDmr,
                    adjPvalProbe = params$adjPvalProbe,
                    meanLassoRadius = params$meanLassoRadius,
                    minDmrSep = params$minDmrSep,
                    minDmrSize = params$minDmrSize,
                    PDFplot = FALSE,
                    Rplot = FALSE,
                    resultsDir = file.path(tempdir(), "champ_probelasso")
                )
                pl_dmr <- result$ProbeLassoDMR
                if (is.null(pl_dmr) || nrow(pl_dmr) == 0L) return(data.frame())
                as.data.frame(pl_dmr)
            },
            args = list(
                beta_mat = beta_mat,
                sample_group = as.character(pheno$Sample_Group),
                params = params
            ),
            show = TRUE
        )
    }, error = function(e) {
        warning(sprintf("probeLasso failed: %s", e$message))
        data.frame()
    })
    if (nrow(pl_dmr) == 0L) return(GRanges(seqinfo = GenomeInfoDb::Seqinfo(genome = params$genome)))
    if (!"seqnames" %in% names(pl_dmr) && "dmrChrom" %in% names(pl_dmr)) pl_dmr$seqnames <- pl_dmr$dmrChrom
    if (!"start" %in% names(pl_dmr) && "dmrStart" %in% names(pl_dmr)) pl_dmr$start <- pl_dmr$dmrStart
    if (!"end" %in% names(pl_dmr) && "dmrEnd" %in% names(pl_dmr)) pl_dmr$end <- pl_dmr$dmrEnd
    makeGRangesFromDataFrame(
        pl_dmr, seqnames.field = "seqnames", start.field = "start", end.field = "end",
        seqinfo = GenomeInfoDb::Seqinfo(genome = params$genome), keep.extra.columns = TRUE
    )
}

run_dmrcate_array <- function(beta_mat, pheno, params) {
    if (!requireNamespace("DMRcate", quietly = TRUE)) return(GRanges(seqinfo = GenomeInfoDb::Seqinfo(genome = params$genome)))
    design <- model.matrix(~ Sample_Group, data = pheno)
    coef_col <- which(colnames(design) == "Sample_Groupcase")
    colnames(design)[1] <- "(Intercept)"
    colnames(design)[coef_col] <- "case_vs_control"
    ann <- DMRcate::cpg.annotate(
        "array",
        logit2(beta_mat),
        what = "M",
        arraytype = params$array_type,
        analysis.type = "differential",
        design = design,
        coef = coef_col,
        fdr = params$fdr
    )
    DMRcate::extractRanges(DMRcate::dmrcate(
        ann,
        C = params$C,
        lambda = params$lambda,
        min.cpgs = params$min.cpgs,
        betacutoff = params$betacutoff
    ), genome = params$genome)
}

run_bumphunter_array <- function(beta_mat, locs, pheno, params) {
    if (!requireNamespace("bumphunter", quietly = TRUE) || !requireNamespace("doParallel", quietly = TRUE)) return(GRanges())
    design <- model.matrix(~ Sample_Group, data = pheno)
    coef_col <- which(colnames(design) == "Sample_Groupcase")
    on.exit({
        try(doParallel::stopImplicitCluster(), silent = TRUE)
        gc()
    }, add = TRUE)
    cluster <- bumphunter::clusterMaker(as.character(locs$chr), locs$start, maxGap = params$max_gap)

    run_fit <- function(cutoff, B) {
        bumphunter::bumphunter(
            beta_mat,
            design = design,
            chr = as.character(locs$chr),
            pos = locs$start,
            cluster = cluster,
            cutoff = cutoff,
            maxGap = params$max_gap,
            nullMethod = "bootstrap",
            B = B,
            type = "Beta",
            coef = coef_col
        )
    }
    has_informative_null <- function(fit) {
        null_values <- unlist(fit$null$value, use.names = FALSE)
        null_lengths <- unlist(fit$null$length, use.names = FALSE)
        length(null_values) > 0L &&
            any(abs(null_values) > 0) &&
            length(null_lengths) > 0L &&
            any(null_lengths > 0)
    }

    cutoff <- params$cutoff_start
    n_candidates <- Inf
    iter <- 0L
    while (n_candidates > params$max_dmrs && iter < params$max_iter) {
        iter <- iter + 1L
        fit0 <- run_fit(cutoff = cutoff, B = 0L)
        n_candidates <- nrow(fit0$table)
        cutoff <- cutoff * 1.2
        rm(fit0)
        gc()
    }
    cutoff <- cutoff / 1.2
    cat(sprintf(
        "Bumphunter: cutoff=%.4f, candidate_bumps=%d, minL=%d, maxGap=%d, cores=%d, B=%d\n",
        cutoff, n_candidates, params$min_l, params$max_gap, njobs, params$B
    ))

    fit <- run_fit(cutoff = cutoff, B = params$B)
    adjust_iter <- 0L
    while (
        nrow(fit$table) > 0L &&
            (all(is.na(fit$table$p.value)) || !has_informative_null(fit)) &&
            cutoff > params$cutoff_floor &&
            adjust_iter < params$max_bootstrap_adjust
    ) {
        prev_cutoff <- cutoff
        cutoff <- max(cutoff / 2, params$cutoff_floor)
        if (identical(cutoff, prev_cutoff)) break
        adjust_iter <- adjust_iter + 1L
        cat(sprintf(
            "Bumphunter: bootstrap null degenerate at cutoff=%.4f; retrying with cutoff=%.4f\n",
            prev_cutoff, cutoff
        ))
        fit <- run_fit(cutoff = cutoff, B = params$B)
    }

    tab <- fit$table
    keep <- tab$L >= params$min_l &
        is.finite(tab$p.value) &
        tab$p.value <= params$region_p &
        is.finite(tab$start) &
        is.finite(tab$end)
    tab <- tab[keep, , drop = FALSE]
    if (nrow(tab) == 0L) return(GRanges())
    makeGRangesFromDataFrame(tab, seqnames.field = "chr", start.field = "start", end.field = "end", keep.extra.columns = TRUE)
}

time_method <- function(expr) {
    start <- Sys.time()
    value <- force(expr)
    list(dmrs = value, time = Sys.time() - start)
}
```

## Run Benchmark

```{r run_benchmark}
run_one_simulation <- function(plan_row) {
    sim <- simulate_one(plan_row)
    truth_dmrs <- as_truth_gr(sim)
    dat <- prepare_array_data(sim)
    sim_dir <- sprintf("sim_%02d", plan_row$simulation_id)
    common <- list(delta_beta_max = plan_row$delta_beta_max, simulation_seed = plan_row$seed)

    method_specs <- list(
        CMEnt = list(
            params = c(common, list(
                sample_group_col = "Sample_Group", casecontrol_col = "casecontrol",
                max_bridge_seeds_gaps = 0L, max_bridge_extension_gaps = 1L,
                min_seeds = 2L, min_sites = 3L, ext_site_delta_beta = plan_row$delta_beta_max / 3,
                max_lookup_dist = base_simulation$max_gap,
                expansion_window = base_simulation$max_dmr_size,
                njobs = njobs,
                max_pval = 0.05, testing_mode = "parametric", entanglement = "weak",
                genome = genome
            )),
            run = function(params) {
                time_method(
                    do.call(
                        CMEnt::buildDMRs, c(
                            list(
                                beta = dat$beta_handler, seeds = dat$seeds, pheno = dat$pheno,
                                verbose = 2, score_dmrs = FALSE, extract_motifs = FALSE, annotate_with_genes = FALSE
                            ),
                            params[setdiff(names(params), c("delta_beta_max", "simulation_seed"))]
                        )
                    )
                )
            }
        ),
        `comb-p` = list(
            params = c(
                common,
                list(
                    seed = base_simulation$max_gap, dist = base_simulation$max_gap, region_filter_p = 0.01,
                    region_filter_n = base_simulation$min_sites, region_sidak_p = 0.05
                )
            ),
            run = function(params) time_method(run_combp(dat$dmps_gr, params))
        ),
        probeLasso = list(
            params = c(
                common,
                list(
                    runner = "callr_attach_champ", array_type = array_type, genome = genome,
                    minProbes = base_simulation$min_sites, adjPvalDmr = 0.05, adjPvalProbe = 0.1,
                    meanLassoRadius = base_simulation$max_gap,
                    minDmrSep = base_simulation$max_gap, minDmrSize = base_simulation$min_dmr_size
                )
            ),
            run = function(params) time_method(run_probelasso_champ(dat$beta_mat, dat$pheno, params))
        ),
        DMRcate = list(
            params = c(
                common,
                list(
                    array_type = array_type, genome = genome, fdr = 0.05,
                    C = 2, lambda = base_simulation$max_gap * 2,
                    min.cpgs = base_simulation$min_sites, betacutoff = 0.05
                )
            ),
            run = function(params) time_method(run_dmrcate_array(dat$beta_mat, dat$pheno, params))
        ),
        Bumphunter = list(
            params = c(common, list(
                runner = "adaptive_bootstrap_v2",
                B = 100L,
                max_dmrs = 100000L,
                max_gap = base_simulation$max_gap,
                min_l = base_simulation$min_sites,
                region_p = 0.05,
                cutoff_start = plan_row$delta_beta_max / 3,
                cutoff_floor = 0.005,
                max_iter = 10L,
                max_bootstrap_adjust = 6L
            )),
            run = function(params) time_method(run_bumphunter_array(dat$beta_mat, dat$locs, dat$pheno, params))
        )
    )

    method_outputs <- lapply(names(method_specs), function(method) {
        spec <- method_specs[[method]]
        cache_file <- getBenchmarkFile(file.path(sim_dir, sprintf("dmrs.%s.rds", gsub("[^A-Za-z0-9]+", "_", method))))
        cached <- read_cache(cache_file, spec$params)
        out <- if (!is.null(cached) && length(cached$value$dmrs) > 0) {
            cached$value
        } else {
            value <- spec$run(spec$params)
            if (!is.null(value) && length(value$dmrs) > 0) {
                save_cache(cache_file, value, spec$params)
            }
            value
        }
        out$dmrs <- force_genome(out$dmrs)
        list(method = method, dmrs = out$dmrs, time = out$time)
    })
    names(method_outputs) <- names(method_specs)

    list(
        simulation_id = plan_row$simulation_id,
        params = plan_row,
        truth = truth_dmrs,
        seeds_n = length(dat$seeds),
        methods = method_outputs
    )
}

benchmark_results <- lapply(seq_len(nrow(simulation_plan)), function(i) {
    run_one_simulation(simulation_plan[i, , drop = FALSE])
})
```

## Comparative Analysis

```{r analysis_helpers}
get_dmr_stats <- function(dmrs, method, simulation_id) {
    data.frame(
        simulation_id = simulation_id,
        Method = method,
        Number_of_DMRs = length(dmrs),
        Median_DMR_Width = ifelse(length(dmrs) == 0L, NA_real_, median(width(dmrs))),
        Total_Coverage_bp = ifelse(length(dmrs) == 0L, 0, sum(width(reduce(dmrs)))),
        Min_Width = ifelse(length(dmrs) == 0L, NA_real_, min(width(dmrs))),
        Max_Width = ifelse(length(dmrs) == 0L, NA_real_, max(width(dmrs)))
    )
}

evaluate_against_truth <- function(pred, truth) {
    pred <- force_genome(pred)
    truth <- force_genome(truth)
    n_pred <- length(pred)
    n_truth <- length(truth)
    pred_hit <- sum(overlapsAny(pred, truth))
    truth_hit <- sum(overlapsAny(truth, pred))
    precision <- ifelse(n_pred == 0L, 0, pred_hit / n_pred)
    recall <- ifelse(n_truth == 0L, 0, truth_hit / n_truth)
    pred_red <- reduce(pred)
    truth_red <- reduce(truth)
    inter_bp <- sum(width(GenomicRanges::intersect(pred_red, truth_red)))
    pred_bp <- sum(width(pred_red))
    truth_bp <- sum(width(truth_red))
    union_bp <- pred_bp + truth_bp - inter_bp
    precision_bp <- ifelse(pred_bp == 0L, 0, inter_bp / pred_bp)
    recall_bp <- ifelse(truth_bp == 0L, 0, inter_bp / truth_bp)
    data.frame(
        TP_regions = truth_hit,
        FP_regions = n_pred - pred_hit,
        FN_regions = n_truth - truth_hit,
        Precision_regions = precision,
        Recall_regions = recall,
        F1_regions = ifelse(precision + recall == 0, 0, 2 * precision * recall / (precision + recall)),
        Precision_bp = precision_bp,
        Recall_bp = recall_bp,
        F1_bp = ifelse(precision_bp + recall_bp == 0, 0, 2 * precision_bp * recall_bp / (precision_bp + recall_bp)),
        IoU_bp = ifelse(union_bp == 0L, 0, inter_bp / union_bp)
    )
}

format_mean_pm <- function(x, digits = 2L) {
    x <- x[is.finite(x)]
    if (length(x) == 0L) return(NA_character_)
    sd_x <- if (length(x) > 1L) stats::sd(x) else 0
    sprintf(paste0("%.", digits, "f (+/- %.", digits, "f)"), mean(x), sd_x)
}

summarise_mean_pm <- function(df, group_cols, value_cols, digits = 2L) {
    metric_digits <- if (length(digits) == 1L) {
        stats::setNames(rep(digits, length(value_cols)), value_cols)
    } else {
        stats::setNames(as.integer(digits), names(digits))
    }
    df |>
        tidyr::pivot_longer(dplyr::all_of(value_cols), names_to = "Metric", values_to = "Value") |>
        dplyr::group_by(dplyr::across(dplyr::all_of(c(group_cols, "Metric")))) |>
        dplyr::summarise(
            Value = format_mean_pm(Value, digits = metric_digits[dplyr::first(Metric)]),
            .groups = "drop"
        ) |>
        tidyr::pivot_wider(names_from = Metric, values_from = Value)
}

summarise_benchmark_table <- function(stats_all, runtime_all) {
    stats_all |>
        dplyr::select(
            simulation_id,
            Method,
            dmrs_number = Number_of_DMRs,
            median_width_bp = Median_DMR_Width,
            min_width_bp = Min_Width,
            max_width_bp = Max_Width,
            Total_Coverage_bp
        ) |>
        dplyr::mutate(total_coverage_kb = Total_Coverage_bp / 1000) |>
        dplyr::select(-Total_Coverage_bp) |>
        dplyr::left_join(
            runtime_all |>
                dplyr::rename(runtime_sec = Seconds),
            by = c("simulation_id", "Method")
        ) |>
        dplyr::rename(method = Method) |>
        dplyr::group_by(method) |>
        dplyr::summarise(
            dmrs_number = format_mean_pm(dmrs_number, digits = 0L),
            median_width_bp = format_mean_pm(median_width_bp, digits = 0L),
            min_width_bp = format_mean_pm(min_width_bp, digits = 0L),
            max_width_bp = format_mean_pm(max_width_bp, digits = 0L),
            total_coverage_kb = format_mean_pm(total_coverage_kb, digits = 1L),
            runtime_sec = format_mean_pm(runtime_sec, digits = 1L),
            .groups = "drop"
        )
}

stats_all <- do.call(rbind, lapply(benchmark_results, function(res) {
    rbind(
        do.call(rbind, lapply(res$methods, function(x) get_dmr_stats(x$dmrs, x$method, res$simulation_id))),
        get_dmr_stats(res$truth, "Truth", res$simulation_id)
    )
}))

truth_eval_all <- do.call(rbind, lapply(benchmark_results, function(res) {
    do.call(rbind, lapply(res$methods, function(x) {
        out <- evaluate_against_truth(x$dmrs, res$truth)
        out$simulation_id <- res$simulation_id
        out$Method <- x$method
        out
    }))
}))

runtime_all <- do.call(rbind, lapply(benchmark_results, function(res) {
    do.call(rbind, lapply(res$methods, function(x) {
        data.frame(simulation_id = res$simulation_id, Method = x$method, Seconds = as.numeric(x$time, units = "secs"))
    }))
}))
```

### Basic Statistics

```{r}
stats_summary <- summarise_benchmark_table(stats_all, runtime_all)
knitr::kable(stats_summary, caption = "Benchmark statistics across simulations, mean (+/- SD)")
```

```{r, fig.width=9, fig.height=4}
stats_plot <- stats_all |>
    dplyr::select(Method, simulation_id, Number_of_DMRs, Total_Coverage_bp) |>
    tidyr::pivot_longer(c(Number_of_DMRs, Total_Coverage_bp), names_to = "Metric", values_to = "Value") |>
    dplyr::group_by(Method, Metric) |>
    dplyr::mutate(Median = median(Value), Min = min(Value), Max = max(Value)) |>
    dplyr::ungroup()

ggplot(stats_plot, aes(Method, Value, colour = Method)) +
    geom_point(alpha = 0.45, position = position_jitter(width = 0.08, height = 0)) +
    stat_summary(fun = median, geom = "point", size = 2.6, colour = "black") +
    stat_summary(fun.min = min, fun.max = max, geom = "errorbar", width = 0.25, colour = "black") +
    facet_wrap(~ Metric, scales = "free_y") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
    labs(title = "DMR counts and coverage across simulations", x = "", y = "Value")
```

## Ground-Truth Evaluation

```{r}
regions_truth_eval <- summarise_mean_pm(
    truth_eval_all,
    "Method",
    c("TP_regions", "FP_regions", "FN_regions", "Precision_regions", "Recall_regions", "F1_regions"),
    digits = c(TP_regions = 0L, FP_regions = 0L, FN_regions = 0L, Precision_regions = 2L, Recall_regions = 2L, F1_regions = 2L)
)
bp_truth_eval <- summarise_mean_pm(
    truth_eval_all,
    "Method",
    c("Precision_bp", "Recall_bp", "F1_bp", "IoU_bp"),
    digits = 2
)
knitr::kable(regions_truth_eval, caption = "Macro region-level performance against ground truth, mean (+/- SD)")
```

```{r}
knitr::kable(bp_truth_eval, caption = "Micro base-pair-level performance against ground truth, mean (+/- SD)")
```

```{r, fig.width=10, fig.height=5}
metric_plot <- truth_eval_all |>
    dplyr::select(Method, simulation_id, F1_regions, F1_bp, IoU_bp) |>
    tidyr::pivot_longer(c(F1_regions, F1_bp, IoU_bp), names_to = "Metric", values_to = "Value")
metric_plot$Metric <- factor(
    metric_plot$Metric,
    levels = c("F1_regions", "F1_bp", "IoU_bp"),
    labels = c("F1 (Region)", "F1 (Base-pair)", "IoU (Base-pair)")
)

ggplot(metric_plot, aes(Method, Value, colour = Method)) +
    geom_point(alpha = 0.5, position = position_jitter(width = 0.08, height = 0)) +
    stat_summary(fun = median, geom = "point", size = 2.6, colour = "black") +
    stat_summary(fun.min = min, fun.max = max, geom = "errorbar", width = 0.25, colour = "black") +
    facet_wrap(~ Metric, nrow = 1) +
    coord_cartesian(ylim = c(0, 1)) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
    labs(title = "Method performance vs ground truth across simulations", x = "", y = "Score")
```

### Runtime

```{r}
runtime_summary <- summarise_mean_pm(runtime_all, "Method", "Seconds", digits = 2)
knitr::kable(runtime_summary, caption = "Runtime across simulations, mean (+/- SD)")
```

```{r, fig.width=8, fig.height=4}
ggplot(runtime_all, aes(Method, Seconds, colour = Method)) +
    geom_point(alpha = 0.45, position = position_jitter(width = 0.08, height = 0)) +
    stat_summary(fun = median, geom = "point", size = 2.6, colour = "black") +
    stat_summary(fun.min = min, fun.max = max, geom = "errorbar", width = 0.25, colour = "black") +
    scale_y_continuous(labels = scales::comma) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
    labs(title = "Runtime across simulations", x = "", y = "Seconds")
```

### Width Distributions

```{r, fig.width=8, fig.height=6}
width_data <- do.call(rbind, lapply(benchmark_results, function(res) {
    sets <- c(lapply(res$methods, `[[`, "dmrs"), list(Truth = res$truth))
    do.call(rbind, lapply(names(sets), function(method) {
        if (length(sets[[method]]) == 0L) return(NULL)
        data.frame(simulation_id = res$simulation_id, Method = method, Width = width(sets[[method]]))
    }))
}))

ggplot(width_data, aes(Method, Width, fill = Method)) +
    geom_violin(alpha = 0.55) +
    geom_boxplot(width = 0.18, alpha = 0.55, outlier.size = 0.4) +
    scale_y_log10(labels = scales::comma) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
    labs(title = "DMR width distributions pooled across simulations", x = "", y = "Width (bp, log scale)")
```

## Summary

```{r}
best_f1 <- truth_eval_all |>
    dplyr::group_by(Method) |>
    dplyr::summarise(Median_F1 = median(F1_regions), Median_IoU = median(IoU_bp), .groups = "drop")
cat(sprintf("- Best median region-level F1: **%s** (%.3f)\n", best_f1$Method[which.max(best_f1$Median_F1)], max(best_f1$Median_F1)))
cat(sprintf("- Best median base-pair IoU: **%s** (%.3f)\n", best_f1$Method[which.max(best_f1$Median_IoU)], max(best_f1$Median_IoU)))
```

## Session Info

```{r}
sessionInfo()
```
