---
title: "Benchmarking CMEnt on Simulated Microarray Data"
subtitle: "CMEnt vs comb-p, probeLasso, DMRcate, and Bumphunter"
author: "CMEnt Package"
date: "`r Sys.Date()`"
output: 
    BiocStyle::html_document:
        code_folding: hide
---

```{r, include=FALSE}
knitr::opts_chunk$set(
    results = "asis",
    echo = TRUE,
    warning = FALSE,
    message = FALSE,
    fig.width = 6,
    fig.height = 4,
    dpi = 300
)
```

## Introduction

This vignette provides an extended benchmark comparing CMEnt against multiple popular DMR detection methods:

- **CMEnt**: Correlation-based seed-guided assembly approach
- **comb-p**: Combines adjacent p-values using Stouffer-Liptak-Kechris correction
- **probeLasso**: Lasso-based feature selection for DMR detection
- **DMRcate**: Gaussian kernel smoothing approach
- **Bumphunter**: Permutation-based regional testing

We evaluate these methods on:

- Number and size distribution of DMRs
- Overlap and concordance between methods
- Biological relevance of detected regions

The dataset used is a simulated dataset, produced by running `simulateDMRs` on ch21 and ch22 subset of preprocessed beta values from 656 450K illumina microarray healthy blood samples found in GEO with identifier GSE40279, with 500 simulated DMRs of varying widths and effect sizes. The simulated dataset is available for download from Zenodo (see code below).
## Setup and Dependencies

```{r, message=FALSE}
try(devtools::load_all(), silent = TRUE)
suppressPackageStartupMessages({
    try(library(CMEnt), silent = TRUE)
    library(minfi)
    library(data.table)
    library(ggplot2)
    library(dplyr)
    library(GenomicRanges)
})

get_benchmark_file <- function(filename) {
    benchmark_file <- ""
    folder <- "simulated_microarray"
    if (getwd() == "notebooks") {
        benchmark_file <- file.path("benchmarks", folder, filename)
    } else if ("NAMESPACE" %in% dir()) {
        benchmark_file <- file.path("notebooks", "benchmarks", folder, filename)
    } else {
        benchmark_file <- file.path("benchmarks", folder, filename)
    }
    dir.create(dirname(benchmark_file), showWarnings = FALSE, recursive = TRUE)
    benchmark_file
}
```

### Install External Dependencies

```{r, eval=FALSE}
r <- getOption("repos")
if (is.null(r) || is.na(r["CRAN"]) || r["CRAN"] %in% c("@CRAN@", "")) {
  options(repos = c(CRAN = "https://cloud.r-project.org"))
}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

if (!requireNamespace("DMRcate", quietly = TRUE)) {
    BiocManager::install("DMRcate", update = FALSE, ask = FALSE)
}

if (!requireNamespace("bumphunter", quietly = TRUE)) {
    BiocManager::install("bumphunter", update = FALSE, ask = FALSE)
}

if (!requireNamespace("doParallel", quietly = TRUE)) {
    install.packages("doParallel")
}

if (!requireNamespace("tidyr", quietly = TRUE)) {
    install.packages("tidyr")
}

if (!requireNamespace("ChAMP", quietly = TRUE)) {
    BiocManager::install("ChAMP", update = FALSE, ask = FALSE)
}
```

**Note:** comb-p requires both the Python tool itself (combined-pvalues) and bedtools, installing using new conda env if not present:

```{bash, engine.opts='-l', eval=FALSE}
{ command -v comb-p >/dev/null 2>&1 && command -v bedtools >/dev/null 2>&1; } || { 
    echo >&2 "comb-p or bedtools not found. Installing with conda...";
    conda create -n cment_microarray_exp python=3.9
    conda activate cment_microarray_exp
    conda install -yc bioconda combined-pvalues bedtools
}
```

## Load Data

```{r}
sim_beta_path <- get_benchmark_file("sim_beta_500.tsv")
sim_truth_path <- get_benchmark_file("sim_dmrs_500.tsv")
array_type <- "450K"

if (!file.exists(sim_beta_path)) {
    curl::curl_download(
        url = "https://zenodo.org/records/19663986/files/microarray_sim_beta_500.tsv?download=1",
        destfile = sim_beta_path
    )
}
if (!file.exists(sim_truth_path)) {
    curl::curl_download(
        url = "https://zenodo.org/records/19663986/files/microarray_sim_dmrs_500.tsv?download=1",
        destfile = sim_truth_path
    )
}


genome <- "hg19"

beta_handler <- CMEnt::getBetaHandler(sim_beta_path, array = array_type, genome = genome)

sample_ids <- beta_handler$getBetaColNames()
pheno <- data.frame(
    Sample_Group = ifelse(
        grepl("Condition1", sample_ids, ignore.case = TRUE),
        "normal",
        "cancer"
    ),
    row.names = sample_ids,
    stringsAsFactors = FALSE
)

beta_mat <- as.matrix(beta_handler$getBeta())
locs <- beta_handler$getBetaLocs()
logit2 <- function(x) log2(x) - log2(1 - x)
mvalues <- logit2(beta_mat)
pheno$casecontrol <- pheno$Sample_Group == "cancer"
```

## Find Differentially Methylated Positions

```{r}
dmps <- suppressWarnings(dmpFinder(mvalues,
    pheno = pheno$casecontrol,
    type = "categorical",
    shrinkVar = TRUE
))
# sort DMPs by location
dmps <- dmps[!is.na(dmps$pval), ]
dmp_locs <- locs[rownames(locs) %in% rownames(dmps), , drop = FALSE]
dmps <- dmps[rownames(dmp_locs), , drop = FALSE]
dmps$pval_adj <- p.adjust(dmps$pval, method = "BH")

case_idx <- which(pheno$casecontrol)
ctrl_idx <- which(!pheno$casecontrol)
dmps$delta_beta <- rowMeans(beta_mat[rownames(dmps), case_idx, drop = FALSE]) -
    rowMeans(beta_mat[rownames(dmps), ctrl_idx, drop = FALSE])

sig_dmps <- dmps[dmps$pval_adj < 0.01 & abs(dmps$delta_beta) >= 0.05, ]
cat(sprintf("Found %d significant DMPs (FDR < 0.01, |delta beta| >= 0.05)\n", nrow(sig_dmps)))
dmps_gr <- makeGRangesFromDataFrame(
    data.frame(
        chr = dmp_locs$chr,
        start = dmp_locs$start,
        end = dmp_locs$end,
        pval = dmps$pval,
        qval = dmps$pval_adj,
        stat = dmps$qval
    ),
    keep.extra.columns = TRUE
)
names(dmps_gr) <- rownames(dmps)
```

## Method 1: CMEnt

```{r CMEnt}
cment_file <-  get_benchmark_file("dmrs.CMEnt.rds")
if (!file.exists(cment_file)) {
    cment_njobs <- max(1L, min(8L, future::availableCores() - 1L))
    options("CMEnt.verbose" = 1)
    options("CMEnt.njobs" = cment_njobs)
    cat(sprintf("CMEnt parallel jobs: %d\n", cment_njobs))
    cat(sprintf("future::availableCores(): %d\n", future::availableCores()))
    cat("Starting CMEnt::findDMRsFromSeeds()...\n")
    flush.console()
    start_time <- Sys.time()
    dmrs_cment <- CMEnt::findDMRsFromSeeds(
        beta = beta_handler,
        seeds = sig_dmps,
        pheno = pheno,
        sample_group_col = "Sample_Group",
        casecontrol_col = "casecontrol",
        max_bridge_seeds_gaps  = 1,
        max_bridge_extension_gaps = 3,
        min_seeds = 2,
        min_sites = 3,
        ext_site_delta_beta = 0.2,
        max_lookup_dist = 1000,
        max_pval = 0.05,
        entanglement = "weak",
        .score_dmrs = TRUE,
        njobs = cment_njobs,
        verbose = 1
    )
    end_time <- Sys.time()
    time_cment <- end_time - start_time
    saveRDS(list(dmrs = dmrs_cment, time = time_cment), cment_file)
    rm(beta_handler)
} else {
    ret <- readRDS(cment_file)
    dmrs_cment <- ret$dmrs
    time_cment <- ret$time
}

cat(sprintf("CMEnt: Found %d DMRs in %d seconds\n", length(dmrs_cment), round(as.numeric(time_cment, units = "secs"))))
```

## Method 2: comb-p

```{r comb-p}
combp_file <- get_benchmark_file("dmrs.combp.rds")

run_combp <- function(dmps_gr, seed = 500, dist = 500) {
    if (Sys.which("bedtools") == "") {
        warning("bedtools not found in PATH. comb-p requires bedtools to be installed.")
        warning("Install with: conda install -yc bioconda bedtools")
        return(GRanges())
    }

    if (Sys.which("comb-p") == "") {
        warning("comb-p not found in PATH.")
        warning("Install with: conda install -yc bioconda combined-pvalues")
        return(GRanges())
    }
    tmpdir <- tempdir()
    bed_file <- file.path(tmpdir, "dmps_for_combp.bed")
    bed_df <- data.frame(
        chrom = as.character(seqnames(dmps_gr)),
        start = start(dmps_gr) - 1,
        end = end(dmps_gr),
        pval = mcols(dmps_gr)$pval
    )
    bed_df <- bed_df[order(bed_df$chrom, bed_df$start), ]

    write.table(bed_df, bed_file,
        quote = FALSE, sep = "\t",
        row.names = FALSE, col.names = TRUE
    )

    cmd <- sprintf(
        "comb-p pipeline -c 4 --seed %d --dist %d -p %s --region-filter-p 0.01 --region-filter-n 3 %s",
        seed, dist, file.path(tmpdir, "output."), bed_file
    )

    tryCatch(
        {
            exit_code <- system(cmd, ignore.stdout = FALSE, ignore.stderr = FALSE)

            if (exit_code != 0) {
                warning(sprintf("comb-p exited with code %d", exit_code))
                return(GRanges())
            }

            regions_file <- paste0(file.path(tmpdir, "output."), "regions-p.bed.gz")
            if (file.exists(regions_file)) {
                regions <- fread(cmd = sprintf("zcat %s", regions_file))
                colnames(regions) <- c("chr", "start", "end", "min_p", "n_probes", "z_p", "z_sidak_p")

                regions <- regions[regions$z_sidak_p < 0.05 & regions$n_probes >= 3, ]

                dmrs_combp <- makeGRangesFromDataFrame(regions,
                    seqnames.field = "chr",
                    start.field = "start",
                    end.field = "end",
                    keep.extra.columns = TRUE
                )
                return(dmrs_combp)
            } else {
                warning("comb-p did not produce output file")
                return(GRanges())
            }
        },
        error = function(e) {
            warning(sprintf("comb-p failed: %s", e$message))
            GRanges()
        }
    )
}

if (!file.exists(combp_file)) {
    start_time <- Sys.time()
    dmrs_combp <- run_combp(dmps_gr)
    end_time <- Sys.time()
    time_combp <- end_time - start_time
    saveRDS(list(dmrs = dmrs_combp, time = time_combp), combp_file)
} else {
    ret <- readRDS(combp_file)
    dmrs_combp <- ret$dmrs
    time_combp <- ret$time
}

cat(sprintf("comb-p: Found %d DMRs in %d seconds\n", length(dmrs_combp), round(as.numeric(time_combp, units = "secs"))))
```

## Method 3: probeLasso
```{r probeLasso}
probelasso_file <- get_benchmark_file("dmrs.probeLasso.rds")

run_probelasso_champ <- function(beta_mat, pheno, array_type = "450K") {
    if (!requireNamespace("ChAMP", quietly = TRUE)) {
        warning("ChAMP not installed. probeLasso requires ChAMP.")
        warning("Install with: BiocManager::install('ChAMP')")
        return(GRanges())
    }

    library(ChAMP)

    tryCatch(
        {
            result <- champ.DMR(
                beta = beta_mat,
                pheno = pheno$Sample_Group,
                arraytype = array_type,
                method = "ProbeLasso",
                minProbes = 3,
                adjPvalDmr = 0.05,
                adjPvalProbe = 0.1,
                meanLassoRadius = 500,
                minDmrSep = 500,
                minDmrSize = 50,
                PDFplot = FALSE,
                Rplot = FALSE,
                resultsDir = file.path(tempdir(), "champ_probelasso")
            )

            pl_dmr <- result$ProbeLassoDMR
            if (is.null(pl_dmr) || nrow(pl_dmr) == 0) {
                return(GRanges())
            }

            # ChAMP output column names can differ by version.
            if (!"seqnames" %in% names(pl_dmr) && "dmrChrom" %in% names(pl_dmr)) {
                pl_dmr$seqnames <- pl_dmr$dmrChrom
            }
            if (!"start" %in% names(pl_dmr) && "dmrStart" %in% names(pl_dmr)) {
                pl_dmr$start <- pl_dmr$dmrStart
            }
            if (!"end" %in% names(pl_dmr) && "dmrEnd" %in% names(pl_dmr)) {
                pl_dmr$end <- pl_dmr$dmrEnd
            }

            return(makeGRangesFromDataFrame(pl_dmr,
                seqnames.field = "seqnames",
                start.field = "start",
                end.field = "end",
                keep.extra.columns = TRUE
            ))
        },
        error = function(e) {
            warning(sprintf("probeLasso (ChAMP) failed: %s", e$message))
            GRanges()
        }
    )
}

if (!file.exists(probelasso_file)) {
    start_time <- Sys.time()
    dmrs_probelasso <- run_probelasso_champ(beta_mat, pheno, array_type = "450K")
    end_time <- Sys.time()
    time_probelasso <- end_time - start_time
    saveRDS(list(dmrs = dmrs_probelasso, time = time_probelasso), probelasso_file)
} else {
    ret <- readRDS(probelasso_file)
    dmrs_probelasso <- ret$dmrs
    time_probelasso <- ret$time
}

cat(sprintf("probeLasso: Found %d DMRs in %d seconds\n", length(dmrs_probelasso), round(as.numeric(time_probelasso, units = "secs"))))
```

## Method 4: DMRcate

```{r dmrcate}
dmrcate_file <- get_benchmark_file("dmrs.dmrcate.rds")
if (!file.exists(dmrcate_file)) {
    library(DMRcate)
    start_time <- Sys.time()
    design <- model.matrix(~Sample_Group, data = pheno)

    dmrs_dmrcate <- tryCatch(
        {
            if (ncol(design) < 2L) {
                warning("DMRcate requires at least two groups in Sample_Group; returning empty set.")
                GenomicRanges::GRanges()
            } else {
                coef_col <- ncol(design)
                colnames(design)[1] <- "(Intercept)"
                if (coef_col == 2L) {
                    colnames(design)[2] <- "cancer_vs_normal"
                }

                myannotation <- cpg.annotate("array", mvalues,
                    what = "M", arraytype = "450K",
                    analysis.type = "differential", design = design,
                    coef = coef_col, fdr = 0.05
                )

                dmrcate_results <- dmrcate(myannotation, C = 2, lambda = 1000, min.cpgs = 3, betacutoff = 0.05)
                extractRanges(dmrcate_results, genome = "hg19")
            }
        },
        error = function(e) {
            warning(sprintf("DMRcate failed: %s. Returning empty set.", e$message))
            GenomicRanges::GRanges()
        }
    )

    end_time <- Sys.time()
    time_dmrcate <- end_time - start_time
    saveRDS(list(dmrs = dmrs_dmrcate, time = time_dmrcate), dmrcate_file)
} else {
    ret <- readRDS(dmrcate_file)
    dmrs_dmrcate <- ret$dmrs
    time_dmrcate <- ret$time
}

cat(sprintf("DMRcate: Found %d DMRs in %d seconds\n", length(dmrs_dmrcate), round(as.numeric(time_dmrcate, units = "secs"))))
```

## Method 5: Bumphunter

```{r bumphunter}
bumphunter_file <- get_benchmark_file("dmrs.bumphunter.rds")
if (!file.exists(bumphunter_file)) {
    suppressPackageStartupMessages({
        library(bumphunter)
        library(doParallel)
    })
    start_time <- Sys.time()
    bh_cores <- max(1L, min(4L, future::availableCores() - 1L))
    bh_b <- 100L
    bh_max_dmrs <- 1000000L
    bh_max_gap <- 2000L
    bh_min_l <- 2L
    bh_region_p <- 0.05

    doParallel::registerDoParallel(cores = bh_cores)
    on.exit({
        try(doParallel::stopImplicitCluster(), silent = TRUE)
        gc()
    }, add = TRUE)

    design <- model.matrix(~Sample_Group, data = pheno)
    bh_cluster <- bumphunter::clusterMaker(
        chr = as.character(locs$chr),
        pos = locs$start,
        maxGap = bh_max_gap
    )
    run_bumphunter_fit <- function(cutoff, B) {
        bumphunter(beta_mat,
            design = design,
            chr = as.character(locs$chr),
            pos = locs$start,
            cluster = bh_cluster,
            cutoff = cutoff,
            maxGap = max_gap,
            nullMethod = "bootstrap",
            B = B,
            type = "Beta",
            coef = 2
        )
    }
    has_informative_bumphunter_null <- function(fit) {
        null_values <- unlist(fit$null$value, use.names = FALSE)
        null_lengths <- unlist(fit$null$length, use.names = FALSE)
        length(null_values) > 0 &&
            any(abs(null_values) > 0) &&
            length(null_lengths) > 0 &&
            any(null_lengths > 0)
    }
    max_gap <- bh_max_gap
    dmrs_cutoff <- 0.1
    dmrs_cnt <- Inf
    iter <- 0L
    max_iter <- 10L
    while (dmrs_cnt > bh_max_dmrs && iter < max_iter) {
        iter <- iter + 1L
        dmrs <- run_bumphunter_fit(cutoff = dmrs_cutoff, B = 0)
        dmrs_cnt <- nrow(dmrs$table)
        dmrs_cutoff <- dmrs_cutoff * 1.2
        rm(dmrs)
        gc()
    }
    final_cutoff <- dmrs_cutoff / 1.2
    cat(sprintf(
        "Bumphunter: cutoff=%.4f, candidate_bumps=%d, minL=%d, maxGap=%d, cores=%d, B=%d\n",
        final_cutoff, dmrs_cnt, bh_min_l, max_gap, bh_cores, bh_b
    ))
    bumphunter_results <- run_bumphunter_fit(cutoff = final_cutoff, B = bh_b)
    bh_cutoff_floor <- 0.005
    bh_bootstrap_adjust_iter <- 0L
    bh_max_bootstrap_adjust <- 6L
    while (
        nrow(bumphunter_results$table) > 0 &&
            (all(is.na(bumphunter_results$table$p.value)) ||
                !has_informative_bumphunter_null(bumphunter_results)) &&
            final_cutoff > bh_cutoff_floor &&
            bh_bootstrap_adjust_iter < bh_max_bootstrap_adjust
    ) {
        prev_cutoff <- final_cutoff
        final_cutoff <- max(final_cutoff / 2, bh_cutoff_floor)
        if (identical(final_cutoff, prev_cutoff)) {
            break
        }
        bh_bootstrap_adjust_iter <- bh_bootstrap_adjust_iter + 1L
        cat(sprintf(
            "Bumphunter: bootstrap null degenerate at cutoff=%.4f; retrying with cutoff=%.4f\n",
            prev_cutoff, final_cutoff
        ))
        bumphunter_results <- run_bumphunter_fit(cutoff = final_cutoff, B = bh_b)
    }

    dmr_table <- bumphunter_results$table
    # bumphunter already estimates region-level p-values from the bootstrap null.
    dmr_table <- dmr_table[dmr_table$L >= bh_min_l & dmr_table$p.value <= bh_region_p, ]

    dmrs_bumphunter <- makeGRangesFromDataFrame(dmr_table,
        seqnames.field = "chr",
        start.field = "start",
        end.field = "end",
        keep.extra.columns = TRUE
    )
    end_time <- Sys.time()
    time_bumphunter <- end_time - start_time
    saveRDS(list(dmrs = dmrs_bumphunter, time = time_bumphunter), bumphunter_file)
} else {
    ret <- readRDS(bumphunter_file)
    dmrs_bumphunter <- ret$dmrs
    time_bumphunter <- ret$time
}

cat(sprintf("Bumphunter: Found %d DMRs in %d seconds\n", length(dmrs_bumphunter), round(as.numeric(time_bumphunter, units = "secs"))))
```

## Comparative Analysis

### Basic Statistics

```{r}
get_dmr_stats <- function(dmrs, method) {
    if (length(dmrs) == 0) {
        return(data.frame(
            Method = method,
            `Number of DMRs` = 0,
            `Mean DMR Width` = NA,
            `Median DMR Width` = NA,
            `Total Coverage bp` = 0,
            `Min Width` = NA,
            `Max Width` = NA
        ))
    }

    data.frame(
        Method = method,
        `Number of DMRs` = length(dmrs),
        `Mean DMR Width` = round(mean(width(dmrs)), 1),
        `Median DMR Width` = median(width(dmrs)),
        `Total Coverage bp` = sum(width(dmrs)),
        `Min Width` = min(width(dmrs)),
        `Max Width` = max(width(dmrs))
    )
}

stats_list <- list(
    get_dmr_stats(dmrs_cment, "CMEnt"),
    get_dmr_stats(dmrs_combp, "comb-p"),
    get_dmr_stats(dmrs_probelasso, "probeLasso"),
    get_dmr_stats(dmrs_dmrcate, "DMRcate"),
    get_dmr_stats(dmrs_bumphunter, "Bumphunter")
)

results_comparison <- do.call(rbind, stats_list)
knitr::kable(results_comparison, digits = 2, caption = "Basic DMR Statistics Across Methods")
```

### Pairwise Overlap Analysis

```{r, results='asis'}
methods_list <- list(
    CMEnt = dmrs_cment,
    `comb-p` = dmrs_combp,
    probeLasso = dmrs_probelasso,
    DMRcate = dmrs_dmrcate,
    Bumphunter = dmrs_bumphunter
)

methods_list <- methods_list[sapply(methods_list, length) > 0]

calculate_overlap <- function(gr1, gr2) {
    if (length(gr1) == 0 || length(gr2) == 0) {
        return(list(count = 0, percentage = 0))
    }
    overlap_count <- length(subsetByOverlaps(gr1, gr2))
    pct <- (overlap_count / length(gr1)) * 100
    list(count = overlap_count, percentage = pct)
}

overlap_matrix <- matrix(0, nrow = length(methods_list), ncol = length(methods_list))
pct_matrix <- matrix(0, nrow = length(methods_list), ncol = length(methods_list))
rownames(overlap_matrix) <- colnames(overlap_matrix) <- names(methods_list)
rownames(pct_matrix) <- colnames(pct_matrix) <- names(methods_list)

for (i in seq_along(methods_list)) {
    for (j in seq_along(methods_list)) {
        if (i == j) {
            overlap_matrix[i, j] <- length(methods_list[[i]])
            pct_matrix[i, j] <- 100
        } else {
            overlap_result <- calculate_overlap(methods_list[[i]], methods_list[[j]])
            overlap_matrix[i, j] <- overlap_result$count
            pct_matrix[i, j] <- overlap_result$percentage
        }
    }
}

cat("\n#### Overlap Counts (rows overlap with columns)\n")
knitr::kable(overlap_matrix, digits = 0, caption = "Number of DMRs from Method A (rows) overlapping with Method B (columns)")

cat("\n#### Overlap Percentages\n")
knitr::kable(pct_matrix, digits = 2, caption = "Percentage of DMRs from Method A (rows) overlapping with Method B (columns)")
```

### Jaccard Index and IoU

```{r}
jaccard_index <- function(gr1, gr2) {
    if (length(gr1) == 0 || length(gr2) == 0) {
        return(0)
    }
    overlap_count <- length(subsetByOverlaps(gr1, gr2))
    total_unique <- length(unique(c(gr1, gr2)))
    if (total_unique == 0) {
        return(0)
    }
    overlap_count / total_unique
}

iou <- function(gr1, gr2) {
    if (length(gr1) == 0 || length(gr2) == 0) {
        return(0)
    }
    intersection <- sum(width(GenomicRanges::intersect(gr1, gr2)))
    union <- sum(width(GenomicRanges::union(gr1, gr2)))
    if (union == 0) {
        return(0)
    }
    intersection / union
}

jaccard_matrix <- matrix(0, nrow = length(methods_list), ncol = length(methods_list))
iou_matrix <- matrix(0, nrow = length(methods_list), ncol = length(methods_list))
rownames(jaccard_matrix) <- colnames(jaccard_matrix) <- names(methods_list)
rownames(iou_matrix) <- colnames(iou_matrix) <- names(methods_list)

for (i in seq_along(methods_list)) {
    for (j in seq_along(methods_list)) {
        if (i < j) {
            jaccard_matrix[i, j] <- jaccard_matrix[j, i] <- jaccard_index(methods_list[[i]], methods_list[[j]])
            iou_matrix[i, j] <- iou_matrix[j, i] <- iou(methods_list[[i]], methods_list[[j]])
        } else if (i == j) {
            jaccard_matrix[i, j] <- 1
            iou_matrix[i, j] <- 1
        }
    }
}

cat("\n#### Jaccard Index\n")
knitr::kable(jaccard_matrix, digits = 4, caption = "Jaccard Index (region count-based similarity)")

cat("\n#### Intersection over Union (IoU)\n")
knitr::kable(iou_matrix, digits = 4, caption = "IoU (genomic coverage-based similarity)")
```

## Visualizations

### DMR Count and Coverage Comparison

```{r, fig.width=8, fig.height=3}
viz_data <- results_comparison |>
    dplyr::select(Method, Number.of.DMRs, Total.Coverage.bp) |>
    tidyr::pivot_longer(
        cols = c(Number.of.DMRs, Total.Coverage.bp),
        names_to = "Metric", values_to = "Value"
    )

viz_data$Metric <- ifelse(viz_data$Metric == "Number.of.DMRs",
    "Number of DMRs",
    "Total Coverage (bp)"
)

p1 <- ggplot(viz_data, aes(x = Method, y = Value, fill = Method)) +
    geom_bar(stat = "identity") +
    facet_wrap(~Metric, scales = "free_y") +
    theme_minimal() +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none",
        panel.grid.major.x = element_blank()
    ) +
    labs(
        title = "DMR Detection: Count and Coverage Comparison",
        y = "Value", x = ""
    )

print(p1)
```

### DMR Width Distribution

```{r}
width_data <- do.call(rbind, lapply(names(methods_list), function(method) {
    if (length(methods_list[[method]]) > 0) {
        data.frame(
            Method = method,
            Width = width(methods_list[[method]])
        )
    }
}))

p2 <- ggplot(width_data, aes(x = Method, y = Width, fill = Method)) +
    geom_violin(alpha = 0.7) +
    geom_boxplot(width = 0.2, alpha = 0.5, outlier.size = 0.5) +
    scale_y_log10(labels = scales::comma) +
    theme_minimal() +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none"
    ) +
    labs(
        title = "Distribution of DMR Widths",
        y = "Width (bp, log scale)", x = ""
    )

print(p2)
```

### Overlap Heatmap

```{r, fig.width=8, fig.height=6}
library(reshape2)

heatmap_text_color <- function(value, low, mid, high, midpoint, limits) {
    value <- pmin(pmax(value, limits[1]), limits[2])

    interpolate_color <- function(x) {
        if (!is.finite(x)) {
            return("#FFFFFF")
        }

        if (x <= midpoint) {
            ratio <- if (midpoint == limits[1]) 1 else (x - limits[1]) / (midpoint - limits[1])
            anchors <- grDevices::col2rgb(c(low, mid)) / 255
        } else {
            ratio <- if (midpoint == limits[2]) 1 else (x - midpoint) / (limits[2] - midpoint)
            anchors <- grDevices::col2rgb(c(mid, high)) / 255
        }

        rgb <- anchors[, 1] + pmin(pmax(ratio, 0), 1) * (anchors[, 2] - anchors[, 1])
        grDevices::rgb(rgb[1], rgb[2], rgb[3])
    }

    fill_colors <- vapply(value, interpolate_color, character(1))
    rgb <- t(grDevices::col2rgb(fill_colors) / 255)
    rgb_linear <- ifelse(rgb <= 0.04045, rgb / 12.92, ((rgb + 0.055) / 1.055)^2.4)
    luminance <- rgb_linear[, 1] * 0.2126 + rgb_linear[, 2] * 0.7152 + rgb_linear[, 3] * 0.0722
    black_contrast <- (luminance + 0.05) / 0.05
    white_contrast <- 1.05 / (luminance + 0.05)
    ifelse(white_contrast > black_contrast, "white", "black")
}

heatmap_levels <- names(methods_list)

overlap_df <- melt(pct_matrix)
colnames(overlap_df) <- c("Method1", "Method2", "Percentage")
overlap_df$Method1 <- factor(overlap_df$Method1, levels = rev(heatmap_levels))
overlap_df$Method2 <- factor(overlap_df$Method2, levels = heatmap_levels)
overlap_df$LabelColor <- heatmap_text_color(
    overlap_df$Percentage,
    low = "white", mid = "lightblue", high = "darkblue",
    midpoint = 50, limits = c(0, 100)
)

p3 <- ggplot(overlap_df, aes(x = Method2, y = Method1, fill = Percentage)) +
    geom_tile() +
    geom_text(aes(label = sprintf("%.1f%%", Percentage), color = LabelColor), size = 3) +
    scale_fill_gradient2(
        low = "white", mid = "lightblue", high = "darkblue",
        midpoint = 50, limits = c(0, 100)
    ) +
    scale_color_identity() +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(
        title = "DMR Overlap Matrix",
        subtitle = "% of row method's DMRs overlapping with column method",
        x = "", y = "", fill = "Overlap %"
    )

print(p3)
```

### Similarity Matrix (IoU)

```{r, fig.width=8, fig.height=6}
iou_df <- melt(iou_matrix)
colnames(iou_df) <- c("Method1", "Method2", "IoU")
iou_df$Method1 <- factor(iou_df$Method1, levels = rev(heatmap_levels))
iou_df$Method2 <- factor(iou_df$Method2, levels = heatmap_levels)
iou_df$LabelColor <- heatmap_text_color(
    iou_df$IoU,
    low = "white", mid = "yellow", high = "red",
    midpoint = 0.5, limits = c(0, 1)
)

p4 <- ggplot(iou_df, aes(x = Method2, y = Method1, fill = IoU)) +
    geom_tile() +
    geom_text(aes(label = sprintf("%.3f", IoU), color = LabelColor), size = 3) +
    scale_fill_gradient2(
        low = "white", mid = "yellow", high = "red",
        midpoint = 0.5, limits = c(0, 1)
    ) +
    scale_color_identity() +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(
        title = "Method Similarity Matrix (IoU)",
        subtitle = "Intersection over Union of genomic coverage",
        x = "", y = "", fill = "IoU"
    )

print(p4)
```

## Biological Validation

### Effect Size Analysis (Delta Beta)

```{r, results='asis'}
calculate_mean_delta_beta <- function(dmrs, beta_mat, pheno, locs, group_col = "Sample_Group", max_dmrs = 5000) {
    if (length(dmrs) == 0) {
        return(NA)
    }

    if (length(dmrs) > max_dmrs) {
        set.seed(42)
        dmrs <- dmrs[sample(length(dmrs), max_dmrs)]
    }

    case_samples <- colnames(beta_mat)[pheno[[group_col]] == "cancer"]
    control_samples <- colnames(beta_mat)[pheno[[group_col]] == "normal"]

    site_gr <- makeGRangesFromDataFrame(locs)
    names(site_gr) <- rownames(beta_mat)

    hits <- findOverlaps(site_gr, dmrs)
    if (length(hits) == 0) {
        return(NA)
    }

    site_idx <- queryHits(hits)
    dmr_idx <- subjectHits(hits)

    case_means <- rowMeans(beta_mat[site_idx, case_samples, drop = FALSE], na.rm = TRUE)
    control_means <- rowMeans(beta_mat[site_idx, control_samples, drop = FALSE], na.rm = TRUE)
    delta_per_site <- abs(case_means - control_means)

    dmr_deltas <- tapply(delta_per_site, dmr_idx, mean, na.rm = TRUE)
    mean(dmr_deltas, na.rm = TRUE)
}

effect_size_list <- lapply(names(methods_list), function(method) {
    data.frame(
        Method = method,
        Mean_Delta_Beta = calculate_mean_delta_beta(methods_list[[method]], beta_mat, pheno, locs)
    )
})

effect_size_df <- do.call(rbind, effect_size_list)
knitr::kable(effect_size_df, digits = 4, caption = "Mean Effect Size (Delta Beta) per Method")

ggplot(effect_size_df, aes(x = reorder(Method, -Mean_Delta_Beta), y = Mean_Delta_Beta, fill = Method)) +
    geom_bar(stat = "identity") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
    labs(title = "Mean Effect Size by Method", x = "", y = "Mean |Delta Beta|")
```

### Gene Annotation Analysis

```{r}
if (requireNamespace("TxDb.Hsapiens.UCSC.hg19.knownGene", quietly = TRUE)) {
    library(TxDb.Hsapiens.UCSC.hg19.knownGene)
    txdb <- TxDb.Hsapiens.UCSC.hg19.knownGene

    promoters_gr <- promoters(txdb, upstream = 2000, downstream = 500)
    genes_gr <- genes(txdb)

    annotate_dmrs <- function(dmrs, method_name) {
        if (length(dmrs) == 0) {
            return(data.frame(
                Method = method_name,
                Total_DMRs = 0,
                In_Promoters = 0,
                In_Gene_Bodies = 0,
                Intergenic = 0,
                Pct_Promoters = 0,
                Pct_Gene_Bodies = 0,
                Pct_Intergenic = 0
            ))
        }

        in_promoters <- sum(overlapsAny(dmrs, promoters_gr))
        in_genes <- sum(overlapsAny(dmrs, genes_gr))
        intergenic <- length(dmrs) - sum(overlapsAny(dmrs, c(promoters_gr, genes_gr)))

        data.frame(
            Method = method_name,
            Total_DMRs = length(dmrs),
            In_Promoters = in_promoters,
            In_Gene_Bodies = in_genes,
            Intergenic = intergenic,
            Pct_Promoters = round(100 * in_promoters / length(dmrs), 1),
            Pct_Gene_Bodies = round(100 * in_genes / length(dmrs), 1),
            Pct_Intergenic = round(100 * intergenic / length(dmrs), 1)
        )
    }

    gene_annotation_list <- lapply(names(methods_list), function(method) {
        annotate_dmrs(methods_list[[method]], method)
    })

    gene_annotation_df <- do.call(rbind, gene_annotation_list)
    knitr::kable(gene_annotation_df, caption = "DMR Gene Context Annotation")

    gene_plot_data <- gene_annotation_df |>
        dplyr::select(Method, Pct_Promoters, Pct_Gene_Bodies, Pct_Intergenic) |>
        tidyr::pivot_longer(cols = -Method, names_to = "Context", values_to = "Percentage")

    gene_plot_data$Context <- gsub("Pct_", "", gene_plot_data$Context)
    gene_plot_data$Context <- gsub("_", " ", gene_plot_data$Context)

    ggplot(gene_plot_data, aes(x = Method, y = Percentage, fill = Context)) +
        geom_bar(stat = "identity", position = "dodge") +
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        labs(title = "DMR Distribution by Gene Context", x = "", y = "Percentage of DMRs")
}
```

### site Context Distribution

```{r}
if (requireNamespace("IlluminaHumanMethylation450kanno.ilmn12.hg19", quietly = TRUE)) {
    library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
    anno <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)

    site_context_gr <- makeGRangesFromDataFrame(
        data.frame(
            chr = anno$chr,
            start = anno$pos,
            end = anno$pos,
            context = anno$Relation_to_Island
        ),
        keep.extra.columns = TRUE
    )
    names(site_context_gr) <- rownames(anno)

    get_site_context_distribution <- function(dmrs, method_name) {
        if (length(dmrs) == 0) {
            return(data.frame(Method = method_name, Context = NA, Count = 0, Percentage = 0))
        }

        overlapping <- subsetByOverlaps(site_context_gr, dmrs)
        context_counts <- table(mcols(overlapping)$context)

        data.frame(
            Method = method_name,
            Context = names(context_counts),
            Count = as.vector(context_counts),
            Percentage = round(100 * as.vector(context_counts) / sum(context_counts), 1)
        )
    }

    site_context_list <- lapply(names(methods_list), function(method) {
        get_site_context_distribution(methods_list[[method]], method)
    })

    site_context_df <- do.call(rbind, site_context_list)
    site_context_df <- site_context_df[!is.na(site_context_df$Context), ]

    ggplot(site_context_df, aes(x = Method, y = Percentage, fill = Context)) +
        geom_bar(stat = "identity", position = "stack") +
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        scale_fill_brewer(palette = "Set2") +
        labs(
            title = "site Context Distribution (Island/Shore/Shelf/OpenSea)",
            x = "", y = "Percentage of sites in DMRs"
        )
}
```

## Consensus Analysis

### UpSet Plot of Method Intersections

```{r, fig.width=8, fig.height=6}
if (requireNamespace("UpSetR", quietly = TRUE) && length(methods_list) >= 2) {
    library(UpSetR)

    all_dmrs_reduced <- GenomicRanges::reduce(do.call(c, unname(methods_list)))

    membership_matrix <- sapply(methods_list, function(method_dmrs) {
        as.integer(overlapsAny(all_dmrs_reduced, method_dmrs))
    })

    membership_df <- as.data.frame(membership_matrix)
    colnames(membership_df) <- names(methods_list)

    upset_plot <- upset(membership_df,
        sets = names(methods_list),
        order.by = "freq",
        main.bar.color = "steelblue",
        sets.bar.color = "darkgreen",
        text.scale = c(1.3, 1.3, 1.3, 1.3, 1.3, 0.9),
        point.size = 3,
        line.size = 1,
        show.numbers = "yes",
        number.angles = 90
    )

    # Recenter rotated number labels over bars (UpSetR fixes vjust = -1 internally).
    panel_idx <- which(upset_plot$Main_bar$layout$name == "panel")
    if (length(panel_idx) == 1) {
        panel_grob <- upset_plot$Main_bar$grobs[[panel_idx]]
        text_children <- names(panel_grob$children)[vapply(panel_grob$children, inherits, logical(1), what = "text")]
        if (length(text_children) >= 1) {
            text_grob <- panel_grob$children[[text_children[1]]]
            text_grob$hjust <- 0.5
            text_grob$vjust <- 0.5
            panel_grob$children[[text_children[1]]] <- text_grob
            upset_plot$Main_bar$grobs[[panel_idx]] <- panel_grob
        }
    }

    print(upset_plot)
}
```

### Consensus DMR Analysis

```{r}
find_consensus <- function(methods_list, min_methods = 2) {
    all_dmrs <- do.call(c, unname(methods_list))
    all_dmrs <- GenomicRanges::reduce(all_dmrs)

    overlap_matrix <- sapply(methods_list, function(method_dmrs) {
        overlapsAny(all_dmrs, method_dmrs)
    })
    consensus_counts <- rowSums(overlap_matrix)

    consensus_dmrs <- all_dmrs[consensus_counts >= min_methods]
    mcols(consensus_dmrs)$n_methods <- consensus_counts[consensus_counts >= min_methods]
    consensus_dmrs
}

consensus_2plus <- find_consensus(methods_list, min_methods = 2)
consensus_3plus <- find_consensus(methods_list, min_methods = 3)
consensus_4plus <- find_consensus(methods_list, min_methods = 4)
cat(sprintf("- Consensus DMRs (2+ methods): %d\n", length(consensus_2plus)))
cat(sprintf("- Consensus DMRs (3+ methods): %d\n", length(consensus_3plus)))
cat(sprintf("- Consensus DMRs (4+ methods): %d\n", length(consensus_4plus)))
```

```{r}
consensus_sets <- list(
    `2+ methods` = consensus_2plus,
    `3+ methods` = consensus_3plus,
    `4+ methods` = consensus_4plus
)

consensus_overlap_trend <- do.call(rbind, lapply(names(methods_list), function(method) {
    method_dmrs <- methods_list[[method]]
    overlap_counts <- sapply(consensus_sets, function(consensus_dmrs) {
        if (length(method_dmrs) == 0 || length(consensus_dmrs) == 0) {
            return(0L)
        }
        round(sum(overlapsAny(consensus_dmrs, method_dmrs)) / length(consensus_dmrs), 2)
    })

    data.frame(
        Method = method,
        Consensus_Level = factor(
            c("", names(consensus_sets)),
            levels = c("", names(consensus_sets))
        ),
        Overlapping_DMRs = c(1, overlap_counts)
    )
}))

ggplot(consensus_overlap_trend, aes(
    x = Consensus_Level,
    y = Overlapping_DMRs,
    color = Method,
    group = Method
)) +
    geom_line(linewidth = 1) +
    geom_point(size = 2) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
    labs(
        title = "Fraction of consensus captured by each method",
        x = "Consensus Level",
        y = "Overlapping method DMRs (%)",
        color = "Method"
    )
```

```{r}
capture_rate <- sapply(names(methods_list), function(method) {
    if (length(consensus_3plus) == 0) {
        return(0)
    }
    sum(overlapsAny(consensus_3plus, methods_list[[method]])) / length(consensus_3plus) * 100
})

capture_df <- data.frame(
    Method = names(capture_rate),
    Capture_Rate_Pct = capture_rate
)
capture_df <- capture_df[order(-capture_df$Capture_Rate_Pct), ]

knitr::kable(capture_df,
    digits = 1, row.names = FALSE,
    caption = "% of High-Confidence (3+ methods) DMRs Captured by Each Method"
)

ggplot(capture_df, aes(x = reorder(Method, Capture_Rate_Pct), y = Capture_Rate_Pct, fill = Method)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    theme_minimal() +
    theme(legend.position = "none") +
    geom_hline(yintercept = 100, linetype = "dashed", color = "red") +
    labs(
        title = "Capture Rate of High-Confidence DMRs",
        subtitle = "% of consensus (3+ methods) DMRs detected by each method",
        x = "", y = "Capture Rate (%)"
    )
```

### Method-Specific vs Consensus DMRs

```{r}
find_specific_dmrs <- function(target_method, methods_list) {
    target_dmrs <- methods_list[[target_method]]
    if (length(target_dmrs) == 0) {
        return(GRanges())
    }

    other_methods <- methods_list[names(methods_list) != target_method]
    all_others <- do.call(c, unname(other_methods))

    if (length(all_others) == 0) {
        return(target_dmrs)
    }

    specific <- target_dmrs[countOverlaps(target_dmrs, all_others) == 0]
    specific
}

specificity_df <- do.call(rbind, lapply(names(methods_list), function(method) {
    specific <- find_specific_dmrs(method, methods_list)
    shared_2plus <- methods_list[[method]][overlapsAny(methods_list[[method]], consensus_2plus)]
    shared_3plus <- methods_list[[method]][overlapsAny(methods_list[[method]], consensus_3plus)]

    data.frame(
        Method = method,
        Total_DMRs = length(methods_list[[method]]),
        Method_Specific = length(specific),
        Shared_2plus = length(shared_2plus),
        Shared_3plus = length(shared_3plus),
        Pct_Specific = round(100 * length(specific) / max(1, length(methods_list[[method]])), 1),
        Pct_Shared_2plus = round(100 * length(shared_2plus) / max(1, length(methods_list[[method]])), 1),
        Pct_Shared_3plus = round(100 * length(shared_3plus) / max(1, length(methods_list[[method]])), 1)
    )
}))

knitr::kable(specificity_df, caption = "Method-Specific vs Shared DMRs")

specificity_plot_data <- specificity_df |>
    dplyr::select(Method, Pct_Specific, Pct_Shared_2plus) |>
    tidyr::pivot_longer(cols = -Method, names_to = "Type", values_to = "Percentage")

specificity_plot_data$Type <- ifelse(specificity_plot_data$Type == "Pct_Specific",
    "Method-Specific", "Shared (2+ methods)"
)

ggplot(specificity_plot_data, aes(x = Method, y = Percentage, fill = Type)) +
    geom_bar(stat = "identity", position = "stack") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_fill_manual(values = c("Method-Specific" = "#E41A1C", "Shared (2+ methods)" = "#377EB8")) +
    labs(title = "Method-Specific vs Shared DMRs", x = "", y = "Percentage of DMRs", fill = "")
```

## Example DMR Locus Visualization

```{r}
if (length(consensus_3plus) > 0) {
    top_consensus <- consensus_3plus[order(-mcols(consensus_3plus)$n_methods)][1]

    extended_region <- GRanges(
        seqnames = seqnames(top_consensus),
        ranges = IRanges(
            start = max(1, start(top_consensus) - 2000),
            end = end(top_consensus) + 2000
        )
    )

    cat(sprintf(
        "\nExample locus: %s:%d-%d (detected by %d methods)\n",
        as.character(seqnames(top_consensus)),
        start(top_consensus), end(top_consensus),
        mcols(top_consensus)$n_methods
    ))

    method_colors <- c(
        "CMEnt" = "#E41A1C", "comb-p" = "#377EB8",
        "probeLasso" = "#4DAF4A", "DMRcate" = "#984EA3",
        "Bumphunter" = "#FF7F00"
    )

    plot_data <- do.call(rbind, lapply(names(methods_list), function(method) {
        overlapping <- subsetByOverlaps(methods_list[[method]], extended_region)
        if (length(overlapping) > 0) {
            data.frame(
                Method = method,
                start = start(overlapping),
                end = end(overlapping),
                y = which(names(methods_list) == method)
            )
        } else {
            NULL
        }
    }))

    if (!is.null(plot_data) && nrow(plot_data) > 0) {
        p_locus <- ggplot(plot_data) +
            geom_rect(aes(
                xmin = start, xmax = end,
                ymin = y - 0.4, ymax = y + 0.4,
                fill = Method
            ), alpha = 0.7) +
            geom_vline(
                xintercept = c(start(top_consensus), end(top_consensus)),
                linetype = "dashed", color = "gray40"
            ) +
            scale_y_continuous(
                breaks = seq_along(methods_list),
                labels = names(methods_list)
            ) +
            scale_fill_manual(values = method_colors) +
            theme_minimal() +
            theme(
                legend.position = "none",
                panel.grid.minor = element_blank()
            ) +
            labs(
                title = sprintf(
                    "Example Consensus DMR: %s:%d-%d",
                    as.character(seqnames(top_consensus)),
                    start(top_consensus), end(top_consensus)
                ),
                subtitle = sprintf("Detected by %d methods", mcols(top_consensus)$n_methods),
                x = "Genomic Position", y = ""
            )

        print(p_locus)
    }
}
```

## Method-Reported Region Significance

```{r}
significance_method_specs <- data.frame(
    Method = c("CMEnt", "comb-p", "probeLasso", "DMRcate", "Bumphunter"),
    Metric = c(
        "Not available",
        "Sidak-adjusted p-value",
        "Reported DMR p-value",
        "HMFDR",
        "FWER"
    ),
    Column = c(NA, "z_sidak_p", "dmrP", "HMFDR", "fwer"),
    stringsAsFactors = FALSE
)

method_dmrs <- list(
    CMEnt = dmrs_cment,
    `comb-p` = dmrs_combp,
    probeLasso = dmrs_probelasso,
    DMRcate = dmrs_dmrcate,
    Bumphunter = dmrs_bumphunter
)
method_colors_sig <- c(
    `comb-p` = "#377EB8",
    probeLasso = "#4DAF4A",
    DMRcate = "#984EA3",
    Bumphunter = "#FF7F00"
)

extract_region_significance <- function(dmrs, method, metric_label, value_col) {
    if (is.na(value_col) || length(dmrs) == 0 || !value_col %in% names(mcols(dmrs))) {
        return(NULL)
    }

    values <- suppressWarnings(as.numeric(mcols(dmrs)[[value_col]]))
    values <- values[is.finite(values)]
    if (length(values) == 0) {
        return(NULL)
    }

    data.frame(
        Method = method,
        Metric = metric_label,
        Reported_Error_Value = pmax(values, .Machine$double.xmin)
    )
}

region_significance_long <- do.call(rbind, Filter(Negate(is.null), lapply(seq_len(nrow(significance_method_specs)), function(i) {
    spec <- significance_method_specs[i, ]
    extract_region_significance(
        dmrs = method_dmrs[[spec$Method]],
        method = spec$Method,
        metric_label = spec$Metric,
        value_col = spec$Column
    )
})))

significance_method_specs$N_DMRs <- vapply(significance_method_specs$Method, function(method) {
    length(method_dmrs[[method]])
}, numeric(1))
significance_method_specs$Included_In_Plots <- vapply(seq_len(nrow(significance_method_specs)), function(i) {
    spec <- significance_method_specs[i, ]
    !is.na(spec$Column) && spec$Column %in% names(mcols(method_dmrs[[spec$Method]]))
}, logical(1))


region_significance_summary <- region_significance_long |>
    group_by(Method, Metric) |>
    summarise(
        N_DMRs = n(),
        Median_Reported_Value = signif(median(Reported_Error_Value), 3),
        Median_NegLog10 = round(median(-log10(Reported_Error_Value)), 2),
        Pct_Le_0_05 = round(100 * mean(Reported_Error_Value <= 0.05), 1),
        Pct_Le_0_01 = round(100 * mean(Reported_Error_Value <= 0.01), 1),
        .groups = "drop"
    )

knitr::kable(
    region_significance_summary,
    caption = "Summary of Method-Reported Region Significance Metrics"
)
```


```{r}
if (exists("region_significance_long") && nrow(region_significance_long) > 0) {
    region_significance_long$NegLog10_Reported_Error <- pmin(-log10(region_significance_long$Reported_Error_Value), 50)

    ggplot(region_significance_long, aes(x = Method, y = NegLog10_Reported_Error, fill = Method)) +
        geom_boxplot(outlier.alpha = 0.2, width = 0.65) +
        scale_fill_manual(values = method_colors_sig) +
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        labs(
            title = "Method-Reported Region Significance Strength",
            x = "", y = expression(paste("Capped ", -log[10], "(Reported Error Metric)")), fill = "Method"
        )
}
```

## Ground-Truth Evaluation

```{r}
evaluate_against_truth <- function(pred, truth) {
    n_pred <- length(pred)
    n_truth <- length(truth)
    pred_overlap_truth <- sum(overlapsAny(pred, truth))
    truth_overlap_pred <- sum(overlapsAny(truth, pred))
    fp <- n_pred - pred_overlap_truth
    fn <- n_truth - truth_overlap_pred

    precision <- ifelse(n_pred == 0, 0, pred_overlap_truth / n_pred)
    recall <- ifelse(n_truth == 0, 0, truth_overlap_pred / n_truth)
    f1 <- ifelse(precision + recall == 0, 0, 2 * precision * recall / (precision + recall))

    pred_red <- reduce(pred)
    truth_red <- reduce(truth)
    inter <- GenomicRanges::intersect(pred_red, truth_red)

    pred_bp <- sum(width(pred_red))
    truth_bp <- sum(width(truth_red))
    inter_bp <- ifelse(length(inter) == 0, 0, sum(width(inter)))
    union_bp <- pred_bp + truth_bp - inter_bp

    precision_bp <- ifelse(pred_bp == 0, 0, inter_bp / pred_bp)
    recall_bp <- ifelse(truth_bp == 0, 0, inter_bp / truth_bp)
    f1_bp <- ifelse(precision_bp + recall_bp == 0, 0, 2 * precision_bp * recall_bp / (precision_bp + recall_bp))
    iou_bp <- ifelse(union_bp == 0, 0, inter_bp / union_bp)

    data.frame(
        TP_regions = truth_overlap_pred,
        FP_regions = fp,
        FN_regions = fn,
        Precision_regions = precision,
        Recall_regions = recall,
        F1_regions = f1,
        Predicted_bp = pred_bp,
        Truth_bp = truth_bp,
        Intersect_bp = inter_bp,
        Precision_bp = precision_bp,
        Recall_bp = recall_bp,
        F1_bp = f1_bp,
        IoU_bp = iou_bp
    )
}

truth_eval <- do.call(rbind, lapply(names(methods_list), function(method) {
    out <- evaluate_against_truth(methods_list[[method]], truth_dmrs)
    out$Method <- method
    out
}))
truth_eval$Method <- factor(truth_eval$Method, levels = method_order)
truth_eval <- truth_eval[order(truth_eval$Method), , drop = FALSE]
truth_eval$Method <- as.character(truth_eval$Method)


truth_eval <- truth_eval |>
    dplyr::select(
        Method,
        TP_regions,
        FP_regions, FN_regions,
        Precision_regions, Recall_regions, F1_regions,
        Precision_bp, Recall_bp, F1_bp, IoU_bp
    )

regions_truth_eval <- truth_eval |>
    dplyr::select(Method, TP_regions, FP_regions, FN_regions, Precision_regions, Recall_regions, F1_regions) |>
    dplyr::rename_with(~ sub("_regions$", "", .), ends_with("_regions"))

bp_truth_eval <- truth_eval |>
    dplyr::select(Method, Precision_bp, Recall_bp, F1_bp, IoU_bp) |>
    dplyr::rename_with(~ sub("_bp$", "", .), ends_with("_bp"))
```
```{r}
knitr::kable(regions_truth_eval, digits = 4, caption = "Macro (Region-level) Performance Against Ground Truth")
```

```{r}
knitr::kable(bp_truth_eval, digits = 4, caption = "Micro (Base-pair-level) Performance Against Ground Truth")
```

### Overlap Matrices (Methods + Truth)

```{r}
all_sets <- c(methods_list, list(`Ground Truth` = truth_dmrs))

calculate_overlap <- function(gr1, gr2) {
    if (length(gr1) == 0 || length(gr2) == 0) {
        return(list(count = 0, percentage = 0))
    }
    overlap_count <- sum(overlapsAny(gr1, gr2))
    pct <- overlap_count / length(gr1) * 100
    list(count = overlap_count, percentage = pct)
}

overlap_count_mat <- matrix(0, nrow = length(all_sets), ncol = length(all_sets))
overlap_pct_mat <- matrix(0, nrow = length(all_sets), ncol = length(all_sets))
rownames(overlap_count_mat) <- colnames(overlap_count_mat) <- names(all_sets)
rownames(overlap_pct_mat) <- colnames(overlap_pct_mat) <- names(all_sets)

for (i in seq_along(all_sets)) {
    for (j in seq_along(all_sets)) {
        if (i == j) {
            overlap_count_mat[i, j] <- length(all_sets[[i]])
            overlap_pct_mat[i, j] <- 100
        } else {
            ov <- calculate_overlap(all_sets[[i]], all_sets[[j]])
            overlap_count_mat[i, j] <- ov$count
            overlap_pct_mat[i, j] <- ov$percentage
        }
    }
}

knitr::kable(overlap_count_mat, digits = 0, caption = "Overlap Counts (query=row, reference=column)")
```
```{r}
knitr::kable(overlap_pct_mat, digits = 2, caption = "Overlap Percentages (query=row, reference=column)")
```

### CMEnt Stage 1 Quality Assessment (Seed Linkage)
Assessing seed-space recoverability and seed capture efficiency.

Metrics explanation:

- Truth_DMRs_Total: Total number of ground truth DMRs.
- Truth_DMRs_With_Seeds: Number of truth DMRs that have at least one overlapping seed site.
- Truth_DMRs_Stage1Recoverable_2plusSeeds: Number of truth DMRs that have at least 2 overlapping seed sites, which is the minimum for CMEnt stage 1 to potentially recover the region.
- Truth_DMRs_Stage1Recoverable_With2plusCapturedSeeds: Number of truth DMRs that have at least 2 overlapping seed sites that also overlap CMEnt-predicted DMRs, indicating they are potentially recoverable and captured by the method.
- Truth_DMRs_Stage1Recoverable_FullyCapturedSeeds: Number of truth DMRs where all overlapping seed sites are captured by CMEnt-predicted DMRs, indicating ideal stage 1 capture.
- Seed_Capture_Rate_Pct_in_Truth: Percentage of seed sites that overlap truth DMRs that are also captured by CMEnt-predicted DMRs.
- Median_Seed_Capture_Pct_Per_Truth: For truth DMRs that have at least one overlapping seed site, the median percentage of those seed sites that are captured by CMEnt-predicted DMRs.
- Truth_DMRs_With_Missed_Seeds: Number of truth DMRs that have at least one overlapping seed site that is not captured by CMEnt-predicted DMRs.
- Median_Max_Missed_Run_Per_Truth: For truth DMRs with missed seeds, the median length of the longest consecutive run of missed seed sites.
- Max_Missed_Run_Observed: The maximum length of consecutive missed seed sites observed across all truth DMRs.

Metrics values:

```{r}
seed_gr <- sort(seed_gr)
mcols(seed_gr)$index <- seq_along(seed_gr)
seeds_in_truth <- subsetByOverlaps(seed_gr, truth_dmrs, ignore.strand = TRUE)

if (length(dmrs_cment) > 0) {
    seeds_in_dmrs <- subsetByOverlaps(seed_gr, dmrs_cment, ignore.strand = TRUE)
    seeds_in_truth_and_dmrs <- subsetByOverlaps(seeds_in_truth, dmrs_cment, ignore.strand = TRUE)
    seeds_in_truth_not_dmrs <- subsetByOverlaps(seeds_in_truth, dmrs_cment, ignore.strand = TRUE, invert = TRUE)
} else {
    seeds_in_dmrs <- seed_gr[0]
    seeds_in_truth_and_dmrs <- seed_gr[0]
    seeds_in_truth_not_dmrs <- seeds_in_truth
}

seed_truth_ov <- findOverlaps(seeds_in_truth, truth_dmrs, ignore.strand = TRUE)
truth_seed_total <- tabulate(subjectHits(seed_truth_ov), nbins = length(truth_dmrs))
seed_truth_hit_ov <- findOverlaps(seeds_in_truth_and_dmrs, truth_dmrs, ignore.strand = TRUE)
truth_seed_captured <- tabulate(subjectHits(seed_truth_hit_ov), nbins = length(truth_dmrs))
truth_seed_capture_pct <- ifelse(truth_seed_total > 0, round(100 * truth_seed_captured / truth_seed_total, 1), NA_real_)

recoverable_truth <- truth_seed_total >= 2L
recoverable_with2captured <- recoverable_truth & truth_seed_captured >= 2L
recoverable_fully_captured <- recoverable_truth & truth_seed_captured == truth_seed_total

stage1_quality <- data.frame(
    Truth_DMRs_Total = length(truth_dmrs),
    Truth_DMRs_With_Seeds = sum(truth_seed_total >= 1L),
    Truth_DMRs_Stage1Recoverable_2plusSeeds = sum(recoverable_truth),
    Truth_DMRs_Stage1Recoverable_With2plusCapturedSeeds = sum(recoverable_with2captured),
    Truth_DMRs_Stage1Recoverable_FullyCapturedSeeds = sum(recoverable_fully_captured),
    Seed_Capture_Rate_Pct_in_Truth = ifelse(length(seeds_in_truth) == 0, NA_real_, round(100 * length(seeds_in_truth_and_dmrs) / length(seeds_in_truth), 1)),
    Median_Seed_Capture_Pct_Per_Truth = ifelse(sum(truth_seed_total > 0) == 0, NA_real_, stats::median(truth_seed_capture_pct[truth_seed_total > 0], na.rm = TRUE))
)

for (col in colnames(stage1_quality)) {
    cat(sprintf("- **%s**: %s\n", col, stage1_quality[[col]]))
}

# Consecutive missed-seed runs are computed per truth DMR using original seed index.
ov_truth <- findOverlaps(seeds_in_truth_not_dmrs, truth_dmrs, ignore.strand = TRUE)
qh_by_truth <- split(queryHits(ov_truth), subjectHits(ov_truth))
run_df <- do.call(rbind, lapply(names(qh_by_truth), function(truth_id_chr) {
    truth_id <- as.integer(truth_id_chr)
    qh <- qh_by_truth[[truth_id_chr]]
    idx <- mcols(seeds_in_truth_not_dmrs)$index[qh]
    db <- as.numeric(mcols(seeds_in_truth_not_dmrs)$delta_beta[qh])

    ord <- order(idx)
    idx <- idx[ord]
    db <- db[ord]

    run_id <- cumsum(c(TRUE, diff(idx) != 1L))
    data.frame(
        Truth_DMR_Index = truth_id,
        run_length = as.integer(tabulate(run_id)),
        mean_delta_beta = as.numeric(tapply(db, run_id, mean, na.rm = TRUE))
    )
}))

run_count <- as.data.frame(table(run_df$run_length))
colnames(run_count) <- c("Consecutive_Seed_Length", "Count")
run_delta <- aggregate(mean_delta_beta ~ run_length, data = run_df, FUN = mean, na.rm = TRUE)
colnames(run_delta) <- c("Consecutive_Seed_Length", "Mean_Delta_Beta")
run_summary <- merge(run_count, run_delta, by = "Consecutive_Seed_Length", all.x = TRUE)

max_run_per_truth <- aggregate(run_length ~ Truth_DMR_Index, data = run_df, FUN = max)
run_overview <- data.frame(
    Truth_DMRs_With_Missed_Seeds = nrow(max_run_per_truth),
    Median_Max_Missed_Run_Per_Truth = stats::median(max_run_per_truth$run_length),
    Max_Missed_Run_Observed = max(max_run_per_truth$run_length)
)
for (col in colnames(run_overview)) {
    cat(sprintf("- **%s**: %s\n", col, run_overview[[col]]))
}

```

### CMEnt Stage 2-4 Quality Assessment (Extension and Merging)
Assessing truth-side recovery and prediction-side fragmentation/purity after extension and merging.
Metrics explanation:

- Pred_DMRs_Total: Total number of CMEnt DMRs.
- Pred_DMRs_Overlapping_Truth: Number of CMEnt DMRs that overlap at least one truth DMR.
- Pred_DMRs_BPPrecision_ge50pct: Number of CMEnt DMRs with base-pair precision >= 50% (i.e., at least 50% of the CMEnt DMR's base pairs overlap with truth DMRs).
- Pred_DMRs_BPPrecision_ge80pct: Number of CMEnt DMRs with base-pair precision >= 80%.
- Truth_DMRs_Total: Total number of truth DMRs.
- Truth_DMRs_Hit: Number of truth DMRs that have at least one overlapping CMEnt DMR.
- Truth_DMRs_BPRecall_ge50pct: Number of truth DMRs with base-pair recall >= 50% (i.e., at least 50% of the truth DMR's base pairs are covered by CMEnt DMRs).
- Truth_DMRs_BPRecall_ge80pct: Number of truth DMRs with base-pair recall >= 80%.
- Median_Truth_BPRecall_Pct_Hit: Median base-pair recall percentage among truth DMRs that have at least one overlapping CMEnt DMR.
- Median_Pred_BPPrecision_Pct_Overlap: Median base-pair precision percentage among CMEnt DMRs that have at least some overlap with truth DMRs.
- Truth_DMRs_With_Multiple_Pred: Number of truth DMRs that have 2 or more overlapping CMEnt DMRs, indicating potential fragmentation.
- Median_Preds_Per_Hit_Truth: Median number of overlapping CMEnt DMRs among truth DMRs that have at least one overlapping CMEnt DMR.
- Max_Preds_On_Truth: Maximum number of overlapping CMEnt DMRs observed on any single truth DMR.

Metrics values:

```{r}
truth_to_pred_hits <- findOverlaps(truth_dmrs, dmrs_cment, ignore.strand = TRUE)
preds_per_truth <- tabulate(queryHits(truth_to_pred_hits), nbins = length(truth_dmrs))

truth_overlap_bp <- numeric(length(truth_dmrs))
if (length(truth_to_pred_hits) > 0) {
    truth_intersections <- pintersect(
        truth_dmrs[queryHits(truth_to_pred_hits)],
        dmrs_cment[subjectHits(truth_to_pred_hits)]
    )
    truth_intersections_by_truth <- split(truth_intersections, queryHits(truth_to_pred_hits))
    truth_overlap_bp[as.integer(names(truth_intersections_by_truth))] <- vapply(
        truth_intersections_by_truth,
        function(gr) sum(width(reduce(gr))),
        numeric(1)
    )
}
truth_bp_recall <- ifelse(width(truth_dmrs) > 0, truth_overlap_bp / width(truth_dmrs), 0)

pred_to_truth_hits <- findOverlaps(dmrs_cment, truth_dmrs, ignore.strand = TRUE)
pred_overlap_bp <- numeric(length(dmrs_cment))
if (length(pred_to_truth_hits) > 0) {
    pred_intersections <- pintersect(
        dmrs_cment[queryHits(pred_to_truth_hits)],
        truth_dmrs[subjectHits(pred_to_truth_hits)]
    )
    pred_intersections_by_pred <- split(pred_intersections, queryHits(pred_to_truth_hits))
    pred_overlap_bp[as.integer(names(pred_intersections_by_pred))] <- vapply(
        pred_intersections_by_pred,
        function(gr) sum(width(reduce(gr))),
        numeric(1)
    )
}
pred_bp_precision <- ifelse(width(dmrs_cment) > 0, pred_overlap_bp / width(dmrs_cment), 0)

stage24_quality <- data.frame(
    Pred_DMRs_Total = length(dmrs_cment),
    Pred_DMRs_Overlapping_Truth = sum(overlapsAny(dmrs_cment, truth_dmrs)),
    Pred_DMRs_BPPrecision_ge50pct = sum(pred_bp_precision >= 0.5),
    Pred_DMRs_BPPrecision_ge80pct = sum(pred_bp_precision >= 0.8),
    Truth_DMRs_Total = length(truth_dmrs),
    Truth_DMRs_Hit = sum(preds_per_truth > 0),
    Truth_DMRs_BPRecall_ge50pct = sum(truth_bp_recall >= 0.5),
    Truth_DMRs_BPRecall_ge80pct = sum(truth_bp_recall >= 0.8),
    Median_Truth_BPRecall_Pct_Hit = ifelse(sum(preds_per_truth > 0) == 0, 0, 100 * stats::median(truth_bp_recall[preds_per_truth > 0])),
    Median_Pred_BPPrecision_Pct_Overlap = ifelse(sum(pred_bp_precision > 0) == 0, 0, 100 * stats::median(pred_bp_precision[pred_bp_precision > 0])),
    Truth_DMRs_With_Multiple_Pred = sum(preds_per_truth >= 2),
    Median_Preds_Per_Hit_Truth = ifelse(sum(preds_per_truth > 0) == 0, 0, stats::median(preds_per_truth[preds_per_truth > 0])),
    Max_Preds_On_Truth = ifelse(length(preds_per_truth) == 0, 0, max(preds_per_truth))
)
for (col in names(stage24_quality)) {
    cat(sprintf("- **%s**: %s\n", col, stage24_quality[[col]]))
}
```

## Visualizations

### Ground-Truth Metrics

```{r, fig.width=10, fig.height=5}
plot_df <- truth_eval |>
    dplyr::select(Method, F1_regions, F1_bp, IoU_bp) |>
    tidyr::pivot_longer(
        cols = c(F1_regions, F1_bp, IoU_bp),
        names_to = "Metric",
        values_to = "Value"
    )

plot_df$Metric <- factor(
    plot_df$Metric,
    levels = c("F1_regions", "F1_bp", "IoU_bp"),
    labels = c("F1 (Region)", "F1 (Base-pair)", "IoU (Base-pair)")
)
plot_df$Method <- factor(plot_df$Method, levels = method_order)

ggplot(plot_df, aes(x = Method, y = Value, fill = Method)) +
    geom_col() +
    facet_wrap(~Metric, nrow = 1) +
    theme_minimal() +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none"
    ) +
    ylim(0, 1) +
    labs(
        title = "Method Performance vs Ground Truth",
        x = "",
        y = "Score"
    )
```

### Width Distributions

```{r, fig.width=8, fig.height=6, dpi=300}
width_data <- do.call(rbind, lapply(names(all_sets), function(method) {
    x <- all_sets[[method]]
    if (length(x) == 0) {
        return(NULL)
    }
    data.frame(Method = method, Width = width(x))
}))
width_data$Method <- factor(width_data$Method, levels = names(all_sets))

ggplot(width_data, aes(x = Method, y = Width, fill = Method)) +
    geom_violin(alpha = 0.7) +
    geom_boxplot(width = 0.2, alpha = 0.5, outlier.size = 0.5) +
    scale_y_log10(labels = scales::comma) +
    theme_minimal() +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none"
    ) +
    labs(
        title = "DMR Width Distributions",
        x = "",
        y = "Width (bp, log scale)"
    )
```

## Summary

```{r}
cat("\n## Key Findings\n\n")

if (nrow(truth_eval) > 0) {
    best_f1 <- truth_eval[which.max(truth_eval$F1_regions), ]
    best_iou <- truth_eval[which.max(truth_eval$IoU_bp), ]

    cat(sprintf(
        "- Best region-level F1: **%s** (%.3f)\n",
        best_f1$Method, best_f1$F1_regions
    ))
    cat(sprintf(
        "- Best base-pair IoU: **%s** (%.3f)\n",
        best_iou$Method, best_iou$IoU_bp
    ))

    if ("CMEnt" %in% truth_eval$Method) {
        seg <- truth_eval[truth_eval$Method == "CMEnt", ]
        cat(sprintf(
            "- **CMEnt** region metrics: Precision = %.3f, Recall = %.3f, F1 = %.3f\n",
            seg$Precision_regions, seg$Recall_regions, seg$F1_regions
        ))
        cat(sprintf(
            "- **CMEnt** base-pair metrics: Precision = %.3f, Recall = %.3f, F1 = %.3f, IoU = %.3f\n",
            seg$Precision_bp, seg$Recall_bp, seg$F1_bp, seg$IoU_bp
        ))
    }
}
```

## Session Info

```{r}
sessionInfo()
```
