---
title: "Benchmarking CMEnt on Repeated Simulated ONT Data"
subtitle: "CMEnt vs DMRcate, DMRseq, and hummingbird across repeated simulateDMRs runs"
author: "CMEnt Package"
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
        code_folding: hide
---

```{r, include=FALSE}
knitr::opts_chunk$set(
    results = "asis",
    cache = TRUE,
    echo = TRUE,
    dev = "png",
    warning = TRUE,
    message = FALSE,
    fig.width = 8,
    fig.height = 6,
    dpi = 300
)
```

## Introduction

This vignette repeats the simulated ONT benchmark over multiple independently
simulated datasets. Each dataset is generated from the same augmented ONT BSseq
input with `CMEnt::simulateDMRs()`, using simulation parameters drawn under a
master seed of 42. Outputs are cached per simulation and method, and summary
plots show medians with observed interval ranges across simulations.

Methods:

- **CMEnt**: seed-guided, correlation-based DMR assembly
- **DMRcate**: Gaussian kernel smoothing-based DMR detection
- **DMRseq**: BSseq-native region-level inference
- **hummingbird**: HMM-based DMR detection

## Setup and Dependencies

```{r load_packages, warning=FALSE, message=FALSE, cache=FALSE}
.loadCMEnt <- function() {
    root <- if (file.exists("NAMESPACE")) "." else if (file.exists(file.path("..", "NAMESPACE"))) ".." else NULL
    if (!is.null(root) && requireNamespace("pkgload", quietly = TRUE)) {
        ok <- tryCatch({
            pkgload::load_all(root, quiet = TRUE)
            TRUE
        }, error = function(e) {
            message("Could not load local CMEnt source: ", conditionMessage(e))
            FALSE
        })
        if (ok) return(invisible(TRUE))
    }
    suppressPackageStartupMessages(library(CMEnt))
}
.loadCMEnt()
suppressPackageStartupMessages({
    library(bsseq)
    library(data.table)
    library(dplyr)
    library(GenomicRanges)
    library(ggplot2)
    library(SummarizedExperiment)
    library(S4Vectors)
})

getBenchmarkFile <- function(filename) {
    folder <- "simulated_ont_multisim"
    benchmark_file <- if (getwd() == "notebooks") {
        file.path("cached_results", folder, filename)
    } else if ("NAMESPACE" %in% dir()) {
        file.path("notebooks", "cached_results", folder, filename)
    } else {
        file.path("cached_results", folder, filename)
    }
    dir.create(dirname(benchmark_file), showWarnings = FALSE, recursive = TRUE)
    benchmark_file
}

getSimulationFile <- function(sim_id, filename) {
    getBenchmarkFile(file.path(sprintf("sim_%02d", sim_id), filename))
}

read_cache <- function(path, params = NULL) {
    if (!file.exists(path)) return(NULL)
    out <- readRDS(path)
    if (is.null(params) || identical(out$params, params)) out else NULL
}

save_cache <- function(path, value, params = NULL) {
    saveRDS(list(value = value, params = params), path)
}


nsim <- suppressWarnings(as.integer(Sys.getenv("CMENT_BENCHMARK_NSIM", "10")))
nsim <- ifelse(is.na(nsim) || nsim < 1L, 10L, nsim)
plan_file <- getBenchmarkFile(paste0("simulation_parameters", nsim, ".tsv"))
if (file.exists(plan_file)) {
    simulation_plan <- data.table::fread(plan_file)
} else {
    set.seed(42)
    simulation_plan <- data.frame(
        simulation_id = seq_len(nsim),
        seed = sample(0:1000, nsim),
        delta_beta_max = round(runif(nsim, 0.2, 0.5), 2),
        expected_corr = round(runif(nsim, 0.3, 0.7), 2)
    )

    data.table::fwrite(simulation_plan, plan_file, sep = "\t")
}
knitr::kable(simulation_plan, caption = "Simulation parameters generated under master seed 42")
```

### Install External Dependencies

```{r install_dependencies, warning=FALSE, message=FALSE, cache=FALSE}
if (!requireNamespace("benchmarkme", quietly = TRUE)) install.packages("benchmarkme")
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
if (!requireNamespace("DMRcate", quietly = TRUE)) BiocManager::install("DMRcate", update = FALSE, ask = FALSE)
if (!requireNamespace("dmrseq", quietly = TRUE)) BiocManager::install("dmrseq", update = FALSE, ask = FALSE)
if (!requireNamespace("hummingbird", quietly = TRUE)) BiocManager::install("hummingbird", update = FALSE, ask = FALSE)
if (!requireNamespace("tidyr", quietly = TRUE)) install.packages("tidyr")
```

## Load Input Data and Simulate

```{r simulate_data}
input_bsseq_path <- getBenchmarkFile("ont_bsseq_augmented.rds")
zenodo_url <- "https://zenodo.org/records/20592944/files/ont_bsseq_augmented.rds?download=1"
if (!file.exists(input_bsseq_path)) {
    curl::curl_download(zenodo_url, input_bsseq_path)
}
input_bsseq <- sort(readRDS(input_bsseq_path))
stopifnot(inherits(input_bsseq, "BSseq"))

base_simulation <- list(
    delta_jitter = 0.1,
    max_gap = 500L,
    min_dmr_size = 500L,
    max_dmr_size = 5000L,
    min_sites = 50L,
    max_sites = 2000L,
    num_dmrs = 1000L,
    genome = "hg38"
)

min_seeds <- 2L
min_sites <- 50L
max_pval <- 0.05
testing_mode <- "parametric"
ext_site_delta_beta <- 0.1
max_bridge_seeds_gaps <- 1L
max_bridge_extension_gaps <- 2L
max_lookup_dist <- 1000L

simulate_one <- function(plan_row) {
    sim_file <- getSimulationFile(plan_row$simulation_id, "simulation.rds")
    params <- c(base_simulation, list(
        seed = plan_row$seed,
        delta_beta_max = plan_row$delta_beta_max,
        expected_corr = plan_row$expected_corr
    ))
    cached <- read_cache(sim_file, params)
    if (!is.null(cached)) return(cached$value)
    cat(sprintf("Simulating dataset %d with seed %d, delta_beta_max %.3f, expected_corr %.3f", plan_row$simulation_id, plan_row$seed, plan_row$delta_beta_max, plan_row$expected_corr))

    set.seed(plan_row$seed)
    sim <- CMEnt::simulateDMRs(
        beta = input_bsseq,
        sample_group_col = "Sample_Group",
        num_dmrs = base_simulation$num_dmrs,
        delta_max0 = plan_row$delta_beta_max,
        delta_jitter = base_simulation$delta_jitter,
        expected_correlation = plan_row$expected_corr,
        max_gap = base_simulation$max_gap,
        min_dmr_size = base_simulation$min_dmr_size,
        max_dmr_size = base_simulation$max_dmr_size,
        min_sites = base_simulation$min_sites,
        max_sites = base_simulation$max_sites,
        verbose = 1
    )
    sim$pheno$Sample_ID <- rownames(sim$pheno)
    sim$pheno$casecontrol <- sim$pheno$Sample_Group == "case"
    SummarizedExperiment::colData(sim$simulated) <- S4Vectors::DataFrame(sim$pheno, row.names = sim$pheno$Sample_ID)
    data.table::fwrite(sim$truth, getSimulationFile(plan_row$simulation_id, "truth.tsv"), sep = "\t")
    data.table::fwrite(sim$pheno, getSimulationFile(plan_row$simulation_id, "samplesheet.tsv"), sep = "\t")
    save_cache(sim_file, sim, params)
    sim
}

NJOBS <- as.integer(
    Sys.getenv(
        "NJOBS",
        unset = as.character(max(1L, min(8L, parallel::detectCores(logical = TRUE) - 1L)))
    )
)
doParallel::registerDoParallel(cores = NJOBS)

benchmark_specs_file <- getBenchmarkFile("benchmark_specs.rds")
if (!file.exists(benchmark_specs_file)) {
    benchmark_specs <- list(
        cpu = benchmarkme::get_cpu(),
        ram = benchmarkme::get_ram(),
        platform = benchmarkme::get_platform_info(),
        rversion = R.version.string,
        njobs = NJOBS
    )
    saveRDS(benchmark_specs, benchmark_specs_file)
} else {
    benchmark_specs <- readRDS(benchmark_specs_file)
}
cpu <- benchmark_specs$cpu
ram <- benchmark_specs$ram
platform <- benchmark_specs$platform
rversion <- benchmark_specs$rversion
njobs <- benchmark_specs$njobs
cat("CPU:", cpu$vendor, cpu$model, "(", cpu$no_of_cores, "cores )\n")
cat("RAM:", capture.output(print(ram)), "\n")
cat("OS:", platform$OS.type, "\n")
cat("R Version:", rversion, "\n")
cat("Using", njobs, "parallel jobs\n")
```

## Method Runners

```{r method_helpers}
as_truth_gr <- function(sim) {
    makeGRangesFromDataFrame(
        sim$truth,
        seqnames.field = "seqnames",
        start.field = "start",
        end.field = "end",
        keep.extra.columns = TRUE,
        seqinfo = Seqinfo(genome = base_simulation$genome)
    )
}

prepare_bsseq_data <- function(sim, plan_row) {
    bs <- sort(sim$simulated)
    pheno <- as.data.frame(SummarizedExperiment::colData(bs))
    pheno$Sample_ID <- rownames(pheno)
    pheno$Sample_Group <- factor(pheno$Sample_Group, levels = c("control", "case"))
    pheno$casecontrol <- pheno$Sample_Group == "case"
    SummarizedExperiment::colData(bs) <- S4Vectors::DataFrame(pheno, row.names = pheno$Sample_ID)

    seeds_file <- getSimulationFile(plan_row$simulation_id, sprintf("seeds.p%s.tsv.gz", gsub("[^0-9]+", "", sprintf("%.3f", max_pval))))
    if (!file.exists(seeds_file)) {
        seeds <- CMEnt::findDMPsBSSeq(
            bsseq = bs,
            samplesheet = pheno,
            sample_group_col = "Sample_Group",
            id_col = "Sample_ID",
            case_group = "case",
            output_file = seeds_file,
            njobs = NJOBS
        )
    } else {
        seeds <- data.table::fread(seeds_file)
    }
    seeds <- as.data.frame(seeds, stringsAsFactors = FALSE)
    seeds <- seeds[seeds$qval < max_pval, , drop = FALSE]
    rownames(seeds) <- seeds$site_id
    list(bs = bs, pheno = pheno, seeds = seeds)
}

run_dmrcate_bsseq <- function(bs_obj, design, max_sites_per_chunk = 500000L, chunk_overlap_sites = 2000L) {
    if (!requireNamespace("DMRcate", quietly = TRUE)) return(GRanges())
    dmrcate_design <- as.matrix(design)
    colnames(dmrcate_design) <- gsub("Sample_Group", "", colnames(dmrcate_design))
    colnames(dmrcate_design)[1] <- "Intercept"
    dmrcate_design <- edgeR::modelMatrixMeth(dmrcate_design)
    coef_names <- colnames(dmrcate_design)
    sample_cols <- grepl("^Sample[0-9]+$", coef_names)
    contrast_candidates <- coef_names[!(sample_cols | coef_names %in% c("Intercept", "(Intercept)"))]
    if (length(contrast_candidates) == 0L) return(GRanges())
    selected_coef <- match(ifelse("case" %in% contrast_candidates, "case", tail(contrast_candidates, 1L)), coef_names)
    ann <- DMRcate::sequencing.annotate(bs_obj, all.cov = TRUE, methdesign = dmrcate_design, coef = selected_coef, fdr = max_pval)
    out <- DMRcate::extractRanges(
        DMRcate::dmrcate(ann, lambda = 1000, C = 2, min.cpgs = min_sites),
        genome = base_simulation$genome
    )
    out


}

run_dmrseq_bsseq <- function(bs_obj, max_sites_per_chunk = 500000L, chunk_overlap_sites = 2000L) {
    if (!requireNamespace("dmrseq", quietly = TRUE)) return(GRanges())
    materialize <- function(x) bsseq::BSseq(
        M = as.matrix(bsseq::getCoverage(x, type = "M")),
        Cov = as.matrix(bsseq::getCoverage(x, type = "Cov")),
        gr = bsseq::granges(x),
        pData = as.data.frame(SummarizedExperiment::colData(x)),
        sampleNames = colnames(x)
    )
    bpparam <- CMEnt:::.makeBiocParallelParam(max(1L, min(8L, parallel::detectCores(logical = TRUE) - 1L)), progressbar = TRUE)
    run_chunk <- function(chunk_bs, core_start_pos = NA_integer_, core_end_pos = NA_integer_) {
        tryCatch({
            fit <- dmrseq::dmrseq(
                bs = chunk_bs,
                testCovariate = "Sample_Group",
                cutoff = 0.1,
                minNumRegion = min_sites,
                verbose = FALSE,
                BPPARAM = bpparam # parallelization is handled at the chunks level, not within dmrseq
            )
            if (is.null(fit) || length(fit) == 0L) return(GRanges())
            qvals <- mcols(fit)$qval
            if (any(is.finite(qvals))) fit <- fit[is.finite(qvals) & qvals < 0.05]
            if (length(fit) > 0L && !is.na(core_start_pos) && !is.na(core_end_pos)) {
                mids <- floor((start(fit) + end(fit)) / 2)
                fit <- fit[mids >= core_start_pos & mids <= core_end_pos]
            }
            fit
        }, error = function(e) {
            warning(sprintf("dmrseq failed: %s", e$message))
            GRanges()
        })
    }
    dmrs_by_chr <- run_chunk(materialize(bs_obj))
    sort(dmrs_by_chr)
}

run_hummingbird_bsseq <- function(bs_obj, max_gap = 300L, min_sites = 50L, min_length = 50L) {
    if (!requireNamespace("hummingbird", quietly = TRUE)) return(GRanges())
    sample_df <- as.data.frame(SummarizedExperiment::colData(bs_obj))
    ctrl_idx <- which(sample_df$Sample_Group == "control")
    case_idx <- which(sample_df$Sample_Group == "case")
    if (length(ctrl_idx) == 0L || length(case_idx) == 0L) return(GRanges())
    row_gr <- bsseq::granges(bs_obj)
    row_gr <- GPos(seqnames = seqnames(row_gr), pos = start(row_gr))
    to_count_matrix <- function(x) {
        x <- as.matrix(x)
        x[!is.finite(x) | x < 0] <- 0
        storage.mode(x) <- "integer"
        x
    }
    all_seqnames <- as.character(seqnames(row_gr))
    chr_vec <- unique(all_seqnames[!is.na(all_seqnames) & nzchar(all_seqnames)])
    dmrs_by_chr <- lapply(chr_vec, function(chr_name) {
        chr_idx <- which(all_seqnames == chr_name)
        if (length(chr_idx) < as.integer(min_sites)) return(NULL)
        chr_bs <- bs_obj[chr_idx, , drop = FALSE]
        meth_mat <- bsseq::getCoverage(chr_bs, type = "M")
        cov_mat <- bsseq::getCoverage(chr_bs, type = "Cov")
        unmeth_mat <- cov_mat - meth_mat
        control_se <- SummarizedExperiment::SummarizedExperiment(
            rowRanges = row_gr[chr_idx],
            assays = list(normM = to_count_matrix(meth_mat[, ctrl_idx, drop = FALSE]), normUM = to_count_matrix(unmeth_mat[, ctrl_idx, drop = FALSE]))
        )
        case_se <- SummarizedExperiment::SummarizedExperiment(
            rowRanges = row_gr[chr_idx],
            assays = list(abnormM = to_count_matrix(meth_mat[, case_idx, drop = FALSE]), abnormUM = to_count_matrix(unmeth_mat[, case_idx, drop = FALSE]))
        )
        tryCatch({
            fit <- hummingbird::hummingbirdEM(experimentInfoControl = control_se, experimentInfoCase = case_se, binSize = 40)
            fit <- hummingbird::hummingbirdPostAdjustment(
                emInfo = fit,
                experimentInfoControl = control_se,
                experimentInfoCase = case_se,
                maxGap = as.integer(max_gap),
                minCpGs = as.integer(min_sites),
                minLength = as.integer(min_length)
            )
            dmrs <- fit$DMRs
            if (!inherits(dmrs, "GenomicRanges") || length(dmrs) == 0L) return(NULL)
            as(dmrs, "GRanges")
        }, error = function(e) {
            warning(sprintf("hummingbird failed (%s): %s", chr_name, e$message))
            NULL
        })
    })
    dmrs_by_chr <- Filter(function(x) !is.null(x) && length(x) > 0L, dmrs_by_chr)
    if (length(dmrs_by_chr) == 0L) {
        GRanges()
    } else {
        sort(unlist(GRangesList(dmrs_by_chr), use.names = FALSE))
    }
}

time_method <- function(expr) {
    start <- Sys.time()
    value <- force(expr)
    list(dmrs = value, time = Sys.time() - start)
}
```

## Run Benchmark

```{r run_benchmark, cache=FALSE}
run_one_simulation <- function(plan_row) {
    sim <- simulate_one(plan_row)
    truth_dmrs <- as_truth_gr(sim)
    dat <- prepare_bsseq_data(sim, plan_row)
    sim_dir <- sprintf("sim_%02d", plan_row$simulation_id)
    common <- list(delta_beta_max = plan_row$delta_beta_max, simulation_seed = plan_row$seed)

    method_specs <- list(
        CMEnt = list(
            params = c(common, list(
                min_seeds = min_seeds, min_sites = min_sites / 2,
                max_pval = max_pval, testing_mode = testing_mode,
                ext_site_delta_beta = ext_site_delta_beta,
                max_bridge_seeds_gaps = max_bridge_seeds_gaps,
                max_bridge_extension_gaps = max_bridge_extension_gaps,
                max_lookup_dist = max_lookup_dist
            )),
            run = function(params) time_method({
                beta_handler <- CMEnt::getBetaHandler(dat$bs)
                CMEnt::buildDMRs(
                    beta = beta_handler,
                    seeds = dat$seeds,
                    pheno = dat$pheno,
                    seeds_id_col = "site_id",
                    sample_group_col = "Sample_Group",
                    casecontrol_col = "casecontrol",
                    genome = base_simulation$genome,
                    array = "NULL",
                    entanglement = "weak",
                    testing_mode = testing_mode,
                    max_bridge_seeds_gaps = max_bridge_seeds_gaps,
                    max_bridge_extension_gaps = max_bridge_extension_gaps,
                    min_seeds = min_seeds,
                    min_sites = min_sites / 2,
                    max_pval = max_pval,
                    ext_site_delta_beta = ext_site_delta_beta,
                    max_lookup_dist = max_lookup_dist,
                    annotate_with_genes = FALSE,
                    extract_motifs = FALSE,
                    njobs = NJOBS
                )
            })
        ),
        DMRcate = list(
            params = c(common, list(min_sites = min_sites, max_pval = max_pval)),
            run = function(params) time_method(run_dmrcate_bsseq(dat$bs, model.matrix(~ Sample_Group, data = dat$pheno)))
        ),
        DMRseq = list(
            params = c(common, list(min_sites = min_sites, qval = 0.05)),
            run = function(params) time_method(run_dmrseq_bsseq(dat$bs))
        ),
        hummingbird = list(
            params = c(common, list(max_gap = 300L, min_sites = min_sites, min_length = 50L)),
            run = function(params) time_method(run_hummingbird_bsseq(dat$bs, max_gap = 300L, min_sites = min_sites, min_length = 50L))
        )
    )

    method_outputs <- lapply(names(method_specs), function(method) {
        spec <- method_specs[[method]]
        cache_file <- getBenchmarkFile(file.path(sim_dir, sprintf("dmrs.%s.rds", gsub("[^A-Za-z0-9]+", "_", method))))
        cached <- read_cache(cache_file, spec$params)
        out <- if (
            !is.null(cached) &&
                inherits(cached$value$dmrs, "GenomicRanges") &&
                length(cached$value$dmrs) > 0L
        ) {
            cached$value
        } else {
            cat(sprintf("Running method %s for simulation %d", method, plan_row$simulation_id))
            value <- spec$run(spec$params)
            if (length(value$dmrs) > 0L) {
                save_cache(cache_file, value, spec$params)
            }
            value
        }
        list(method = method, dmrs = out$dmrs, time = out$time)
    })
    names(method_outputs) <- names(method_specs)

    list(
        simulation_id = plan_row$simulation_id,
        params = plan_row,
        truth = truth_dmrs,
        seeds_n = nrow(dat$seeds),
        methods = method_outputs
    )
}

benchmark_results <- lapply(seq_len(nrow(simulation_plan)), function(i) {
    run_one_simulation(simulation_plan[i, , drop = FALSE])
})
```

```{r analysis_cache, include=FALSE}
knitr::opts_chunk$set(cache = FALSE)
```

## Comparative Analysis

```{r analysis_helpers, cache=FALSE}
get_dmr_stats <- function(dmrs, method, simulation_id) {
    data.frame(
        simulation_id = simulation_id,
        Method = method,
        Number_of_DMRs = length(dmrs),
        Median_DMR_Width = ifelse(length(dmrs) == 0L, NA_real_, median(width(dmrs))),
        Total_Coverage_bp = ifelse(length(dmrs) == 0L, 0, sum(width(reduce(dmrs)))),
        Min_Width = ifelse(length(dmrs) == 0L, NA_real_, min(width(dmrs))),
        Max_Width = ifelse(length(dmrs) == 0L, NA_real_, max(width(dmrs)))
    )
}

evaluate_against_truth <- function(pred, truth) {
    n_pred <- length(pred)
    n_truth <- length(truth)
    pred_hit <- sum(overlapsAny(pred, truth))
    truth_hit <- sum(overlapsAny(truth, pred))
    precision <- ifelse(n_pred == 0L, 0, pred_hit / n_pred)
    recall <- ifelse(n_truth == 0L, 0, truth_hit / n_truth)
    pred_red <- reduce(pred)
    truth_red <- reduce(truth)
    inter_bp <- sum(width(GenomicRanges::intersect(pred_red, truth_red)))
    pred_bp <- sum(width(pred_red))
    truth_bp <- sum(width(truth_red))
    union_bp <- pred_bp + truth_bp - inter_bp
    precision_bp <- ifelse(pred_bp == 0L, 0, inter_bp / pred_bp)
    recall_bp <- ifelse(truth_bp == 0L, 0, inter_bp / truth_bp)
    data.frame(
        TP_regions = truth_hit,
        FP_regions = n_pred - pred_hit,
        FN_regions = n_truth - truth_hit,
        Precision_regions = precision,
        Recall_regions = recall,
        F1_regions = ifelse(precision + recall == 0, 0, 2 * precision * recall / (precision + recall)),
        Precision_bp = precision_bp,
        Recall_bp = recall_bp,
        F1_bp = ifelse(precision_bp + recall_bp == 0, 0, 2 * precision_bp * recall_bp / (precision_bp + recall_bp)),
        IoU_bp = ifelse(union_bp == 0L, 0, inter_bp / union_bp)
    )
}

format_mean_pm <- function(x, digits = 2L) {
    x <- x[is.finite(x)]
    if (length(x) == 0L) return(NA_character_)
    sd_x <- if (length(x) > 1L) stats::sd(x) else 0
    sprintf(paste0("%.", digits, "f (+/- %.", digits, "f)"), mean(x), sd_x)
}

summarise_mean_pm <- function(df, group_cols, value_cols, digits = 2L) {
    metric_digits <- if (length(digits) == 1L) {
        stats::setNames(rep(digits, length(value_cols)), value_cols)
    } else {
        stats::setNames(as.integer(digits), names(digits))
    }
    df |>
        tidyr::pivot_longer(dplyr::all_of(value_cols), names_to = "Metric", values_to = "Value") |>
        dplyr::group_by(dplyr::across(dplyr::all_of(c(group_cols, "Metric")))) |>
        dplyr::summarise(
            Value = format_mean_pm(Value, digits = metric_digits[dplyr::first(Metric)]),
            .groups = "drop"
        ) |>
        tidyr::pivot_wider(names_from = Metric, values_from = Value)
}

summarise_benchmark_table <- function(stats_all, runtime_all) {
    stats_all |>
        dplyr::select(
            simulation_id,
            Method,
            dmrs_number = Number_of_DMRs,
            median_width_bp = Median_DMR_Width,
            min_width_bp = Min_Width,
            max_width_bp = Max_Width,
            Total_Coverage_bp
        ) |>
        dplyr::mutate(total_coverage_kb = Total_Coverage_bp / 1000) |>
        dplyr::select(-Total_Coverage_bp) |>
        dplyr::left_join(
            runtime_all |>
                dplyr::rename(runtime_sec = Seconds),
            by = c("simulation_id", "Method")
        ) |>
        dplyr::rename(method = Method) |>
        dplyr::group_by(method) |>
        dplyr::summarise(
            dmrs_number = format_mean_pm(dmrs_number, digits = 0L),
            median_width_bp = format_mean_pm(median_width_bp, digits = 0L),
            min_width_bp = format_mean_pm(min_width_bp, digits = 0L),
            max_width_bp = format_mean_pm(max_width_bp, digits = 0L),
            total_coverage_kb = format_mean_pm(total_coverage_kb, digits = 1L),
            runtime_sec = format_mean_pm(runtime_sec, digits = 1L),
            .groups = "drop"
        )
}

stats_all <- do.call(rbind, lapply(benchmark_results, function(res) {
    rbind(
        do.call(rbind, lapply(res$methods, function(x) get_dmr_stats(x$dmrs, x$method, res$simulation_id))),
        get_dmr_stats(res$truth, "Truth", res$simulation_id)
    )
}))

truth_eval_all <- do.call(rbind, lapply(benchmark_results, function(res) {
    do.call(rbind, lapply(res$methods, function(x) {
        out <- evaluate_against_truth(x$dmrs, res$truth)
        out$simulation_id <- res$simulation_id
        out$Method <- x$method
        out
    }))
}))

runtime_all <- do.call(rbind, lapply(benchmark_results, function(res) {
    do.call(rbind, lapply(res$methods, function(x) {
        data.frame(simulation_id = res$simulation_id, Method = x$method, Seconds = as.numeric(x$time, units = "secs"))
    }))
}))
```

### Basic Statistics

```{r}
stats_summary <- summarise_benchmark_table(stats_all, runtime_all)
knitr::kable(stats_summary, caption = "Benchmark statistics across simulations, mean (+/- SD)")
```

```{r, fig.width=9, fig.height=4}
stats_plot <- stats_all |>
    dplyr::select(Method, simulation_id, Number_of_DMRs, Total_Coverage_bp) |>
    tidyr::pivot_longer(c(Number_of_DMRs, Total_Coverage_bp), names_to = "Metric", values_to = "Value")

ggplot(stats_plot, aes(Method, Value, colour = Method)) +
    geom_point(alpha = 0.45, position = position_jitter(width = 0.08, height = 0)) +
    stat_summary(fun = median, geom = "point", size = 2.6, colour = "black") +
    stat_summary(fun.min = min, fun.max = max, geom = "errorbar", width = 0.25, colour = "black") +
    facet_wrap(~ Metric, scales = "free_y") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
    labs(title = "DMR counts and coverage across simulations", x = "", y = "Value")
```

## Ground-Truth Evaluation

```{r}
regions_truth_eval <- summarise_mean_pm(
    truth_eval_all,
    "Method",
    c("TP_regions", "FP_regions", "FN_regions", "Precision_regions", "Recall_regions", "F1_regions"),
    digits = c(TP_regions = 0L, FP_regions = 0L, FN_regions = 0L, Precision_regions = 2L, Recall_regions = 2L, F1_regions = 2L)
)
bp_truth_eval <- summarise_mean_pm(
    truth_eval_all,
    "Method",
    c("Precision_bp", "Recall_bp", "F1_bp", "IoU_bp"),
    digits = 2
)
knitr::kable(regions_truth_eval, caption = "Macro region-level performance against ground truth, mean (+/- SD)")
```

```{r}
knitr::kable(bp_truth_eval, caption = "Micro base-pair-level performance against ground truth, mean (+/- SD)")
```

```{r, fig.width=10, fig.height=5}
metric_plot <- truth_eval_all |>
    dplyr::select(Method, simulation_id, F1_regions, F1_bp, IoU_bp) |>
    tidyr::pivot_longer(c(F1_regions, F1_bp, IoU_bp), names_to = "Metric", values_to = "Value")
metric_plot$Metric <- factor(metric_plot$Metric, levels = c("F1_regions", "F1_bp", "IoU_bp"), labels = c("F1 (Region)", "F1 (Base-pair)", "IoU (Base-pair)"))

ggplot(metric_plot, aes(Method, Value, colour = Method)) +
    geom_point(alpha = 0.5, position = position_jitter(width = 0.08, height = 0)) +
    stat_summary(fun = median, geom = "point", size = 2.6, colour = "black") +
    stat_summary(fun.min = min, fun.max = max, geom = "errorbar", width = 0.25, colour = "black") +
    facet_wrap(~ Metric, nrow = 1) +
    coord_cartesian(ylim = c(0, 1)) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
    labs(title = "Method performance vs ground truth across simulations", x = "", y = "Score")
```

### Seed Recovery

```{r}
seed_summary <- do.call(rbind, lapply(benchmark_results, function(res) {
    data.frame(simulation_id = res$simulation_id, Seeds = res$seeds_n, Truth_DMRs = length(res$truth))
}))
knitr::kable(summarise_mean_pm(seed_summary, character(0), c("Seeds", "Truth_DMRs"), digits = 2), caption = "Seed and truth-set sizes, mean (+/- SD)")
```

```{r, fig.width=7, fig.height=4}
ggplot(seed_summary, aes(simulation_id, Seeds)) +
    geom_line() +
    geom_point() +
    theme_minimal() +
    labs(title = "CMEnt seed counts by simulation", x = "Simulation", y = "Seeds passing FDR threshold")
```

### Runtime

```{r}
runtime_summary <- summarise_mean_pm(runtime_all, "Method", "Seconds", digits = 2)
knitr::kable(runtime_summary, caption = "Runtime across simulations, mean (+/- SD)")
```

```{r, fig.width=8, fig.height=4}
ggplot(runtime_all, aes(Method, Seconds, colour = Method)) +
    geom_point(alpha = 0.45, position = position_jitter(width = 0.08, height = 0)) +
    stat_summary(fun = median, geom = "point", size = 2.6, colour = "black") +
    stat_summary(fun.min = min, fun.max = max, geom = "errorbar", width = 0.25, colour = "black") +
    scale_y_continuous(labels = scales::comma) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
    labs(title = "Runtime across simulations", x = "", y = "Seconds")
```

### Width Distributions

```{r, fig.width=8, fig.height=6}
width_data <- do.call(rbind, lapply(benchmark_results, function(res) {
    sets <- c(lapply(res$methods, `[[`, "dmrs"), list(Truth = res$truth))
    do.call(rbind, lapply(names(sets), function(method) {
        if (length(sets[[method]]) == 0L) return(NULL)
        data.frame(simulation_id = res$simulation_id, Method = method, Width = width(sets[[method]]))
    }))
}))

ggplot(width_data, aes(Method, Width, fill = Method)) +
    geom_violin(alpha = 0.55) +
    geom_boxplot(width = 0.18, alpha = 0.55, outlier.size = 0.4) +
    scale_y_log10(labels = scales::comma) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
    labs(title = "DMR width distributions pooled across simulations", x = "", y = "Width (bp, log scale)")
```

## Summary

```{r}
best_f1 <- truth_eval_all |>
    dplyr::group_by(Method) |>
    dplyr::summarise(Median_F1 = median(F1_regions), Median_IoU = median(IoU_bp), .groups = "drop")
cat(sprintf("- Best median region-level F1: **%s** (%.3f)\n", best_f1$Method[which.max(best_f1$Median_F1)], max(best_f1$Median_F1)))
cat(sprintf("- Best median base-pair IoU: **%s** (%.3f)\n", best_f1$Method[which.max(best_f1$Median_IoU)], max(best_f1$Median_IoU)))
```

## Session Info

```{r}
sessionInfo()
```
