---
title: "Benchmarking CMEnt on Simulated Microarray Data"
subtitle: "CMEnt vs comb-p, probeLasso, DMRcate, and Bumphunter"
author: "CMEnt Package"
date: "`r Sys.Date()`"
output: 
    BiocStyle::html_document:
        code_folding: hide
---

```{r, include=FALSE}
knitr::opts_chunk$set(
    results = "asis",
    cache = TRUE,
    echo = TRUE,
    dev = "png",
    warning = FALSE,
    message = FALSE,
    fig.width = 8,
    fig.height = 6,
    dpi = 300
)
```

## Introduction

This vignette provides an extended benchmark comparing CMEnt against multiple popular DMR detection methods on a simulated microarray dataset with known ground truth. The methods compared include:

- **CMEnt**: Correlation-based seed-guided assembly approach
- **comb-p**: Combines adjacent p-values using Stouffer-Liptak-Kechris correction
- **probeLasso**: Lasso-based feature selection for DMR detection
- **DMRcate**: Gaussian kernel smoothing approach
- **Bumphunter**: Permutation-based regional testing

The dataset used is a simulated dataset, produced by running `simulateDMRs`, on a chr20 and chr21 subset of preprocessed beta values from 656 450K illumina microarray healthy blood samples found in GEO with identifier GSE40279, with 500 simulated DMRs of varying widths and effect sizes. The simulated dataset is available for download from Zenodo (see code below).

## Setup and Dependencies

```{r, message=FALSE}
try(devtools::load_all(), silent = TRUE)
suppressPackageStartupMessages({
    try(library(CMEnt), silent = TRUE)
    library(minfi)
    library(data.table)
    library(ggplot2)
    library(dplyr)
    library(GenomicRanges)
})

getBenchmarkFile <- function(filename) {
    benchmark_file <- ""
    folder <- "simulated_microarray450k"
    if (getwd() == "notebooks") {
        benchmark_file <- file.path("cached_results", folder, filename)
    } else if ("NAMESPACE" %in% dir()) {
        benchmark_file <- file.path("notebooks", "cached_results", folder, filename)
    } else {
        benchmark_file <- file.path("cached_results", folder, filename)
    }
    dir.create(dirname(benchmark_file), showWarnings = FALSE, recursive = TRUE)
    benchmark_file
}

get_method_cache_file <- function(method) {
    getBenchmarkFile(sprintf("dmrs.%s.rds", method))
}

read_method_cache <- function(cache_file, params) {
    if (!file.exists(cache_file)) {
        return(NULL)
    }
    ret <- readRDS(cache_file)
    if (identical(ret$params, params)) ret else NULL
}

save_method_cache <- function(cache_file, dmrs, time, params) {
    saveRDS(list(dmrs = dmrs, time = time, params = params), cache_file)
}
```

### Install External Dependencies

```{r setup_dependencies, message=FALSE, warning=FALSE}
r <- getOption("repos")
if (is.null(r) || is.na(r["CRAN"]) || r["CRAN"] %in% c("@CRAN@", "")) {
  options(repos = c(CRAN = "https://cloud.r-project.org"))
}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

if (!requireNamespace("DMRcate", quietly = TRUE)) {
    BiocManager::install("DMRcate", update = FALSE, ask = FALSE)
}

if (!requireNamespace("bumphunter", quietly = TRUE)) {
    BiocManager::install("bumphunter", update = FALSE, ask = FALSE)
}

if (!requireNamespace("doParallel", quietly = TRUE)) {
    install.packages("doParallel")
}

if (!requireNamespace("tidyr", quietly = TRUE)) {
    install.packages("tidyr")
}

if (!requireNamespace("ChAMP", quietly = TRUE)) {
    BiocManager::install("ChAMP", update = FALSE, ask = FALSE)
}
```

**Note:** comb-p requires both the Python tool itself (combined-pvalues) and bedtools, installing using new conda env if not present:

```{bash, engine.opts='-l', eval=FALSE}
{ command -v comb-p >/dev/null 2>&1 && command -v bedtools >/dev/null 2>&1; } || { 
    echo >&2 "comb-p or bedtools not found. Installing with conda...";
    conda create -n cment_microarray_exp python=3.9
    conda activate cment_microarray_exp
    conda install -yc bioconda combined-pvalues bedtools
}
```

## Load Data

```{r}
sim_beta_path <- getBenchmarkFile("GSE40279_sim_beta_values.tsv.gz")
sim_truth_path <- getBenchmarkFile("GSE40279_sim_dmrs.tsv")
sim_pheno_path <- getBenchmarkFile("GSE40279_sim_samplesheet.tsv")
array_type <- "450K"
genome <- "hg19"
simulation <- list(
    delta_max0 = 0.3,
    expected_correlation = 0.7,
    max_gap = 500L,
    min_dmr_size = 50L,
    max_dmr_size = 2000L,
    min_sites = 5L,
    max_sites = 500L,
    num_dmrs = 500L,
    genome = genome
)
cat("Simulation parameters:\n")
for (param in names(simulation)) {
    cat(sprintf("- %s: %s\n", param, simulation[[param]]))
}
zenodo_base_url <- "https://zenodo.org/records/20592944/files/"
if (!file.exists(sim_beta_path)) {
    curl::curl_download(
        url = paste0(zenodo_base_url, basename(sim_beta_path), "?download=1"),
        destfile = sim_beta_path
    )
}
if (!file.exists(sim_truth_path)) {
    curl::curl_download(
        url = paste0(zenodo_base_url, basename(sim_truth_path), "?download=1"),
        destfile = sim_truth_path
    )
}
if (!file.exists(sim_pheno_path)) {
    curl::curl_download(
        url = paste0(zenodo_base_url, basename(sim_pheno_path), "?download=1"),
        destfile = sim_pheno_path
    )
}

truth_df <- data.table::fread(sim_truth_path)
truth_dmrs <- makeGRangesFromDataFrame(
    truth_df,
    seqnames.field = "seqnames",
    start.field = "start",
    end.field = "end",
    keep.extra.columns = TRUE
)

beta_handler <- CMEnt::getBetaHandler(sim_beta_path, array = array_type, genome = genome)

sample_ids <- beta_handler$getBetaColNames()
pheno <- data.table::fread(sim_pheno_path, data.table = FALSE)
rownames(pheno) <- pheno$V1

if (!all(sample_ids %in% rownames(pheno))) {
    stop("The simulated phenotype file does not contain all simulated beta samples.")
}
pheno <- pheno[sample_ids, , drop = FALSE]
pheno$Sample_Group <- factor(pheno$Sample_Group, levels = c("control", "case"))

beta_mat <- as.matrix(beta_handler$getBeta())
locs <- beta_handler$getBetaLocs()
truth_probe_counts <- countOverlaps(truth_dmrs, makeGRangesFromDataFrame(
    as.data.frame(locs),
    seqnames.field = "chr",
    start.field = "start",
    end.field = "end",
    keep.extra.columns = TRUE
))
truth_overlap_n <- sum(truth_probe_counts > 0L)
if (truth_overlap_n < 0.9 * length(truth_dmrs)) {
    stop(
        sprintf(
            paste(
                "Only %d/%d truth DMRs overlap probe loci under genome='%s'.",
                "This simulated microarray truth set was generated on hg19;",
                "delete stale cache files and rerun with genome='hg19'."
            ),
            truth_overlap_n,
            length(truth_dmrs),
            genome
        )
    )
}
logit2 <- function(x) log2(x) - log2(1 - x)
mvalues <- logit2(beta_mat)
pheno$casecontrol <- pheno$Sample_Group == "case"
```

## Find Differentially Methylated Positions

```{r}
dmps <- CMEnt::findDMPsArray(beta_handler,
    pheno,
    sample_group_col = "Sample_Group",
    id_col = "V1"
)
rownames(dmps) <- dmps$site_id
# sort DMPs by location
dmps <- dmps[!is.na(dmps$pval), ]
dmp_locs <- locs[rownames(locs) %in% rownames(dmps), , drop = FALSE]
dmps <- dmps[rownames(dmp_locs), , drop = FALSE]

case_idx <- which(pheno$casecontrol)
ctrl_idx <- which(!pheno$casecontrol)
dmps$delta_beta <- rowMeans(beta_mat[rownames(dmps), case_idx, drop = FALSE]) -
    rowMeans(beta_mat[rownames(dmps), ctrl_idx, drop = FALSE])

dmps$qval <- stats::p.adjust(dmps$pval, method = "fdr")

dmps_gr <- makeGRangesFromDataFrame(
    data.frame(
        chr = dmp_locs$chr,
        start = dmp_locs$start,
        end = dmp_locs$end,
        id = rownames(dmps),
        pval = dmps$pval,
        qval = dmps$qval,
        delta_beta = dmps$delta_beta
    ),
    keep.extra.columns = TRUE
)
dmps_gr <- sort(dmps_gr)
names(dmps_gr) <- rownames(dmps)

seeds_gr <- dmps_gr[mcols(dmps_gr)$qval < 0.1]
seeds <- mcols(seeds_gr)$id
cat(sprintf("Found %d significant DMPs to be used as seeds (FDR < 0.1)\n", length(seeds)))

```



## Method 1: CMEnt

```{r cment}
cment_params <- list(
    sample_group_col = "Sample_Group",
    casecontrol_col = "casecontrol",
    max_bridge_seeds_gaps = 0,
    max_bridge_extension_gaps = 1,
    min_seeds = 2,
    min_sites = 3,
    ext_site_delta_beta = simulation$delta_max0 / 3,
    max_lookup_dist = simulation$max_gap,
    expansion_window = simulation$max_dmr_size,
    max_pval = 0.05,
    testing_mode = "parametric",
    entanglement = "weak"
)
cment_njobs <- max(1L, min(8L, parallel::detectCores(logical = TRUE) - 1L))
options("CMEnt.njobs" = cment_njobs)
cment_file <- get_method_cache_file("CMEnt")

ret <- read_method_cache(cment_file, cment_params)
if (!is.null(ret)) {
    dmrs_cment <- ret$dmrs
    time_cment <- ret$time
} else {
    cment_njobs <- max(1L, min(8L, parallel::detectCores(logical = TRUE) - 1L))
    options("CMEnt.njobs" = cment_njobs)
    cat(sprintf("CMEnt parallel jobs: %d\n", cment_njobs))
    start_time <- Sys.time()
    dmrs_cment <- do.call(CMEnt::buildDMRs,
        c(list(
            beta = beta_handler,
            seeds = seeds,
            pheno = pheno,
            verbose = 2,
            .score_dmrs = FALSE,
            extract_motifs = FALSE,
            annotate_with_genes = FALSE
        ), cment_params)
    )
    end_time <- Sys.time()
    time_cment <- end_time - start_time
    save_method_cache(cment_file, dmrs_cment, time_cment, cment_params)
}

cat(sprintf("CMEnt: Found %d DMRs in %d seconds\n", length(dmrs_cment), round(as.numeric(time_cment, units = "secs"))))
```

## Method 2: comb-p

```{r comb-p}
combp_file <- get_method_cache_file("combp")
combp_params <- list(
    seed = simulation$max_gap,
    dist = simulation$max_gap,
    region_filter_p = 0.01,
    region_filter_n = simulation$min_sites,
    region_sidak_p = 0.05
)

run_combp <- function(dmps_gr, params) {
    if (Sys.which("bedtools") == "") {
        warning("bedtools not found in PATH. comb-p requires bedtools to be installed.")
        warning("Install with: conda install -yc bioconda bedtools")
        return(GRanges())
    }

    if (Sys.which("comb-p") == "") {
        warning("comb-p not found in PATH.")
        warning("Install with: conda install -yc bioconda combined-pvalues")
        return(GRanges())
    }
    tmpdir <- tempdir()
    bed_file <- file.path(tmpdir, "dmps_for_combp.bed")
    bed_df <- data.frame(
        chrom = as.character(seqnames(dmps_gr)),
        start = start(dmps_gr) - 1,
        end = end(dmps_gr),
        pval = mcols(dmps_gr)$pval
    )
    bed_df <- bed_df[order(bed_df$chrom, bed_df$start), ]

    write.table(bed_df, bed_file,
        quote = FALSE, sep = "\t",
        row.names = FALSE, col.names = TRUE
    )

    cmd <- sprintf(
        paste(
            "comb-p pipeline -c 4 --seed %d --dist %d -p %s",
            "--region-filter-p %s --region-filter-n %d %s"
        ),
        params$seed, params$dist, file.path(tmpdir, "output."),
        params$region_filter_p, params$region_filter_n, bed_file
    )

    tryCatch(
        {
            exit_code <- system(cmd, ignore.stdout = FALSE, ignore.stderr = FALSE)

            if (exit_code != 0) {
                warning(sprintf("comb-p exited with code %d", exit_code))
                return(GRanges())
            }

            regions_file <- paste0(file.path(tmpdir, "output."), "regions-p.bed.gz")
            if (file.exists(regions_file)) {
                regions <- fread(cmd = sprintf("zcat %s", regions_file))
                colnames(regions) <- c("chr", "start", "end", "min_p", "n_probes", "z_p", "z_sidak_p")

                regions <- regions[regions$z_sidak_p < params$region_sidak_p & regions$n_probes >= params$region_filter_n, ]

                dmrs_combp <- makeGRangesFromDataFrame(regions,
                    seqnames.field = "chr",
                    start.field = "start",
                    end.field = "end",
                    keep.extra.columns = TRUE
                )
                return(dmrs_combp)
            } else {
                warning("comb-p did not produce output file")
                return(GRanges())
            }
        },
        error = function(e) {
            warning(sprintf("comb-p failed: %s", e$message))
            GRanges()
        }
    )
}

ret <- read_method_cache(combp_file, combp_params)
if (!is.null(ret)) {
    dmrs_combp <- ret$dmrs
    time_combp <- ret$time
} else {
    start_time <- Sys.time()
    dmrs_combp <- run_combp(dmps_gr, combp_params)
    end_time <- Sys.time()
    time_combp <- end_time - start_time
    save_method_cache(combp_file, dmrs_combp, time_combp, combp_params)
}

cat(sprintf("comb-p: Found %d DMRs in %d seconds\n", length(dmrs_combp), round(as.numeric(time_combp, units = "secs"))))
```

## Method 3: probeLasso
```{r probeLasso}
probelasso_file <- get_method_cache_file("probeLasso")
probelasso_params <- list(
    array_type = array_type,
    genome = genome,
    minProbes = simulation$min_sites,
    adjPvalDmr = 0.05,
    adjPvalProbe = 0.1,
    meanLassoRadius = simulation$max_gap,
    minDmrSep = simulation$max_gap,
    minDmrSize = simulation$min_dmr_size
)

run_probelasso_champ <- function(beta_mat, pheno, params) {
    if (!requireNamespace("ChAMP", quietly = TRUE)) {
        warning("ChAMP not installed. probeLasso requires ChAMP.")
        warning("Install with: BiocManager::install('ChAMP')")
        return(GRanges(seqinfo = GenomeInfoDb::Seqinfo(genome = params$genome)))
    }

    library(ChAMP)

    tryCatch(
        {
            result <- champ.DMR(
                beta = beta_mat,
                pheno = pheno$Sample_Group,
                arraytype = params$array_type,
                method = "ProbeLasso",
                minProbes = params$minProbes,
                adjPvalDmr = params$adjPvalDmr,
                adjPvalProbe = params$adjPvalProbe,
                meanLassoRadius = params$meanLassoRadius,
                minDmrSep = params$minDmrSep,
                minDmrSize = params$minDmrSize,
                PDFplot = FALSE,
                Rplot = FALSE,
                resultsDir = file.path(tempdir(), "champ_probelasso")
            )

            pl_dmr <- result$ProbeLassoDMR
            if (is.null(pl_dmr) || nrow(pl_dmr) == 0) {
                return(GRanges())
            }

            # ChAMP output column names can differ by version.
            if (!"seqnames" %in% names(pl_dmr) && "dmrChrom" %in% names(pl_dmr)) {
                pl_dmr$seqnames <- pl_dmr$dmrChrom
            }
            if (!"start" %in% names(pl_dmr) && "dmrStart" %in% names(pl_dmr)) {
                pl_dmr$start <- pl_dmr$dmrStart
            }
            if (!"end" %in% names(pl_dmr) && "dmrEnd" %in% names(pl_dmr)) {
                pl_dmr$end <- pl_dmr$dmrEnd
            }

            return(makeGRangesFromDataFrame(pl_dmr,
                seqnames.field = "seqnames",
                start.field = "start",
                end.field = "end",
                seqinfo = GenomeInfoDb::Seqinfo(genome = params$genome),
                keep.extra.columns = TRUE
            ))
        },
        error = function(e) {
            warning(sprintf("probeLasso (ChAMP) failed: %s", e$message))
            GRanges(seqinfo = GenomeInfoDb::Seqinfo(genome = params$genome))
        }
    )
}

ret <- read_method_cache(probelasso_file, probelasso_params)
if (!is.null(ret)) {
    dmrs_probelasso <- ret$dmrs
    time_probelasso <- ret$time
} else {
    start_time <- Sys.time()
    dmrs_probelasso <- run_probelasso_champ(beta_mat, pheno, probelasso_params)
    end_time <- Sys.time()
    time_probelasso <- end_time - start_time
    save_method_cache(probelasso_file, dmrs_probelasso, time_probelasso, probelasso_params)
}

cat(sprintf("probeLasso: Found %d DMRs in %d seconds\n", length(dmrs_probelasso), round(as.numeric(time_probelasso, units = "secs"))))
```

## Method 4: DMRcate

```{r dmrcate}
dmrcate_file <- get_method_cache_file("dmrcate")
dmrcate_params <- list(
    array_type = array_type,
    genome = genome,
    fdr = 0.05,
    C = 2,
    lambda = simulation$max_gap * 2,
    min.cpgs = simulation$min_sites,
    betacutoff = 0.05
)

ret <- read_method_cache(dmrcate_file, dmrcate_params)
if (!is.null(ret)) {
    dmrs_dmrcate <- ret$dmrs
    time_dmrcate <- ret$time
} else {
    library(DMRcate)
    start_time <- Sys.time()
    formula_str <- "~Sample_Group"
    design <- model.matrix(stats::as.formula(formula_str), data = pheno)
    coef_col <- which(colnames(design) == "Sample_Groupcase")
    colnames(design)[1] <- "(Intercept)"
    colnames(design)[coef_col] <- "case_vs_control"

    myannotation <- cpg.annotate("array", mvalues,
        what = "M", arraytype = dmrcate_params$array_type,
        analysis.type = "differential", design = design,
        coef = coef_col, fdr = dmrcate_params$fdr
    )

    dmrcate_results <- dmrcate(myannotation,
        C = dmrcate_params$C,
        lambda = dmrcate_params$lambda,
        min.cpgs = dmrcate_params$min.cpgs,
        betacutoff = dmrcate_params$betacutoff
    )
    dmrs_dmrcate <- extractRanges(dmrcate_results, genome = dmrcate_params$genome)


    end_time <- Sys.time()
    time_dmrcate <- end_time - start_time
    save_method_cache(dmrcate_file, dmrs_dmrcate, time_dmrcate, dmrcate_params)
}

cat(sprintf("DMRcate: Found %d DMRs in %d seconds\n", length(dmrs_dmrcate), round(as.numeric(time_dmrcate, units = "secs"))))
```

## Method 5: Bumphunter

```{r bumphunter}
bumphunter_file <- get_method_cache_file("bumphunter")
bumphunter_params <- list(
    B = 100L,
    max_dmrs = 100000L,
    max_gap = simulation$max_gap,
    min_l = simulation$min_sites,
    region_p = 0.05,
    cutoff_start = simulation$delta_max0 / 3,
    cutoff_floor = 0.005,
    max_iter = 10L,
    max_bootstrap_adjust = 6L
)

ret <- read_method_cache(bumphunter_file, bumphunter_params)
if (!is.null(ret)) {
    dmrs_bumphunter <- ret$dmrs
    time_bumphunter <- ret$time
} else {
    suppressPackageStartupMessages({
        library(bumphunter)
        library(doParallel)
    })
    start_time <- Sys.time()
    bh_cores <- max(1L, min(8L, parallel::detectCores(logical = TRUE) - 1L))
    formula_str <- "~Sample_Group"
    design <- model.matrix(stats::as.formula(formula_str), data = pheno)
    coef_col <- which(colnames(design) == "Sample_Groupcase")
    colnames(design)[1] <- "(Intercept)"
    colnames(design)[coef_col] <- "case_vs_control"

    doParallel::registerDoParallel(cores = bh_cores)
    on.exit({
        try(doParallel::stopImplicitCluster(), silent = TRUE)
        gc()
    }, add = TRUE)

    bh_cluster <- bumphunter::clusterMaker(
        chr = as.character(locs$chr),
        pos = locs$start,
        maxGap = bumphunter_params$max_gap
    )
    run_bumphunter_fit <- function(cutoff, B) {
        bumphunter(beta_mat,
            design = design,
            chr = as.character(locs$chr),
            pos = locs$start,
            cluster = bh_cluster,
            cutoff = cutoff,
            maxGap = bumphunter_params$max_gap,
            nullMethod = "bootstrap",
            B = B,
            type = "Beta",
            coef = coef_col
        )
    }
    has_informative_bumphunter_null <- function(fit) {
        null_values <- unlist(fit$null$value, use.names = FALSE)
        null_lengths <- unlist(fit$null$length, use.names = FALSE)
        length(null_values) > 0 &&
            any(abs(null_values) > 0) &&
            length(null_lengths) > 0 &&
            any(null_lengths > 0)
    }
    dmrs_cutoff <- bumphunter_params$cutoff_start
    dmrs_cnt <- Inf
    iter <- 0L
    while (dmrs_cnt > bumphunter_params$max_dmrs && iter < bumphunter_params$max_iter) {
        iter <- iter + 1L
        dmrs <- run_bumphunter_fit(cutoff = dmrs_cutoff, B = 0)
        dmrs_cnt <- nrow(dmrs$table)
        dmrs_cutoff <- dmrs_cutoff * 1.2
        rm(dmrs)
        gc()
    }
    final_cutoff <- dmrs_cutoff / 1.2
    cat(sprintf(
        "Bumphunter: cutoff=%.4f, candidate_bumps=%d, minL=%d, maxGap=%d, cores=%d, B=%d\n",
        final_cutoff, dmrs_cnt, bumphunter_params$min_l, bumphunter_params$max_gap,
        bh_cores, bumphunter_params$B
    ))
    bumphunter_results <- run_bumphunter_fit(cutoff = final_cutoff, B = bumphunter_params$B)
    bh_bootstrap_adjust_iter <- 0L
    while (
        nrow(bumphunter_results$table) > 0 &&
            (all(is.na(bumphunter_results$table$p.value)) ||
                !has_informative_bumphunter_null(bumphunter_results)) &&
            final_cutoff > bumphunter_params$cutoff_floor &&
            bh_bootstrap_adjust_iter < bumphunter_params$max_bootstrap_adjust
    ) {
        prev_cutoff <- final_cutoff
        final_cutoff <- max(final_cutoff / 2, bumphunter_params$cutoff_floor)
        if (identical(final_cutoff, prev_cutoff)) {
            break
        }
        bh_bootstrap_adjust_iter <- bh_bootstrap_adjust_iter + 1L
        cat(sprintf(
            "Bumphunter: bootstrap null degenerate at cutoff=%.4f; retrying with cutoff=%.4f\n",
            prev_cutoff, final_cutoff
        ))
        bumphunter_results <- run_bumphunter_fit(cutoff = final_cutoff, B = bumphunter_params$B)
    }

    dmr_table <- bumphunter_results$table
    # bumphunter already estimates region-level p-values from the bootstrap null.
    dmr_table <- dmr_table[dmr_table$L >= bumphunter_params$min_l & dmr_table$p.value <= bumphunter_params$region_p, ]

    dmrs_bumphunter <- makeGRangesFromDataFrame(dmr_table,
        seqnames.field = "chr",
        start.field = "start",
        end.field = "end",
        keep.extra.columns = TRUE
    )
    end_time <- Sys.time()
    time_bumphunter <- end_time - start_time
    save_method_cache(bumphunter_file, dmrs_bumphunter, time_bumphunter, bumphunter_params)
}

cat(sprintf("Bumphunter: Found %d DMRs in %d seconds\n", length(dmrs_bumphunter), round(as.numeric(time_bumphunter, units = "secs"))))
```

## Comparative Analysis

### Basic Statistics

```{r}
get_dmr_stats <- function(dmrs, method) {
    if (length(dmrs) == 0) {
        return(data.frame(
            Method = method,
            `Number of DMRs` = 0,
            `Mean DMR Width` = NA,
            `Median DMR Width` = NA,
            `Total Coverage bp` = 0,
            `Min Width` = NA,
            `Max Width` = NA
        ))
    }

    data.frame(
        Method = method,
        `Number of DMRs` = length(dmrs),
        `Mean DMR Width` = round(mean(width(dmrs)), 1),
        `Median DMR Width` = median(width(dmrs)),
        `Total Coverage bp` = sum(width(dmrs)),
        `Min Width` = min(width(dmrs)),
        `Max Width` = max(width(dmrs))
    )
}

stats_list <- list(
    get_dmr_stats(dmrs_cment, "CMEnt"),
    get_dmr_stats(dmrs_combp, "comb-p"),
    get_dmr_stats(dmrs_probelasso, "probeLasso"),
    get_dmr_stats(dmrs_dmrcate, "DMRcate"),
    get_dmr_stats(dmrs_bumphunter, "Bumphunter")
)

results_comparison <- do.call(rbind, stats_list)
knitr::kable(results_comparison, digits = 2, caption = "Basic DMR Statistics Across Methods")
```

### Pairwise Overlap Analysis

```{r, results='asis'}
methods_list <- list(
    CMEnt = dmrs_cment,
    `comb-p` = dmrs_combp,
    probeLasso = dmrs_probelasso,
    DMRcate = dmrs_dmrcate,
    Bumphunter = dmrs_bumphunter
)

methods_list <- methods_list[sapply(methods_list, length) > 0]
method_order <- names(methods_list)

calculate_overlap <- function(gr1, gr2) {
    if (length(gr1) == 0 || length(gr2) == 0) {
        return(list(count = 0, percentage = 0))
    }
    overlap_count <- length(subsetByOverlaps(gr1, gr2))
    pct <- (overlap_count / length(gr1)) * 100
    list(count = overlap_count, percentage = pct)
}

overlap_matrix <- matrix(0, nrow = length(methods_list), ncol = length(methods_list))
pct_matrix <- matrix(0, nrow = length(methods_list), ncol = length(methods_list))
rownames(overlap_matrix) <- colnames(overlap_matrix) <- names(methods_list)
rownames(pct_matrix) <- colnames(pct_matrix) <- names(methods_list)

for (i in seq_along(methods_list)) {
    for (j in seq_along(methods_list)) {
        if (i == j) {
            overlap_matrix[i, j] <- length(methods_list[[i]])
            pct_matrix[i, j] <- 100
        } else {
            overlap_result <- calculate_overlap(methods_list[[i]], methods_list[[j]])
            overlap_matrix[i, j] <- overlap_result$count
            pct_matrix[i, j] <- overlap_result$percentage
        }
    }
}

cat("\n#### Overlap Counts (rows overlap with columns)\n")
knitr::kable(overlap_matrix, digits = 0, caption = "Number of DMRs from Method A (rows) overlapping with Method B (columns)")

cat("\n#### Overlap Percentages\n")
knitr::kable(pct_matrix, digits = 2, caption = "Percentage of DMRs from Method A (rows) overlapping with Method B (columns)")
```

### Jaccard Index and IoU

```{r}
jaccard_index <- function(gr1, gr2) {
    if (length(gr1) == 0 || length(gr2) == 0) {
        return(0)
    }
    overlap_count <- length(subsetByOverlaps(gr1, gr2))
    total_unique <- length(unique(c(gr1, gr2)))
    if (total_unique == 0) {
        return(0)
    }
    overlap_count / total_unique
}

iou <- function(gr1, gr2) {
    if (length(gr1) == 0 || length(gr2) == 0) {
        return(0)
    }
    intersection <- sum(width(GenomicRanges::intersect(gr1, gr2)))
    union <- sum(width(GenomicRanges::union(gr1, gr2)))
    if (union == 0) {
        return(0)
    }
    intersection / union
}

jaccard_matrix <- matrix(0, nrow = length(methods_list), ncol = length(methods_list))
iou_matrix <- matrix(0, nrow = length(methods_list), ncol = length(methods_list))
rownames(jaccard_matrix) <- colnames(jaccard_matrix) <- names(methods_list)
rownames(iou_matrix) <- colnames(iou_matrix) <- names(methods_list)

for (i in seq_along(methods_list)) {
    for (j in seq_along(methods_list)) {
        if (i < j) {
            jaccard_matrix[i, j] <- jaccard_matrix[j, i] <- jaccard_index(methods_list[[i]], methods_list[[j]])
            iou_matrix[i, j] <- iou_matrix[j, i] <- iou(methods_list[[i]], methods_list[[j]])
        } else if (i == j) {
            jaccard_matrix[i, j] <- 1
            iou_matrix[i, j] <- 1
        }
    }
}

cat("\n#### Jaccard Index\n")
knitr::kable(jaccard_matrix, digits = 4, caption = "Jaccard Index (region count-based similarity)")

cat("\n#### Intersection over Union (IoU)\n")
knitr::kable(iou_matrix, digits = 4, caption = "IoU (genomic coverage-based similarity)")
```

## Visualizations

### DMR Count and Coverage Comparison

```{r, fig.width=8, fig.height=3}
viz_data <- results_comparison |>
    dplyr::select(Method, Number.of.DMRs, Total.Coverage.bp) |>
    tidyr::pivot_longer(
        cols = c(Number.of.DMRs, Total.Coverage.bp),
        names_to = "Metric", values_to = "Value"
    )

viz_data$Metric <- ifelse(viz_data$Metric == "Number.of.DMRs",
    "Number of DMRs",
    "Total Coverage (bp)"
)

p1 <- ggplot(viz_data, aes(x = Method, y = Value, fill = Method)) +
    geom_bar(stat = "identity") +
    facet_wrap(~Metric, scales = "free_y") +
    theme_minimal() +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none",
        panel.grid.major.x = element_blank()
    ) +
    labs(
        title = "DMR Detection: Count and Coverage Comparison",
        y = "Value", x = ""
    )

print(p1)
```

## Ground-Truth Evaluation

```{r}
evaluate_against_truth <- function(pred, truth) {
    n_pred <- length(pred)
    n_truth <- length(truth)
    pred_overlap_truth <- sum(overlapsAny(pred, truth))
    truth_overlap_pred <- sum(overlapsAny(truth, pred))
    fp <- n_pred - pred_overlap_truth
    fn <- n_truth - truth_overlap_pred

    precision <- ifelse(n_pred == 0, 0, pred_overlap_truth / n_pred)
    recall <- ifelse(n_truth == 0, 0, truth_overlap_pred / n_truth)
    f1 <- ifelse(precision + recall == 0, 0, 2 * precision * recall / (precision + recall))

    pred_red <- reduce(pred)
    truth_red <- reduce(truth)
    inter <- GenomicRanges::intersect(pred_red, truth_red)

    pred_bp <- sum(width(pred_red))
    truth_bp <- sum(width(truth_red))
    inter_bp <- ifelse(length(inter) == 0, 0, sum(width(inter)))
    union_bp <- pred_bp + truth_bp - inter_bp

    precision_bp <- ifelse(pred_bp == 0, 0, inter_bp / pred_bp)
    recall_bp <- ifelse(truth_bp == 0, 0, inter_bp / truth_bp)
    f1_bp <- ifelse(precision_bp + recall_bp == 0, 0, 2 * precision_bp * recall_bp / (precision_bp + recall_bp))
    iou_bp <- ifelse(union_bp == 0, 0, inter_bp / union_bp)

    data.frame(
        TP_regions = truth_overlap_pred,
        FP_regions = fp,
        FN_regions = fn,
        Precision_regions = precision,
        Recall_regions = recall,
        F1_regions = f1,
        Predicted_bp = pred_bp,
        Truth_bp = truth_bp,
        Intersect_bp = inter_bp,
        Precision_bp = precision_bp,
        Recall_bp = recall_bp,
        F1_bp = f1_bp,
        IoU_bp = iou_bp
    )
}

truth_eval <- do.call(rbind, lapply(names(methods_list), function(method) {
    out <- evaluate_against_truth(methods_list[[method]], truth_dmrs)
    out$Method <- method
    out
}))
truth_eval$Method <- factor(truth_eval$Method, levels = method_order)
truth_eval <- truth_eval[order(truth_eval$Method), , drop = FALSE]
truth_eval$Method <- as.character(truth_eval$Method)


truth_eval <- truth_eval |>
    dplyr::select(
        Method,
        TP_regions,
        FP_regions, FN_regions,
        Precision_regions, Recall_regions, F1_regions,
        Precision_bp, Recall_bp, F1_bp, IoU_bp
    )

regions_truth_eval <- truth_eval |>
    dplyr::select(Method, TP_regions, FP_regions, FN_regions, Precision_regions, Recall_regions, F1_regions) |>
    dplyr::rename_with(~ sub("_regions$", "", .), ends_with("_regions"))

bp_truth_eval <- truth_eval |>
    dplyr::select(Method, Precision_bp, Recall_bp, F1_bp, IoU_bp) |>
    dplyr::rename_with(~ sub("_bp$", "", .), ends_with("_bp"))
```
```{r}
knitr::kable(regions_truth_eval, digits = 4, caption = "Macro (Region-level) Performance Against Ground Truth")
```

```{r}
knitr::kable(bp_truth_eval, digits = 4, caption = "Micro (Base-pair-level) Performance Against Ground Truth")
```

### Overlap Matrices (Methods + Truth)

```{r}
all_sets <- c(methods_list, list(`Ground Truth` = truth_dmrs))

calculate_overlap <- function(gr1, gr2) {
    if (length(gr1) == 0 || length(gr2) == 0) {
        return(list(count = 0, percentage = 0))
    }
    overlap_count <- sum(overlapsAny(gr1, gr2))
    pct <- overlap_count / length(gr1) * 100
    list(count = overlap_count, percentage = pct)
}

overlap_count_mat <- matrix(0, nrow = length(all_sets), ncol = length(all_sets))
overlap_pct_mat <- matrix(0, nrow = length(all_sets), ncol = length(all_sets))
rownames(overlap_count_mat) <- colnames(overlap_count_mat) <- names(all_sets)
rownames(overlap_pct_mat) <- colnames(overlap_pct_mat) <- names(all_sets)

for (i in seq_along(all_sets)) {
    for (j in seq_along(all_sets)) {
        if (i == j) {
            overlap_count_mat[i, j] <- length(all_sets[[i]])
            overlap_pct_mat[i, j] <- 100
        } else {
            ov <- calculate_overlap(all_sets[[i]], all_sets[[j]])
            overlap_count_mat[i, j] <- ov$count
            overlap_pct_mat[i, j] <- ov$percentage
        }
    }
}

knitr::kable(overlap_count_mat, digits = 0, caption = "Overlap Counts (query=row, reference=column)")
```
```{r}
knitr::kable(overlap_pct_mat, digits = 2, caption = "Overlap Percentages (query=row, reference=column)")
```

## Visualizations

### Ground-Truth Metrics

```{r, fig.width=10, fig.height=5}
plot_df <- truth_eval |>
    dplyr::select(Method, F1_regions, F1_bp, IoU_bp) |>
    tidyr::pivot_longer(
        cols = c(F1_regions, F1_bp, IoU_bp),
        names_to = "Metric",
        values_to = "Value"
    )

plot_df$Metric <- factor(
    plot_df$Metric,
    levels = c("F1_regions", "F1_bp", "IoU_bp"),
    labels = c("F1 (Region)", "F1 (Base-pair)", "IoU (Base-pair)")
)
plot_df$Method <- factor(plot_df$Method, levels = method_order)

ggplot(plot_df, aes(x = Method, y = Value, fill = Method)) +
    geom_col() +
    facet_wrap(~Metric, nrow = 1) +
    theme_minimal() +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none"
    ) +
    ylim(0, 1) +
    labs(
        title = "Method Performance vs Ground Truth",
        x = "",
        y = "Score"
    )
```

### Width Distributions

```{r, fig.width=8, fig.height=6, dpi=300}
width_data <- do.call(rbind, lapply(names(all_sets), function(method) {
    x <- all_sets[[method]]
    if (length(x) == 0) {
        return(NULL)
    }
    data.frame(Method = method, Width = width(x))
}))
width_data$Method <- factor(width_data$Method, levels = names(all_sets))

ggplot(width_data, aes(x = Method, y = Width, fill = Method)) +
    geom_violin(alpha = 0.7) +
    geom_boxplot(width = 0.2, alpha = 0.5, outlier.size = 0.5) +
    scale_y_log10(labels = scales::comma) +
    theme_minimal() +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none"
    ) +
    labs(
        title = "DMR Width Distributions",
        x = "",
        y = "Width (bp, log scale)"
    )
```

## Summary

```{r}
cat("\n## Key Findings\n\n")

if (nrow(truth_eval) > 0) {
    best_f1 <- truth_eval[which.max(truth_eval$F1_regions), ]
    best_iou <- truth_eval[which.max(truth_eval$IoU_bp), ]

    cat(sprintf(
        "- Best region-level F1: **%s** (%.3f)\n",
        best_f1$Method, best_f1$F1_regions
    ))
    cat(sprintf(
        "- Best base-pair IoU: **%s** (%.3f)\n",
        best_iou$Method, best_iou$IoU_bp
    ))

    if ("CMEnt" %in% truth_eval$Method) {
        seg <- truth_eval[truth_eval$Method == "CMEnt", ]
        cat(sprintf(
            "- **CMEnt** region metrics: Precision = %.3f, Recall = %.3f, F1 = %.3f\n",
            seg$Precision_regions, seg$Recall_regions, seg$F1_regions
        ))
        cat(sprintf(
            "- **CMEnt** base-pair metrics: Precision = %.3f, Recall = %.3f, F1 = %.3f, IoU = %.3f\n",
            seg$Precision_bp, seg$Recall_bp, seg$F1_bp, seg$IoU_bp
        ))
    }
}
```

## Session Info

```{r}
sessionInfo()
```
