---
title: "CMEnt Meta-Analysis on GSE244696"
subtitle: "Building DMRs from the published DMP list and comparing with the supplied Bumphunter regions"
author: "CMEnt Package"
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
        code_folding: hide
---

```{r, include=FALSE}
knitr::opts_chunk$set(
    results = "asis",
    cache = TRUE,
    echo = TRUE,
    dev = "png",
    warning = FALSE,
    message = FALSE,
    fig.width = 8,
    fig.height = 6,
    dpi = 300
)

try(assignInNamespace("plot_crop", function(x, quiet = TRUE) x, ns = "knitr"), silent = TRUE)
```

## Introduction

GSE244696 (doi: https://doi.org/10.1016/j.trsl.2024.05.005) is a PDAC MethylationEPIC study comparing tumor-cell enriched
("Tumor") and stroma-cell enriched ("Stroma") compartments. This notebook uses CMEnt in a meta-analysis mode: the
published DMP list (Supplementary table 1) is treated as the external evidence, and CMEnt uses the
available beta matrix from GEO to assemble coherent DMRs around those known
signals. The only external DMR comparison retained here is the Bumphunter DMR table (Supplementary table 2)
provided by the original researchers. 

## Setup

```{r setup, message=FALSE}
ensure_cran_packages <- function(pkgs) {
    repos <- getOption("repos")
    if (is.null(repos) || is.na(repos["CRAN"]) || repos["CRAN"] %in% c("@CRAN@", "")) {
        options(repos = c(CRAN = "https://cloud.r-project.org"))
    }
    missing <- pkgs[!vapply(pkgs, requireNamespace, logical(1), quietly = TRUE)]
    if (length(missing) > 0L) {
        install.packages(missing)
    }
}

ensure_bioc_packages <- function(pkgs) {
    if (!requireNamespace("BiocManager", quietly = TRUE)) {
        ensure_cran_packages("BiocManager")
    }
    missing <- pkgs[!vapply(pkgs, requireNamespace, logical(1), quietly = TRUE)]
    if (length(missing) > 0L) {
        BiocManager::install(missing, update = FALSE, ask = FALSE)
    }
}
suppressWarnings(
    {
        ensure_cran_packages(c("readxl", "data.table", "devtools", "dplyr", "ggplot2", "tidyr", "scales", "svglite"))
        ensure_bioc_packages(c(
            "BiocStyle", "minfi", "GenomicRanges", "annotatr", "missMethyl",
            "org.Hs.eg.db", "TxDb.Hsapiens.UCSC.hg19.knownGene",
            "IlluminaHumanMethylationEPICanno.ilm10b4.hg19", "clusterProfiler",
            "ReactomePA", "pathview", "KEGGREST"
        ))
    }
)

try(devtools::load_all(), silent = TRUE)
suppressPackageStartupMessages({
    try(library(CMEnt), silent = TRUE)
    library(data.table)
    library(dplyr)
    library(ggplot2)
    library(GenomicRanges)
    library(minfi)
})

getBenchmarkFile <- function(filename) {
    folder <- "metastudyEpic"
    if (getwd() == "notebooks") {
        benchmark_file <- file.path("cached_results", folder, filename)
    } else if ("NAMESPACE" %in% dir()) {
        benchmark_file <- file.path("notebooks", "cached_results", folder, filename)
    } else {
        benchmark_file <- file.path("cached_results", folder, filename)
    }
    dir.create(dirname(benchmark_file), showWarnings = FALSE, recursive = TRUE)
    benchmark_file
}

short_label <- function(x, width = 65L) {
    x <- as.character(x)
    ifelse(nchar(x) > width, paste0(substr(x, 1L, width - 3L), "..."), x)
}

has_text <- function(x) {
    !is.na(x) & nzchar(as.character(x))
}

theme_meta <- function() {
    theme_minimal(base_size = 11) +
        theme(
            panel.grid.minor = element_blank(),
            axis.text.x = element_text(angle = 35, hjust = 1),
            legend.position = "bottom"
        )
}

method_palette <- c(CMEnt = "#0072B2", `Published Bumphunter` = "#D55E00")
array_type <- "EPIC"
genome <- "hg19"
```

## Load Data and Published Evidence

```{r load-inputs}
beta_file <- getBenchmarkFile("GSE244696_beta_values.tsv.gz")
samplesheet <- getBenchmarkFile("GSE244696_samplesheet.tsv")
dmp_file <- getBenchmarkFile("GSE244696_DMPs.xlsx")
published_dmr_file <- getBenchmarkFile("GSE244696_DMRs.xlsx")


zenodo_base_url <- "https://zenodo.org/records/20592944/files/"
if (!file.exists(beta_file)) {
    curl::curl_download(
        url = paste0(zenodo_base_url, basename(beta_file), "?download=1"),
        destfile = beta_file
    )
}
if (!file.exists(samplesheet)) {
    curl::curl_download(
        url = paste0(zenodo_base_url, basename(samplesheet), "?download=1"),
        destfile = samplesheet
    )
}
if (!file.exists(dmp_file)) {
    curl::curl_download(
        url = paste0(zenodo_base_url, basename(dmp_file), "?download=1"),
        destfile = dmp_file
    )
}
if (!file.exists(published_dmr_file)) {
    curl::curl_download(
        url = paste0(zenodo_base_url, basename(published_dmr_file), "?download=1"),
        destfile = published_dmr_file
    )
}


pheno <- utils::read.table(samplesheet, header = TRUE, stringsAsFactors = FALSE, sep = "\t", check.names = FALSE)
if (all(rownames(pheno) == as.character(seq_len(nrow(pheno)))) && nzchar(names(pheno)[1])) {
    rownames(pheno) <- pheno[[1]]
}
if (all(rownames(pheno) == as.character(seq_len(nrow(pheno)))) && !nzchar(names(pheno)[1])) {
    rownames(pheno) <- pheno[[1]]
}
pheno$casecontrol <- pheno$tissue.ch1 == "Tumor"

beta_handler <- CMEnt::getBetaHandler(beta_file, array = array_type, genome = genome)
beta_mat <- as.matrix(beta_handler$getBeta())
locs <- beta_handler$getBetaLocs()
beta_row_names <- beta_handler$getBetaRowNames()

common_samples <- intersect(colnames(beta_mat), rownames(pheno))
beta_mat <- beta_mat[, common_samples, drop = FALSE]
pheno <- pheno[common_samples, , drop = FALSE]
case_samples <- rownames(pheno)[pheno$tissue.ch1 == "Tumor"]
control_samples <- rownames(pheno)[pheno$tissue.ch1 == "Stroma"]

reported_dmps_raw <- suppressWarnings(suppressMessages(readxl::read_xlsx(
    dmp_file,
    sheet = "Suppl_Table 1",
    range = readxl::cell_cols(1:21),
    .name_repair = "unique"
)))
reported_dmps <- data.frame(
    site_id = as.character(reported_dmps_raw[[1]]),
    pval = reported_dmps_raw[["Stroma_to_Tumor.P.Value"]],
    qval = reported_dmps_raw[["Stroma_to_Tumor.adj.P.Val"]],
    logFC_reported = reported_dmps_raw[["Stroma_to_Tumor.logFC"]],
    delta_beta_reported = reported_dmps_raw[["Stroma_to_Tumor.deltaBeta"]],
    stroma_avg_reported = reported_dmps_raw[["Stroma_to_Tumor.Stroma_AVG"]],
    tumor_avg_reported = reported_dmps_raw[["Stroma_to_Tumor.Tumor_AVG"]],
    chr_reported = paste0("chr", reported_dmps_raw[["Stroma_to_Tumor.CHR"]]),
    mapinfo_reported = reported_dmps_raw[["Stroma_to_Tumor.MAPINFO"]],
    gene_reported = reported_dmps_raw[["Stroma_to_Tumor.gene"]],
    feature_reported = reported_dmps_raw[["Stroma_to_Tumor.feature"]],
    cgi_reported = reported_dmps_raw[["Stroma_to_Tumor.cgi"]],
    stringsAsFactors = FALSE
)
reported_dmps <- reported_dmps[!is.na(reported_dmps$site_id) & nzchar(reported_dmps$site_id), ]
reported_dmps <- reported_dmps[!duplicated(reported_dmps$site_id), ]

sig_dmps <- reported_dmps[reported_dmps$site_id %in% beta_row_names, , drop = FALSE]
sig_dmps$delta_beta_available <- rowMeans(beta_mat[sig_dmps$site_id, case_samples, drop = FALSE], na.rm = TRUE) -
    rowMeans(beta_mat[sig_dmps$site_id, control_samples, drop = FALSE], na.rm = TRUE)

dmp_locs <- locs[sig_dmps$site_id, , drop = FALSE]
dmp_gr <- makeGRangesFromDataFrame(
    data.frame(
        chr = dmp_locs$chr,
        start = dmp_locs$start,
        end = dmp_locs$end,
        site_id = sig_dmps$site_id,
        pval = sig_dmps$pval,
        qval = sig_dmps$qval
    ),
    keep.extra.columns = TRUE
)
names(dmp_gr) <- sig_dmps$site_id

published_dmr_raw <- suppressWarnings(suppressMessages(readxl::read_xlsx(
    published_dmr_file,
    sheet = "Suppl_DMR",
    range = readxl::cell_cols(1:32),
    .name_repair = "unique"
)))
published_bumphunter <- makeGRangesFromDataFrame(
    as.data.frame(published_dmr_raw, check.names = FALSE),
    seqnames.field = "seqnames",
    start.field = "start",
    end.field = "end",
    keep.extra.columns = TRUE
)
names(published_bumphunter) <- make.unique(as.character(published_dmr_raw[[1]]))
genome(published_bumphunter) <- genome
mcols(published_bumphunter)$dmr_id <- names(published_bumphunter)

input_summary <- data.frame(
    Metric = c(
        "Samples",
        "Tumor samples",
        "Stroma samples",
        "Published DMPs",
        "Published DMPs present in beta matrix",
        "Published Bumphunter DMRs"
    ),
    Value = c(
        nrow(pheno),
        length(case_samples),
        length(control_samples),
        nrow(reported_dmps),
        nrow(sig_dmps),
        length(published_bumphunter)
    )
)
knitr::kable(input_summary, row.names = FALSE, caption = "Meta-study input summary")
```

```{r}
dmp_direction <- data.frame(
    Direction = c("Higher in Tumor", "Higher in Stroma"),
    Count = c(sum(sig_dmps$delta_beta_available > 0, na.rm = TRUE), sum(sig_dmps$delta_beta_available < 0, na.rm = TRUE))
)
knitr::kable(dmp_direction, row.names = FALSE, caption = "Direction of published DMPs")
```
```{r dmp-diagnostics, fig.width=8, fig.height=6}
sig_dmps_delta_beta_medians <- sig_dmps |>
    group_by(Direction = ifelse(delta_beta_available > 0, "Higher in Tumor", "Higher in Stroma")) |>
    summarise(
        Median = median(delta_beta_available, na.rm = TRUE),
        y = 1.1 * max(hist(delta_beta_available, breaks = 60, plot = FALSE)$counts, na.rm = TRUE)
    )
ggplot(sig_dmps, aes(x = delta_beta_available)) +
    geom_histogram(bins = 60, fill = "#0072B2", color = "white", linewidth = 0.15) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey35") +
    geom_text(
        data = sig_dmps_delta_beta_medians,
        aes(x = Median, y = y, label = paste0("Median: ", round(Median, 3))),
        inherit.aes = FALSE,
        color = "black",
        fontface = "bold",
        size = 3.5
    ) +
    theme_meta() +
    labs(
        title = "Effect Size for Published DMPs",
        x = "Delta beta", y = "Published DMPs"
    )
```

## CMEnt Meta-Analysis DMRs

```{r cment-run}
cment_file <- getBenchmarkFile("dmrs.CMEnt.meta_reported_dmps.rds")
if (!file.exists(cment_file)) {
    cment_njobs <- max(1L, min(8L, parallel::detectCores(logical = TRUE) - 1L))
    options("CMEnt.verbose" = 1)
    options("CMEnt.njobs" = cment_njobs)
    cat(sprintf("CMEnt parallel jobs: %d\n", cment_njobs))
    start_time <- Sys.time()
    dmrs_cment <- CMEnt::buildDMRs(
        beta = beta_handler,
        seeds = sig_dmps,
        pheno = pheno,
        seeds_id_col = "site_id",
        sample_group_col = "tissue.ch1",
        covariates = "predictedSex",
        max_bridge_seeds_gaps = 1,
        max_bridge_extension_gaps = 3,
        min_seeds = 2,
        min_sites = 3,
        ext_site_delta_beta = 0.2,
        max_lookup_dist = 1000,
        max_pval = 0.05,
        entanglement = "weak",
        score_dmrs = TRUE,
        annotate_with_genes = TRUE,
        extract_motifs = TRUE,
        njobs = cment_njobs,
        verbose = 1
    )
    time_cment <- Sys.time() - start_time
    saveRDS(list(dmrs = dmrs_cment, time = time_cment), cment_file)
} else {
    ret <- readRDS(cment_file)
    dmrs_cment <- ret$dmrs
    time_cment <- ret$time
}

cat(sprintf("CMEnt found %d DMRs in %d seconds.\n", length(dmrs_cment), round(as.numeric(time_cment, units = "secs"))))
if (is.null(dmrs_cment)) {
    dmrs_cment <- GRanges()
}
if (length(dmrs_cment) > 0L && !"in_promoter_of" %in% colnames(mcols(dmrs_cment))) {
    dmrs_cment <- CMEnt::annotateDMRsWithGenes(dmrs_cment, genome = genome)
}
published_bumphunter <- CMEnt::annotateDMRsWithGenes(published_bumphunter, genome = genome)
methods_list <- list(CMEnt = dmrs_cment, `Published Bumphunter` = published_bumphunter)
```


```{r cment_dmr_scatter_plot, fig.width=12, fig.height=6}
dmr_effect_df <- bind_rows(
    data.frame(
        Region = "Promoter-overlapping DMRs",
        Width = width(dmrs_cment),
        Delta_beta = as.numeric(mcols(dmrs_cment)$delta_beta_promoter)
    ),
    data.frame(
        Region = "Gene-body-overlapping DMRs",
        Width = width(dmrs_cment),
        Delta_beta = as.numeric(mcols(dmrs_cment)$delta_beta_gene_body)
    )
) |>
    filter(is.finite(Delta_beta)) |>
    mutate(
        Region = factor(Region, levels = unique(Region)),
        Direction = factor(
            ifelse(Delta_beta > 0, "Hyper", "Hypo"),
            levels = c("Hyper", "Hypo")
        )
    )

library(dplyr)
library(ggplot2)
library(scales)

ignore_abs_delta <- !is.finite(dmr_effect_df$Delta_beta) | abs(dmr_effect_df$Delta_beta) < 0.05
dmr_effect_df <- dmr_effect_df[!ignore_abs_delta, ]
# Summary per Region and Direction
dmr_summary <- dmr_effect_df |>
    group_by(Region, Direction) |>
    summarise(
        median_delta_beta = median(Delta_beta, na.rm = TRUE),
        median_width_bp = median(Width, na.rm = TRUE),
        n_dmrs = n(),
        .groups = "drop"
    )

mood_median_p_value <- function(width, direction) {
    widths <- split(width, direction)
    if (!all(c("Hyper", "Hypo") %in% names(widths)) || any(lengths(widths[c("Hyper", "Hypo")]) < 2L)) {
        return(NA_real_)
    }
    median_width <- median(width, na.rm = TRUE)
    keep <- is.finite(width) & width != median_width
    tab <- table(
        factor(direction[keep], levels = c("Hypo", "Hyper")),
        factor(width[keep] > median_width, levels = c(FALSE, TRUE))
    )
    if (any(rowSums(tab) == 0L) || any(colSums(tab) == 0L)) {
        return(NA_real_)
    }
    suppressWarnings(stats::chisq.test(tab, correct = FALSE)$p.value)
}

p_to_stars <- function(p) {
    case_when(
        is.na(p) ~ "",
        p < 0.001 ~ "***",
        p < 0.01 ~ "**",
        p < 0.05 ~ "*",
        TRUE ~ ""
    )
}

x_range <- range(dmr_effect_df$Delta_beta, na.rm = TRUE)
x_min <- x_range[[1]]

region_levels <- unique(dmr_effect_df$Region)
dmr_mood <- dmr_summary |>
    group_by(Region) |>
    summarise(
        star_y = exp(mean(log(median_width_bp))), .groups = "drop",
        segy_min = min(median_width_bp), segy_max = max(median_width_bp)
    ) |>
    left_join(
        dmr_effect_df |>
            group_by(Region) |>
            summarise(mood_p = mood_median_p_value(Width, Direction), .groups = "drop"),
        by = "Region"
    ) |>
    mutate(
        stars = p_to_stars(mood_p),
        star_x = x_min - 0.05 * abs(x_min),
        hjust = 0.5
    )

# Define two y-levels for the two summary bars
y_max <- max(dmr_effect_df$Width, na.rm = TRUE)

dmr_summary <- dmr_summary |>
    mutate(
        bar_y = case_when(
            Direction == "Hyper" ~ y_max * 2.0,
            Direction == "Hypo" ~ y_max * 3.2
        ),
        label_y = bar_y * 1.12
    )

# Expand y-range so the bars and labels fit
y_upper <- max(dmr_summary$label_y) * 1.15

ggplot(dmr_effect_df, aes(x = Delta_beta, y = Width, color = Direction)) +
    geom_hline(
        data = dmr_summary,
        aes(yintercept = median_width_bp, color = Direction),
        inherit.aes = FALSE,
        linewidth = 0.2
    ) +
    geom_vline(xintercept = 0, linewidth = 0.2, color = "grey70") +
    geom_point(alpha = 0.55, size = 1.2) +
    geom_segment(
        data = dmr_mood,
        aes(x = star_x + 0.05 * abs(x_min), xend = star_x + 0.05 * abs(x_min), y = segy_min, yend = segy_max),
        inherit.aes = FALSE,
        linewidth = 0.4,
        arrow = arrow(length = unit(0.15, "cm"), type = "closed", ends = "both"),
        color = "black",
        linetype = "dashed"
    ) +
    geom_text(
        data = dmr_mood,
        aes(x = star_x, y = star_y, label = stars, hjust = hjust),
        inherit.aes = FALSE,
        angle = 90,
        color = "black",
        fontface = "bold",
        size = 5
    ) +

    # Separate horizontal bars for hypo and hyper
    geom_segment(
        data = dmr_summary,
        aes(
            x = 0,
            xend = median_delta_beta,
            y = bar_y,
            yend = bar_y,
            color = Direction
        ),
        inherit.aes = FALSE,
        linewidth = 5,
        lineend = "butt"
    ) +

    # Labels for each bar
    geom_text(
        data = dmr_summary,
        aes(
            x = median_delta_beta,
            y = label_y,
            label = paste0(
                "median Δβ = ", round(median_delta_beta, 3),
                "\nmedian width = ", round(median_width_bp, 1), " bp",
                "\nDMRs = ", n_dmrs
            )
        ),
        inherit.aes = FALSE,
        color = "black",
        size = 4,
        hjust = ifelse(dmr_summary$median_delta_beta >= 0, -0.05, 1.05)
    ) +
    facet_wrap(~Region, nrow = 1) +
    scale_y_log10(
        limits = c(min(dmr_effect_df$Width[dmr_effect_df$Width > 0], na.rm = TRUE), y_upper),
        labels = comma
    ) +
    scale_color_manual(
        values = c(
            "Hyper" = "#B8405E",
            "Hypo" = "#247BA0"
        )
    ) +
    coord_cartesian(clip = "off") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom", strip.text.x = element_text(size = 12)) +
    labs(
        title = "DMR Effect Size and Width by Gene Context",
        x = "Context-specific delta beta (Tumor - Stroma)",
        y = "DMR width (bp, log scale)",
        color = ""
    )
```

```{r}
metadata_numeric_summary <- function(gr, fields) {
    fields <- intersect(fields, colnames(mcols(gr)))
    if (length(fields) == 0L || length(gr) == 0L) {
        return(data.frame())
    }
    base::do.call(rbind, lapply(fields, function(field) {
        x <- suppressWarnings(as.numeric(mcols(gr)[[field]]))
        data.frame(
            Field = field,
            Min = min(x, na.rm = TRUE),
            Median = stats::median(x, na.rm = TRUE),
            Mean = mean(x, na.rm = TRUE),
            Max = max(x, na.rm = TRUE)
        )
    }))
}

dmr_summary <- data.frame(
    Set = names(methods_list),
    DMRs = vapply(methods_list, length, integer(1)),
    Median_width_bp = vapply(methods_list, function(gr) stats::median(width(gr)), numeric(1)),
    Total_covered_bp = vapply(methods_list, function(gr) sum(width(reduce(gr))), numeric(1)),
    stringsAsFactors = FALSE
)
knitr::kable(dmr_summary, row.names = FALSE, caption = "DMR set summary")
```
```{r}
cment_fields <- c("n_sites", "n_seeds", "pval", "qval", "score", "cv_accuracy", "mean_delta_beta", "median_delta_beta")
cment_meta_summary <- metadata_numeric_summary(dmrs_cment, cment_fields)
if (nrow(cment_meta_summary) > 0L) {
    knitr::kable(cment_meta_summary, digits = 3, row.names = FALSE, caption = "CMEnt numeric metadata summary")
}
```
```{r dmr-width-comparison, fig.width=9, fig.height=4}
width_df <- base::do.call(rbind, lapply(names(methods_list), function(method) {
    data.frame(Method = method, Width = width(methods_list[[method]]))
}))
ggplot(width_df, aes(x = Method, y = Width, fill = Method)) +
    geom_boxplot(outlier.alpha = 0.35, width = 0.55) +
    scale_y_log10(labels = scales::comma) +
    scale_fill_manual(values = method_palette, drop = FALSE) +
    theme_meta() +
    labs(title = "DMR Width Distribution", x = "", y = "Width (bp, log10 scale)")
```
## CpG and Gene Context

```{r gene_context}
gene_context_df <- data.frame(
    Context = c("Promoter", "Gene body", "Promoter and gene body", "No annotated gene overlap"),
    DMRs = c(
        sum(has_text(mcols(dmrs_cment)$in_promoter_of)),
        sum(has_text(mcols(dmrs_cment)$in_gene_body_of)),
        sum(has_text(mcols(dmrs_cment)$in_promoter_of) & has_text(mcols(dmrs_cment)$in_gene_body_of)),
        sum(!has_text(mcols(dmrs_cment)$in_promoter_of) & !has_text(mcols(dmrs_cment)$in_gene_body_of))
    )
) |>
    mutate(Percentage = round(100 * DMRs / length(dmrs_cment), 1))

knitr::kable(gene_context_df, caption = "Gene context of CMEnt DMRs")
```

```{r gene_context_plot, fig.width=6, fig.height=3.5}
ggplot(gene_context_df, aes(x = reorder(Context, DMRs), y = DMRs, fill = Context)) +
    geom_col(width = 0.7) +
    coord_flip() +
    scale_y_continuous(labels = scales::comma) +
    theme_minimal() +
    theme(legend.position = "none") +
    labs(title = "DMR Gene Context", x = "", y = "DMRs")
```

```{r cpg_island_context}
split_ids <- function(x) {
    ids <- unlist(strsplit(stats::na.omit(as.character(x)), ","), use.names = FALSE)
    unique(ids[nzchar(ids)])
}

if (requireNamespace("IlluminaHumanMethylationEPICanno.ilm10b4.hg19", quietly = TRUE)) {
    library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
    anno <- as.data.frame(minfi::getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19))
    anno$probe_id <- rownames(anno)

    dmr_site_ids <- split_ids(mcols(dmrs_cment)$sites)
    dmr_site_context <- anno[intersect(dmr_site_ids, rownames(anno)), c("probe_id", "Relation_to_Island")]
    dmp_site_context <- anno[intersect(sig_dmps$site_id, rownames(anno)), c("probe_id", "Relation_to_Island")]

    site_context_df <- bind_rows(
        dmp_site_context |> mutate(Source = "Significant DMPs"),
        dmr_site_context |> mutate(Source = "Sites inside DMRs")
    ) |>
        dplyr::count(Source, Relation_to_Island, name = "Sites") |>
        group_by(Source) |>
        mutate(Percentage = round(100 * Sites / sum(Sites), 1)) |>
        ungroup()

    knitr::kable(site_context_df, caption = "CpG island context of significant DMPs and sites inside DMRs")
}
```

```{r cpg_island_context_plot, fig.width=8, fig.height=4}
if (exists("site_context_df")) {
    context_levels <- site_context_df |>
        group_by(Relation_to_Island) |>
        summarise(Max_sites = max(Sites), .groups = "drop") |>
        arrange(Max_sites) |>
        pull(Relation_to_Island)
    plot_context_df <- site_context_df |>
        mutate(
            Relation_to_Island = factor(Relation_to_Island, levels = context_levels),
            Source = factor(Source, levels = c("Significant DMPs", "Sites inside DMRs"))
        )
    diff_df <- plot_context_df |>
        dplyr::select(Source, Relation_to_Island, Sites) |>
        tidyr::pivot_wider(names_from = Source, values_from = Sites, values_fill = 0) |>
        dplyr::mutate(
            Difference = `Sites inside DMRs` - `Significant DMPs`,
            Label = paste0("Delta=", scales::comma(Difference)),
            Label_x = pmax(`Significant DMPs`, `Sites inside DMRs`)
        )

    ggplot(plot_context_df, aes(x = Sites, y = Relation_to_Island, fill = Source)) +
        geom_col(data = dplyr::filter(plot_context_df, Source == "Significant DMPs"), width = 0.78, alpha = 0.45) +
        geom_col(data = dplyr::filter(plot_context_df, Source == "Sites inside DMRs"), width = 0.48, alpha = 0.9) +
        geom_text(data = diff_df, aes(x = Label_x, y = Relation_to_Island, label = Label), inherit.aes = FALSE, hjust = -0.05, size = 3) +
        scale_x_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.2))) +
        scale_fill_manual(values = c("Significant DMPs" = "grey65", "Sites inside DMRs" = "#0072B2")) +
        theme_minimal() +
        theme(legend.position = "bottom") +
        labs(title = "CpG Island Context", x = "Sites", y = "", fill = "")
}
```


## Comparison With the Supplied Bumphunter DMRs

```{r}
overlap_hits <- findOverlaps(dmrs_cment, published_bumphunter, ignore.strand = TRUE)
cment_overlapping <- unique(queryHits(overlap_hits))
published_overlapping <- unique(subjectHits(overlap_hits))

intersection_bp <- sum(width(intersect(reduce(dmrs_cment), reduce(published_bumphunter), ignore.strand = TRUE)))
union_bp <- sum(width(union(reduce(dmrs_cment), reduce(published_bumphunter), ignore.strand = TRUE)))
overlap_summary <- data.frame(
    Metric = c(
        "CMEnt DMRs overlapping supplied Bumphunter",
        "Supplied Bumphunter DMRs overlapping CMEnt",
        "Fraction of CMEnt DMRs overlapping supplied Bumphunter",
        "Fraction of supplied Bumphunter DMRs overlapping CMEnt",
        "Base-pair Jaccard index"
    ),
    Value = c(
        length(cment_overlapping),
        length(published_overlapping),
        length(cment_overlapping) / max(1, length(dmrs_cment)),
        length(published_overlapping) / max(1, length(published_bumphunter)),
        intersection_bp / max(1, union_bp)
    )
)
knitr::kable(overlap_summary, digits = 3, row.names = FALSE, caption = "CMEnt vs supplied Bumphunter overlap")
```
```{r}
capture_by_set <- data.frame(
    Set = names(methods_list),
    Captured_DMPs = vapply(methods_list, function(gr) sum(overlapsAny(dmp_gr, gr, ignore.strand = TRUE)), integer(1)),
    Total_DMPs = length(dmp_gr),
    stringsAsFactors = FALSE
)
capture_by_set$Captured_Pct <- 100 * capture_by_set$Captured_DMPs / capture_by_set$Total_DMPs
knitr::kable(capture_by_set, digits = 1, row.names = FALSE, caption = "Published DMPs captured by each DMR set")
```
```{r dmp-capture-plot, fig.width=8, fig.height=4}
ggplot(capture_by_set, aes(x = Set, y = Captured_Pct, fill = Set)) +
    geom_col(width = 0.6) +
    geom_text(aes(label = paste0(scales::comma(Captured_DMPs), " probes")), vjust = -0.4, size = 3.3) +
    scale_fill_manual(values = method_palette, drop = FALSE) +
    scale_y_continuous(labels = scales::percent_format(scale = 1), limits = c(0, max(capture_by_set$Captured_Pct) * 1.18)) +
    theme_meta() +
    labs(title = "Capture of Published DMPs", x = "", y = "Published DMPs captured (%)")
```

```{r}
best_overlap_table <- data.frame()
if (length(overlap_hits) > 0L) {
    qi <- queryHits(overlap_hits)
    si <- subjectHits(overlap_hits)
    inter_width <- pmax(
        0,
        pmin(end(dmrs_cment)[qi], end(published_bumphunter)[si]) -
            pmax(start(dmrs_cment)[qi], start(published_bumphunter)[si]) + 1
    )
    pair_jaccard <- inter_width / (width(dmrs_cment)[qi] + width(published_bumphunter)[si] - inter_width)
    best_overlap_table <- data.frame(
        CMEnt_region = paste0(as.character(seqnames(dmrs_cment))[qi], ":", start(dmrs_cment)[qi], "-", end(dmrs_cment)[qi]),
        Bumphunter_region = paste0(as.character(seqnames(published_bumphunter))[si], ":", start(published_bumphunter)[si], "-", end(published_bumphunter)[si]),
        Bumphunter_id = names(published_bumphunter)[si],
        Overlap_bp = inter_width,
        Pair_Jaccard = pair_jaccard,
        Bumphunter_fwer = suppressWarnings(as.numeric(mcols(published_bumphunter)$fwer[si])),
        Bumphunter_gene = as.character(mcols(published_bumphunter)$genename[si]),
        stringsAsFactors = FALSE
    )
    best_overlap_table <- best_overlap_table |>
        arrange(CMEnt_region, desc(Pair_Jaccard)) |>
        group_by(CMEnt_region) |>
        slice_head(n = 1) |>
        ungroup() |>
        arrange(desc(Pair_Jaccard))
    knitr::kable(
        head(best_overlap_table, 15),
        digits = 3,
        row.names = FALSE,
        caption = "Best supplied Bumphunter match for each overlapping CMEnt DMR"
    )
} else {
    cat("No direct genomic overlap was found between CMEnt and the supplied Bumphunter DMRs.\n")
}
```
## CMEnt-specific plots
```{r cment-specific-plots-plot-dmr, fig.width=8, fig.height=6}
dmr_rank_table <- as.data.frame(dmrs_cment)
CMEnt::plotDMRs(
    dmrs_cment,
    top_n = 1,
    score_by = "score",
    plot_motif = FALSE,
    beta = beta_handler,
    pheno = pheno,
    sample_group_col = "tissue.ch1",
    array = "EPIC",
    genome = "hg19"
)
```
```{r cment-specific-plots-plot-manhattan, fig.width=10, fig.height=5}
cat("CMEnt DMR Manhattan plot:\n")
CMEnt::plotDMRsManhattan(
    dmrs_cment,
    genome = "hg19"
)
```
```{r cment-specific-plots-circos-interactions, fig.width=8, fig.height=8}
cat("CMEnt DMR Circos interactions plot:\n")
CMEnt::plotDMRsCircos(
    dmrs_cment,
    genome = "hg19",
    array = "EPIC"
)
```

## Biological Interpretation

`annotatr` is used for CpG and gene-region context,
while `missMethyl` is applied to promoter-overlapping DMRs for array-aware GO/KEGG enrichment because it accounts
for probe-per-gene bias in methylation arrays, and it was also included in the original publication.

```{r top_scoring_dmrs_genes}
top_10_cment_dmrs <- as.data.frame(dmrs_cment) |>
    arrange(by = -score) |>
    head(10) |>
    dplyr::select(id, score, cv_accuracy, in_promoter_of, in_gene_body_of)
knitr::kable(
    top_10_cment_dmrs,
    digits = 3,
    row.names = FALSE,
    caption = "Top 10 CMEnt DMRs with gene annotations"
)
```



```{r}

normalize_missmethyl_result <- function(x, method, collection) {
    if (is.null(x) || nrow(x) == 0L) {
        return(NULL)
    }
    x <- as.data.frame(x)
    id_col <- intersect(c("ID", "PathwayID", "Pathway.ID", "KEGGID"), colnames(x))[1]
    term_col <- intersect(c("Term", "TERM", "Description", "Pathway"), colnames(x))[1]
    fdr_col <- intersect(c("FDR", "adj.P.Val", "P.DE", "P.Value"), colnames(x))[1]
    p_col <- intersect(c("P.DE", "P.Value", "P.DE.weighted"), colnames(x))[1]
    de_col <- intersect(c("DE", "N_DE", "Count"), colnames(x))[1]
    n_col <- intersect(c("N", "Total", "Size"), colnames(x))[1]
    ont_col <- intersect(c("Ont", "Ontology", "Category"), colnames(x))[1]
    if (is.na(term_col) || is.na(fdr_col)) {
        return(NULL)
    }
    pathway_id <- if (!is.na(id_col)) as.character(x[[id_col]]) else rownames(x)
    pathway_id <- ifelse(collection == "KEGG" & grepl("^(path:)?hsa[0-9]{5}$|^[0-9]{5}$", pathway_id), pathway_id, NA_character_)
    data.frame(
        Method = method,
        Collection = collection,
        PathwayID = pathway_id,
        Term = as.character(x[[term_col]]),
        Ontology = if (!is.na(ont_col)) as.character(x[[ont_col]]) else collection,
        N = if (!is.na(n_col)) suppressWarnings(as.numeric(x[[n_col]])) else NA_real_,
        DE = if (!is.na(de_col)) suppressWarnings(as.numeric(x[[de_col]])) else NA_real_,
        PValue = if (!is.na(p_col)) suppressWarnings(as.numeric(x[[p_col]])) else NA_real_,
        FDR = suppressWarnings(as.numeric(x[[fdr_col]])),
        stringsAsFactors = FALSE
    )
}
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
missmethyl_long <- NULL
missmethyl_array <- if (toupper(array_type) == "EPICV2") "EPIC_V2" else toupper(array_type)
missmethyl_regions <- lapply(names(methods_list), function(method) {
    gr <- methods_list[[method]]
    if (!"in_promoter_of" %in% colnames(mcols(gr))) {
        return(gr[integer()])
    }
    gr[has_text(mcols(gr)$in_promoter_of)]
})
names(missmethyl_regions) <- names(methods_list)
missmethyl_region_summary <- data.frame(
    Method = names(methods_list),
    Promoter_overlapping_DMRs = vapply(missmethyl_regions, length, integer(1)),
    Total_DMRs = vapply(methods_list, length, integer(1)),
    stringsAsFactors = FALSE
)
missmethyl_region_summary$Percentage <- 100 * missmethyl_region_summary$Promoter_overlapping_DMRs / pmax(1L, missmethyl_region_summary$Total_DMRs)
knitr::kable(
    missmethyl_region_summary,
    digits = 1,
    row.names = FALSE,
    caption = "CMEnt-annotated promoter-overlapping DMRs supplied to missMethyl"
)

cached <- getBenchmarkFile("missmethyl_cment_gene_annotation_promoter_enrichment_results.rds")
if (file.exists(cached)) {
    missmethyl_long <- readRDS(cached)
    if (!all(c("PathwayID", "RegionScope") %in% colnames(missmethyl_long))) {
        unlink(cached)
        missmethyl_long <- NULL
    }
}
if (is.null(missmethyl_long)) {
    missmethyl_results <- Filter(Negate(is.null), unlist(lapply(names(missmethyl_regions), function(method) {
        regions <- missmethyl_regions[[method]]
        if (length(regions) == 0L) {
            cat(sprintf("Skipping missMethyl enrichment for %s: no promoter-overlapping DMRs.\n", method))
            return(NULL)
        }
        lapply(c("GO", "KEGG"), function(collection) {
            cat(sprintf("Running missMethyl %s enrichment for %s CMEnt-annotated promoter-overlapping DMRs (n = %d)...\n", collection, method, length(regions)))
            ret <- tryCatch(
                missMethyl::goregion(
                    regions = regions,
                    all.cpg = beta_row_names,
                    collection = collection,
                    array.type = missmethyl_array,
                    plot.bias = FALSE,
                    prior.prob = TRUE
                ),
                error = function(e) {
                    warning(sprintf("missMethyl %s enrichment failed for %s: %s", collection, method, e$message))
                    NULL
                }
            )
            ret <- normalize_missmethyl_result(ret, method, collection)
            if (!is.null(ret)) {
                ret$RegionScope <- "CMEnt-annotated promoter-overlapping DMRs"
            }
            ret
        })
    }), recursive = FALSE))
    missmethyl_long <- if (length(missmethyl_results) > 0L) {
        base::do.call(rbind, missmethyl_results)
    } else {
        NULL
    }
    saveRDS(missmethyl_long, cached)
}
```
```{r}
if (!is.null(missmethyl_long) && nrow(missmethyl_long) > 0L) {
    missmethyl_long <- missmethyl_long[is.finite(missmethyl_long$FDR), , drop = FALSE]
    missmethyl_long$NegLog10FDR <- -log10(pmax(missmethyl_long$FDR, .Machine$double.xmin))

    missmethyl_top <- missmethyl_long |>
        group_by(Method, Collection) |>
        arrange(FDR, .by_group = TRUE) |>
        slice_head(n = 10) |>
        ungroup()
    missmethyl_top$Term_Label <- short_label(missmethyl_top$Term, 70)
    missmethyl_top$Facet <- paste(missmethyl_top$Method, missmethyl_top$Collection, sep = " - ")
    shared_missmethyl_terms <- missmethyl_top |>
        distinct(Collection, Term, Method) |>
        dplyr::count(Collection, Term, name = "Methods") |>
        filter(Methods > 1L) |>
        transmute(Term_Key = paste(Collection, Term, sep = "___")) |>
        pull(Term_Key)
    missmethyl_top$Shared_Term <- paste(missmethyl_top$Collection, missmethyl_top$Term, sep = "___") %in% shared_missmethyl_terms
    missmethyl_top$Term_Facet <- factor(
        paste(missmethyl_top$Term_Label, missmethyl_top$Facet, sep = "___"),
        levels = rev(unique(paste(missmethyl_top$Term_Label, missmethyl_top$Facet, sep = "___")))
    )
    missmethyl_axis_term_labels <- gsub("\\\\", "\\\\\\\\", missmethyl_top$Term_Label)
    missmethyl_axis_term_labels <- gsub('"', '\\"', missmethyl_axis_term_labels, fixed = TRUE)
    missmethyl_axis_term_labels <- paste0('"', missmethyl_axis_term_labels, '"')
    missmethyl_axis_labels <- setNames(
        ifelse(missmethyl_top$Shared_Term, paste0("bold(", missmethyl_axis_term_labels, ")"), missmethyl_axis_term_labels),
        as.character(missmethyl_top$Term_Facet)
    )

    knitr::kable(
        missmethyl_top |>
            dplyr::select(Method, Collection, PathwayID, Ontology, Term, N, DE, PValue, FDR) |>
            arrange(Collection, Method, FDR),
        digits = 3,
        row.names = FALSE,
        caption = "Top missMethyl GO/KEGG enrichments from promoter-overlapping DMRs"
    )
}
```
```{r biology-missmethyl, fig.width=15, fig.height=7}
if (!is.null(missmethyl_long) && nrow(missmethyl_long) > 0L) {
    ggplot(missmethyl_top, aes(x = NegLog10FDR, y = Term_Facet, size = DE)) +
        geom_point(alpha = 0.9) +
        facet_wrap(~Facet, scales = "free_y", ncol = 2) +
        scale_y_discrete(labels = function(x) parse(text = unname(missmethyl_axis_labels[x]))) +
        theme_minimal(base_size = 17) +
        labs(
            x = expression(-log[10]("FDR")), y = "", color = "", size = "DM genes"
        )
} else {
    cat("missMethyl returned no enrichment results.\n")
}
```

```{r missmethyl-kegg-pathview, fig.width=10, fig.height=7}
if (!is.null(missmethyl_long) && nrow(missmethyl_long) > 0L &&
    requireNamespace("pathview", quietly = TRUE) &&
    requireNamespace("KEGGREST", quietly = TRUE)) {
    library(pathview)
    kegg_candidates <- missmethyl_long |>
        filter(Method == "CMEnt", Collection == "KEGG", !is.na(PathwayID), is.finite(PValue)) |>
        arrange(FDR, PValue) |>
        mutate(
            PathwayID = sub("^path:", "", PathwayID),
            PathwayID = ifelse(grepl("^hsa", PathwayID), PathwayID, paste0("hsa", PathwayID))
        )
    promoter_gene_delta <- as.data.frame(dmrs_cment) |>
        arrange(desc(score)) |>
        filter(is.finite(as.numeric(delta_beta_promoter)), !is.na(in_promoter_of), nzchar(in_promoter_of)) |>
        mutate(delta_beta_promoter = as.numeric(delta_beta_promoter)) |>
        dplyr::select(id, score, delta_beta_promoter, in_promoter_of) |>
        tidyr::separate_rows(in_promoter_of, sep = "\\s*,\\s*") |>
        mutate(Gene = trimws(in_promoter_of)) |>
        filter(nzchar(Gene)) |>
        arrange(desc(abs(delta_beta_promoter))) |>
        distinct(Gene, .keep_all = TRUE)

    if (nrow(kegg_candidates) > 0L && nrow(promoter_gene_delta) > 0L) {
        gene_map <- suppressMessages(AnnotationDbi::select(
            org.Hs.eg.db,
            keys = promoter_gene_delta$Gene,
            keytype = "SYMBOL",
            columns = c("SYMBOL", "ENTREZID")
        )) |>
            filter(!is.na(ENTREZID)) |>
            distinct(SYMBOL, ENTREZID)

        promoter_gene_delta <- promoter_gene_delta |>
            inner_join(gene_map, by = c("Gene" = "SYMBOL")) |>
            arrange(desc(abs(delta_beta_promoter))) |>
            distinct(ENTREZID, .keep_all = TRUE)

        if (nrow(promoter_gene_delta) == 0L) {
            cat("No promoter genes could be mapped to Entrez IDs for pathview.\n")
        } else {
            kegg_gene_map_file <- getBenchmarkFile("kegg_hsa_pathway_gene_map.rds")
            kegg_gene_map <- if (file.exists(kegg_gene_map_file)) {
                readRDS(kegg_gene_map_file)
            } else {
                links <- tryCatch(
                    KEGGREST::keggLink("hsa", "pathway"),
                    error = function(e) {
                        warning(sprintf("Could not retrieve KEGG pathway-gene links: %s", e$message))
                        NULL
                    }
                )
                if (is.null(links)) {
                    data.frame(PathwayID = character(), ENTREZID = character())
                } else {
                    data.frame(
                        PathwayID = sub("^path:", "", names(links)),
                        ENTREZID = sub("^hsa:", "", unname(links)),
                        stringsAsFactors = FALSE
                    ) |>
                        distinct()
                }
            }
            if (!file.exists(kegg_gene_map_file)) {
                saveRDS(kegg_gene_map, kegg_gene_map_file)
            }

            candidate_overlaps <- kegg_candidates |>
                left_join(kegg_gene_map, by = "PathwayID") |>
                filter(ENTREZID %in% promoter_gene_delta$ENTREZID) |>
                group_by(PathwayID) |>
                summarise(OverlapEntrez = list(unique(ENTREZID)), .groups = "drop")

            matched_pathway <- match(kegg_candidates$PathwayID, candidate_overlaps$PathwayID)
            matched_pathway <- which(!is.na(matched_pathway))[1]

            if (is.na(matched_pathway)) {
                cat("No missMethyl KEGG pathway contained the top promoter-overlapping DMR genes for pathview coloring.\n")
            } else {
                top_kegg <- kegg_candidates |> dplyr::slice(matched_pathway)
                matched_genes <- candidate_overlaps$OverlapEntrez[[match(top_kegg$PathwayID[1], candidate_overlaps$PathwayID)]]
                promoter_pathway_delta <- promoter_gene_delta |>
                    filter(ENTREZID %in% matched_genes)
                gene_delta <- promoter_pathway_delta$delta_beta_promoter
                names(gene_delta) <- promoter_pathway_delta$ENTREZID
                kegg_id <- sub("^hsa", "", top_kegg$PathwayID[1])
                pathview_dir <- getBenchmarkFile("pathview")
                dir.create(pathview_dir, showWarnings = FALSE, recursive = TRUE)
                pathview_suffix <- "promoter_delta_beta"
                pathview_png <- file.path(pathview_dir, paste0("hsa", kegg_id, ".", pathview_suffix, ".png"))
                gene_limit <- max(0.001, max(abs(gene_delta), na.rm = TRUE))
                unlink(c(pathview_png, basename(pathview_png)))

                pathview_result <- tryCatch(
                    pathview::pathview(
                        gene.data = gene_delta,
                        gene.idtype = "entrez",
                        pathway.id = kegg_id,
                        species = "hsa",
                        out.suffix = pathview_suffix,
                        kegg.dir = pathview_dir,
                        limit = list(gene = gene_limit, cpd = 1),
                        low = list(gene = "#2F80ED", cpd = "#2F80ED"),
                        mid = list(gene = "#FFFFFF", cpd = "#FFFFFF"),
                        high = list(gene = "#FF3B30", cpd = "#FF3B30")
                    ),
                    error = function(e) {
                        warning(sprintf("pathview failed for KEGG pathway %s: %s", kegg_id, e$message))
                        NULL
                    }
                )
                if (file.exists(basename(pathview_png))) {
                    file.rename(basename(pathview_png), pathview_png)
                }
                pathview_mapped <- if (!is.null(pathview_result$plot.data.gene)) {
                    sum(!is.na(pathview_result$plot.data.gene[["mol.data"]]))
                } else {
                    NA_integer_
                }

                knitr::kable(
                    promoter_pathway_delta |>
                        dplyr::select(Gene, ENTREZID, delta_beta_promoter, score, id) |>
                        arrange(desc(abs(delta_beta_promoter))) |>
                        head(20),
                    digits = 3,
                    row.names = FALSE,
                    caption = paste0(
                        "Promoter delta beta values supplied to pathview for the top missMethyl KEGG pathway with promoter-gene overlap: ",
                        top_kegg$Term[1],
                        ". Genes overlapping this KEGG pathway: ", nrow(promoter_pathway_delta),
                        ". Genes mapped by pathview: ", pathview_mapped,
                        ". Red indicates higher methylation in Tumor; blue indicates lower methylation in Tumor."
                    )
                )

                if (file.exists(pathview_png)) {
                    knitr::include_graphics(pathview_png)
                } else {
                    cat("pathview did not produce a KEGG pathway image.\n")
                }
            }
        }
    } else {
        cat("No CMEnt KEGG pathway with ID or promoter gene delta-beta values were available for pathview.\n")
    }
} else {
    cat("pathview and KEGGREST are required for the KEGG pathway diagram.\n")
}
```

## Conclusions

In the GSE244696 PDAC tumor-stroma comparison, CMEnt was applied in meta-analysis mode by using the published DMPs
as external evidence and the available beta matrix to assemble spatially coherent DMRs around those signals. 
The resulting DMRs recovered biological themes that are consistent with the original study and with PDAC tumor
microenvironment biology. Study-context terms highlighted extracellular matrix, myeloid leukocyte differentiation, 
steroid-hormone response and hormone-regulatory processes. These signals are plausible because PDAC is characterized
by a dense desmoplastic extracellular matrix and because the original study reported compartment-specific immune 
and estrogen receptor-related epigenetic dysregulation, with tumor-associated macrophages enriched in tumor-cell
regions and associated with patient outcome. The annotation profile further supports a regulatory interpretation:
CMEnt DMRs overlap promoters, exons, introns and 5′ UTRs, indicating that the method organizes CpG-level evidence
into regions connected to gene-regulatory architecture rather than producing isolated CpG hits. 
The genome-wide circos view shows both hypermethylated and hypomethylated DMRs, consistent with a
compartment-specific remodeling process. A motif-supported DMR interaction involving AHR::ARNT provides an
additional hypothesis-generating link to xenobiotic and cytochrome P450-related metabolism,
although this motif result should not be overinterpreted given the small number of motif-supported interactions.
Overall, these results suggest that CMEnt preserves the biological context of the published DMP evidence while
emphasizing coherent regulatory, extracellular-matrix, myeloid and hormone-associated tumor-stroma programs.

## Session Info
```{r}
sessionInfo()
```
