---
title: "CMEnt Meta-Analysis on GSE244696"
subtitle: "Building DMRs from the published DMP list and comparing with the supplied Bumphunter regions"
author: "CMEnt Package"
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
        code_folding: hide
---

```{r, include=FALSE}
knitr::opts_chunk$set(
    results = "asis",
    cache = TRUE,
    echo = TRUE,
    dev = "pdf",
    warning = FALSE,
    message = FALSE,
    fig.width = 8,
    fig.height = 6,
    dpi = 300
)

try(assignInNamespace("plot_crop", function(x, quiet = TRUE) x, ns = "knitr"), silent = TRUE)
```

## Introduction

GSE244696 (doi: https://doi.org/10.1016/j.trsl.2024.05.005) is a PDAC MethylationEPIC study comparing tumor-cell enriched
("Tumor") and stroma-cell enriched ("Stroma") compartments. The GEO deposition
does not include the complete raw IDAT set used by the original study, so a
head-to-head DMR caller benchmark on the locally available beta matrix would be
unfair. This notebook therefore uses CMEnt in a meta-analysis mode: the
published DMP list (Supplementary table 1) is treated as the external evidence, and CMEnt uses the
available beta matrix to assemble coherent DMRs around those known
signals. The only external DMR comparison retained here is the Bumphunter DMR table (Supplementary table 2)
provided by the original researchers. 

## Setup

```{r setup, message=FALSE}
ensure_cran_packages <- function(pkgs) {
    repos <- getOption("repos")
    if (is.null(repos) || is.na(repos["CRAN"]) || repos["CRAN"] %in% c("@CRAN@", "")) {
        options(repos = c(CRAN = "https://cloud.r-project.org"))
    }
    missing <- pkgs[!vapply(pkgs, requireNamespace, logical(1), quietly = TRUE)]
    if (length(missing) > 0L) {
        install.packages(missing)
    }
}

ensure_bioc_packages <- function(pkgs) {
    if (!requireNamespace("BiocManager", quietly = TRUE)) {
        ensure_cran_packages("BiocManager")
    }
    missing <- pkgs[!vapply(pkgs, requireNamespace, logical(1), quietly = TRUE)]
    if (length(missing) > 0L) {
        BiocManager::install(missing, update = FALSE, ask = FALSE)
    }
}

ensure_cran_packages(c("readxl", "data.table", "devtools", "dplyr", "ggplot2", "tidyr", "scales", "svglite"))
ensure_bioc_packages(c(
    "BiocStyle", "minfi", "GenomicRanges", "annotatr", "missMethyl",
    "org.Hs.eg.db", "TxDb.Hsapiens.UCSC.hg19.knownGene",
    "IlluminaHumanMethylationEPICanno.ilm10b4.hg19", "clusterProfiler",
    "ReactomePA"
))

try(devtools::load_all(), silent = TRUE)
suppressPackageStartupMessages({
    try(library(CMEnt), silent = TRUE)
    library(data.table)
    library(dplyr)
    library(ggplot2)
    library(GenomicRanges)
    library(minfi)
})

getBenchmarkFile <- function(filename) {
    folder <- "metastudyEpic"
    if (getwd() == "notebooks") {
        benchmark_file <- file.path("cached_results", folder, filename)
    } else if ("NAMESPACE" %in% dir()) {
        benchmark_file <- file.path("notebooks", "cached_results", folder, filename)
    } else {
        benchmark_file <- file.path("cached_results", folder, filename)
    }
    dir.create(dirname(benchmark_file), showWarnings = FALSE, recursive = TRUE)
    benchmark_file
}

short_label <- function(x, width = 65L) {
    x <- as.character(x)
    ifelse(nchar(x) > width, paste0(substr(x, 1L, width - 3L), "..."), x)
}

theme_meta <- function() {
    theme_minimal(base_size = 11) +
        theme(
            panel.grid.minor = element_blank(),
            axis.text.x = element_text(angle = 35, hjust = 1),
            legend.position = "bottom"
        )
}

method_palette <- c(CMEnt = "#0072B2", `Published Bumphunter` = "#D55E00")
array_type <- "EPIC"
genome <- "hg19"
```

## Load Local Data and Published Evidence

```{r load-inputs}
beta_file <- getBenchmarkFile("GSE244696_beta_values.tsv.gz")
samplesheet <- getBenchmarkFile("GSE244696_samplesheet.tsv")
dmp_file <- getBenchmarkFile("GSE244696_DMPs.xlsx")
published_dmr_file <- getBenchmarkFile("GSE244696_DMRs.xlsx")



zenodo_base_url <- "https://zenodo.org/records/20592944/files/"
if (!file.exists(beta_file)) {
    curl::curl_download(
        url = paste0(zenodo_base_url, basename(beta_file), "?download=1"),
        destfile = beta_file
    )
}
if (!file.exists(samplesheet)) {
    curl::curl_download(
        url = paste0(zenodo_base_url, basename(samplesheet), "?download=1"),
        destfile = samplesheet
    )
}
if (!file.exists(dmp_file)) {
    curl::curl_download(
        url = paste0(zenodo_base_url, basename(dmp_file), "?download=1"),
        destfile = dmp_file
    )
}
if (!file.exists(published_dmr_file)) {
    curl::curl_download(
        url = paste0(zenodo_base_url, basename(published_dmr_file), "?download=1"),
        destfile = published_dmr_file
    )
}



pheno <- utils::read.table(samplesheet, header = TRUE, stringsAsFactors = FALSE, sep = "\t", check.names = FALSE)
if (all(rownames(pheno) == as.character(seq_len(nrow(pheno)))) && nzchar(names(pheno)[1])) {
    rownames(pheno) <- pheno[[1]]
}
if (all(rownames(pheno) == as.character(seq_len(nrow(pheno)))) && !nzchar(names(pheno)[1])) {
    rownames(pheno) <- pheno[[1]]
}
pheno$casecontrol <- pheno$tissue.ch1 == "Tumor"

beta_handler <- CMEnt::getBetaHandler(beta_file, array = array_type, genome = genome)
beta_mat <- as.matrix(beta_handler$getBeta())
locs <- beta_handler$getBetaLocs()
beta_row_names <- beta_handler$getBetaRowNames()

common_samples <- intersect(colnames(beta_mat), rownames(pheno))
beta_mat <- beta_mat[, common_samples, drop = FALSE]
pheno <- pheno[common_samples, , drop = FALSE]
case_samples <- rownames(pheno)[pheno$tissue.ch1 == "Tumor"]
control_samples <- rownames(pheno)[pheno$tissue.ch1 == "Stroma"]

reported_dmps_raw <- suppressWarnings(suppressMessages(readxl::read_xlsx(
    dmp_file,
    sheet = "Suppl_Table 1",
    range = readxl::cell_cols(1:21),
    .name_repair = "unique"
)))
reported_dmps <- data.frame(
    site_id = as.character(reported_dmps_raw[[1]]),
    pval = reported_dmps_raw[["Stroma_to_Tumor.P.Value"]],
    qval = reported_dmps_raw[["Stroma_to_Tumor.adj.P.Val"]],
    logFC_reported = reported_dmps_raw[["Stroma_to_Tumor.logFC"]],
    delta_beta_reported = reported_dmps_raw[["Stroma_to_Tumor.deltaBeta"]],
    stroma_avg_reported = reported_dmps_raw[["Stroma_to_Tumor.Stroma_AVG"]],
    tumor_avg_reported = reported_dmps_raw[["Stroma_to_Tumor.Tumor_AVG"]],
    chr_reported = paste0("chr", reported_dmps_raw[["Stroma_to_Tumor.CHR"]]),
    mapinfo_reported = reported_dmps_raw[["Stroma_to_Tumor.MAPINFO"]],
    gene_reported = reported_dmps_raw[["Stroma_to_Tumor.gene"]],
    feature_reported = reported_dmps_raw[["Stroma_to_Tumor.feature"]],
    cgi_reported = reported_dmps_raw[["Stroma_to_Tumor.cgi"]],
    stringsAsFactors = FALSE
)
reported_dmps <- reported_dmps[!is.na(reported_dmps$site_id) & nzchar(reported_dmps$site_id), ]
reported_dmps <- reported_dmps[!duplicated(reported_dmps$site_id), ]

sig_dmps <- reported_dmps[reported_dmps$site_id %in% beta_row_names, , drop = FALSE]
sig_dmps$delta_beta_available <- rowMeans(beta_mat[sig_dmps$site_id, case_samples, drop = FALSE], na.rm = TRUE) -
    rowMeans(beta_mat[sig_dmps$site_id, control_samples, drop = FALSE], na.rm = TRUE)

dmp_locs <- locs[sig_dmps$site_id, , drop = FALSE]
dmp_gr <- makeGRangesFromDataFrame(
    data.frame(
        chr = dmp_locs$chr,
        start = dmp_locs$start,
        end = dmp_locs$end,
        site_id = sig_dmps$site_id,
        pval = sig_dmps$pval,
        qval = sig_dmps$qval
    ),
    keep.extra.columns = TRUE
)
names(dmp_gr) <- sig_dmps$site_id

published_dmr_raw <- suppressWarnings(suppressMessages(readxl::read_xlsx(
    published_dmr_file,
    sheet = "Suppl_DMR",
    range = readxl::cell_cols(1:32),
    .name_repair = "unique"
)))
published_bumphunter <- makeGRangesFromDataFrame(
    as.data.frame(published_dmr_raw, check.names = FALSE),
    seqnames.field = "seqnames",
    start.field = "start",
    end.field = "end",
    keep.extra.columns = TRUE
)
names(published_bumphunter) <- make.unique(as.character(published_dmr_raw[[1]]))
genome(published_bumphunter) <- genome
mcols(published_bumphunter)$dmr_id <- names(published_bumphunter)

input_summary <- data.frame(
    Metric = c(
        "Local samples",
        "Local Tumor samples",
        "Local Stroma samples",
        "Published DMPs",
        "Published DMPs present in local beta matrix",
        "Published Bumphunter DMRs"
    ),
    Value = c(
        nrow(pheno),
        length(case_samples),
        length(control_samples),
        nrow(reported_dmps),
        nrow(sig_dmps),
        length(published_bumphunter)
    )
)
knitr::kable(input_summary, row.names = FALSE, caption = "Meta-study input summary")
```

```{r}
dmp_direction <- data.frame(
    Direction = c("Higher in Tumor", "Higher in Stroma"),
    Count = c(sum(sig_dmps$delta_beta_available > 0, na.rm = TRUE), sum(sig_dmps$delta_beta_available < 0, na.rm = TRUE))
)
knitr::kable(dmp_direction, row.names = FALSE, caption = "Direction of published DMPs in the locally available samples")
```
```{r dmp-diagnostics, fig.width=9, fig.height=4}
ggplot(sig_dmps, aes(x = delta_beta_available)) +
    geom_histogram(bins = 60, fill = "#0072B2", color = "white", linewidth = 0.15) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey35") +
    theme_meta() +
    labs(
        title = "Local Effect Size for Published DMPs",
        subtitle = "Tumor minus Stroma beta difference, computed only from locally available samples",
        x = "Delta beta in local samples", y = "Published DMPs"
    )
```

## CMEnt Meta-Analysis DMRs

```{r cment-run}
cment_file <- getBenchmarkFile("dmrs.CMEnt.meta_reported_dmps.rds")
if (!file.exists(cment_file)) {
    cment_njobs <- max(1L, min(8L, parallel::detectCores(logical = TRUE) - 1L))
    options("CMEnt.verbose" = 1)
    options("CMEnt.njobs" = cment_njobs)
    cat(sprintf("CMEnt parallel jobs: %d\n", cment_njobs))
    start_time <- Sys.time()
    dmrs_cment <- CMEnt::buildDMRs(
        beta = beta_handler,
        seeds = sig_dmps,
        pheno = pheno,
        seeds_id_col = "site_id",
        sample_group_col = "tissue.ch1",
        covariates = "predictedSex",
        max_bridge_seeds_gaps = 1,
        max_bridge_extension_gaps = 3,
        min_seeds = 2,
        min_sites = 3,
        ext_site_delta_beta = 0.2,
        max_lookup_dist = 1000,
        max_pval = 0.05,
        entanglement = "weak",
        .score_dmrs = TRUE,
        annotate_with_genes = TRUE,
        extract_motifs = TRUE,
        njobs = cment_njobs,
        verbose = 1
    )
    time_cment <- Sys.time() - start_time
    saveRDS(list(dmrs = dmrs_cment, time = time_cment), cment_file)
} else {
    ret <- readRDS(cment_file)
    dmrs_cment <- ret$dmrs
    time_cment <- ret$time
}

cat(sprintf("CMEnt found %d DMRs in %d seconds.\n", length(dmrs_cment), round(as.numeric(time_cment, units = "secs"))))
if (is.null(dmrs_cment)) {
    dmrs_cment <- GRanges()
}
methods_list <- list(CMEnt = dmrs_cment, `Published Bumphunter` = published_bumphunter)
```

```{r}
metadata_numeric_summary <- function(gr, fields) {
    fields <- intersect(fields, colnames(mcols(gr)))
    if (length(fields) == 0L || length(gr) == 0L) {
        return(data.frame())
    }
    base::do.call(rbind, lapply(fields, function(field) {
        x <- suppressWarnings(as.numeric(mcols(gr)[[field]]))
        data.frame(
            Field = field,
            Min = min(x, na.rm = TRUE),
            Median = stats::median(x, na.rm = TRUE),
            Mean = mean(x, na.rm = TRUE),
            Max = max(x, na.rm = TRUE)
        )
    }))
}

dmr_summary <- data.frame(
    Set = names(methods_list),
    DMRs = vapply(methods_list, length, integer(1)),
    Median_width_bp = vapply(methods_list, function(gr) stats::median(width(gr)), numeric(1)),
    Total_covered_bp = vapply(methods_list, function(gr) sum(width(reduce(gr))), numeric(1)),
    stringsAsFactors = FALSE
)
knitr::kable(dmr_summary, row.names = FALSE, caption = "DMR set summary")
```
```{r}
cment_fields <- c("n_sites", "n_seeds", "pval", "qval", "score", "cv_accuracy", "mean_delta_beta", "median_delta_beta")
cment_meta_summary <- metadata_numeric_summary(dmrs_cment, cment_fields)
if (nrow(cment_meta_summary) > 0L) {
    knitr::kable(cment_meta_summary, digits = 3, row.names = FALSE, caption = "CMEnt numeric metadata summary")
}
```
```{r dmr-width-comparison, fig.width=9, fig.height=4}
width_df <- base::do.call(rbind, lapply(names(methods_list), function(method) {
    data.frame(Method = method, Width = width(methods_list[[method]]))
}))
ggplot(width_df, aes(x = Method, y = Width, fill = Method)) +
    geom_boxplot(outlier.alpha = 0.35, width = 0.55) +
    scale_y_log10(labels = scales::comma) +
    scale_fill_manual(values = method_palette, drop = FALSE) +
    theme_meta() +
    labs(title = "DMR Width Distribution", x = "", y = "Width (bp, log10 scale)")
```

## Comparison With the Supplied Bumphunter DMRs

```{r}
overlap_hits <- findOverlaps(dmrs_cment, published_bumphunter, ignore.strand = TRUE)
cment_overlapping <- unique(queryHits(overlap_hits))
published_overlapping <- unique(subjectHits(overlap_hits))

intersection_bp <- sum(width(intersect(reduce(dmrs_cment), reduce(published_bumphunter), ignore.strand = TRUE)))
union_bp <- sum(width(union(reduce(dmrs_cment), reduce(published_bumphunter), ignore.strand = TRUE)))
overlap_summary <- data.frame(
    Metric = c(
        "CMEnt DMRs overlapping supplied Bumphunter",
        "Supplied Bumphunter DMRs overlapping CMEnt",
        "Fraction of CMEnt DMRs overlapping supplied Bumphunter",
        "Fraction of supplied Bumphunter DMRs overlapping CMEnt",
        "Base-pair Jaccard index"
    ),
    Value = c(
        length(cment_overlapping),
        length(published_overlapping),
        length(cment_overlapping) / max(1, length(dmrs_cment)),
        length(published_overlapping) / max(1, length(published_bumphunter)),
        intersection_bp / max(1, union_bp)
    )
)
knitr::kable(overlap_summary, digits = 3, row.names = FALSE, caption = "CMEnt vs supplied Bumphunter overlap")
```
```{r}
capture_by_set <- data.frame(
    Set = names(methods_list),
    Captured_DMPs = vapply(methods_list, function(gr) sum(overlapsAny(dmp_gr, gr, ignore.strand = TRUE)), integer(1)),
    Total_DMPs = length(dmp_gr),
    stringsAsFactors = FALSE
)
capture_by_set$Captured_Pct <- 100 * capture_by_set$Captured_DMPs / capture_by_set$Total_DMPs
knitr::kable(capture_by_set, digits = 1, row.names = FALSE, caption = "Published DMPs captured by each DMR set")
```
```{r dmp-capture-plot, fig.width=8, fig.height=4}
ggplot(capture_by_set, aes(x = Set, y = Captured_Pct, fill = Set)) +
    geom_col(width = 0.6) +
    geom_text(aes(label = paste0(scales::comma(Captured_DMPs), " probes")), vjust = -0.4, size = 3.3) +
    scale_fill_manual(values = method_palette, drop = FALSE) +
    scale_y_continuous(labels = scales::percent_format(scale = 1), limits = c(0, max(capture_by_set$Captured_Pct) * 1.18)) +
    theme_meta() +
    labs(title = "Capture of Published DMPs", x = "", y = "Published DMPs captured (%)")
```

```{r}
best_overlap_table <- data.frame()
if (length(overlap_hits) > 0L) {
    qi <- queryHits(overlap_hits)
    si <- subjectHits(overlap_hits)
    inter_width <- pmax(
        0,
        pmin(end(dmrs_cment)[qi], end(published_bumphunter)[si]) -
            pmax(start(dmrs_cment)[qi], start(published_bumphunter)[si]) + 1
    )
    pair_jaccard <- inter_width / (width(dmrs_cment)[qi] + width(published_bumphunter)[si] - inter_width)
    best_overlap_table <- data.frame(
        CMEnt_region = paste0(as.character(seqnames(dmrs_cment))[qi], ":", start(dmrs_cment)[qi], "-", end(dmrs_cment)[qi]),
        Bumphunter_region = paste0(as.character(seqnames(published_bumphunter))[si], ":", start(published_bumphunter)[si], "-", end(published_bumphunter)[si]),
        Bumphunter_id = names(published_bumphunter)[si],
        Overlap_bp = inter_width,
        Pair_Jaccard = pair_jaccard,
        Bumphunter_fwer = suppressWarnings(as.numeric(mcols(published_bumphunter)$fwer[si])),
        Bumphunter_gene = as.character(mcols(published_bumphunter)$genename[si]),
        stringsAsFactors = FALSE
    )
    best_overlap_table <- best_overlap_table |>
        arrange(CMEnt_region, desc(Pair_Jaccard)) |>
        group_by(CMEnt_region) |>
        slice_head(n = 1) |>
        ungroup() |>
        arrange(desc(Pair_Jaccard))
    knitr::kable(
        head(best_overlap_table, 15),
        digits = 3,
        row.names = FALSE,
        caption = "Best supplied Bumphunter match for each overlapping CMEnt DMR"
    )
} else {
    cat("No direct genomic overlap was found between CMEnt and the supplied Bumphunter DMRs.\n")
}
```
## CMEnt-specific plots
```{r cment-specific-plots-plot-dmr, fig.width=10, fig.height=5}
dmr_rank_table <- as.data.frame(dmrs_cment)
top_dmr_idx <- if ("qval" %in% colnames(dmr_rank_table) && any(is.finite(dmr_rank_table$qval))) {
    which.min(dmr_rank_table$qval)
} else {
    delta_col <- intersect(c("median_delta_beta", "mean_delta_beta", "delta_beta"), colnames(dmr_rank_table))[1]
    if (is.na(delta_col)) 1L else order(abs(dmr_rank_table[[delta_col]]), decreasing = TRUE)[1]
}
cat("Top CMEnt DMR prioritized by DMR q-value when available; SVM score is complementary:\n")
CMEnt::plotDMRs(
    dmrs_cment,
    dmr_indices = top_dmr_idx,
    beta = beta_handler,
    pheno = pheno,
    sample_group_col = "tissue.ch1",
    array = "EPIC",
    genome = "hg19"
)
```
```{r cment-specific-plots-plot-manhattan, fig.width=10, fig.height=5}
cat("CMEnt DMR Manhattan plot:\n")
CMEnt::plotDMRsManhattan(
    dmrs_cment,
    genome = "hg19"
)
```
```{r cment-specific-plots-circos-interactions, fig.width=10, fig.height=10}
cat("CMEnt DMR Circos interactions plot:\n")
CMEnt::plotDMRsCircos(
    dmrs_cment,
    genome = "hg19",
    array = "EPIC"
)
```
## Biological Interpretation

`annotatr` is used for CpG and gene-region context,
while `missMethyl` is used for array-aware GO/KEGG enrichment because it accounts
for probe-per-gene bias in methylation arrays, and it was also included in the original publication.

```{r}
context_summary <- NULL
annotatr_gene_ids_by_method <- list()
annotatr_types <- c(
    paste0(genome, "_cpg_islands"),
    paste0(genome, "_cpg_shores"),
    paste0(genome, "_cpg_shelves"),
    paste0(genome, "_cpg_inter"),
    paste0(genome, "_genes_promoters"),
    paste0(genome, "_genes_exons"),
    paste0(genome, "_genes_introns"),
    paste0(genome, "_genes_3UTRs"),
    paste0(genome, "_genes_5UTRs"),
    paste0(genome, "_genes_intergenic")
)
annotatr_types <- annotatr_types[annotatr_types %in% annotatr::builtin_annotations()]
annotatr_ann <- annotatr::build_annotations(genome = genome, annotations = annotatr_types)
annotation_type_labels <- c(
    setNames(
        c("CpG island", "CpG shore", "CpG shelf", "Open sea"),
        paste0(genome, c("_cpg_islands", "_cpg_shores", "_cpg_shelves", "_cpg_inter"))
    ),
    setNames(
        c("Promoter", "Exon", "Intron", "3' UTR", "5' UTR", "Intergenic"),
        paste0(genome, c("_genes_promoters", "_genes_exons", "_genes_introns", "_genes_3UTRs", "_genes_5UTRs", "_genes_intergenic"))
    )
)

context_summary <- base::do.call(rbind, lapply(names(methods_list), function(method) {
    gr <- methods_list[[method]]
    if (length(gr) == 0L) {
        return(NULL)
    }
    gr <- granges(gr)
    mcols(gr)$dmr_id <- seq_along(gr)
    names(gr) <- NULL
    ann_gr <- annotatr::annotate_regions(regions = gr, annotations = annotatr_ann, quiet = TRUE)
    ann_mcols <- as.data.frame(S4Vectors::mcols(ann_gr), optional = TRUE)
    rownames(ann_mcols) <- NULL
    ann_df <- data.frame(
        seqnames = as.character(seqnames(ann_gr)),
        start = start(ann_gr),
        end = end(ann_gr),
        ann_mcols,
        check.names = FALSE,
        stringsAsFactors = FALSE
    )
    if ("annot.gene_id" %in% colnames(ann_df)) {
        annotatr_gene_ids_by_method[[method]] <<- unique(stats::na.omit(as.character(ann_df$annot.gene_id)))
    }
    ann_df <- ann_df[ann_df$annot.type %in% names(annotation_type_labels), , drop = FALSE]
    if (nrow(ann_df) == 0L) {
        return(NULL)
    }
    ann_df <- unique(ann_df[, c("dmr_id", "annot.type"), drop = FALSE])
    counts <- as.data.frame(table(ann_df$annot.type), stringsAsFactors = FALSE)
    colnames(counts) <- c("annot.type", "DMRs")
    counts$Method <- method
    counts$Annotation <- unname(annotation_type_labels[as.character(counts$annot.type)])
    counts$Class <- ifelse(grepl("_cpg_", counts$annot.type), "CpG context", "Gene context")
    counts$Percentage <- 100 * counts$DMRs / length(gr)
    counts
}))
if (!is.null(context_summary) && nrow(context_summary) > 0L) {
    context_summary$Annotation <- factor(
        context_summary$Annotation,
        levels = c("CpG island", "CpG shore", "CpG shelf", "Open sea", "Promoter", "Exon", "Intron", "5' UTR", "3' UTR", "Intergenic")
    )
    knitr::kable(
        context_summary |>
            dplyr::select(Method, Class, Annotation, DMRs, Percentage) |>
            arrange(Class, Annotation, Method),
        digits = 1,
        row.names = FALSE,
        caption = "Region context summary. Categories can overlap."
    )
}
```
```{r biology-context, fig.width=10, fig.height=5}
if (!is.null(context_summary) && nrow(context_summary) > 0L) {
    ggplot(context_summary, aes(x = Annotation, y = Percentage, fill = Method)) +
        geom_col(position = position_dodge(width = 0.8), width = 0.7) +
        facet_wrap(~Class, scales = "free_x") +
        scale_fill_manual(values = method_palette, drop = FALSE) +
        theme_meta() +
        labs(title = "DMR Annotation Context", x = "", y = "DMRs overlapping annotation (%)", fill = "")
}
```

```{r}
pdac_term_pattern <- paste(
    c(
        "immune", "immun", "macrophage", "myeloid", "leukocyte", "lymphocyte",
        "cytokine", "chemokine", "inflamm", "antigen", "interferon",
        "estrogen", "oestrogen", "hormone", "stromal", "extracellular matrix",
        "collagen", "fibroblast", "pancrea"
    ),
    collapse = "|"
)

normalize_missmethyl_result <- function(x, method, collection) {
    if (is.null(x) || nrow(x) == 0L) {
        return(NULL)
    }
    x <- as.data.frame(x)
    term_col <- intersect(c("Term", "TERM", "Description", "Pathway"), colnames(x))[1]
    fdr_col <- intersect(c("FDR", "adj.P.Val", "P.DE", "P.Value"), colnames(x))[1]
    p_col <- intersect(c("P.DE", "P.Value", "P.DE.weighted"), colnames(x))[1]
    de_col <- intersect(c("DE", "N_DE", "Count"), colnames(x))[1]
    n_col <- intersect(c("N", "Total", "Size"), colnames(x))[1]
    ont_col <- intersect(c("Ont", "Ontology", "Category"), colnames(x))[1]
    if (is.na(term_col) || is.na(fdr_col)) {
        return(NULL)
    }
    data.frame(
        Method = method,
        Collection = collection,
        Term = as.character(x[[term_col]]),
        Ontology = if (!is.na(ont_col)) as.character(x[[ont_col]]) else collection,
        N = if (!is.na(n_col)) suppressWarnings(as.numeric(x[[n_col]])) else NA_real_,
        DE = if (!is.na(de_col)) suppressWarnings(as.numeric(x[[de_col]])) else NA_real_,
        PValue = if (!is.na(p_col)) suppressWarnings(as.numeric(x[[p_col]])) else NA_real_,
        FDR = suppressWarnings(as.numeric(x[[fdr_col]])),
        stringsAsFactors = FALSE
    )
}
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
missmethyl_long <- NULL
missmethyl_array <- if (toupper(array_type) == "EPICV2") "EPIC_V2" else toupper(array_type)
cached <- getBenchmarkFile("missmethyl_enrichment_results.rds")
if (file.exists(cached)) {
    missmethyl_long <- readRDS(cached)
} else {
    missmethyl_long <- base::do.call(rbind, Filter(Negate(is.null), unlist(lapply(names(methods_list), function(method) {
        lapply(c("GO", "KEGG"), function(collection) {
            ret <- tryCatch(
                missMethyl::goregion(
                    regions = methods_list[[method]],
                    all.cpg = beta_row_names,
                    collection = collection,
                    array.type = missmethyl_array,
                    plot.bias = FALSE,
                    prior.prob = TRUE
                ),
                error = function(e) {
                    warning(sprintf("missMethyl %s enrichment failed for %s: %s", collection, method, e$message))
                    NULL
                }
            )
            normalize_missmethyl_result(ret, method, collection)
        })
    }), recursive = FALSE)))
    saveRDS(missmethyl_long, cached)
}
```
```{r}
if (!is.null(missmethyl_long) && nrow(missmethyl_long) > 0L) {
    missmethyl_long <- missmethyl_long[is.finite(missmethyl_long$FDR), , drop = FALSE]
    missmethyl_long$NegLog10FDR <- -log10(pmax(missmethyl_long$FDR, .Machine$double.xmin))
    missmethyl_long$PDAC_Context <- ifelse(grepl(pdac_term_pattern, missmethyl_long$Term, ignore.case = TRUE), "Study-context term", "Other top term")

    missmethyl_top <- missmethyl_long |>
        group_by(Method, Collection) |>
        arrange(FDR, .by_group = TRUE) |>
        slice_head(n = 10) |>
        ungroup()
    missmethyl_top$Term_Label <- short_label(missmethyl_top$Term, 70)
    missmethyl_top$Facet <- paste(missmethyl_top$Method, missmethyl_top$Collection, sep = " - ")
    missmethyl_top$Term_Facet <- factor(
        paste(missmethyl_top$Term_Label, missmethyl_top$Facet, sep = "___"),
        levels = rev(unique(paste(missmethyl_top$Term_Label, missmethyl_top$Facet, sep = "___")))
    )

    knitr::kable(
        missmethyl_top |>
            dplyr::select(Method, Collection, Ontology, Term, N, DE, PValue, FDR) |>
            arrange(Collection, Method, FDR),
        digits = 3,
        row.names = FALSE,
        caption = "Top missMethyl GO/KEGG enrichments"
    )
}
```
```{r biology-missmethyl, fig.width=12, fig.height=7}
if (!is.null(missmethyl_long) && nrow(missmethyl_long) > 0L) {

    ggplot(missmethyl_top, aes(x = NegLog10FDR, y = Term_Facet, size = DE)) +
        geom_point(alpha = 0.9) +
        facet_wrap(~Facet, scales = "free_y", ncol = 2) +
        scale_y_discrete(labels = function(x) sub("___.*$", "", x)) +
        theme_meta() +
        theme(axis.text.y = element_text(size = 7)) +
        labs(title = "missMethyl Enrichment", x = expression(-log[10]("FDR")), y = "", color = "", size = "DM genes")
} else {
    cat("missMethyl returned no enrichment results.\n")
}

```

```{r}
focused_terms <- missmethyl_long |>
    filter(grepl(pdac_term_pattern, Term, ignore.case = TRUE)) |>
    group_by(Method, Collection) |>
    arrange(FDR, .by_group = TRUE) |>
    slice_head(n = 8) |>
    ungroup()

if (nrow(focused_terms) > 0L) {
    focused_terms$Term_Label <- short_label(focused_terms$Term, 62)
    focused_terms$Term_Method <- factor(
        paste(focused_terms$Term_Label, focused_terms$Method, focused_terms$Collection, sep = "___"),
        levels = rev(unique(paste(focused_terms$Term_Label, focused_terms$Method, focused_terms$Collection, sep = "___")))
    )
    knitr::kable(
        focused_terms |>
            dplyr::select(Method, Collection, Ontology, Term, N, DE, PValue, FDR) |>
            arrange(Collection, Method, FDR),
        digits = 3,
        row.names = FALSE,
        caption = "Top immune, macrophage, stromal, pancreatic, and hormone/estrogen-related terms"
    )
}
```
```{r biology-focused-terms-plot, fig.width=10, fig.height=5}
if (nrow(focused_terms) > 0L) {
    ggplot(focused_terms, aes(x = -log10(pmax(FDR, .Machine$double.xmin)), y = Term_Method, fill = Method)) +
        geom_col(width = 0.7) +
        facet_wrap(~Collection, scales = "free_y") +
        scale_y_discrete(labels = function(x) sub("___.*$", "", x)) +
        scale_fill_manual(values = method_palette, drop = FALSE) +
        theme_meta() +
        theme(axis.text.y = element_text(size = 8)) +
        labs(title = "Study-Relevant Enrichment Terms", x = expression(-log[10]("FDR")), y = "", fill = "")
} else {
    cat("No missMethyl terms matched the study-focused keyword set.\n")
}
```


## Conclusions

In the GSE244696 PDAC tumor-stroma comparison, CMEnt was applied in meta-analysis mode by using the published DMPs
as external evidence and the available beta matrix to assemble spatially coherent DMRs around those signals. 
The resulting DMRs recovered biological themes that are consistent with the original study and with PDAC tumor
microenvironment biology. Study-context terms highlighted extracellular matrix, myeloid leukocyte differentiation, 
steroid-hormone response and hormone-regulatory processes. These signals are plausible because PDAC is characterized
by a dense desmoplastic extracellular matrix and because the original study reported compartment-specific immune 
and estrogen receptor-related epigenetic dysregulation, with tumor-associated macrophages enriched in tumor-cell
regions and associated with patient outcome. The annotation profile further supports a regulatory interpretation:
CMEnt DMRs overlap promoters, exons, introns and 5′ UTRs, indicating that the method organizes CpG-level evidence
into regions connected to gene-regulatory architecture rather than producing isolated CpG hits. 
The genome-wide circos view shows both hypermethylated and hypomethylated DMRs, consistent with a
compartment-specific remodeling process. A motif-supported DMR interaction involving AHR::ARNT provides an
additional hypothesis-generating link to xenobiotic and cytochrome P450-related metabolism,
although this motif result should not be overinterpreted given the small number of motif-supported interactions.
Overall, these results suggest that CMEnt preserves the biological context of the published DMP evidence while
emphasizing coherent regulatory, extracellular-matrix, myeloid and hormone-associated tumor-stroma programs.

## Session Info
```{r}
sessionInfo()
```
