---
title: "CMEnt Analysis of AML 450K Methylation Data"
subtitle: "Biological interpretation of GSE63409"
author: "CMEnt Package"
date: "`r Sys.Date()`"
output: 
    BiocStyle::html_document:
        code_folding: hide
---

```{r, include=FALSE}
knitr::opts_chunk$set(
    results = "asis",
    cache = TRUE,
    echo = TRUE,
    dev = "png",
    warning = FALSE,
    message = FALSE,
    fig.width = 8,
    fig.height = 6,
    dpi = 300
)
```

## Introduction

This vignette analyzes GSE63409, an Illumina 450K methylation dataset from acute myeloid leukemia (AML).
The dataset contains 15 AML patients represented by leukemia stem and blast cell samples, and normal
hematopoietic stem and progenitor cell controls from healthy bone marrow.

The original study, Hu et al. 2019
([doi:10.21037/atm.2019.11.122](https://dx.doi.org/10.21037/atm.2019.11.122)),
used GSE63409 together with an expression dataset to identify AML-relevant methylation/expression
signals, immune and differentiation pathways, and candidate prognostic genes. Because this dataset does not contain a region-level ground truth, the notebook focuses on
biological interpretation of CMEnt regions rather than comparisons with other DMR callers.

- CMEnt DMR size, promoter/gene-body effect size, and CpG context
- AML marker and pathway DMR signals
- GO and KEGG enrichment of CMEnt DMRs
- Candidate gene-overlapping loci for follow-up biological interpretation
 


## Setup and Dependencies

```{r, message=FALSE}
try(devtools::load_all(), silent = TRUE)
suppressPackageStartupMessages({
    try(library(CMEnt), silent = TRUE)
    library(minfi)
    library(data.table)
    library(ggplot2)
    library(dplyr)
    library(GenomicRanges)
})

getBenchmarkFile <- function(filename) {
    benchmark_file <- ""
    folder <- "microarray450k"
    if (getwd() == "notebooks") {
        benchmark_file <- file.path("cached_results", folder, filename)
    } else if ("NAMESPACE" %in% dir()) {
        benchmark_file <- file.path("notebooks", "cached_results", folder, filename)
    } else {
        benchmark_file <- file.path("cached_results", folder, filename)
    }
    dir.create(dirname(benchmark_file), showWarnings = FALSE, recursive = TRUE)
    benchmark_file
}

split_ids <- function(x) {
    ids <- unlist(strsplit(stats::na.omit(as.character(x)), ","), use.names = FALSE)
    unique(ids[nzchar(ids)])
}

has_text <- function(x) {
    !is.na(x) & nzchar(as.character(x))
}

short_label <- function(x, width = 65L) {
    ifelse(nchar(x) > width, paste0(substr(x, 1L, width - 3L), "..."), x)
}
```

### Install External Dependencies

```{r, eval=FALSE}
r <- getOption("repos")
if (is.null(r) || is.na(r["CRAN"]) || r["CRAN"] %in% c("@CRAN@", "")) {
  options(repos = c(CRAN = "https://cloud.r-project.org"))
}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

if (!requireNamespace("tidyr", quietly = TRUE)) {
    install.packages("tidyr")
}

if (!requireNamespace("scales", quietly = TRUE)) {
    install.packages("scales")
}

if (!requireNamespace("IlluminaHumanMethylation450kanno.ilmn12.hg19", quietly = TRUE)) {
    BiocManager::install("IlluminaHumanMethylation450kanno.ilmn12.hg19", update = FALSE, ask = FALSE)
}

if (!requireNamespace("TxDb.Hsapiens.UCSC.hg19.knownGene", quietly = TRUE)) {
    BiocManager::install("TxDb.Hsapiens.UCSC.hg19.knownGene", update = FALSE, ask = FALSE)
}

if (!requireNamespace("org.Hs.eg.db", quietly = TRUE)) {
    BiocManager::install("org.Hs.eg.db", update = FALSE, ask = FALSE)
}

if (!requireNamespace("missMethyl", quietly = TRUE)) {
    BiocManager::install("missMethyl", update = FALSE, ask = FALSE)
}
```

## Load Data
Covariates are:
 - predictedSex: the sex predicted from methylation data, which can be a strong confounder in methylation analyses.
 - subject.id.ch1: the patient ID
 - source_name_ch1: the cell type
```{r}
covariates <- c("predictedSex", "subject.id.ch1", "source_name_ch1")
sample_group_col <- "subject.status.ch1"

beta <- getBenchmarkFile("GSE63409_beta_values.tsv.gz")
samplesheet <- getBenchmarkFile("GSE63409_samplesheet.tsv")

zenodo_base_url <- "https://zenodo.org/records/20592944/files/"
if (!file.exists(beta)) {
    curl::curl_download(
        url = paste0(zenodo_base_url, basename(beta), "?download=1"),
        destfile = beta
    )
}
if (!file.exists(samplesheet)) {
    curl::curl_download(
        url = paste0(zenodo_base_url, basename(samplesheet), "?download=1"),
        destfile = samplesheet
    )
}


pheno <- fread(samplesheet)
pheno[[sample_group_col]] <- factor(pheno[[sample_group_col]], levels = c("normal", "AML"))
pheno <- as.data.frame(pheno)
rownames(pheno) <- pheno$V1
array_type <- "450K"
genome <- "hg19"

beta_handler <- CMEnt::getBetaHandler(beta, array = array_type, genome = genome)
beta_mat <- as.matrix(beta_handler$getBeta())
locs <- beta_handler$getBetaLocs()
logit2 <- function(x) log2(x) - log2(1 - x)
mvalues <- logit2(beta_mat)
pheno$casecontrol <- pheno[[sample_group_col]] == "AML"
```

## Find Differentially Methylated Positions

```{r}
dmps_file <- getBenchmarkFile("dmps.rds")
if (!file.exists(dmps_file)) {
    dmps <- CMEnt::findDMPsArray(
        beta_handler,
        pheno,
        sample_group_col = sample_group_col,
        id_col = "V1",
        covariates = covariates,
        array = array_type,
        genome = genome,
        njobs = max(1L, min(8L, parallel::detectCores(logical = TRUE) - 1L))
    )
    # sort DMPs by location
    rownames(dmps) <- dmps$site_id
    dmps <- dmps[rownames(locs), , drop = FALSE]
    dmps <- dmps[!is.na(dmps$pval), ]
    dmps$qval <- p.adjust(dmps$pval, method = "BH")

    case_idx <- which(pheno$casecontrol)
    ctrl_idx <- which(!pheno$casecontrol)
    dmps$delta_beta <- rowMeans(beta_mat[rownames(dmps), case_idx, drop = FALSE]) -
        rowMeans(beta_mat[rownames(dmps), ctrl_idx, drop = FALSE])

    dmps_gr <- makeGRangesFromDataFrame(
        data.frame(
            chr = locs$chr,
            start = locs$start,
            end = locs$end,
            pval = dmps$pval,
            qval = dmps$qval,
            stat = dmps$qval,
            delta_beta = dmps$delta_beta
        ),
        keep.extra.columns = TRUE
    )
    names(dmps_gr) <- rownames(dmps)
    saveRDS(dmps_gr, dmps_file)
} else {
    dmps_gr <- readRDS(dmps_file)
}
dmps <- as.data.frame(dmps_gr)
sig_dmps <- dmps[dmps$qval < 0.05 & abs(dmps$delta_beta) >= 0.05, ]
cat(sprintf("Found %d significant DMPs (FDR < 0.05, |delta beta| >= 0.05)\n", nrow(sig_dmps)))
```

## Method 1: CMEnt

```{r CMEnt}
cment_file <- getBenchmarkFile("dmrs.CMEnt.rds")
if (!file.exists(cment_file)) {
    cment_njobs <- max(1L, min(8L, parallel::detectCores(logical = TRUE) - 1L))
    options("CMEnt.verbose" = 1)
    options("CMEnt.njobs" = cment_njobs)
    cat(sprintf("CMEnt parallel jobs: %d\n", cment_njobs))
    cat(sprintf("parallel::detectCores(): %d\n", parallel::detectCores(logical = TRUE)))
    cat("Starting CMEnt::buildDMRs()...\n")
    flush.console()
    start_time <- Sys.time()
    dmrs_cment <- CMEnt::buildDMRs(
        beta = beta_handler,
        seeds = sig_dmps,
        pheno = pheno,
        sample_group_col = sample_group_col,
        casecontrol_col = "casecontrol",
        covariates = covariates,
        max_bridge_seeds_gaps  = 1,
        max_bridge_extension_gaps = 3,
        min_seeds = 2,
        min_sites = 3,
        ext_site_delta_beta = 0.2,
        max_lookup_dist = 1000,
        max_pval = 0.05,
        entanglement = "weak",
        njobs = cment_njobs,
        verbose = 1
    )
    end_time <- Sys.time()
    time_cment <- end_time - start_time
    saveRDS(list(dmrs = dmrs_cment, time = time_cment), cment_file)
} else {
    ret <- readRDS(cment_file)
    dmrs_cment <- ret$dmrs
    time_cment <- ret$time
}

cat(sprintf("CMEnt: Found %d DMRs in %d seconds\n", length(dmrs_cment), round(as.numeric(time_cment, units = "secs"))))
```

## CMEnt Region Summary

```{r}
if (!all(c("in_promoter_of", "in_gene_body_of", "delta_beta_promoter", "delta_beta_gene_body") %in% colnames(mcols(dmrs_cment))) &&
    requireNamespace("TxDb.Hsapiens.UCSC.hg19.knownGene", quietly = TRUE) &&
    requireNamespace("org.Hs.eg.db", quietly = TRUE)) {
    dmrs_cment <- CMEnt::annotateDMRsWithGenes(dmrs_cment, genome = genome)
}
for (gene_col in c("in_promoter_of", "in_gene_body_of")) {
    if (!gene_col %in% colnames(mcols(dmrs_cment))) {
        mcols(dmrs_cment)[[gene_col]] <- NA_character_
    }
}
for (delta_col in c("delta_beta_promoter", "delta_beta_gene_body")) {
    if (!delta_col %in% colnames(mcols(dmrs_cment))) {
        mcols(dmrs_cment)[[delta_col]] <- NA_real_
    }
}

dmr_delta_col <- if ("sites_delta_beta" %in% colnames(mcols(dmrs_cment))) {
    "sites_delta_beta"
} else {
    "delta_beta"
}
dmr_delta <- as.numeric(mcols(dmrs_cment)[[dmr_delta_col]])
dmr_direction <- ifelse(dmr_delta > 0, "Hypermethylated in AML", "Hypomethylated in AML")
dmr_summary <- data.frame(
    DMRs = length(dmrs_cment),
    Median_width_bp = median(width(dmrs_cment)),
    Mean_width_bp = round(mean(width(dmrs_cment)), 1),
    Total_coverage_bp = sum(width(dmrs_cment)),
    Median_delta_beta = round(median(dmr_delta, na.rm = TRUE), 4),
    Mean_abs_delta_beta = round(mean(abs(dmr_delta), na.rm = TRUE), 4),
    Hypermethylated_DMRs = sum(dmr_delta > 0, na.rm = TRUE),
    Hypomethylated_DMRs = sum(dmr_delta < 0, na.rm = TRUE)
)

knitr::kable(dmr_summary, caption = "CMEnt DMR summary for AML versus normal samples")
```

```{r, fig.width=9, fig.height=4}
dmrs_cment_only_promoter_or_gene_body <- dmrs_cment[!(is.finite(mcols(dmrs_cment)$delta_beta_promoter) & is.finite(mcols(dmrs_cment)$delta_beta_gene_body))]
dmr_effect_df <- bind_rows(
    data.frame(
        Region = "Promoter-overlapping DMRs",
        Width = width(dmrs_cment_only_promoter_or_gene_body),
        Delta_beta = as.numeric(mcols(dmrs_cment_only_promoter_or_gene_body)$delta_beta_promoter)
    ),
    data.frame(
        Region = "Gene-body-overlapping DMRs",
        Width = width(dmrs_cment_only_promoter_or_gene_body),
        Delta_beta = as.numeric(mcols(dmrs_cment_only_promoter_or_gene_body)$delta_beta_gene_body)
    )
) |>
    filter(is.finite(Delta_beta)) |>
    mutate(Direction = factor(
        ifelse(Delta_beta > 0, "Hypermethylated in AML", "Hypomethylated in AML"),
        levels = c("Hypermethylated in AML", "Hypomethylated in AML")
    ))

library(dplyr)
library(ggplot2)
library(scales)

ignore_abs_delta <- !is.finite(dmr_effect_df$Delta_beta) | abs(dmr_effect_df$Delta_beta) < 0.05
dmr_effect_df <- dmr_effect_df[!ignore_abs_delta,]
# Summary per Region and Direction
dmr_summary <- dmr_effect_df %>%
  group_by(Region, Direction) %>%
  summarise(
    median_delta_beta = median(Delta_beta, na.rm = TRUE),
    n_dmrs = n(),
    .groups = "drop"
  )

# Define two y-levels for the two summary bars
y_max <- max(dmr_effect_df$Width, na.rm = TRUE)

dmr_summary <- dmr_summary %>%
  mutate(
    bar_y = case_when(
      Direction == "Hypermethylated in AML" ~ y_max * 2.0,
      Direction == "Hypomethylated in AML"  ~ y_max * 3.2
    ),
    label_y = bar_y * 1.12
  )

# Expand y-range so the bars and labels fit
y_upper <- max(dmr_summary$label_y) * 1.15

ggplot(dmr_effect_df, aes(x = Delta_beta, y = Width, color = Direction)) +
  geom_hline(
    yintercept = median(dmr_effect_df$Width, na.rm = TRUE),
    linewidth = 0.2,
    color = "grey70"
  ) +
  geom_vline(xintercept = 0, linewidth = 0.2, color = "grey70") +
  geom_point(alpha = 0.55, size = 1.2) +

  # Separate horizontal bars for hypo and hyper
  geom_segment(
    data = dmr_summary,
    aes(
      x = 0,
      xend = median_delta_beta,
      y = bar_y,
      yend = bar_y,
      color = Direction
    ),
    inherit.aes = FALSE,
    linewidth = 5,
    lineend = "butt"
  ) +

  # Labels for each bar
  geom_text(
    data = dmr_summary,
    aes(
      x = median_delta_beta,
      y = label_y,
      label = paste0(
        "median Δβ = ", round(median_delta_beta, 3),
        "\nDMRs = ", n_dmrs
      )
    ),
    inherit.aes = FALSE,
    color = "black",
    size = 3,
    hjust = ifelse(dmr_summary$median_delta_beta >= 0, -0.05, 1.05)
  ) +

  facet_wrap(~Region, nrow = 1) +
  scale_y_log10(
    limits = c(min(dmr_effect_df$Width[dmr_effect_df$Width > 0], na.rm = TRUE), y_upper),
    labels = comma
  ) +
  scale_color_manual(
    values = c(
      "Hypermethylated in AML" = "#B8405E",
      "Hypomethylated in AML" = "#247BA0"
    )
  ) +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  labs(
    title = "CMEnt DMR Effect Size and Width by Gene Context",
    x = "Context-specific delta beta (AML - normal)",
    y = "DMR width (bp, log scale)",
    color = ""
  )
```

## CpG and Gene Context

```{r}
gene_context_df <- data.frame(
    Context = c("Promoter", "Gene body", "Promoter and gene body", "No annotated gene overlap"),
    DMRs = c(
        sum(has_text(mcols(dmrs_cment)$in_promoter_of)),
        sum(has_text(mcols(dmrs_cment)$in_gene_body_of)),
        sum(has_text(mcols(dmrs_cment)$in_promoter_of) & has_text(mcols(dmrs_cment)$in_gene_body_of)),
        sum(!has_text(mcols(dmrs_cment)$in_promoter_of) & !has_text(mcols(dmrs_cment)$in_gene_body_of))
    )
) |>
    mutate(Percentage = round(100 * DMRs / length(dmrs_cment), 1))

knitr::kable(gene_context_df, caption = "Gene context of CMEnt DMRs")
```

```{r, fig.width=6, fig.height=3.5}
ggplot(gene_context_df, aes(x = reorder(Context, DMRs), y = DMRs, fill = Context)) +
    geom_col(width = 0.7) +
    coord_flip() +
    scale_y_continuous(labels = scales::comma) +
    theme_minimal() +
    theme(legend.position = "none") +
    labs(title = "CMEnt DMR Gene Context", x = "", y = "DMRs")
```

```{r}
if (requireNamespace("IlluminaHumanMethylation450kanno.ilmn12.hg19", quietly = TRUE)) {
    library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
    anno <- as.data.frame(minfi::getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19))
    anno$probe_id <- rownames(anno)

    dmr_site_ids <- split_ids(mcols(dmrs_cment)$sites)
    dmr_site_context <- anno[intersect(dmr_site_ids, rownames(anno)), c("probe_id", "Relation_to_Island")]
    dmp_site_context <- anno[intersect(rownames(sig_dmps), rownames(anno)), c("probe_id", "Relation_to_Island")]

    site_context_df <- bind_rows(
        dmp_site_context |> mutate(Source = "Significant DMPs"),
        dmr_site_context |> mutate(Source = "Sites inside CMEnt DMRs")
    ) |>
        count(Source, Relation_to_Island, name = "Sites") |>
        group_by(Source) |>
        mutate(Percentage = round(100 * Sites / sum(Sites), 1)) |>
        ungroup()

    knitr::kable(site_context_df, caption = "CpG island context of significant DMPs and sites inside CMEnt DMRs")
}
```

```{r, fig.width=8, fig.height=4}
if (exists("site_context_df")) {
    context_levels <- site_context_df |>
        group_by(Relation_to_Island) |>
        summarise(Max_sites = max(Sites), .groups = "drop") |>
        arrange(Max_sites) |>
        pull(Relation_to_Island)
    plot_context_df <- site_context_df |>
        mutate(
            Relation_to_Island = factor(Relation_to_Island, levels = context_levels),
            Source = factor(Source, levels = c("Significant DMPs", "Sites inside CMEnt DMRs"))
        )
    diff_df <- plot_context_df |>
        dplyr::select(Source, Relation_to_Island, Sites) |>
        tidyr::pivot_wider(names_from = Source, values_from = Sites, values_fill = 0) |>
        dplyr::mutate(
            Difference = `Sites inside CMEnt DMRs` - `Significant DMPs`,
            Label = paste0("Delta=", scales::comma(Difference)),
            Label_x = pmax(`Significant DMPs`, `Sites inside CMEnt DMRs`)
        )

    ggplot(plot_context_df, aes(x = Sites, y = Relation_to_Island, fill = Source)) +
        geom_col(data = dplyr::filter(plot_context_df, Source == "Significant DMPs"), width = 0.78, alpha = 0.45) +
        geom_col(data = dplyr::filter(plot_context_df, Source == "Sites inside CMEnt DMRs"), width = 0.48, alpha = 0.9) +
        geom_text(data = diff_df, aes(x = Label_x, y = Relation_to_Island, label = Label), inherit.aes = FALSE, hjust = -0.05, size = 3) +
        scale_x_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.2))) +
        scale_fill_manual(values = c("Significant DMPs" = "grey65", "Sites inside CMEnt DMRs" = "#0072B2")) +
        theme_minimal() +
        labs(title = "CpG Island Context", x = "Sites", y = "", fill = "")
}
```

## AML Pathway and Marker Gene Scan

The table below is a targeted DMR scan of AML genetics, signaling, and hematopoietic/myeloid marker genes.

```{r}
gene_pattern <- function(gene) {
    paste0("(^|,)", gene, "($|,)")
}

dmr_gene_summary <- function(gene) {
    promoter_hits <- grepl(gene_pattern(gene), as.character(mcols(dmrs_cment)$in_promoter_of))
    body_hits <- grepl(gene_pattern(gene), as.character(mcols(dmrs_cment)$in_gene_body_of))
    dmr_hits <- promoter_hits | body_hits
    data.frame(
        Gene = gene,
        CMEnt_DMRs = sum(dmr_hits),
        Promoter_DMRs = sum(promoter_hits),
        Gene_body_DMRs = sum(body_hits),
        Mean_DMR_delta_beta = round(mean(dmr_delta[dmr_hits], na.rm = TRUE), 4)
    )
}

aml_prior_genes <- data.frame(
    Gene = c(
        "NPM1", "FLT3", "KIT", "RUNX1", "TP53", "TET2", "DNMT3A", "CXCR4", "NAT10",
        "JAK2", "STAT3", "STAT5A", "STAT5B", "HOXA9", "HOXA10",
        "CD34", "ITGAM", "SPI1", "CEBPA", "GATA2"
    ),
    Theme = c(
        rep("AML genetics/prognosis", 9),
        rep("JAK-STAT/HOXA signaling", 6),
        rep("Hematopoietic and myeloid differentiation", 5)
    )
)

prior_dmr_summary <- bind_rows(lapply(aml_prior_genes$Gene, dmr_gene_summary))

prior_gene_table <- aml_prior_genes |>
    left_join(prior_dmr_summary, by = "Gene") |>
    mutate(
        CMEnt_DMRs = tidyr::replace_na(CMEnt_DMRs, 0L),
        Methylation_signal = CMEnt_DMRs > 0
    ) |>
    arrange(desc(Methylation_signal), desc(CMEnt_DMRs), Gene)

knitr::kable(prior_gene_table, digits = 4, caption = "Targeted AML gene scan using CMEnt DMRs")
```

```{r, fig.width=8, fig.height=4}
prior_plot_df <- prior_gene_table |>
    filter(Methylation_signal) |>
    mutate(Gene = reorder(Gene, CMEnt_DMRs))

if (nrow(prior_plot_df) > 0) {
    ggplot(prior_plot_df, aes(x = Gene, y = CMEnt_DMRs, fill = Theme)) +
        geom_col(width = 0.7) +
        coord_flip() +
        theme_minimal() +
        labs(title = "AML Genes Overlapping CMEnt DMRs", x = "", y = "CMEnt DMRs", fill = "")
}
```

## GO and KEGG Enrichment

`missMethyl` performs methylation-array-aware enrichment for CMEnt DMRs while correcting for
probe-number bias across genes.

```{r}
aml_term_pattern <- paste(
    c(
        "acute myeloid", "leukemia", "leukaemia", "myeloid", "hematopoiet", "haematopoiet",
        "stem cell", "differentiation", "immune", "leukocyte", "lymphocyte", "granulocyte",
        "monocyte", "macrophage", "cytokine", "chemokine", "phagosome", "JAK", "STAT",
        "HOX", "FLT3", "NPM1", "DNA methylation", "apoptosis", "cell cycle",
        "transcriptional misregulation"
    ),
    collapse = "|"
)

normalize_missmethyl_result <- function(x, collection) {
    if (is.null(x) || nrow(x) == 0L) {
        return(NULL)
    }
    x <- as.data.frame(x)
    term_col <- intersect(c("Term", "TERM", "Description", "Pathway"), colnames(x))[1]
    fdr_col <- intersect(c("FDR", "adj.P.Val", "P.DE", "P.Value"), colnames(x))[1]
    p_col <- intersect(c("P.DE", "P.Value", "P.DE.weighted"), colnames(x))[1]
    de_col <- intersect(c("DE", "N_DE", "Count"), colnames(x))[1]
    n_col <- intersect(c("N", "Total", "Size"), colnames(x))[1]
    ont_col <- intersect(c("Ont", "Ontology", "Category"), colnames(x))[1]
    if (is.na(term_col) || is.na(fdr_col)) {
        return(NULL)
    }
    data.frame(
        Collection = collection,
        Term = as.character(x[[term_col]]),
        Ontology = if (!is.na(ont_col)) as.character(x[[ont_col]]) else collection,
        N = if (!is.na(n_col)) suppressWarnings(as.numeric(x[[n_col]])) else NA_real_,
        DE = if (!is.na(de_col)) suppressWarnings(as.numeric(x[[de_col]])) else NA_real_,
        PValue = if (!is.na(p_col)) suppressWarnings(as.numeric(x[[p_col]])) else NA_real_,
        FDR = suppressWarnings(as.numeric(x[[fdr_col]])),
        stringsAsFactors = FALSE
    )
}

missmethyl_long <- NULL
if (requireNamespace("missMethyl", quietly = TRUE) &&
    requireNamespace("IlluminaHumanMethylation450kanno.ilmn12.hg19", quietly = TRUE)) {
    library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
    cached <- getBenchmarkFile("missmethyl_cment_enrichment_results.rds")
    if (file.exists(cached)) {
        missmethyl_long <- readRDS(cached)
    } else {
        missmethyl_long <- base::do.call(rbind, Filter(Negate(is.null), lapply(c("GO", "KEGG"), function(collection) {
            ret <- tryCatch(
                missMethyl::goregion(
                    regions = dmrs_cment,
                    all.cpg = rownames(locs),
                    collection = collection,
                    array.type = "450K",
                    plot.bias = FALSE,
                    prior.prob = TRUE
                ),
                error = function(e) {
                    warning(sprintf("missMethyl %s enrichment failed for CMEnt DMRs: %s", collection, e$message))
                    NULL
                }
            )
            normalize_missmethyl_result(ret, collection)
        })))
        saveRDS(missmethyl_long, cached)
    }
} else {
    cat("missMethyl or the 450K annotation package is required for GO/KEGG enrichment.\n")
}
```

```{r}
if (!is.null(missmethyl_long) && nrow(missmethyl_long) > 0L) {
    missmethyl_long <- missmethyl_long[is.finite(missmethyl_long$FDR), , drop = FALSE]
    missmethyl_long$NegLog10FDR <- -log10(pmax(missmethyl_long$FDR, .Machine$double.xmin))
    missmethyl_long$AML_Context <- ifelse(grepl(aml_term_pattern, missmethyl_long$Term, ignore.case = TRUE), "AML-context term", "Other top term")

    missmethyl_top <- missmethyl_long |>
        group_by(Collection) |>
        arrange(FDR, .by_group = TRUE) |>
        slice_head(n = 10) |>
        ungroup()
    missmethyl_top$Term_Label <- short_label(missmethyl_top$Term, 70)
    missmethyl_top$Term_Collection <- factor(
        paste(missmethyl_top$Term_Label, missmethyl_top$Collection, sep = "___"),
        levels = rev(unique(paste(missmethyl_top$Term_Label, missmethyl_top$Collection, sep = "___")))
    )

    knitr::kable(
        missmethyl_top |>
            dplyr::select(Collection, Ontology, Term, N, DE, PValue, FDR, AML_Context) |>
            arrange(Collection, FDR),
        digits = 3,
        row.names = FALSE,
        caption = "Top missMethyl GO/KEGG enrichments from CMEnt DMRs"
    )
}
```

```{r, fig.width=10, fig.height=6}
if (!is.null(missmethyl_long) && nrow(missmethyl_long) > 0L) {
    ggplot(missmethyl_top, aes(x = NegLog10FDR, y = Term_Collection, size = DE, color = AML_Context)) +
        geom_point(alpha = 0.9) +
        facet_wrap(~Collection, scales = "free_y") +
        scale_y_discrete(labels = function(x) sub("___.*$", "", x)) +
        scale_color_manual(values = c("AML-context term" = "#B8405E", "Other top term" = "grey50")) +
        theme_minimal() +
        theme(axis.text.y = element_text(size = 7)) +
        labs(title = "missMethyl Enrichment", x = expression(-log[10]("FDR")), y = "", color = "", size = "DM genes")
} else {
    cat("missMethyl returned no enrichment results.\n")
}
```

```{r}
focused_terms <- if (!is.null(missmethyl_long) && nrow(missmethyl_long) > 0L) {
    missmethyl_long |>
        filter(grepl(aml_term_pattern, Term, ignore.case = TRUE)) |>
        group_by(Collection) |>
        arrange(FDR, .by_group = TRUE) |>
        slice_head(n = 8) |>
        ungroup()
} else {
    data.frame()
}

if (nrow(focused_terms) > 0L) {
    focused_terms$Term_Label <- short_label(focused_terms$Term, 62)
    focused_terms$Term_Collection <- factor(
        paste(focused_terms$Term_Label, focused_terms$Collection, sep = "___"),
        levels = rev(unique(paste(focused_terms$Term_Label, focused_terms$Collection, sep = "___")))
    )
    knitr::kable(
        focused_terms |>
            dplyr::select(Collection, Ontology, Term, N, DE, PValue, FDR) |>
            arrange(Collection, FDR),
        digits = 3,
        row.names = FALSE,
        caption = "Top AML-relevant GO/KEGG terms"
    )
}
```

```{r, fig.width=10, fig.height=5}
if (nrow(focused_terms) > 0L) {
    ggplot(focused_terms, aes(x = -log10(pmax(FDR, .Machine$double.xmin)), y = Term_Collection, fill = Collection)) +
        geom_col(width = 0.7) +
        facet_wrap(~Collection, scales = "free_y") +
        scale_y_discrete(labels = function(x) sub("___.*$", "", x)) +
        theme_minimal() +
        theme(axis.text.y = element_text(size = 8), legend.position = "none") +
        labs(title = "AML-Relevant Enrichment Terms", x = expression(-log[10]("FDR")), y = "")
} else {
    cat("No missMethyl terms matched the AML-focused keyword set.\n")
}
```

## Candidate CMEnt Gene Overlapping Loci

```{r}
dmr_table <- as.data.frame(dmrs_cment)
dmr_table$DMR <- paste0(dmr_table$seqnames, ":", dmr_table$start, "-", dmr_table$end)
dmr_table$delta_beta_for_ranking <- dmr_delta
dmr_table$abs_delta_beta <- abs(dmr_delta)
dmr_table$svm_score_for_ranking <- if ("score" %in% colnames(dmr_table)) dmr_table$score else 0
if (!"qval" %in% colnames(dmr_table)) {
    dmr_table$qval <- NA_real_
}
if (!"pval" %in% colnames(dmr_table)) {
    dmr_table$pval <- NA_real_
}
if (!"score" %in% colnames(dmr_table)) {
    dmr_table$score <- NA_real_
}

top_dmrs <- dmr_table |>
    filter(has_text(in_promoter_of) | has_text(in_gene_body_of)) |>
    arrange(desc(svm_score_for_ranking), qval, desc(abs_delta_beta)) |>
    dplyr::select(
        DMR,
        width,
        score,
        qval,
        pval,
        delta_beta_for_ranking,
        any_of(c("delta_beta_promoter", "delta_beta_gene_body", "seeds_num", "sites_num", "in_promoter_of", "in_gene_body_of"))
    ) |>
    head(20)

knitr::kable(top_dmrs, digits = 4, caption = "Top 20 promoter or gene-body overlapping CMEnt DMRs ranked by score")
```

## Top CMEnt DMR Visualization
```{r, fig.width=6, fig.height=6}
CMEnt::plotDMRs(
    dmrs_cment,
    top_n = 1,
    score_by = "score",
    beta = beta_handler,
    pheno = pheno,
    sample_group_col = sample_group_col,
    genome = "hg19",
    array = array_type
)
```

## CMEnt DMR Manhattan Plot
```{r, fig.width=10, fig.height=4}
CMEnt::plotDMRsManhattan(
    dmrs_cment,
    genome = "hg19")
```

## CMEnt DMR Interaction Circos Visualization
```{r, fig.width=10, fig.height=10}
CMEnt::plotDMRsCircos(
    dmrs_cment,
    genome = "hg19",
    array = array_type
)
```


## Insights and Interpretation
This CMEnt re-analysis of GSE63409 extends the original Hu et al. 
study from a gene-level methylation and expression biomarker analysis into a region-level interpretation 
of AML-associated methylation architecture. Whereas the original publication 
integrated methylation and expression data to nominate AML-relevant pathways
and three prognostic genes, ATP11A, ITGAM, and ZNRF2, this notebook shows that 
the same methylation dataset contains a broad and spatially coherent regional
signal: 72,556 significant DMPs were assembled into 8,654 CMEnt DMRs,
covering about 15.1 Mb of the genome, with a slight excess of hypermethylated regions in AML.

A main new observation is that the regional signal is not randomly distributed across
gene annotation space. Most CMEnt DMRs overlap annotated genes, especially gene bodies,
and many overlap both promoter and gene-body annotations. This suggests that AML-associated 
methylation changes in this dataset are not limited to isolated promoter CpGs, but frequently 
form extended regulatory blocks spanning transcriptional units. The effect-size plot
further indicates that both promoter-overlapping and gene-body-overlapping DMRs contain
clear hypermethylated and hypomethylated subclasses, with gene-body DMRs contributing
the larger number of region-level events.

Although significant DMPs are most frequent in OpenSea regions,
CpGs retained inside CMEnt DMRs are strongly enriched in CpG islands and shores.
This indicates that CMEnt preferentially captures locally coherent methylation
changes in CpG-dense regulatory neighborhoods, rather than simply preserving the
genome-wide distribution of significant single CpGs. In practical terms,
the analysis reframes part of the AML methylation signal as coordinated regional
dysregulation around CpG island-associated loci.

The targeted AML gene scan also supports biological relevance while adding
positional detail. CMEnt DMRs overlap several AML-associated regulatory
and differentiation genes, including HOXA9, HOXA10, RUNX1, STAT3, CD34,
and GATA2. This partly recapitulates the original study’s emphasis
on hematopoietic differentiation and AML-relevant signaling, 
but adds information about whether these signals occur as promoter,
gene-body, or mixed regional methylation events. Notably, RUNX1 and STAT3
show strong hypomethylated DMR signals, while HOXA10 and CD34 show
hypermethylated regional signals, suggesting that AML-associated methylation
changes are directionally heterogeneous across disease-relevant loci.

The enrichment results further reinforce the biological interpretation. 
The strongest GO terms are broad developmental, cell-periphery, cell-signaling,
and differentiation categories, while AML-context terms include cell differentiation,
regulation of differentiation, stem-cell pluripotency signaling, hematopoietic cell lineage,
and cytokine-related pathways. Compared with the original study’s pathway-level analysis, 
CMEnt adds the observation that these biological themes are supported by coherent DMR structures 
rather than only by individually significant CpGs or methylation-expression gene intersections.

Finally, the candidate-region and visualization figures provide concrete follow-up loci.
The top-scoring DMR overlaps CD247 on chromosome 1, shows clear group separation with
cross-validation accuracy of 0.85, and is hypomethylated in AML.
Other high-ranking regions include PVT1, RUNX1-associated regulatory candidates
through the AML scan, POU5F1, SH3PXD2A, TLE7, KCNQ1, PLIN5, and MAFG/SIRT7-associated loci.
The Manhattan and circos plots show that high-scoring DMRs are distributed across many chromosomes,
with 115 identified DMR blocks and recurrent motif-interaction patterns involving factors such as
KLF-family motifs and AHR::ARNT-like motifs.

Overall, this analysis supports the original conclusion that AML is associated with dysregulated
methylation in genes and pathways linked to hematopoietic differentiation,
immune signaling, and development. Its main added value is that it converts
the single-CpG signal into interpretable regional methylation structures,
revealing CpG island-enriched DMR organization, promoter and gene-body-specific
effect patterns, and candidate loci that can be prioritized for biological validation.

## Session Info

```{r}
sessionInfo()
```
