---
title: "CMEnt Analysis of ONT Methylation Data on brain tumors"
subtitle: "Biological interpretation of GSE289246"
author: "CMEnt Package"
date: "`r Sys.Date()`"
output: 
    BiocStyle::html_document:
        code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    results = "asis",
    cache = TRUE,
    echo = TRUE,
    dev = "png",
    warning = FALSE,
    message = FALSE,
    fig.width = 8,
    fig.height = 6,
    dpi = 300
)

try(devtools::load_all(), silent = TRUE)
suppressPackageStartupMessages({
    try(library(CMEnt), silent = TRUE)
    library(curl)
    library(GenomicRanges)
    library(GEOquery)
    library(SummarizedExperiment)
    library(data.table)
    library(bsseq)
    library(HDF5Array)
    library(ggplot2)
    library(dplyr)
    library(scales)
})


getBenchmarkFile <- function(filename) {
    benchmark_file <- ""
    folder <- "ont"
    if (getwd() == "notebooks") {
        benchmark_file <- file.path("cached_results", folder, filename)
    } else if ("NAMESPACE" %in% dir()) {
        benchmark_file <- file.path("notebooks", "cached_results", folder, filename)
    } else {
        benchmark_file <- file.path("cached_results", folder, filename)
    }
    dir.create(dirname(benchmark_file), showWarnings = FALSE, recursive = TRUE)
    benchmark_file
}

split_ids <- function(x) {
    ids <- unlist(strsplit(stats::na.omit(as.character(x)), ","), use.names = FALSE)
    unique(ids[nzchar(ids)])
}

has_text <- function(x) {
    !is.na(x) & nzchar(as.character(x))
}

short_label <- function(x, width = 65L) {
    ifelse(nchar(x) > width, paste0(substr(x, 1L, width - 3L), "..."), x)
}
```

## Introduction

This vignette analyzes GSE289246, a long-read ONT methylation dataset from brain tumors.
We compare Pilocytic astrocytoma, used here as the benign reference group, against
Glioblastoma, IDH-wildtype, used here as the aggressive cancer group. The goal is to
identify coherent methylation regions and summarize which gene contexts and candidate
biology distinguish the two tumor types.

- CMEnt DMR size, score, and methylation effect summaries
- promoter and gene-body overlap patterns
- DMR-level methylation heatmaps separating benign and aggressive samples
- cancer-focused candidate gene loci and genome-wide visualization

## Setup and Dependencies

```{r install, eval=FALSE}
r <- getOption("repos")
if (is.null(r) || is.na(r["CRAN"]) || r["CRAN"] %in% c("@CRAN@", "")) {
    options(repos = c(CRAN = "https://cloud.r-project.org"))
}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}
for (pkg in c("DSS", "TxDb.Hsapiens.UCSC.hg19.knownGene", "org.Hs.eg.db", "BSgenome.Hsapiens.UCSC.hg19")) {
    if (!requireNamespace(pkg, quietly = TRUE)) {
        BiocManager::install(pkg, update = FALSE, ask = FALSE)
    }
}
```
## Load Data

Downloading and parsing the GEO dataset samplesheet.
We select 2 cancers, 5 samples of Pilocytic astrocytoma, a benign cancer,
and 21 samples of Glioblastoma, IDH-wildtype, an aggressive cancer, all sequenced with PromethION with the R10.4.1 chemistry.
```{r samplesheet_creation}

sample_group_col <- "tumor_type"
casecontrol_col <- "aggressive"
genome <- "hg19" # The authors of the original study used hg19.
benign_label <- "Pilocytic astrocytoma"
aggressive_label <- "Glioblastoma, IDH-wildtype"

URL <- "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE289nnn/GSE289246/soft/GSE289246_family.soft.gz"
samplesheet_file <- getBenchmarkFile("samplesheet.tsv")
if (!file.exists(samplesheet_file)) {
    message("Downloading and parsing the GEO dataset samplesheet: ", URL)
    soft_file <- getBenchmarkFile("GSE289246_family.soft.gz")
    if (!file.exists(soft_file)) {
        curl::curl_download(URL, soft_file)
    }
    soft_data <- getGEO(filename = soft_file)
    gsms <- attr(soft_data, "gsms")
    samplesheet <- list()
    colnames <- unique(unlist(sapply(names(gsms), function(x) names(attr(gsms[[x]], "header")))))
    for (gsm in names(gsms)) {
        samplesheet[[gsm]] <- sapply(
            attr(gsms[[gsm]], "header"), function(x) {
                if (length(x) > 1) {
                    x <- paste(x, collapse = "; ")
                    x <- gsub("\\s+", " ", x)
                }
                x
            }
        )
        missing_cols <- setdiff(colnames, names(samplesheet[[gsm]]))
        if (length(missing_cols) > 0) {
            samplesheet[[gsm]][missing_cols] <- NA
            samplesheet[[gsm]] <- samplesheet[[gsm]][colnames]
        }
    }
    samplesheet <- do.call(rbind, samplesheet)
    pheno <- samplesheet[samplesheet[, "library_strategy"] == "OTHER", ]
    instrument_models <- table(pheno[, "instrument_model"])
    pheno <- pheno[pheno[, "instrument_model"] == "PromethION", ]
    case1 <- "tissue: tumor; tumor type: Pilocytic astrocytoma"
    case2 <- "tissue: tumor; tumor type: Glioblastoma, IDH-wildtype"
    pheno <- pheno[pheno[, "characteristics_ch1"] %in% c(case1, case2), ]
    unlink(soft_file)
    write.table(pheno, samplesheet_file, row.names = TRUE, col.names = NA, sep = "\t", quote = FALSE)
} else {
    pheno <- read.table(samplesheet_file, header = TRUE, row.names = 1, sep = "\t", quote = "", stringsAsFactors = FALSE, check.names = FALSE)
}
pheno <- as.data.frame(pheno)
pheno$tumor_type <- ifelse(
    grepl(aggressive_label, pheno$characteristics_ch1, fixed = TRUE),
    aggressive_label,
    benign_label
)
pheno$tumor_type <- factor(pheno$tumor_type, levels = c(benign_label, aggressive_label))
pheno$aggressive <- pheno$tumor_type == aggressive_label
pheno$Sample_ID <- rownames(pheno)

knitr::kable(
    as.data.frame(table(Tumor_type = pheno$tumor_type)),
    caption = "Selected GSE289246 ONT samples"
)
```
The long-read ONT methylation data has low coverage, being low depth sequencing, so we will softly filter out sites with coverage < 2.
```{r bsseq_creation}
min_coverage <- 2
input_bsseq_h5 <- getBenchmarkFile("input_h5")
if (file.exists(input_bsseq_h5)) {
    message("Loading BSseq object from ", input_bsseq_h5)
    bsseq_obj <- HDF5Array::loadHDF5SummarizedExperiment(input_bsseq_h5)
} else {
    source_bed_files <- pheno[, "supplementary_file_1"]
    names(source_bed_files) <- rownames(pheno)
    target_bed_files <- sapply(names(source_bed_files), function(x) {
        getBenchmarkFile(file.path("downloaded", paste0(x, ".bed.gz")))
    })
    # download the bed files to the raw_data directory
    for (sample_name in names(source_bed_files)) {
        if (!file.exists(target_bed_files[[sample_name]])) {
            message("Downloading ", source_bed_files[[sample_name]])
            curl::curl_download(source_bed_files[[sample_name]], target_bed_files[[sample_name]])
        }
    }

    # Extract the methylation data from the bed files and create a BSseq object
    bsseq_objects <- list()
    progress_bar <- txtProgressBar(min = 0, max = length(target_bed_files), style = 3)
    target_h5_dirs <- sapply(names(target_bed_files), function(x) {
        getBenchmarkFile(file.path("h5", paste0(x)))
    })
    for (sample_name in names(target_bed_files)) {
        target_h5_folder <- target_h5_dirs[[sample_name]]
        if (file.exists(file.path(target_h5_folder, "assays.h5"))) {
            setTxtProgressBar(progress_bar, which(names(target_bed_files) == sample_name))
            next
        }
        bed_file <- target_bed_files[[sample_name]]
        df <- data.table::fread(bed_file, header = TRUE, sep = "\t", stringsAsFactors = FALSE)
        if (!exists("gr_list")) {
            gr_list <- list()
        }
        gr <- GRanges(
            seqnames = df$chrom,
            ranges = IRanges(start = df$start + 1L, end = df$end),
            strand = "*"
        )
        M <- as.matrix(as.integer(df$percentMeth * df$coverage / 100), ncol = 1)
        Cov <- as.matrix(as.integer(df$coverage), ncol = 1)


        bsseq_obj <- BSseq(
            gr = gr,
            M = M,
            Cov = Cov,
            sampleNames = sample_name
        )
        to_keep <- Cov >= min_coverage
        if (sum(to_keep) == 0) {
            warning("No sites with coverage >= ", min_coverage, " for sample ", sample_name)
            next
        }
        if (sum(to_keep) < nrow(bsseq_obj)) {
            nfiltered <- nrow(bsseq_obj) - sum(to_keep)
            message(
                "Keeping ", sum(to_keep), " sites out of ", nrow(bsseq_obj), " (",
                round(sum(to_keep) / nrow(bsseq_obj) * 100, 2),
                "%) with coverage >= ", min_coverage, " for sample ", sample_name
            )
        }
        bsseq_obj <- bsseq_obj[to_keep, ]
        HDF5Array::saveHDF5SummarizedExperiment(bsseq_obj, dir = target_h5_folder)
        setTxtProgressBar(progress_bar, which(names(target_bed_files) == sample_name))
    }

    bsseq_objects <- list()
    for (sample_name in names(target_bed_files)) {
        target_h5_folder <- target_h5_dirs[[sample_name]]
        bsseq_obj <- HDF5Array::loadHDF5SummarizedExperiment(target_h5_folder)
        bsseq_objects[[sample_name]] <- bsseq_obj
    }
    suppressWarnings({
        bsseq_obj <- do.call(bsseq::combineList, list(bsseq_objects, BACKEND = "HDF5Array"))
    })
    bsseq_obj <- HDF5Array::saveHDF5SummarizedExperiment(bsseq_obj, dir = input_bsseq_h5)
}
```
## Find Differentially Methylated Positions

```{r seeds_calculation}
seeds_file <- getBenchmarkFile("seeds.rds")
if (!file.exists(seeds_file)) {
    dmps_file <- getBenchmarkFile("dmps.rds")
    if (!file.exists(dmps_file)) {
        message("Calculating DMPs...")
        dmps <- CMEnt::findDMPsBSSeq(
            bsseq_obj,
            samplesheet = pheno,
            sample_group_col = sample_group_col,
            id_col = "Sample_ID",
            case_group = aggressive_label,
            njobs = max(1L, min(8L, parallel::detectCores(logical = TRUE) - 1L))
        )
        saveRDS(dmps, dmps_file)
    } else {
        message("Loading DMPs from ", dmps_file)
        dmps <- readRDS(dmps_file)
    }
    seeds <- dmps[dmps$pval < 0.05, ]
    saveRDS(seeds, seeds_file)
} else {
    seeds <- readRDS(seeds_file)
}
rownames(seeds) <- seeds$site_id
seeds$Row.names <- NULL
seed_dist <- unlist(lapply(split(seeds$start, seeds$chr), function(x) {
    if (length(x) < 2) return(numeric(0))
    diff(x)
}))
seeds_qs <- round(stats::quantile(seed_dist, probs = c(0, 0.25, 0.5, 0.75, 0.9, 0.99, 1)))
cat(sprintf("Number of seeds for CMEnt: %d\n", nrow(seeds)))
knitr::kable(
    seeds_qs,
    digits = 2,
    col.names = "Distance (bp)",
    caption = sprintf("Seed distance quantiles")
)
cat("BSseq object dimensions: ", dim(bsseq_obj), "\n")
```

## Method 1: CMEnt

```{r cment}
cment_file <- getBenchmarkFile("dmrs.rds")
if (!file.exists(cment_file)) {
    # Reduce the BSseq object around the seeds with the specified expansion window,
    # so that to minimize memory usage and speed up the DMR building process.
    reduced_bsseq_obj <- CMEnt::reduceBSseqAroundSeeds(bsseq_obj, seeds, expansion_window = 10000)
    cat("Reduced BSseq object dimensions: ", dim(reduced_bsseq_obj), "\n")
    start_time <- Sys.time()
    dmrs_cment <- CMEnt::buildDMRs(
        reduced_bsseq_obj,
        seeds,
        pheno,
        genome = genome,
        sample_group_col = sample_group_col,
        casecontrol_col = casecontrol_col,
        entanglement  = "weak",
        testing_mode  = "parametric",
        expansion_window = 10000, # The maximum distance (in bp) to expand the DMRs from the seeds
        min_seeds = 2, # The minimum number of seeds required to form a DMR
        min_sites = 50, # The minimum number of sites required to form a DMR
        max_pval = 0.05,
        ext_site_delta_beta = 0.1,
        max_lookup_dist = 1000,
        njobs = max(1L, min(8L, parallel::detectCores(logical = TRUE) - 1L)),
        verbose = 2
    )
    end_time <- Sys.time()
    time_cment <- end_time - start_time
    saveRDS(list(dmrs = dmrs_cment, time = time_cment), cment_file)
} else {
    message("Loading CMEnt DMRs from ", cment_file)
    cment_results <- readRDS(cment_file)
    dmrs_cment <- cment_results$dmrs
    time_cment <- cment_results$time
}
cat(sprintf("CMEnt: Found %d DMRs in %d seconds\n", length(dmrs_cment), round(as.numeric(time_cment, units = "secs"))))
```

## CMEnt Region Summary

```{r cment_summary}
if (length(dmrs_cment) > 0L &&
    !all(c("in_promoter_of", "in_gene_body_of", "delta_beta_promoter", "delta_beta_gene_body") %in% colnames(mcols(dmrs_cment))) &&
    requireNamespace("TxDb.Hsapiens.UCSC.hg19.knownGene", quietly = TRUE) &&
    requireNamespace("org.Hs.eg.db", quietly = TRUE)) {
    dmrs_cment <- CMEnt::annotateDMRsWithGenes(dmrs_cment, genome = genome)
}
for (gene_col in c("in_promoter_of", "in_gene_body_of")) {
    if (!gene_col %in% colnames(mcols(dmrs_cment))) {
        mcols(dmrs_cment)[[gene_col]] <- NA_character_
    }
}
for (delta_col in c("delta_beta_promoter", "delta_beta_gene_body")) {
    if (!delta_col %in% colnames(mcols(dmrs_cment))) {
        mcols(dmrs_cment)[[delta_col]] <- NA_real_
    }
}
if (!"qval" %in% colnames(mcols(dmrs_cment))) {
    mcols(dmrs_cment)$qval <- NA_real_
}

dmr_delta_col <- if ("sites_delta_beta" %in% colnames(mcols(dmrs_cment))) "sites_delta_beta" else "delta_beta"
dmr_delta <- as.numeric(mcols(dmrs_cment)[[dmr_delta_col]])
dmr_summary <- data.frame(
    Contrast = paste(aggressive_label, "minus", benign_label),
    DMRs = length(dmrs_cment),
    Median_width_bp = median(width(dmrs_cment)),
    Mean_width_bp = round(mean(width(dmrs_cment)), 1),
    Total_coverage_bp = sum(width(dmrs_cment)),
    Median_delta_beta = round(median(dmr_delta, na.rm = TRUE), 4),
    Mean_abs_delta_beta = round(mean(abs(dmr_delta), na.rm = TRUE), 4),
    Hypermethylated_DMRs = sum(dmr_delta > 0, na.rm = TRUE),
    Hypomethylated_DMRs = sum(dmr_delta < 0, na.rm = TRUE)
)

knitr::kable(dmr_summary, caption = "CMEnt DMR summary for aggressive versus benign brain tumors")
```

```{r cment_effect_plot, fig.width=8, fig.height=4}
if (length(dmrs_cment) > 0L) {
    dmr_effect_df <- data.frame(
        Width = width(dmrs_cment),
        Delta_beta = dmr_delta,
        Score = if ("score" %in% colnames(mcols(dmrs_cment))) as.numeric(mcols(dmrs_cment)$score) else NA_real_
    ) |>
        filter(is.finite(Delta_beta), is.finite(Width), Width > 0) |>
        mutate(Direction = factor(ifelse(Delta_beta > 0, "Hyper in aggressive", "Hypo in aggressive")))

    ggplot(dmr_effect_df, aes(x = Delta_beta, y = Width, color = Direction, size = Score)) +
        geom_vline(xintercept = 0, linewidth = 0.2, color = "grey70") +
        geom_point(alpha = 0.65) +
        scale_y_log10(labels = comma) +
        scale_size_continuous(range = c(0.8, 3)) +
        scale_color_manual(values = c("Hyper in aggressive" = "#B8405E", "Hypo in aggressive" = "#247BA0")) +
        theme_minimal() +
        theme(legend.position = "bottom") +
        labs(
            title = "DMR Effect Size and Width",
            x = paste0("Delta beta (", aggressive_label, " - ", benign_label, ")"),
            y = "DMR width (bp, log scale)",
            color = "",
            size = "Score"
        )
}
```

## Gene Context

```{r gene_context}
gene_context_df <- data.frame(
    Context = c("Promoter", "Gene body", "Promoter and gene body", "No annotated gene overlap"),
    DMRs = c(
        sum(has_text(mcols(dmrs_cment)$in_promoter_of)),
        sum(has_text(mcols(dmrs_cment)$in_gene_body_of)),
        sum(has_text(mcols(dmrs_cment)$in_promoter_of) & has_text(mcols(dmrs_cment)$in_gene_body_of)),
        sum(!has_text(mcols(dmrs_cment)$in_promoter_of) & !has_text(mcols(dmrs_cment)$in_gene_body_of))
    )
) |>
    mutate(Percentage = round(100 * DMRs / max(1L, length(dmrs_cment)), 1))

knitr::kable(gene_context_df, caption = "Gene context of CMEnt DMRs")
```

```{r gene_context_plot, fig.width=6, fig.height=3.5}
ggplot(gene_context_df, aes(x = reorder(Context, DMRs), y = DMRs, fill = Context)) +
    geom_col(width = 0.7) +
    coord_flip() +
    scale_y_continuous(labels = comma) +
    theme_minimal() +
    theme(legend.position = "none") +
    labs(title = "DMR Gene Context", x = "", y = "DMRs")
```

## Candidate CMEnt Gene-Overlapping Loci

```{r candidate_dmrs}
dmr_table <- as.data.frame(dmrs_cment)
dmr_table$dmr_index <- seq_along(dmrs_cment)
dmr_table$DMR <- paste0(dmr_table$seqnames, ":", dmr_table$start, "-", dmr_table$end)
dmr_table$delta_beta_for_ranking <- dmr_delta
dmr_table$abs_delta_beta <- abs(dmr_delta)
dmr_table$score_for_ranking <- if ("score" %in% colnames(dmr_table)) dmr_table$score else 0
if (!"score" %in% colnames(dmr_table)) dmr_table$score <- NA_real_
if (!"pval" %in% colnames(dmr_table)) dmr_table$pval <- NA_real_
if (!"qval" %in% colnames(dmr_table)) dmr_table$qval <- NA_real_
gene_cols <- c("in_promoter_of", "in_gene_body_of")
dmr_table$genes <- vapply(seq_len(nrow(dmr_table)), function(i) {
    paste(split_ids(unlist(dmr_table[i, gene_cols], use.names = FALSE)), collapse = ",")
}, character(1))

top_dmr_index <- dmr_table |>
    arrange(desc(score_for_ranking), qval, desc(abs_delta_beta)) |>
    slice_head(n = 1) |>
    pull(dmr_index)

top_dmrs <- dmr_table |>
    filter(has_text(in_promoter_of) | has_text(in_gene_body_of)) |>
    arrange(desc(score_for_ranking), qval, desc(abs_delta_beta)) |>
    dplyr::select(
        dmr_index,
        DMR,
        width,
        score,
        qval,
        pval,
        delta_beta_for_ranking,
        any_of(c("delta_beta_promoter", "delta_beta_gene_body", "seeds_num", "sites_num", "in_promoter_of", "in_gene_body_of"))
    ) |>
    head(20)

knitr::kable(top_dmrs, digits = 4, caption = "Top 20 promoter or gene-body overlapping CMEnt DMRs ranked by score")
```

```{r top_dmr_genes_plot, fig.width=8, fig.height=4}
gene_dmr_df <- if (nrow(dmr_table) > 0L) {
    do.call(rbind, lapply(seq_len(nrow(dmr_table)), function(i) {
        genes <- split_ids(dmr_table$genes[i])
        if (length(genes) == 0L) return(NULL)
        data.frame(
            Gene = genes,
            Delta_beta = dmr_table$delta_beta_for_ranking[i],
            Score = dmr_table$score_for_ranking[i],
            stringsAsFactors = FALSE
        )
    }))
} else {
    NULL
}

if (!is.null(gene_dmr_df) && nrow(gene_dmr_df) > 0L) {
    top_gene_df <- gene_dmr_df |>
        group_by(Gene) |>
        summarise(
            DMRs = n(),
            Median_delta_beta = median(Delta_beta, na.rm = TRUE),
            Max_score = max(Score, na.rm = TRUE),
            .groups = "drop"
        ) |>
        arrange(desc(DMRs), desc(Max_score)) |>
        slice_head(n = 20) |>
        mutate(Direction = ifelse(Median_delta_beta > 0, "Hyper in aggressive", "Hypo in aggressive"))

    ggplot(top_gene_df, aes(x = reorder(Gene, DMRs), y = DMRs, fill = Direction)) +
        geom_col(width = 0.7) +
        coord_flip() +
        scale_y_continuous(labels = comma) +
        scale_fill_manual(values = c("Hyper in aggressive" = "#B8405E", "Hypo in aggressive" = "#247BA0")) +
        theme_minimal() +
        theme(legend.position = "bottom") +
        labs(title = "Genes Most Frequently Overlapped by CMEnt DMRs", x = "", y = "DMRs", fill = "")
}
```

```{r cancer_gene_candidates}
glioma_genes <- c(
    "EGFR", "TERT", "CDKN2A", "CDKN2B", "PTEN", "TP53", "PDGFRA", "NF1",
    "RB1", "MGMT", "IDH1", "IDH2", "ATRX", "OLIG2", "SOX2", "SOX10",
    "NES", "GFAP", "VEGFA", "MMP2", "MMP9", "CD44", "PROM1", "EZH2",
    "BMI1", "CDK4", "MDM2", "MET", "HIF1A"
)

cancer_dmrs <- dmr_table |>
    filter(vapply(strsplit(genes, ",", fixed = TRUE), function(x) any(x %in% glioma_genes), logical(1))) |>
    arrange(desc(abs_delta_beta), qval, desc(score_for_ranking)) |>
    dplyr::select(
        DMR,
        genes,
        width,
        score,
        qval,
        delta_beta_for_ranking,
        any_of(c("delta_beta_promoter", "delta_beta_gene_body", "seeds_num", "sites_num"))
    ) |>
    head(20)

if (nrow(cancer_dmrs) > 0L) {
    knitr::kable(cancer_dmrs, digits = 4, caption = "DMRs overlapping glioma- and cancer-relevant genes")
} else {
    cat("No DMRs overlapped the focused glioma/cancer gene set.\n")
}
```

```{r cancer_gene_candidates_plot, fig.width=8, fig.height=4}
if (nrow(cancer_dmrs) > 0L) {
    plot_cancer_dmrs <- cancer_dmrs |>
        mutate(
            Label = short_label(paste0(genes, " | ", DMR), 70),
            Direction = ifelse(delta_beta_for_ranking > 0, "Hyper in aggressive", "Hypo in aggressive")
        )

    ggplot(plot_cancer_dmrs, aes(x = reorder(Label, abs(delta_beta_for_ranking)), y = delta_beta_for_ranking, fill = Direction)) +
        geom_col(width = 0.7) +
        coord_flip() +
        scale_fill_manual(values = c("Hyper in aggressive" = "#B8405E", "Hypo in aggressive" = "#247BA0")) +
        theme_minimal() +
        theme(legend.position = "bottom", axis.text.y = element_text(size = 7)) +
        labs(
            title = "Cancer-Relevant DMR Effect Sizes",
            x = "",
            y = paste0("Delta beta (", aggressive_label, " - ", benign_label, ")"),
            fill = ""
        )
}
```

## DMR-Level Methylation Separation

```{r top_dmr_heatmap, fig.width=9, fig.height=6}
beta_handler <- CMEnt::getBetaHandler(beta = bsseq_obj, genome = genome)
heatmap_dmrs <- dmr_table |>
    arrange(desc(score_for_ranking), qval, desc(abs_delta_beta)) |>
    slice_head(n = 15) |>
    dplyr::select(dmr_index, DMR, genes)

if (nrow(heatmap_dmrs) > 0L) {
    region_beta <- lapply(seq_len(nrow(heatmap_dmrs)), function(i) {
        site_ids <- split_ids(mcols(dmrs_cment)$sites[heatmap_dmrs$dmr_index[i]])
        beta <- beta_handler$getBeta(row_names = site_ids, col_names = rownames(pheno), allow_missing = TRUE)
        colMeans(beta, na.rm = TRUE)
    })
    region_beta <- do.call(rbind, region_beta)
    rownames(region_beta) <- short_label(ifelse(
        has_text(heatmap_dmrs$genes),
        paste0(heatmap_dmrs$genes, " | ", heatmap_dmrs$DMR),
        heatmap_dmrs$DMR
    ), 75)
    sample_order <- order(pheno$tumor_type)
    region_beta <- region_beta[, rownames(pheno)[sample_order], drop = FALSE]
    heatmap_df <- as.data.frame(as.table(region_beta), stringsAsFactors = FALSE)
    colnames(heatmap_df) <- c("DMR", "Sample", "Beta")
    heatmap_df$Tumor_type <- pheno[heatmap_df$Sample, "tumor_type"]
    heatmap_df$Sample <- factor(heatmap_df$Sample, levels = colnames(region_beta))
    heatmap_df$DMR <- factor(heatmap_df$DMR, levels = rev(rownames(region_beta)))

    ggplot(heatmap_df, aes(x = Sample, y = DMR, fill = Beta)) +
        geom_tile() +
        facet_grid(. ~ Tumor_type, scales = "free_x", space = "free_x") +
        scale_fill_gradientn(colors = c("#247BA0", "white", "#B8405E"), limits = c(0, 1), na.value = "grey90") +
        theme_minimal() +
        theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), axis.text.y = element_text(size = 7)) +
        labs(title = "Top CMEnt DMR Methylation Across Samples", x = "Samples", y = "", fill = "Beta")
}
```

## Top CMEnt DMR Visualization

```{r top_dmr, fig.width=6, fig.height=6}
if (length(dmrs_cment) > 0L) {
    CMEnt::plotDMR(
        dmrs_cment,
        dmr_index = top_dmr_index,
        beta = beta_handler,
        pheno = pheno,
        sample_group_col = sample_group_col,
        genome = genome,
        plot_motif = FALSE
    )
}
```

## CMEnt DMR Manhattan Plot

```{r manhattan, fig.width=10, fig.height=4}
if (length(dmrs_cment) > 0L && "score" %in% colnames(mcols(dmrs_cment))) {
    CMEnt::plotDMRsManhattan(
        dmrs_cment,
        genome = genome
    )
}
```

## CMEnt DMR Interaction Circos Visualization

```{r circos, fig.width=10, fig.height=10}
if (length(dmrs_cment) > 0L) {
    CMEnt::plotDMRsCircos(
        dmrs_cment,
        genome = genome
    )
}
```

## Insights and Interpretation

This ONT analysis treats positive delta beta values as hypermethylation in Glioblastoma,
IDH-wildtype relative to Pilocytic astrocytoma. The most useful readout is therefore not
only whether CMEnt finds many DMRs, but whether the highest-scoring regions show coherent
sample-level methylation patterns that separate the aggressive and benign tumor groups.

The gene-context and candidate-locus plots help prioritize biological follow-up. Promoter
DMRs are natural candidates for transcriptional regulation, while gene-body DMRs can mark
broader regulatory or cell-state-associated methylation changes. The focused candidate table
does not prove causality; it flags DMRs overlapping genes with known relevance to glioma,
stemness, invasion, hypoxia, growth-factor signaling, and tumor suppression, which are
plausible axes separating pilocytic astrocytoma from IDH-wildtype glioblastoma.

The top-DMR heatmap is the sanity check for classification biology: strong candidate regions
should show consistent methylation within tumor type and clear differences between the benign
and aggressive groups. Regions that score highly but do not separate samples visually should
be treated cautiously, especially given the low-depth ONT coverage and the small benign cohort.

## Session Info

```{r}
sessionInfo()
```
