---
title: "CMEnt Analysis of ONT Methylation Data in Brain Tumors"
subtitle: "GSE289246: pilocytic astrocytoma (PA) versus IDH-wildtype glioblastoma (GBM)"
author: "CMEnt Package"
date: "`r Sys.Date()`"
output: 
    BiocStyle::html_document:
        code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    results = "asis",
    cache = FALSE,
    echo = TRUE,
    dev = "png",
    warning = FALSE,
    message = FALSE,
    fig.width = 8,
    fig.height = 6,
    dpi = 300
)

try(devtools::load_all(), silent = TRUE)
suppressPackageStartupMessages({
    try(library(CMEnt), silent = TRUE)
    library(curl)
    library(GenomicRanges)
    library(GEOquery)
    library(SummarizedExperiment)
    library(data.table)
    library(bsseq)
    library(HDF5Array)
    library(ggplot2)
    library(dplyr)
    library(scales)
})


getBenchmarkFile <- function(filename) {
    benchmark_file <- ""
    folder <- "ont.old"
    if (getwd() == "notebooks") {
        benchmark_file <- file.path("cached_results", folder, filename)
    } else if ("NAMESPACE" %in% dir()) {
        benchmark_file <- file.path("notebooks", "cached_results", folder, filename)
    } else {
        benchmark_file <- file.path("cached_results", folder, filename)
    }
    dir.create(dirname(benchmark_file), showWarnings = FALSE, recursive = TRUE)
    benchmark_file
}

split_ids <- function(x) {
    ids <- unlist(strsplit(stats::na.omit(as.character(x)), ","), use.names = FALSE)
    unique(ids[nzchar(ids)])
}

has_text <- function(x) {
    !is.na(x) & nzchar(as.character(x))
}

short_label <- function(x, width = 65L) {
    ifelse(nchar(x) > width, paste0(substr(x, 1L, width - 3L), "..."), x)
}
```

## Introduction

This report uses CMEnt to summarize long-read ONT methylation differences in GSE289246.
The contrast is intentionally simple: Pilocytic astrocytoma (PA) is used as the lower-grade
reference group and Glioblastoma, IDH-wildtype (GBM) as the aggressive high-grade group. Rather
than treating single CpGs as the final readout, the analysis asks whether significant
methylation seeds can be assembled into coherent DMRs with interpretable genomic context.

- CMEnt DMR size, score, and methylation effect summaries
- promoter and gene-body overlap patterns
- DMR-level methylation heatmaps separating benign and aggressive samples
- cancer-focused candidate gene loci and genome-wide visualization

## Setup and Dependencies

```{r install, eval=FALSE}
r <- getOption("repos")
if (is.null(r) || is.na(r["CRAN"]) || r["CRAN"] %in% c("@CRAN@", "")) {
    options(repos = c(CRAN = "https://cloud.r-project.org"))
}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}
for (pkg in c("DSS", "TxDb.Hsapiens.UCSC.hg19.knownGene", "org.Hs.eg.db", "BSgenome.Hsapiens.UCSC.hg19")) {
    if (!requireNamespace(pkg, quietly = TRUE)) {
        BiocManager::install(pkg, update = FALSE, ask = FALSE)
    }
}
```
## Load Data

The GEO SOFT file is used to reconstruct the sample sheet, then restricted to tumor
samples with ONT methylation calls generated on PromethION. The selected comparison keeps
PA and GBM samples so that all downstream summaries
refer to one clinically meaningful low-grade versus high-grade brain tumor contrast.
```{r samplesheet_creation}

sample_group_col <- "tumor_type"
casecontrol_col <- "aggressive"
genome <- "hg19" # The authors of the original study used hg19.
benign_label <- "PA"
aggressive_label <- "GBM"

URL <- "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE289nnn/GSE289246/soft/GSE289246_family.soft.gz"
samplesheet_file <- getBenchmarkFile("samplesheet.tsv")
if (!file.exists(samplesheet_file)) {
    message("Downloading and parsing the GEO dataset samplesheet: ", URL)
    soft_file <- getBenchmarkFile("GSE289246_family.soft.gz")
    if (!file.exists(soft_file)) {
        curl::curl_download(URL, soft_file)
    }
    soft_data <- getGEO(filename = soft_file)
    gsms <- attr(soft_data, "gsms")
    samplesheet <- list()
    colnames <- unique(unlist(sapply(names(gsms), function(x) names(attr(gsms[[x]], "header")))))
    for (gsm in names(gsms)) {
        samplesheet[[gsm]] <- sapply(
            attr(gsms[[gsm]], "header"), function(x) {
                if (length(x) > 1) {
                    x <- paste(x, collapse = "; ")
                    x <- gsub("\\s+", " ", x)
                }
                x
            }
        )
        missing_cols <- setdiff(colnames, names(samplesheet[[gsm]]))
        if (length(missing_cols) > 0) {
            samplesheet[[gsm]][missing_cols] <- NA
            samplesheet[[gsm]] <- samplesheet[[gsm]][colnames]
        }
    }
    samplesheet <- do.call(rbind, samplesheet)
    pheno <- samplesheet[samplesheet[, "library_strategy"] == "OTHER", ]
    instrument_models <- table(pheno[, "instrument_model"])
    pheno <- pheno[pheno[, "instrument_model"] == "PromethION", ]
    case1 <- "tissue: tumor; tumor type: Pilocytic astrocytoma"
    case2 <- "tissue: tumor; tumor type: Glioblastoma, IDH-wildtype"
    pheno <- pheno[pheno[, "characteristics_ch1"] %in% c(case1, case2), ]
    unlink(soft_file)
    write.table(pheno, samplesheet_file, row.names = TRUE, col.names = NA, sep = "\t", quote = FALSE)
} else {
    pheno <- read.table(samplesheet_file, header = TRUE, row.names = 1, sep = "\t", quote = "", stringsAsFactors = FALSE, check.names = FALSE)
}
pheno <- as.data.frame(pheno)
pheno$tumor_type <- ifelse(
    grepl("Glioblastoma, IDH-wildtype", pheno$characteristics_ch1, fixed = TRUE),
    aggressive_label,
    benign_label
)
pheno$tumor_type <- factor(pheno$tumor_type, levels = c(benign_label, aggressive_label))
pheno$aggressive <- pheno$tumor_type == aggressive_label
pheno$Sample_ID <- rownames(pheno)

knitr::kable(
    as.data.frame(table(Tumor_type = pheno$tumor_type)),
    caption = "Selected GSE289246 ONT samples"
)
```
Because these ONT methylation calls come from low-depth sequencing, CpG-level evidence is
uneven across samples. A minimum coverage of 2 per sample keeps the filter permissive enough for
long-read data while removing single-read methylation calls that are too fragile for DMR
construction. We also apply a position filter based on the intra-group coverage, of at least 2 samples per group.
```{r bsseq_creation}
min_sample_coverage <- 2
min_group_coverage <- 2
group_filter_mode <- "samples_with_min_coverage"

input_bsseq_h5 <- getBenchmarkFile("input_h5")
filtering_summary_rds <- getBenchmarkFile("filtering_summary.rds")
filtering_summary_exists <- file.exists(filtering_summary_rds)
sample_filtering_matches <- FALSE
if (filtering_summary_exists) {
    filtering_summary <- readRDS(filtering_summary_rds)
    sample_filtering_matches <- "min_sample_coverage" %in% names(filtering_summary) &&
        filtering_summary$min_sample_coverage == min_sample_coverage
    if (
        !all(c("min_sample_coverage", "min_group_coverage") %in% names(filtering_summary)) ||
            filtering_summary$min_sample_coverage != min_sample_coverage ||
            filtering_summary$min_group_coverage != min_group_coverage ||
            !identical(filtering_summary$group_filter_mode, group_filter_mode)
    ) {
        warning("Filtering summary exists but does not match the current coverage thresholds. Recalculating BSseq object.")
        filtering_summary_exists <- FALSE
    }
}
if (file.exists(input_bsseq_h5) && filtering_summary_exists) {
    message("Loading BSseq object from ", input_bsseq_h5)
    bsseq_obj <- HDF5Array::loadHDF5SummarizedExperiment(input_bsseq_h5)
} else {
    filtering_summary <- list()
    filtering_summary$min_sample_coverage <- min_sample_coverage
    filtering_summary$min_group_coverage <- min_group_coverage
    filtering_summary$group_filter_mode <- group_filter_mode
    source_bed_files <- pheno[, "supplementary_file_1"]
    names(source_bed_files) <- rownames(pheno)
    target_bed_files <- sapply(names(source_bed_files), function(x) {
        getBenchmarkFile(file.path("downloaded", paste0(x, ".bed.gz")))
    })
    message("Downloading ", length(target_bed_files), " bed files to ", dirname(target_bed_files[1]), "...")
    pb <- txtProgressBar(min = 0, max = length(target_bed_files), style = 3)
    # download the bed files to the raw_data directory
    for (sample_name in names(source_bed_files)) {
        if (!file.exists(target_bed_files[[sample_name]])) {
            message("Downloading ", source_bed_files[[sample_name]])
            max_attempts <- 5
            attempt <- 1
            while (TRUE) {
                tryCatch({
                    curl::curl_download(source_bed_files[[sample_name]], target_bed_files[[sample_name]])
                    # get checksum of the downloaded file
                    downloaded_checksum <- tools::md5sum(target_bed_files[[sample_name]])

                    # verify that the file is a valid gzip file
                    # try to unzip the file and check for errors
                    unzip_success <- tryCatch({
                        R.utils::gunzip(target_bed_files[[sample_name]], remove = FALSE, overwrite = TRUE)
                        TRUE
                    }, error = function(e) {
                        FALSE
                    })
                    if (unzip_success) {
                        message("Downloaded ", source_bed_files[[sample_name]], " successfully.")
                        break
                    } else {
                        attempt <<- attempt + 1
                        if (attempt > max_attempts) {
                            stop("Failed to download ", source_bed_files[[sample_name]], " after ", max_attempts, " attempts.")
                        }
                        message("Downloaded file ", target_bed_files[[sample_name]], " is not a valid gzip file. Retrying...")
                    }
                }, error = function(e) {
                    attempt <<- attempt + 1
                    if (attempt > max_attempts) {
                        stop("Failed to download ", source_bed_files[[sample_name]])
                    }
                    message("Error downloading ", source_bed_files[[sample_name]], ": ", e$message)
                    message("Retrying in 5 seconds...")
                    Sys.sleep(5)
                })
            }
        }
        setTxtProgressBar(pb, which(names(target_bed_files) == sample_name))
    }

    # Extract the methylation data from the bed files and create a BSseq object
    bsseq_objects <- list()
    message("Creating BSseq objects and saving to HDF5...")
    progress_bar <- txtProgressBar(min = 0, max = length(target_bed_files), style = 3)
    target_h5_dirs <- sapply(names(target_bed_files), function(x) {
        getBenchmarkFile(file.path("h5", paste0(x)))
    })
    sample_ratio <- list()
    for (sample_name in names(target_bed_files)) {
        target_h5_folder <- target_h5_dirs[[sample_name]]
        if (file.exists(file.path(target_h5_folder, "assays.h5")) && sample_filtering_matches) {
            setTxtProgressBar(progress_bar, which(names(target_bed_files) == sample_name))
            next
        }
        if (dir.exists(target_h5_folder)) {
            unlink(target_h5_folder, recursive = TRUE)
        }
        bed_file <- target_bed_files[[sample_name]]
        df <- data.table::fread(bed_file, header = TRUE, sep = "\t", stringsAsFactors = FALSE)
        if (!exists("gr_list")) {
            gr_list <- list()
        }
        gr <- GRanges(
            seqnames = df$chrom,
            ranges = IRanges(start = df$start + 1L, end = df$end),
            strand = "*"
        )
        M <- as.matrix(as.integer(df$percentMeth * df$coverage / 100), ncol = 1)
        Cov <- as.matrix(as.integer(df$coverage), ncol = 1)


        bsseq_obj <- BSseq(
            gr = gr,
            M = M,
            Cov = Cov,
            sampleNames = sample_name
        )
        to_keep <- Cov >= min_sample_coverage
        sample_ratio[[sample_name]] <- mean(to_keep)
        if (sum(to_keep) == 0) {
            warning("No sites with coverage >= ", min_sample_coverage, " for sample ", sample_name)
            next
        }
        if (sum(to_keep) < nrow(bsseq_obj)) {
            nfiltered <- nrow(bsseq_obj) - sum(to_keep)
            message(
                "Keeping ", sum(to_keep), " sites out of ", nrow(bsseq_obj), " (",
                round(sum(to_keep) / nrow(bsseq_obj) * 100, 2),
                "%) with coverage >= ", min_sample_coverage, " for sample ", sample_name
            )
        }
        bsseq_obj <- bsseq_obj[to_keep, ]
        HDF5Array::saveHDF5SummarizedExperiment(bsseq_obj, dir = target_h5_folder)
        setTxtProgressBar(progress_bar, which(names(target_bed_files) == sample_name))
    }
    filtering_summary$sample_filter_ratio <- sample_ratio
    message("Loading BSseq objects from HDF5 and combining into a single BSseq object...")
    bsseq_objects <- list()
    for (sample_name in names(target_bed_files)) {
        target_h5_folder <- target_h5_dirs[[sample_name]]
        bsseq_obj <- HDF5Array::loadHDF5SummarizedExperiment(target_h5_folder)
        bsseq_objects[[sample_name]] <- bsseq_obj
    }
    suppressWarnings({
        bsseq_obj <- do.call(bsseq::combineList, list(bsseq_objects, BACKEND = "HDF5Array"))
    })
    message("Filtering BSseq object to keep only sites with information in at least ", min_group_coverage, " sample(s) per group...")
    group_filter <- rep(TRUE, nrow(bsseq_obj))
    coverage <- SummarizedExperiment::assay(bsseq_obj, "Cov")

    for (group in unique(pheno[[sample_group_col]])) {
        group_samples <- rownames(pheno)[pheno[[sample_group_col]] == group]
        group_pass <- rowSums(coverage[, group_samples, drop = FALSE] > 0, na.rm = TRUE) >= min_group_coverage
        group_filter <- group_filter & group_pass
    }
    filtering_summary$group_filter_ratio <- mean(group_filter)
    bsseq_obj <- bsseq_obj[group_filter, ]
    message("Saving filtered BSseq object to HDF5...")
    if (dir.exists(input_bsseq_h5)) {
        unlink(input_bsseq_h5, recursive = TRUE)
    }
    bsseq_obj <- HDF5Array::saveHDF5SummarizedExperiment(bsseq_obj, dir = paste0(input_bsseq_h5))
    saveRDS(filtering_summary, filtering_summary_rds)
}


```
## Find Differentially Methylated Positions

CMEnt starts from DMPs as regional anchors. Here, seeds are CpG sites with nominal evidence
for differential methylation between the two tumor groups; the distance distribution below
shows whether those anchors are dense enough to support regional expansion.
CMEnt reduces the BSseq object to windows around the seed sites before building regions.
This keeps memory use practical for ONT-scale data while preserving nearby CpGs that can
support or reject regional expansion around each seed.
We are using a rather lenient selection threshold, of **non-adjusted** p-value being less than 0.05.
```{r seeds_calculation}
pval_thres <- 0.05
seeds_file <- getBenchmarkFile("seeds.rds")
if (!file.exists(seeds_file)) {
    dmps_file <- getBenchmarkFile("dmps.rds")
    if (!file.exists(dmps_file)) {
        message("Calculating DMPs...")
        dmps <- CMEnt::findDMPsBSSeq(
            bsseq_obj,
            samplesheet = pheno,
            sample_group_col = sample_group_col,
            id_col = "Sample_ID",
            case_group = aggressive_label,
            njobs = max(1L, min(8L, parallel::detectCores(logical = TRUE) - 1L))
        )
        saveRDS(dmps, dmps_file)
    } else {
        message("Loading DMPs from ", dmps_file)
        dmps <- readRDS(dmps_file)
    }
    seeds <- dmps[dmps$pval < pval_thres, ]
    saveRDS(seeds, seeds_file)
} else {
    seeds <- readRDS(seeds_file)
}
rownames(seeds) <- seeds$site_id
seeds$Row.names <- NULL
seed_dist <- unlist(lapply(split(seeds$start, seeds$chr), function(x) {
    if (length(x) < 2) return(numeric(0))
    diff(x)
}))
seeds_qs <- round(stats::quantile(seed_dist, probs = c(0, 0.25, 0.5, 0.75, 0.9, 0.99, 1)))
cat(sprintf("Number of seeds for CMEnt: %d\n", nrow(seeds)))
cat("BSseq object dimensions: ", dim(bsseq_obj), "\n")
reduced_bsseq_obj <- CMEnt::reduceBSseqAroundSeeds(bsseq_obj, seeds, expansion_window = 10000)
beta_handler <- CMEnt::getBetaHandler(beta = reduced_bsseq_obj, genome = genome)
cat("Reduced BSseq object dimensions: ", dim(reduced_bsseq_obj), "\n")
knitr::kable(
    seeds_qs,
    digits = 2,
    col.names = "Distance (bp)",
    caption = sprintf("Seed distance quantiles")
)
```

This leaves `r format(nrow(seeds), big.mark = ",")` seed CpGs across an input BSseq matrix
with `r format(nrow(bsseq_obj), big.mark = ",")` loci and `r ncol(bsseq_obj)` samples.
The median within-chromosome seed spacing is `r format(seeds_qs["50%"], big.mark = ",")`
bp, with a long tail of isolated seeds; this is why the DMR builder later requires multiple
seeds and enough covered sites before accepting a region.

## Method 1: CMEnt
```{r cment}
aggfun <- "mean"
cment_file <- getBenchmarkFile(paste0("dmrs.", aggfun, ".rds"))
if (!file.exists(cment_file)) {
    # Reduce the BSseq object around the seeds with the specified expansion window,
    # so that to minimize memory usage and speed up the DMR building process.
    cat("Reduced BSseq object dimensions: ", dim(reduced_bsseq_obj), "\n")
    start_time <- Sys.time()
    dmrs_cment <- CMEnt::buildDMRs(
        beta_handler,
        seeds,
        pheno,
        genome = genome,
        sample_group_col = sample_group_col,
        casecontrol_col = casecontrol_col,
        entanglement  = "weak",
        testing_mode  = "parametric",
        aggfun = aggfun, # Switching from median, to be more sensitive for larger effect size
        expansion_window = 10000, # The maximum distance (in bp) to expand the DMRs from the seeds
        min_seeds = 2, # The minimum number of seeds required to form a DMR
        min_sites = 50, # The minimum number of sites required to form a DMR
        max_pval = 0.05,
        ext_site_delta_beta = 0.1,
        max_lookup_dist = 1000,
        njobs = max(1L, min(8L, parallel::detectCores(logical = TRUE) - 1L)),
        verbose = 2
    )
    end_time <- Sys.time()
    time_cment <- end_time - start_time
    saveRDS(list(dmrs = dmrs_cment, time = time_cment), cment_file)
} else {
    message("Loading CMEnt DMRs from ", cment_file)
    cment_results <- readRDS(cment_file)
    dmrs_cment <- cment_results$dmrs
    time_cment <- cment_results$time
}
cat(sprintf("CMEnt: Found %d DMRs in %d seconds\n", length(dmrs_cment), round(as.numeric(time_cment, units = "secs"))))
```

The resulting call set contains `r format(length(dmrs_cment), big.mark = ",")` DMRs. In
the following summaries, positive delta beta means higher methylation in GBM
relative to PA.

## CMEnt Region Summary

```{r cment_summary}
if (!"qval" %in% colnames(mcols(dmrs_cment))) {
    mcols(dmrs_cment)$qval <- NA_real_
}

dmr_delta_col <- "sites_delta_beta"
dmr_delta <- as.numeric(mcols(dmrs_cment)[[dmr_delta_col]])
dmr_summary <- data.frame(
    Contrast = paste(aggressive_label, "minus", benign_label),
    DMRs = length(dmrs_cment),
    Median_width_bp = median(width(dmrs_cment)),
    Mean_width_bp = round(mean(width(dmrs_cment)), 1),
    Total_coverage_bp = sum(width(dmrs_cment)),
    Median_delta_beta = round(median(dmr_delta, na.rm = TRUE), 4),
    Mean_abs_delta_beta = round(mean(abs(dmr_delta), na.rm = TRUE), 4),
    Hypermethylated_DMRs = sum(dmr_delta > 0, na.rm = TRUE),
    Hypomethylated_DMRs = sum(dmr_delta < 0, na.rm = TRUE)
)

knitr::kable(dmr_summary, caption = "CMEnt DMR summary for aggressive versus benign brain tumors")
```

The called regions are mostly kilobase-scale: the median DMR spans
`r format(round(dmr_summary$Median_width_bp), big.mark = ",")` bp and together the calls
cover `r format(dmr_summary$Total_coverage_bp, big.mark = ",")` bp. The mean absolute
delta beta of `r dmr_summary$Mean_abs_delta_beta` gives a compact effect-size summary, but
individual loci should still be inspected because sparse ONT coverage can flatten aggregate
regional means.

```{r cment_effect_plot, fig.width=8, fig.height=4, fig.cap="DMR width versus methylation effect. The discrete effect-size bands are expected with sparse ONT beta estimates and should be interpreted together with coverage and sample-level plots."}
if (length(dmrs_cment) > 0L) {
    dmr_effect_df <- data.frame(
        Width = width(dmrs_cment),
        Delta_beta = dmr_delta,
        Score = if ("score" %in% colnames(mcols(dmrs_cment))) as.numeric(mcols(dmrs_cment)$score) else NA_real_
    ) |>
        filter(is.finite(Delta_beta), is.finite(Width), Width > 0) |>
        mutate(Direction = factor(ifelse(Delta_beta > 0, "Hyper in aggressive", "Hypo in aggressive")))

    ggplot(dmr_effect_df, aes(x = Delta_beta, y = Width, color = Direction, size = Score)) +
        geom_vline(xintercept = 0, linewidth = 0.2, color = "grey70") +
        geom_point(alpha = 0.65) +
        scale_y_log10(labels = comma) +
        scale_size_continuous(range = c(0.8, 3)) +
        scale_color_manual(values = c("Hyper in aggressive" = "#B8405E", "Hypo in aggressive" = "#247BA0")) +
        theme_minimal() +
        theme(legend.position = "bottom") +
        labs(
            title = "DMR Effect Size and Width",
            x = paste0("Delta beta (", aggressive_label, " - ", benign_label, ")"),
            y = "DMR width (bp, log scale)",
            color = "",
            size = "Score"
        )
}
```

## Gene Context

Gene annotation is used as a prioritization layer, not as proof of mechanism. Promoter
overlaps are useful for hypotheses about transcriptional regulation, while gene-body
overlaps can reflect broader cell-state or regulatory differences.

```{r gene_context}
gene_context_df <- data.frame(
    Context = c("Promoter", "Gene body", "Promoter and gene body", "No annotated gene overlap"),
    DMRs = c(
        sum(has_text(mcols(dmrs_cment)$in_promoter_of)),
        sum(has_text(mcols(dmrs_cment)$in_gene_body_of)),
        sum(has_text(mcols(dmrs_cment)$in_promoter_of) & has_text(mcols(dmrs_cment)$in_gene_body_of)),
        sum(!has_text(mcols(dmrs_cment)$in_promoter_of) & !has_text(mcols(dmrs_cment)$in_gene_body_of))
    )
) |>
    mutate(Percentage = round(100 * DMRs / max(1L, length(dmrs_cment)), 1))

knitr::kable(gene_context_df, caption = "Gene context of CMEnt DMRs")
```

In this run, `r gene_context_df$Percentage[gene_context_df$Context == "Gene body"]`% of DMRs
overlap gene bodies and `r gene_context_df$Percentage[gene_context_df$Context == "Promoter"]`%
overlap promoters. The sizeable unannotated fraction is expected for genome-wide methylation
data and should not be discarded automatically.

```{r gene_context_plot, fig.width=6, fig.height=3.5}
ggplot(gene_context_df, aes(x = reorder(Context, DMRs), y = DMRs, fill = Context)) +
    geom_col(width = 0.7) +
    coord_flip() +
    scale_y_continuous(labels = comma) +
    theme_minimal() +
    theme(legend.position = "none") +
    labs(title = "DMR Gene Context", x = "", y = "DMRs")
```

## Candidate CMEnt Gene-Overlapping Loci

The next tables move from global summaries to loci that are easier to review manually.
DMRs are ranked by CMEnt score, adjusted significance, and absolute methylation effect, then
split into broad gene-overlapping candidates and a focused glioma/cancer gene panel.

```{r candidate_dmrs}
dmr_table <- as.data.frame(dmrs_cment)
dmr_table$dmr_index <- seq_along(dmrs_cment)
dmr_table$DMR <- paste0(dmr_table$start, "-", dmr_table$end)
dmr_table$seed_delta_beta <- if ("delta_beta" %in% colnames(dmr_table)) dmr_table$delta_beta else NA_real_
dmr_table$region_delta_beta <- dmr_delta
dmr_table$abs_region_delta_beta <- abs(dmr_table$region_delta_beta)
dmr_table$score_for_ranking <- if ("score" %in% colnames(dmr_table)) dmr_table$score else 0
dmr_table$cv_accuracy <- if ("cv_accuracy" %in% colnames(dmr_table)) dmr_table$cv_accuracy else NA_real_

top_dmr_index <- dmr_table |>
    arrange(desc(score_for_ranking), qval, desc(abs_region_delta_beta)) |>
    slice_head(n = 1) |>
    pull(dmr_index)

top_dmrs <- dmr_table |>
    filter(has_text(in_promoter_of) | has_text(in_gene_body_of)) |>
    arrange(desc(score_for_ranking), qval, desc(abs_region_delta_beta)) |>
    dplyr::select(
        dmr_index,
        DMR,
        width,
        score,
        qval,
        pval,
        seed_delta_beta,
        region_delta_beta,
        any_of(c("delta_beta_promoter", "delta_beta_gene_body", "seeds_num", "sites_num", "in_promoter_of", "in_gene_body_of"))
    ) |>
    head(20)

knitr::kable(top_dmrs, digits = 4, caption = "Top 20 promoter or gene-body overlapping CMEnt DMRs ranked by score")
```

```{r top_dmr_genes_plot, fig.width=4, fig.height=6}
gene_dmr_df <- if (nrow(dmr_table) > 0L) {
    do.call(rbind, lapply(seq_len(nrow(dmr_table)), function(i) {
        in_promoter_of <- split_ids(dmr_table$in_promoter_of[i])
        if (length(in_promoter_of) == 0L) return(NULL)
        data.frame(
            Gene = in_promoter_of,
            Delta_beta = dmr_table$region_delta_beta[i],
            Score = dmr_table$score_for_ranking[i],
            stringsAsFactors = FALSE
        )
    }))
} else {
    NULL
}

if (!is.null(gene_dmr_df) && nrow(gene_dmr_df) > 0L) {
    top_gene_df <- gene_dmr_df |>
        group_by(Gene) |>
        summarise(
            delta_beta = median(Delta_beta, na.rm = TRUE),
            score = median(Score, na.rm = TRUE),
            .groups = "drop"
        ) |>
        arrange(desc(score)) |>
        slice_head(n = 20)

    ggplot(top_gene_df, aes(x = reorder(Gene, score), y = delta_beta, fill = score)) +
        geom_col(width = 0.7) +
        coord_flip() +
        scale_fill_continuous(type = "viridis", option = "C") +
        guides(fill = guide_colorbar(barwidth = 0.5, barheight = 10, title = "Score", position="right")) +
        scale_y_continuous(labels = comma) +
        theme_minimal(base_size=12) +
        theme(legend.position = "bottom") +
        labs(title = "Genes Promoters Overlapped by top-scoring CMEnt DMRs", x = "", y = "Delta Beta", fill = "")
}
```

## Top CMEnt DMR Visualization

```{r top_dmr, fig.width=10, fig.height=8, fig.cap="Structure-level view of the top-ranked DMR, including seed placement and gene context. Sample methylation is summarized in the heatmap above."}
CMEnt::plotDMRs(
    dmrs_cment, top_n = 1, score_by = "score",
    beta = beta_handler, pheno = pheno, genome = genome,
    sample_group_col = sample_group_col,
    max_samples_per_group = Inf
)
```

## CMEnt DMR Manhattan Plot

The Manhattan plot places the full DMR set back onto the genome, making it easier to see
whether high-scoring calls are isolated, clustered, or concentrated in specific chromosomes.
Because the score range is narrow, the most useful signal here is genomic distribution and
annotation class rather than large score outliers.

```{r manhattan, fig.width=10, fig.height=4, fig.cap="Genome-wide DMR score distribution by primary annotation class. Scores cluster tightly, while DMRs are spread across many chromosomes."}
if (length(dmrs_cment) > 0L && "score" %in% colnames(mcols(dmrs_cment))) {
    CMEnt::plotDMRsManhattan(
        dmrs_cment,
        genome = genome
    )
}
```

## CMEnt DMR Interaction Circos Visualization

```{r circos, fig.width=10, fig.height=10, fig.cap="Circos summary of DMR positions, methylation direction, and detected DMR interactions. Grey links denote interactions without matched JASPAR motif support."}
if (length(dmrs_cment) > 0L) {
    CMEnt::plotDMRsCircos(
        dmrs_cment,
        genome = genome,
        beta = beta_handler,
        sample_group_col = sample_group_col
    )
}
```

## Session Info

```{r}
sessionInfo()
```
