---
title: "CMEnt Analysis of ONT Methylation Data in Brain Tumors"
subtitle: "GSE289246: pilocytic astrocytoma versus IDH-wildtype glioblastoma"
author: "CMEnt Package"
date: "`r Sys.Date()`"
output: 
    BiocStyle::html_document:
        code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    results = "asis",
    cache = TRUE,
    echo = TRUE,
    dev = "png",
    warning = FALSE,
    message = FALSE,
    fig.width = 8,
    fig.height = 6,
    dpi = 300
)

try(devtools::load_all(), silent = TRUE)
suppressPackageStartupMessages({
    try(library(CMEnt), silent = TRUE)
    library(curl)
    library(GenomicRanges)
    library(GEOquery)
    library(SummarizedExperiment)
    library(data.table)
    library(bsseq)
    library(HDF5Array)
    library(ggplot2)
    library(dplyr)
    library(scales)
})


getBenchmarkFile <- function(filename) {
    benchmark_file <- ""
    folder <- "ont"
    if (getwd() == "notebooks") {
        benchmark_file <- file.path("cached_results", folder, filename)
    } else if ("NAMESPACE" %in% dir()) {
        benchmark_file <- file.path("notebooks", "cached_results", folder, filename)
    } else {
        benchmark_file <- file.path("cached_results", folder, filename)
    }
    dir.create(dirname(benchmark_file), showWarnings = FALSE, recursive = TRUE)
    benchmark_file
}

split_ids <- function(x) {
    ids <- unlist(strsplit(stats::na.omit(as.character(x)), ","), use.names = FALSE)
    unique(ids[nzchar(ids)])
}

has_text <- function(x) {
    !is.na(x) & nzchar(as.character(x))
}

short_label <- function(x, width = 65L) {
    ifelse(nchar(x) > width, paste0(substr(x, 1L, width - 3L), "..."), x)
}
```

## Introduction

This report uses CMEnt to summarize long-read ONT methylation differences in GSE289246.
The contrast is intentionally simple: Pilocytic astrocytoma is used as the lower-grade
reference group and Glioblastoma, IDH-wildtype as the aggressive high-grade group. Rather
than treating single CpGs as the final readout, the analysis asks whether significant
methylation seeds can be assembled into coherent DMRs with interpretable genomic context.

- CMEnt DMR size, score, and methylation effect summaries
- promoter and gene-body overlap patterns
- DMR-level methylation heatmaps separating benign and aggressive samples
- cancer-focused candidate gene loci and genome-wide visualization

## Setup and Dependencies

```{r install, eval=FALSE}
r <- getOption("repos")
if (is.null(r) || is.na(r["CRAN"]) || r["CRAN"] %in% c("@CRAN@", "")) {
    options(repos = c(CRAN = "https://cloud.r-project.org"))
}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}
for (pkg in c("DSS", "TxDb.Hsapiens.UCSC.hg19.knownGene", "org.Hs.eg.db", "BSgenome.Hsapiens.UCSC.hg19")) {
    if (!requireNamespace(pkg, quietly = TRUE)) {
        BiocManager::install(pkg, update = FALSE, ask = FALSE)
    }
}
```
## Load Data

The GEO SOFT file is used to reconstruct the sample sheet, then restricted to tumor
samples with ONT methylation calls generated on PromethION. The selected comparison keeps
Pilocytic astrocytoma and IDH-wildtype glioblastoma samples so that all downstream summaries
refer to one clinically meaningful low-grade versus high-grade brain tumor contrast.
```{r samplesheet_creation}

sample_group_col <- "tumor_type"
casecontrol_col <- "aggressive"
genome <- "hg19" # The authors of the original study used hg19.
benign_label <- "Pilocytic astrocytoma"
aggressive_label <- "Glioblastoma, IDH-wildtype"

URL <- "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE289nnn/GSE289246/soft/GSE289246_family.soft.gz"
samplesheet_file <- getBenchmarkFile("samplesheet.tsv")
if (!file.exists(samplesheet_file)) {
    message("Downloading and parsing the GEO dataset samplesheet: ", URL)
    soft_file <- getBenchmarkFile("GSE289246_family.soft.gz")
    if (!file.exists(soft_file)) {
        curl::curl_download(URL, soft_file)
    }
    soft_data <- getGEO(filename = soft_file)
    gsms <- attr(soft_data, "gsms")
    samplesheet <- list()
    colnames <- unique(unlist(sapply(names(gsms), function(x) names(attr(gsms[[x]], "header")))))
    for (gsm in names(gsms)) {
        samplesheet[[gsm]] <- sapply(
            attr(gsms[[gsm]], "header"), function(x) {
                if (length(x) > 1) {
                    x <- paste(x, collapse = "; ")
                    x <- gsub("\\s+", " ", x)
                }
                x
            }
        )
        missing_cols <- setdiff(colnames, names(samplesheet[[gsm]]))
        if (length(missing_cols) > 0) {
            samplesheet[[gsm]][missing_cols] <- NA
            samplesheet[[gsm]] <- samplesheet[[gsm]][colnames]
        }
    }
    samplesheet <- do.call(rbind, samplesheet)
    pheno <- samplesheet[samplesheet[, "library_strategy"] == "OTHER", ]
    instrument_models <- table(pheno[, "instrument_model"])
    pheno <- pheno[pheno[, "instrument_model"] == "PromethION", ]
    case1 <- "tissue: tumor; tumor type: Pilocytic astrocytoma"
    case2 <- "tissue: tumor; tumor type: Glioblastoma, IDH-wildtype"
    pheno <- pheno[pheno[, "characteristics_ch1"] %in% c(case1, case2), ]
    unlink(soft_file)
    write.table(pheno, samplesheet_file, row.names = TRUE, col.names = NA, sep = "\t", quote = FALSE)
} else {
    pheno <- read.table(samplesheet_file, header = TRUE, row.names = 1, sep = "\t", quote = "", stringsAsFactors = FALSE, check.names = FALSE)
}
pheno <- as.data.frame(pheno)
pheno$tumor_type <- ifelse(
    grepl(aggressive_label, pheno$characteristics_ch1, fixed = TRUE),
    aggressive_label,
    benign_label
)
pheno$tumor_type <- factor(pheno$tumor_type, levels = c(benign_label, aggressive_label))
pheno$aggressive <- pheno$tumor_type == aggressive_label
pheno$Sample_ID <- rownames(pheno)

knitr::kable(
    as.data.frame(table(Tumor_type = pheno$tumor_type)),
    caption = "Selected GSE289246 ONT samples"
)
```
Because these ONT methylation calls come from low-depth sequencing, CpG-level evidence is
uneven across samples. A minimum coverage of 2 keeps the filter permissive enough for
long-read data while removing single-read methylation calls that are too fragile for DMR
construction.
```{r bsseq_creation}
min_coverage <- 2
input_bsseq_h5 <- getBenchmarkFile("input_h5")
if (file.exists(input_bsseq_h5)) {
    message("Loading BSseq object from ", input_bsseq_h5)
    bsseq_obj <- HDF5Array::loadHDF5SummarizedExperiment(input_bsseq_h5)
} else {
    source_bed_files <- pheno[, "supplementary_file_1"]
    names(source_bed_files) <- rownames(pheno)
    target_bed_files <- sapply(names(source_bed_files), function(x) {
        getBenchmarkFile(file.path("downloaded", paste0(x, ".bed.gz")))
    })
    # download the bed files to the raw_data directory
    for (sample_name in names(source_bed_files)) {
        if (!file.exists(target_bed_files[[sample_name]])) {
            message("Downloading ", source_bed_files[[sample_name]])
            curl::curl_download(source_bed_files[[sample_name]], target_bed_files[[sample_name]])
        }
    }

    # Extract the methylation data from the bed files and create a BSseq object
    bsseq_objects <- list()
    progress_bar <- txtProgressBar(min = 0, max = length(target_bed_files), style = 3)
    target_h5_dirs <- sapply(names(target_bed_files), function(x) {
        getBenchmarkFile(file.path("h5", paste0(x)))
    })
    for (sample_name in names(target_bed_files)) {
        target_h5_folder <- target_h5_dirs[[sample_name]]
        if (file.exists(file.path(target_h5_folder, "assays.h5"))) {
            setTxtProgressBar(progress_bar, which(names(target_bed_files) == sample_name))
            next
        }
        bed_file <- target_bed_files[[sample_name]]
        df <- data.table::fread(bed_file, header = TRUE, sep = "\t", stringsAsFactors = FALSE)
        if (!exists("gr_list")) {
            gr_list <- list()
        }
        gr <- GRanges(
            seqnames = df$chrom,
            ranges = IRanges(start = df$start + 1L, end = df$end),
            strand = "*"
        )
        M <- as.matrix(as.integer(df$percentMeth * df$coverage / 100), ncol = 1)
        Cov <- as.matrix(as.integer(df$coverage), ncol = 1)


        bsseq_obj <- BSseq(
            gr = gr,
            M = M,
            Cov = Cov,
            sampleNames = sample_name
        )
        to_keep <- Cov >= min_coverage
        if (sum(to_keep) == 0) {
            warning("No sites with coverage >= ", min_coverage, " for sample ", sample_name)
            next
        }
        if (sum(to_keep) < nrow(bsseq_obj)) {
            nfiltered <- nrow(bsseq_obj) - sum(to_keep)
            message(
                "Keeping ", sum(to_keep), " sites out of ", nrow(bsseq_obj), " (",
                round(sum(to_keep) / nrow(bsseq_obj) * 100, 2),
                "%) with coverage >= ", min_coverage, " for sample ", sample_name
            )
        }
        bsseq_obj <- bsseq_obj[to_keep, ]
        HDF5Array::saveHDF5SummarizedExperiment(bsseq_obj, dir = target_h5_folder)
        setTxtProgressBar(progress_bar, which(names(target_bed_files) == sample_name))
    }

    bsseq_objects <- list()
    for (sample_name in names(target_bed_files)) {
        target_h5_folder <- target_h5_dirs[[sample_name]]
        bsseq_obj <- HDF5Array::loadHDF5SummarizedExperiment(target_h5_folder)
        bsseq_objects[[sample_name]] <- bsseq_obj
    }
    suppressWarnings({
        bsseq_obj <- do.call(bsseq::combineList, list(bsseq_objects, BACKEND = "HDF5Array"))
    })
    bsseq_obj <- HDF5Array::saveHDF5SummarizedExperiment(bsseq_obj, dir = input_bsseq_h5)
}
```
## Find Differentially Methylated Positions

CMEnt starts from DMPs as regional anchors. Here, seeds are CpG sites with nominal evidence
for differential methylation between the two tumor groups; the distance distribution below
shows whether those anchors are dense enough to support regional expansion.
CMEnt reduces the BSseq object to windows around the seed sites before building regions.
This keeps memory use practical for ONT-scale data while preserving nearby CpGs that can
support or reject regional expansion around each seed.

```{r seeds_calculation}
seeds_file <- getBenchmarkFile("seeds.rds")
if (!file.exists(seeds_file)) {
    dmps_file <- getBenchmarkFile("dmps.rds")
    if (!file.exists(dmps_file)) {
        message("Calculating DMPs...")
        dmps <- CMEnt::findDMPsBSSeq(
            bsseq_obj,
            samplesheet = pheno,
            sample_group_col = sample_group_col,
            id_col = "Sample_ID",
            case_group = aggressive_label,
            njobs = max(1L, min(8L, parallel::detectCores(logical = TRUE) - 1L))
        )
        saveRDS(dmps, dmps_file)
    } else {
        message("Loading DMPs from ", dmps_file)
        dmps <- readRDS(dmps_file)
    }
    seeds <- dmps[dmps$pval < 0.05, ]
    saveRDS(seeds, seeds_file)
} else {
    seeds <- readRDS(seeds_file)
}
rownames(seeds) <- seeds$site_id
seeds$Row.names <- NULL
seed_dist <- unlist(lapply(split(seeds$start, seeds$chr), function(x) {
    if (length(x) < 2) return(numeric(0))
    diff(x)
}))
seeds_qs <- round(stats::quantile(seed_dist, probs = c(0, 0.25, 0.5, 0.75, 0.9, 0.99, 1)))
cat(sprintf("Number of seeds for CMEnt: %d\n", nrow(seeds)))
knitr::kable(
    seeds_qs,
    digits = 2,
    col.names = "Distance (bp)",
    caption = sprintf("Seed distance quantiles")
)
cat("BSseq object dimensions: ", dim(bsseq_obj), "\n")
reduced_bsseq_obj <- CMEnt::reduceBSseqAroundSeeds(bsseq_obj, seeds, expansion_window = 10000)
beta_handler <- CMEnt::getBetaHandler(beta = reduced_bsseq_obj, genome = genome)
cat("Reduced BSseq object dimensions: ", dim(reduced_bsseq_obj), "\n")
```

This leaves `r format(nrow(seeds), big.mark = ",")` seed CpGs across an input BSseq matrix
with `r format(nrow(bsseq_obj), big.mark = ",")` loci and `r ncol(bsseq_obj)` samples.
The median within-chromosome seed spacing is `r format(seeds_qs["50%"], big.mark = ",")`
bp, with a long tail of isolated seeds; this is why the DMR builder later requires multiple
seeds and enough covered sites before accepting a region.

## Method 1: CMEnt
```{r cment}
cment_file <- getBenchmarkFile("dmrs.rds")
if (!file.exists(cment_file)) {
    # Reduce the BSseq object around the seeds with the specified expansion window,
    # so that to minimize memory usage and speed up the DMR building process.
    cat("Reduced BSseq object dimensions: ", dim(reduced_bsseq_obj), "\n")
    start_time <- Sys.time()
    dmrs_cment <- CMEnt::buildDMRs(
        beta_handler,
        seeds,
        pheno,
        genome = genome,
        sample_group_col = sample_group_col,
        casecontrol_col = casecontrol_col,
        entanglement  = "weak",
        testing_mode  = "parametric",
        expansion_window = 10000, # The maximum distance (in bp) to expand the DMRs from the seeds
        min_seeds = 2, # The minimum number of seeds required to form a DMR
        min_sites = 50, # The minimum number of sites required to form a DMR
        max_pval = 0.05,
        ext_site_delta_beta = 0.1,
        max_lookup_dist = 1000,
        njobs = max(1L, min(8L, parallel::detectCores(logical = TRUE) - 1L)),
        verbose = 2
    )
    end_time <- Sys.time()
    time_cment <- end_time - start_time
    saveRDS(list(dmrs = dmrs_cment, time = time_cment), cment_file)
} else {
    message("Loading CMEnt DMRs from ", cment_file)
    cment_results <- readRDS(cment_file)
    dmrs_cment <- cment_results$dmrs
    time_cment <- cment_results$time
}
cat(sprintf("CMEnt: Found %d DMRs in %d seconds\n", length(dmrs_cment), round(as.numeric(time_cment, units = "secs"))))
```

The resulting call set contains `r format(length(dmrs_cment), big.mark = ",")` DMRs. In
the following summaries, positive delta beta means higher methylation in IDH-wildtype
glioblastoma relative to pilocytic astrocytoma.

## CMEnt Region Summary

```{r cment_summary}
if (length(dmrs_cment) > 0L &&
    !all(c("in_promoter_of", "in_gene_body_of", "delta_beta_promoter", "delta_beta_gene_body") %in% colnames(mcols(dmrs_cment))) &&
    requireNamespace("TxDb.Hsapiens.UCSC.hg19.knownGene", quietly = TRUE) &&
    requireNamespace("org.Hs.eg.db", quietly = TRUE)) {
    dmrs_cment <- CMEnt::annotateDMRsWithGenes(dmrs_cment, genome = genome)
}
for (gene_col in c("in_promoter_of", "in_gene_body_of")) {
    if (!gene_col %in% colnames(mcols(dmrs_cment))) {
        mcols(dmrs_cment)[[gene_col]] <- NA_character_
    }
}
for (delta_col in c("delta_beta_promoter", "delta_beta_gene_body")) {
    if (!delta_col %in% colnames(mcols(dmrs_cment))) {
        mcols(dmrs_cment)[[delta_col]] <- NA_real_
    }
}
if (!"qval" %in% colnames(mcols(dmrs_cment))) {
    mcols(dmrs_cment)$qval <- NA_real_
}

dmr_delta_col <- if ("sites_delta_beta" %in% colnames(mcols(dmrs_cment))) "sites_delta_beta" else "delta_beta"
dmr_delta <- as.numeric(mcols(dmrs_cment)[[dmr_delta_col]])
dmr_summary <- data.frame(
    Contrast = paste(aggressive_label, "minus", benign_label),
    DMRs = length(dmrs_cment),
    Median_width_bp = median(width(dmrs_cment)),
    Mean_width_bp = round(mean(width(dmrs_cment)), 1),
    Total_coverage_bp = sum(width(dmrs_cment)),
    Median_delta_beta = round(median(dmr_delta, na.rm = TRUE), 4),
    Mean_abs_delta_beta = round(mean(abs(dmr_delta), na.rm = TRUE), 4),
    Hypermethylated_DMRs = sum(dmr_delta > 0, na.rm = TRUE),
    Hypomethylated_DMRs = sum(dmr_delta < 0, na.rm = TRUE)
)

knitr::kable(dmr_summary, caption = "CMEnt DMR summary for aggressive versus benign brain tumors")
```

The called regions are mostly kilobase-scale: the median DMR spans
`r format(round(dmr_summary$Median_width_bp), big.mark = ",")` bp and together the calls
cover `r format(dmr_summary$Total_coverage_bp, big.mark = ",")` bp. The mean absolute
delta beta of `r dmr_summary$Mean_abs_delta_beta` gives a compact effect-size summary, but
individual loci should still be inspected because sparse ONT coverage can flatten aggregate
regional means.

```{r cment_effect_plot, fig.width=8, fig.height=4, fig.cap="DMR width versus methylation effect. The discrete effect-size bands are expected with sparse ONT beta estimates and should be interpreted together with coverage and sample-level plots."}
if (length(dmrs_cment) > 0L) {
    dmr_effect_df <- data.frame(
        Width = width(dmrs_cment),
        Delta_beta = dmr_delta,
        Score = if ("score" %in% colnames(mcols(dmrs_cment))) as.numeric(mcols(dmrs_cment)$score) else NA_real_
    ) |>
        filter(is.finite(Delta_beta), is.finite(Width), Width > 0) |>
        mutate(Direction = factor(ifelse(Delta_beta > 0, "Hyper in aggressive", "Hypo in aggressive")))

    ggplot(dmr_effect_df, aes(x = Delta_beta, y = Width, color = Direction, size = Score)) +
        geom_vline(xintercept = 0, linewidth = 0.2, color = "grey70") +
        geom_point(alpha = 0.65) +
        scale_y_log10(labels = comma) +
        scale_size_continuous(range = c(0.8, 3)) +
        scale_color_manual(values = c("Hyper in aggressive" = "#B8405E", "Hypo in aggressive" = "#247BA0")) +
        theme_minimal() +
        theme(legend.position = "bottom") +
        labs(
            title = "DMR Effect Size and Width",
            x = paste0("Delta beta (", aggressive_label, " - ", benign_label, ")"),
            y = "DMR width (bp, log scale)",
            color = "",
            size = "Score"
        )
}
```

## Gene Context

Gene annotation is used as a prioritization layer, not as proof of mechanism. Promoter
overlaps are useful for hypotheses about transcriptional regulation, while gene-body
overlaps can reflect broader cell-state or regulatory differences.

```{r gene_context}
gene_context_df <- data.frame(
    Context = c("Promoter", "Gene body", "Promoter and gene body", "No annotated gene overlap"),
    DMRs = c(
        sum(has_text(mcols(dmrs_cment)$in_promoter_of)),
        sum(has_text(mcols(dmrs_cment)$in_gene_body_of)),
        sum(has_text(mcols(dmrs_cment)$in_promoter_of) & has_text(mcols(dmrs_cment)$in_gene_body_of)),
        sum(!has_text(mcols(dmrs_cment)$in_promoter_of) & !has_text(mcols(dmrs_cment)$in_gene_body_of))
    )
) |>
    mutate(Percentage = round(100 * DMRs / max(1L, length(dmrs_cment)), 1))

knitr::kable(gene_context_df, caption = "Gene context of CMEnt DMRs")
```

In this run, `r gene_context_df$Percentage[gene_context_df$Context == "Gene body"]`% of DMRs
overlap gene bodies and `r gene_context_df$Percentage[gene_context_df$Context == "Promoter"]`%
overlap promoters. The sizeable unannotated fraction is expected for genome-wide methylation
data and should not be discarded automatically.

```{r gene_context_plot, fig.width=6, fig.height=3.5, fig.cap="Gene-body overlaps are the most common annotation class, followed by unannotated regions and promoter-overlapping DMRs."}
ggplot(gene_context_df, aes(x = reorder(Context, DMRs), y = DMRs, fill = Context)) +
    geom_col(width = 0.7) +
    coord_flip() +
    scale_y_continuous(labels = comma) +
    theme_minimal() +
    theme(legend.position = "none") +
    labs(title = "DMR Gene Context", x = "", y = "DMRs")
```

## Candidate CMEnt Gene-Overlapping Loci

The next tables move from global summaries to loci that are easier to review manually.
DMRs are ranked by CMEnt score, adjusted significance, and absolute methylation effect, then
split into broad gene-overlapping candidates and a focused glioma/cancer gene panel.

```{r candidate_dmrs}
dmr_table <- as.data.frame(dmrs_cment)
dmr_table$dmr_index <- seq_along(dmrs_cment)
dmr_table$DMR <- paste0(dmr_table$seqnames, ":", dmr_table$start, "-", dmr_table$end)
dmr_table$delta_beta_for_ranking <- dmr_delta
dmr_table$abs_delta_beta <- abs(dmr_delta)
dmr_table$score_for_ranking <- if ("score" %in% colnames(dmr_table)) dmr_table$score else 0
if (!"score" %in% colnames(dmr_table)) dmr_table$score <- NA_real_
if (!"pval" %in% colnames(dmr_table)) dmr_table$pval <- NA_real_
if (!"qval" %in% colnames(dmr_table)) dmr_table$qval <- NA_real_
gene_cols <- c("in_promoter_of", "in_gene_body_of")
dmr_table$genes <- vapply(seq_len(nrow(dmr_table)), function(i) {
    paste(split_ids(unlist(dmr_table[i, gene_cols], use.names = FALSE)), collapse = ",")
}, character(1))

top_dmr_index <- dmr_table |>
    arrange(desc(score_for_ranking), qval, desc(abs_delta_beta)) |>
    slice_head(n = 1) |>
    pull(dmr_index)

top_dmrs <- dmr_table |>
    filter(has_text(in_promoter_of) | has_text(in_gene_body_of)) |>
    arrange(desc(score_for_ranking), qval, desc(abs_delta_beta)) |>
    dplyr::select(
        dmr_index,
        DMR,
        width,
        score,
        qval,
        pval,
        delta_beta_for_ranking,
        any_of(c("delta_beta_promoter", "delta_beta_gene_body", "seeds_num", "sites_num", "in_promoter_of", "in_gene_body_of"))
    ) |>
    head(20)

knitr::kable(top_dmrs, digits = 4, caption = "Top 20 promoter or gene-body overlapping CMEnt DMRs ranked by score")
```

```{r top_dmr_genes_plot, fig.width=8, fig.height=4, fig.cap="Genes most often overlapped by CMEnt DMRs. Recurrent gene-body hits are mostly hypomethylated in the aggressive group, while EGFR appears among the recurrent hypermethylated candidates."}
gene_dmr_df <- if (nrow(dmr_table) > 0L) {
    do.call(rbind, lapply(seq_len(nrow(dmr_table)), function(i) {
        genes <- split_ids(dmr_table$genes[i])
        if (length(genes) == 0L) return(NULL)
        data.frame(
            Gene = genes,
            Delta_beta = dmr_table$delta_beta_for_ranking[i],
            Score = dmr_table$score_for_ranking[i],
            stringsAsFactors = FALSE
        )
    }))
} else {
    NULL
}

if (!is.null(gene_dmr_df) && nrow(gene_dmr_df) > 0L) {
    top_gene_df <- gene_dmr_df |>
        group_by(Gene) |>
        summarise(
            DMRs = n(),
            Median_delta_beta = median(Delta_beta, na.rm = TRUE),
            Max_score = max(Score, na.rm = TRUE),
            .groups = "drop"
        ) |>
        arrange(desc(DMRs), desc(Max_score)) |>
        slice_head(n = 20) |>
        mutate(Direction = ifelse(Median_delta_beta > 0, "Hyper in aggressive", "Hypo in aggressive"))

    ggplot(top_gene_df, aes(x = reorder(Gene, DMRs), y = DMRs, fill = Direction)) +
        geom_col(width = 0.7) +
        coord_flip() +
        scale_y_continuous(labels = comma) +
        scale_fill_manual(values = c("Hyper in aggressive" = "#B8405E", "Hypo in aggressive" = "#247BA0")) +
        theme_minimal() +
        theme(legend.position = "bottom") +
        labs(title = "Genes Most Frequently Overlapped by CMEnt DMRs", x = "", y = "DMRs", fill = "")
}
```

```{r cancer_gene_candidates}
glioma_genes <- c(
    "EGFR", "TERT", "CDKN2A", "CDKN2B", "PTEN", "TP53", "PDGFRA", "NF1",
    "RB1", "MGMT", "IDH1", "IDH2", "ATRX", "OLIG2", "SOX2", "SOX10",
    "NES", "GFAP", "VEGFA", "MMP2", "MMP9", "CD44", "PROM1", "EZH2",
    "BMI1", "CDK4", "MDM2", "MET", "HIF1A"
)

cancer_dmrs <- dmr_table |>
    filter(vapply(strsplit(genes, ",", fixed = TRUE), function(x) any(x %in% glioma_genes), logical(1))) |>
    arrange(desc(abs_delta_beta), qval, desc(score_for_ranking)) |>
    dplyr::select(
        DMR,
        genes,
        width,
        score,
        qval,
        delta_beta_for_ranking,
        any_of(c("delta_beta_promoter", "delta_beta_gene_body", "seeds_num", "sites_num"))
    ) |>
    head(20)

if (nrow(cancer_dmrs) > 0L) {
    knitr::kable(cancer_dmrs, digits = 4, caption = "DMRs overlapping glioma- and cancer-relevant genes")
} else {
    cat("No DMRs overlapped the focused glioma/cancer gene set.\n")
}
```

The focused cancer table is deliberately hypothesis-generating. It highlights DMRs that
intersect genes with known glioma relevance, but the ranking still comes from the methylation
data rather than from prior gene labels.

```{r cancer_gene_candidates_plot, fig.width=8, fig.height=4, fig.cap="Non-zero effect sizes among DMRs overlapping the focused glioma/cancer gene set. The full table above keeps zero-effect overlaps for transparency."}
if (nrow(cancer_dmrs) > 0L) {
    plot_cancer_dmrs <- cancer_dmrs |>
        filter(abs(delta_beta_for_ranking) > 0) |>
        mutate(
            Label = short_label(paste0(genes, " | ", DMR), 70),
            Direction = ifelse(delta_beta_for_ranking > 0, "Hyper in aggressive", "Hypo in aggressive")
        )

    if (nrow(plot_cancer_dmrs) > 0L) {
        ggplot(plot_cancer_dmrs, aes(x = reorder(Label, abs(delta_beta_for_ranking)), y = delta_beta_for_ranking, fill = Direction)) +
            geom_col(width = 0.7) +
            coord_flip() +
            scale_fill_manual(values = c("Hyper in aggressive" = "#B8405E", "Hypo in aggressive" = "#247BA0")) +
            theme_minimal() +
            theme(legend.position = "bottom", axis.text.y = element_text(size = 7)) +
            labs(
                title = "Cancer-Relevant DMR Effect Sizes",
                x = "",
                y = paste0("Delta beta (", aggressive_label, " - ", benign_label, ")"),
                fill = ""
            )
    } else {
        cat("No non-zero effect-size DMRs overlapped the focused glioma/cancer gene set.\n")
    }
}
```

## DMR-Level Methylation Separation

The heatmap averages beta values over the highest-ranked DMRs for each sample. It is a
sample-level sanity check rather than a classifier: grey cells indicate missing regional beta
estimates, which are common in low-depth ONT data. Regions with patchy coverage or inconsistent
within-group signal should be treated more cautiously than regions with clear blocks of beta
differences.

```{r top_dmr_heatmap, fig.width=9, fig.height=6, fig.cap="Top-ranked CMEnt DMR methylation by sample. Missing beta values are shown in grey, so visual separation should be judged only where coverage is present."}
heatmap_dmrs <- dmr_table |>
    arrange(desc(score_for_ranking), qval, desc(abs_delta_beta)) |>
    slice_head(n = 15) |>
    dplyr::select(dmr_index, DMR, genes)

if (nrow(heatmap_dmrs) > 0L) {
    region_beta <- lapply(seq_len(nrow(heatmap_dmrs)), function(i) {
        site_ids <- split_ids(mcols(dmrs_cment)$sites[heatmap_dmrs$dmr_index[i]])
        beta <- beta_handler$getBeta(row_names = site_ids, col_names = rownames(pheno), allow_missing = TRUE)
        colMeans(beta, na.rm = TRUE)
    })
    region_beta <- do.call(rbind, region_beta)
    rownames(region_beta) <- short_label(ifelse(
        has_text(heatmap_dmrs$genes),
        paste0(heatmap_dmrs$genes, " | ", heatmap_dmrs$DMR),
        heatmap_dmrs$DMR
    ), 55)
    sample_order <- order(pheno$tumor_type)
    region_beta <- region_beta[, rownames(pheno)[sample_order], drop = FALSE]
    heatmap_df <- as.data.frame(as.table(region_beta), stringsAsFactors = FALSE)
    colnames(heatmap_df) <- c("DMR", "Sample", "Beta")
    heatmap_df$Tumor_type <- factor(
        ifelse(pheno[heatmap_df$Sample, "tumor_type"] == benign_label, "Pilocytic", "IDH-wt GBM"),
        levels = c("Pilocytic", "IDH-wt GBM")
    )
    heatmap_df$Sample <- factor(heatmap_df$Sample, levels = colnames(region_beta))
    heatmap_df$DMR <- factor(heatmap_df$DMR, levels = rev(rownames(region_beta)))

    ggplot(heatmap_df, aes(x = Sample, y = DMR, fill = Beta)) +
        geom_tile() +
        facet_grid(. ~ Tumor_type, scales = "free_x", space = "free_x") +
        scale_fill_gradientn(colors = c("#247BA0", "white", "#B8405E"), limits = c(0, 1), na.value = "grey90") +
        theme_minimal() +
        theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), axis.text.y = element_text(size = 7)) +
        labs(title = "Top CMEnt DMR Methylation Across Samples", x = "Samples", y = "", fill = "Beta")
}
```

## Top CMEnt DMR Visualization

The top-ranked DMR is then inspected at locus level. The sample-level methylation view above
already carries the beta-value evidence, so this compact panel focuses on the DMR span, seed
placement, supporting CpGs, and gene context of the call.

```{r top_dmr, fig.width=8, fig.height=5, fig.cap="Structure-level view of the top-ranked DMR, including seed placement and gene context. Sample methylation is summarized in the heatmap above."}
if (length(dmrs_cment) > 0L) {
    top_dmr_gr <- dmrs_cment[top_dmr_index]
    top_dmr_sites <- split_ids(mcols(top_dmr_gr)$sites)
    top_dmr_seeds <- split_ids(mcols(top_dmr_gr)$seeds)
    beta_locs <- beta_handler$getBetaLocs()
    top_site_ids <- intersect(unique(c(top_dmr_sites, top_dmr_seeds)), rownames(beta_locs))
    top_locs <- as.data.frame(beta_locs[top_site_ids, ])
    top_locs$start <- as.integer(top_locs$start)
    top_locs$end <- as.integer(top_locs$end)
    top_locs$site_id <- top_site_ids
    top_locs$Role <- ifelse(top_locs$site_id %in% top_dmr_seeds, "Seed", "Supporting CpG")
    top_genes <- paste(split_ids(c(mcols(top_dmr_gr)$in_promoter_of, mcols(top_dmr_gr)$in_gene_body_of)), collapse = ", ")
    if (!nzchar(top_genes)) top_genes <- "No annotated gene overlap"

    ggplot() +
        annotate(
            "rect",
            xmin = start(top_dmr_gr),
            xmax = end(top_dmr_gr),
            ymin = 0.85,
            ymax = 1.15,
            fill = "#F6C7C7",
            color = "#B8405E",
            alpha = 0.65
        ) +
        geom_linerange(data = top_locs, aes(x = start, ymin = 0.78, ymax = 1.22, color = Role), linewidth = 0.35) +
        geom_point(data = top_locs, aes(x = start, y = 1, color = Role), size = 1.6) +
        scale_x_continuous(labels = comma) +
        scale_y_continuous(NULL, breaks = 1, labels = "Top DMR") +
        scale_color_manual(values = c("Seed" = "#B8405E", "Supporting CpG" = "#247BA0")) +
        theme_minimal() +
        theme(legend.position = "bottom", axis.text.y = element_text(size = 10)) +
        labs(
            title = paste0("Top CMEnt DMR: ", as.character(seqnames(top_dmr_gr)), ":", comma(start(top_dmr_gr)), "-", comma(end(top_dmr_gr))),
            subtitle = paste0("Score: ", round(mcols(top_dmr_gr)$score, 3), "; genes: ", top_genes),
            x = "Genomic position (bp)",
            color = ""
        )
}
```

## CMEnt DMR Manhattan Plot

The Manhattan plot places the full DMR set back onto the genome, making it easier to see
whether high-scoring calls are isolated, clustered, or concentrated in specific chromosomes.
Because the score range is narrow, the most useful signal here is genomic distribution and
annotation class rather than large score outliers.

```{r manhattan, fig.width=10, fig.height=4, fig.cap="Genome-wide DMR score distribution by primary annotation class. Scores cluster tightly, while DMRs are spread across many chromosomes."}
if (length(dmrs_cment) > 0L && "score" %in% colnames(mcols(dmrs_cment))) {
    CMEnt::plotDMRsManhattan(
        dmrs_cment,
        genome = genome
    )
}
```

## CMEnt DMR Interaction Circos Visualization

The circos view summarizes putative relationships among DMRs across the genome. It is best
read as an exploratory structure plot after candidate regions have already passed the simpler
effect-size, annotation, and heatmap checks above; grey links without JASPAR matches should
not be over-interpreted as regulatory mechanisms.

```{r circos, fig.width=10, fig.height=10, fig.cap="Circos summary of DMR positions, methylation direction, and detected DMR interactions. Grey links denote interactions without matched JASPAR motif support."}
if (length(dmrs_cment) > 0L) {
    CMEnt::plotDMRsCircos(
        dmrs_cment,
        genome = genome
    )
}
```

## Insights and Interpretation

CMEnt identifies a broad DMR landscape for the pilocytic astrocytoma versus IDH-wildtype
glioblastoma contrast, with many regions falling in gene bodies and a smaller but important
promoter-overlapping subset. The most useful candidates are not simply the highest-scoring
regions; they are the regions that combine statistical support, interpretable effect size,
gene context, and visible sample-level separation in the heatmap or locus plot.

The cancer-focused table provides a biological entry point for follow-up by flagging DMRs
near genes involved in glioma, growth-factor signaling, stemness, invasion, hypoxia, or tumor
suppression. These overlaps should be treated as prioritization evidence rather than causal
claims, especially because the ONT data are low depth and the pilocytic astrocytoma group is
small.

Overall, this report supports using CMEnt as a regional summarization layer for sparse ONT
methylation data: DMP seeds provide sensitivity, DMR construction adds local coherence, and
the downstream plots separate technical confidence checks from biological interpretation.

## Session Info

```{r}
sessionInfo()
```
