---
title: "CMEnt Analysis of In-House ONT PanNET Methylation Data"
subtitle: "Pancreatic neuroendocrine tumors versus neuroendocrine islets"
author: "CMEnt Package"
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
        code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    results = "asis",
    cache = TRUE,
    echo = TRUE,
    dev = "png",
    warning = FALSE,
    message = FALSE,
    fig.width = 8,
    fig.height = 6,
    dpi = 300
)

suppressPackageStartupMessages({
    cment_source_dir <- if (file.exists("DESCRIPTION") && dir.exists("R")) {
        "."
    } else if (file.exists(file.path("..", "DESCRIPTION")) && dir.exists(file.path("..", "R"))) {
        ".."
    } else {
        NA_character_
    }
    loaded_cment_source <- FALSE
    if (!is.na(cment_source_dir) && requireNamespace("pkgload", quietly = TRUE)) {
        loaded_cment_source <- isTRUE(tryCatch({
            pkgload::load_all(cment_source_dir, quiet = TRUE)
            TRUE
        }, error = function(e) FALSE))
    }
    if (!loaded_cment_source) {
        library(CMEnt)
    }
    library(bsseq)
    library(curl)
    library(data.table)
    library(dplyr)
    library(GenomicRanges)
    library(ggplot2)
    library(HDF5Array)
    library(scales)
    library(SummarizedExperiment)
})

getBenchmarkFile <- function(filename) {
    folder <- "ont"
    if (getwd() == "notebooks") {
        benchmark_file <- file.path("cached_results", folder, filename)
    } else if ("NAMESPACE" %in% dir()) {
        benchmark_file <- file.path("notebooks", "cached_results", folder, filename)
    } else {
        benchmark_file <- file.path("cached_results", folder, filename)
    }
    dir.create(dirname(benchmark_file), showWarnings = FALSE, recursive = TRUE)
    benchmark_file
}

split_ids <- function(x) {
    ids <- unlist(strsplit(stats::na.omit(as.character(x)), ","), use.names = FALSE)
    unique(ids[nzchar(ids)])
}

has_text <- function(x) {
    !is.na(x) & nzchar(as.character(x))
}

short_label <- function(x, width = 65L) {
    ifelse(nchar(x) > width, paste0(substr(x, 1L, width - 3L), "..."), x)
}

has_hdf5_se <- function(path) {
    dir.exists(path) && file.exists(file.path(path, "assays.h5"))
}
samplesheet_file <- getBenchmarkFile("pannet_samplesheet.csv")
bsseq_archive <- getBenchmarkFile("pannet.h5.tar.gz")
bsseq_h5_dir <- getBenchmarkFile("pannet.h5")
pheno_file <- getBenchmarkFile("pheno.tsv")
input_summary_file <- getBenchmarkFile("input_summary.tsv")
filtered_bsseq_dir <- getBenchmarkFile("filtered_bsseq.h5")
filtered_summary_file <- getBenchmarkFile("filtered_summary.tsv")
dmps_file <- getBenchmarkFile("dmps.rds")
seeds_file <- getBenchmarkFile("seeds.rds")
seeds_summary_file <- getBenchmarkFile("seeds_summary.tsv")
seeds_qs_file <- getBenchmarkFile("seeds_quantiles.tsv")
reduced_bsseq_dir <- getBenchmarkFile("reduced_bsseq.h5")
reduced_bsseq_summary_file <- getBenchmarkFile("reduced_bsseq_summary.tsv")
sparsified_seeds_summary_file <- getBenchmarkFile("sparsified_seeds_summary.tsv")
sparsified_seeds_qs_file <- getBenchmarkFile("sparsified_seeds_quantiles.tsv")
sparsified_seeds_file <- getBenchmarkFile("sparsified_seeds.rds")

aggfun <- "median"
cment_file <- getBenchmarkFile(sprintf("dmrs.CMEnt.%s.rds", aggfun))
dmr_vis_file <- getBenchmarkFile("top_dmr_visualization.png")
heatmap_file <- getBenchmarkFile("top_dmrs_heatmap.png")


seed_probs <- c(0, 0.25, 0.5, 0.75, 0.9, 0.99, 1)

need_cment <- !file.exists(cment_file)
need_heatmap <- !file.exists(heatmap_file)
need_dmr_vis <- !file.exists(dmr_vis_file)
can_build_large_inputs <- (
    has_hdf5_se(reduced_bsseq_dir) ||
        has_hdf5_se(filtered_bsseq_dir) ||
        has_hdf5_se(bsseq_h5_dir) ||
        file.exists(bsseq_archive)
)

need_beta_handler <- need_cment || ((need_heatmap || need_dmr_vis) && can_build_large_inputs)
need_reduced_bsseq_obj <- need_beta_handler
need_reduced_bsseq_file <- need_reduced_bsseq_obj && !has_hdf5_se(reduced_bsseq_dir)
need_sparsified_seeds_obj <- (
    need_cment ||
        need_reduced_bsseq_file ||
        !file.exists(sparsified_seeds_summary_file) ||
        !file.exists(sparsified_seeds_qs_file)
)
need_sparsified_seeds <- need_sparsified_seeds_obj && !file.exists(sparsified_seeds_file)
need_seeds_obj <- (
    need_sparsified_seeds ||
        !file.exists(seeds_summary_file) ||
        !file.exists(seeds_qs_file)
)
need_seeds <- need_seeds_obj && !file.exists(seeds_file)
need_filtered_bsseq_obj <- (
    need_seeds ||
        need_reduced_bsseq_file ||
        !file.exists(filtered_summary_file)
)
need_raw_bsseq_obj <- need_filtered_bsseq_obj && !has_hdf5_se(filtered_bsseq_dir)
need_bsseq <- (
    need_raw_bsseq_obj ||
        !file.exists(pheno_file) ||
        !file.exists(input_summary_file)
)
need_download <- need_bsseq && (!file.exists(samplesheet_file) || !has_hdf5_se(bsseq_h5_dir))

```

## Introduction

This report uses CMEnt to summarize long-read ONT methylation differences in an
in-house pancreatic neuroendocrine cohort. The comparison is between 15 pancreatic
neuroendocrine tumor samples (PNET) and 5 neuroendocrine islet samples (ISLT).


## Setup and Dependencies

```{r install, eval=FALSE}
r <- getOption("repos")
if (is.null(r) || is.na(r["CRAN"]) || r["CRAN"] %in% c("@CRAN@", "")) {
    options(repos = c(CRAN = "https://cloud.r-project.org"))
}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}
for (pkg in c("DSS", "TxDb.Hsapiens.UCSC.hg38.knownGene", "org.Hs.eg.db", "BSgenome.Hsapiens.UCSC.hg38")) {
    if (!requireNamespace(pkg, quietly = TRUE)) {
        BiocManager::install(pkg, update = FALSE, ask = FALSE)
    }
}
```

## Load Data

The sample sheet and BSseq HDF5 archive are downloaded from Zenodo. The archive is
already a tarred HDF5-backed BSseq object.

```{r load_data}
sample_group_col <- "Sample_Group"
casecontrol_col <- "casecontrol"
control_label <- "ISLT"
case_label <- "PNET"
genome <- "hg38"

zenodo_base_url <- "https://zenodo.org/records/20592944/files/"
samplesheet_url <- paste0(zenodo_base_url, "pannet_samplesheet.csv?download=1")
bsseq_archive_url <- paste0(zenodo_base_url, "pannet.h5.tar.gz?download=1")

if (!file.exists(samplesheet_file)) {
    curl::curl_download(samplesheet_url, samplesheet_file)
}
pheno <- data.table::fread(samplesheet_file, data.table = FALSE, check.names = FALSE)
stopifnot(all(c("Sample_ID", sample_group_col) %in% colnames(pheno)))
pheno <- pheno[pheno[[sample_group_col]] %in% c(control_label, case_label), , drop = FALSE]
pheno[[sample_group_col]] <- factor(pheno[[sample_group_col]], levels = c(control_label, case_label))
pheno[[casecontrol_col]] <- pheno[[sample_group_col]] == case_label
rownames(pheno) <- pheno$Sample_ID

if (need_download) {


    if (!dir.exists(bsseq_h5_dir) || !file.exists(file.path(bsseq_h5_dir, "assays.h5"))) {
        if (!file.exists(bsseq_archive)) {
            curl::curl_download(bsseq_archive_url, bsseq_archive)
        }
        utils::untar(bsseq_archive, exdir = dirname(bsseq_h5_dir))
    }
    h5_candidates <- list.dirs(dirname(bsseq_h5_dir), recursive = TRUE, full.names = TRUE)
    h5_candidates <- h5_candidates[file.exists(file.path(h5_candidates, "assays.h5"))]
    if (!file.exists(file.path(bsseq_h5_dir, "assays.h5")) && length(h5_candidates) > 0L) {
        bsseq_h5_dir <- h5_candidates[1]
    }
    stopifnot(dir.exists(bsseq_h5_dir), file.exists(file.path(bsseq_h5_dir, "assays.h5")))
}
if (need_bsseq) {
    bsseq_obj <- HDF5Array::loadHDF5SummarizedExperiment(bsseq_h5_dir)
    stopifnot(inherits(bsseq_obj, "BSseq"))

    common_samples <- intersect(colnames(bsseq_obj), rownames(pheno))
    if (length(common_samples) != ncol(bsseq_obj)) {
        missing_pheno <- setdiff(colnames(bsseq_obj), rownames(pheno))
        stop("Missing samples in phenotype table: ", paste(missing_pheno, collapse = ", "))
    }
    bsseq_obj <- bsseq_obj[, common_samples]
    pheno <- pheno[common_samples, , drop = FALSE]
    SummarizedExperiment::colData(bsseq_obj) <- S4Vectors::DataFrame(pheno, row.names = pheno$Sample_ID)
    write.table(pheno, pheno_file, sep = "\t", row.names = FALSE, quote = FALSE)
    bsseq_genome <- unique(as.character(GenomeInfoDb::genome(GenomeInfoDb::seqinfo(bsseq_obj))))
    bsseq_genome <- bsseq_genome[!is.na(bsseq_genome) & nzchar(bsseq_genome)]
    if (length(bsseq_genome) == 1L && bsseq_genome %in% c("hg19", "hg38", "hs1")) {
        genome <- bsseq_genome
    }

    input_summary <- data.frame(
        Metric = c("Samples", "PNET samples", "Islet samples", "CpG sites", "Genome"),
        Value = c(
            nrow(pheno),
            sum(pheno[[sample_group_col]] == case_label),
            sum(pheno[[sample_group_col]] == control_label),
            format(nrow(bsseq_obj), big.mark = ","),
            genome
        )
    )
    write.table(input_summary, input_summary_file, sep = "\t", row.names = FALSE, quote = FALSE)
} else {
    pheno <- read.table(pheno_file, sep = "\t", header = TRUE, stringsAsFactors = FALSE, check.names = FALSE, quote = "")
    input_summary <- read.table(input_summary_file, sep = "\t", header = TRUE, stringsAsFactors = FALSE, check.names = FALSE, quote = "")
}
rownames(pheno) <- pheno$Sample_ID

genome <- input_summary$Value[input_summary$Metric == "Genome"]

knitr::kable(input_summary, row.names = FALSE, caption = "PanNET ONT input summary")
```

```{r sample_overview}
overview_cols <- intersect(
    c("Sample_ID", "Sample_Group", "Diagnosis", "Biological sex", "Age", "WHO grade", "Ki67", "Tumor cell percentage"),
    colnames(pheno)
)
knitr::kable(
    pheno[, overview_cols, drop = FALSE],
    row.names = FALSE,
    caption = "Samples used for the PanNET versus islet contrast"
)
```

## Filtering
Filter the BSseq object to retain only CpG sites with sufficient coverage across samples. This step ensures that downstream analyses focus on reliable methylation measurements.
Ensure that at least half of the samples within each group report coverage for a given CpG site. This is a common practice to reduce noise and improve the robustness of differential methylation analysis.
```{r filter_bsseq}
cat(
    sprintf(
        "Original BSseq object dimensions: %s loci, %s samples\n",
        input_summary[input_summary$Metric == "CpG sites", "Value"],
        input_summary[input_summary$Metric == "Samples", "Value"]
    )
)
if (need_filtered_bsseq_obj) {
    if (!has_hdf5_se(filtered_bsseq_dir)) {
        group_counts <- table(pheno[[sample_group_col]])
        min_samples_per_group <- ceiling(group_counts / 2)
        coverage_matrix <- getCoverage(bsseq_obj, type = "Cov")
        to_keep <- rep(TRUE, nrow(coverage_matrix))
        for (group in levels(pheno[[sample_group_col]])) {
            group_samples <- pheno$Sample_ID[pheno[[sample_group_col]] == group]
            to_keep <- to_keep & rowSums(coverage_matrix[, group_samples] > 0) >= min_samples_per_group[group]
        }
        filtered_bsseq_obj <- bsseq_obj[to_keep, ]
        filtered_bsseq_obj <- HDF5Array::saveHDF5SummarizedExperiment(filtered_bsseq_obj, filtered_bsseq_dir)

        rm(bsseq_obj)
        gc(verbose = FALSE)
    } else {
        filtered_bsseq_obj <- HDF5Array::loadHDF5SummarizedExperiment(filtered_bsseq_dir)
    }
    filtered_summary <- data.frame(
        Metric = c("Filtered CpG sites"),
        Value = c(format(nrow(filtered_bsseq_obj), big.mark = ","))
    )
    write.table(filtered_summary, filtered_summary_file, sep = "\t", row.names = FALSE)
} else {
    filtered_summary <- read.table(filtered_summary_file, sep = "\t", header = TRUE, stringsAsFactors = FALSE, check.names = FALSE, quote = "")
}
cat(sprintf("Filtered BSseq remaining CpG sites: %s\n", filtered_summary$Value[filtered_summary$Metric == "Filtered CpG sites"]))


```

## Find Differentially Methylated Positions

CMEnt starts from DMPs as regional anchors. Here, seeds are CpG sites with nominal
evidence for differential methylation between PNET and islet samples. The BSseq
object is reduced to windows around those seeds before region building, keeping the
subsequent DMR construction focused and memory-conscious.

```{r seeds_calculation}
qval_thres <- 0.05

if (need_seeds_obj) {
    if (!file.exists(seeds_file)) {
        if (!file.exists(dmps_file)) {
            dmps <- CMEnt::findDMPsBSSeq(
                filtered_bsseq_obj,
                samplesheet = pheno,
                sample_group_col = sample_group_col,
                id_col = "Sample_ID",
                case_group = case_label,
                njobs = as.integer(Sys.getenv("NJOBS", unset=as.character(max(1L, min(8L, parallel::detectCores(logical = TRUE) - 1L)))))
            )
            saveRDS(dmps, dmps_file)
        } else {
            dmps <- readRDS(dmps_file)
        }
        seeds <- dmps[dmps$qval < qval_thres, , drop = FALSE]
        rownames(seeds) <- seeds$site_id


        stopifnot(nrow(seeds) > 0L)
        rm(dmps)
        gc(verbose = FALSE)
        saveRDS(seeds, seeds_file)
    } else {
        seeds <- readRDS(seeds_file)
    }
}
if (exists("seeds")) {
    if ("site_id" %in% colnames(seeds)) {
        rownames(seeds) <- seeds$site_id
    }
    seeds$Row.names <- NULL
}
if (!file.exists(seeds_summary_file)) {
    seeds_summary <- data.frame(
        Metric = c("Number of Seeds"),
        Value = c(format(nrow(seeds), big.mark = ","))
    )
    write.table(seeds_summary, seeds_summary_file, sep = "\t", row.names = FALSE, quote = FALSE)
} else {
    seeds_summary <- read.table(seeds_summary_file, sep = "\t", header = TRUE, stringsAsFactors = FALSE, check.names = FALSE, quote = "")
}
if (!file.exists(seeds_qs_file)) {
    seed_dist <- unlist(lapply(split(seeds$start, seeds$chr), function(x) {
        if (length(x) < 2L) return(numeric(0))
        diff(sort(x))
    }))
    seeds_qs <- if (length(seed_dist) > 0L) {
        round(stats::quantile(seed_dist, probs = seed_probs))
    } else {
        stats::setNames(rep(NA_real_, length(seed_probs)), paste0(seed_probs * 100, "%"))
    }
    write.table(
        data.frame(Quantile = names(seeds_qs), Distance_bp = seeds_qs),
        seeds_qs_file,
        sep = "\t",
        row.names = FALSE,
        quote = FALSE
    )
} else {
    seeds_qs <- read.table(seeds_qs_file, sep = "\t", header = TRUE, stringsAsFactors = FALSE, check.names = FALSE, quote = "")
}

cat(sprintf("Number of seeds for CMEnt: %s\n", seeds_summary$Value[seeds_summary$Metric == "Number of Seeds"]))

knitr::kable(
    seeds_qs,
    digits = 2,
    col.names = c("Quantile", "Distance (bp)"),
    caption = "Within-chromosome seed distance quantiles"
)
```
It is observed that half of the seeds are within 45 bps from each other.
We further reduce them to improve computations by sparsifying them using `CMEnt::sparsifySeeds`.
```{r sparsify_seeds}

if (need_sparsified_seeds_obj) {

    if (!file.exists(sparsified_seeds_file)) {
        sparsified_seeds <- CMEnt::sparsifySeeds(seeds, min_distance = 50)
        saveRDS(sparsified_seeds, sparsified_seeds_file)
    } else {
        sparsified_seeds <- readRDS(sparsified_seeds_file)
    }
}
if (exists("seeds")) rm(seeds)
if (!file.exists(sparsified_seeds_summary_file)) {
    sparsified_seeds_summary <- data.frame(
        Metric = c("Number of Sparsified Seeds"),
        Value = c(format(nrow(sparsified_seeds), big.mark = ","))
    )
    write.table(sparsified_seeds_summary, sparsified_seeds_summary_file, sep = "\t", row.names = FALSE, quote = FALSE)
} else {
    sparsified_seeds_summary <- read.table(
        sparsified_seeds_summary_file, sep = "\t", header = TRUE,
        stringsAsFactors = FALSE, check.names = FALSE, quote = ""
    )
}
if (!file.exists(sparsified_seeds_qs_file)) {
    sparsified_seeds_dist <- unlist(lapply(split(sparsified_seeds$start, sparsified_seeds$chr), function(x) {
        if (length(x) < 2L) return(numeric(0))
        diff(sort(x))
    }))
    sparsified_seeds_qs <- if (length(sparsified_seeds_dist) > 0L) {
        round(stats::quantile(sparsified_seeds_dist, probs = seed_probs))
    } else {
        stats::setNames(rep(NA_real_, length(seed_probs)), paste0(seed_probs * 100, "%"))
    }
    write.table(
        data.frame(Quantile = names(sparsified_seeds_qs), Distance_bp = sparsified_seeds_qs),
        sparsified_seeds_qs_file,
        sep = "\t",
        row.names = FALSE,
        quote = FALSE
    )
} else {
    sparsified_seeds_qs <- read.table(
        sparsified_seeds_qs_file, sep = "\t", header = TRUE, stringsAsFactors = FALSE,
        check.names = FALSE, quote = ""
    )
}
cat(
    sprintf(
        "Number of sparsified seeds for CMEnt: %s\n",
        sparsified_seeds_summary$Value[sparsified_seeds_summary$Metric == "Number of Sparsified Seeds"]
    )
)
knitr::kable(
    sparsified_seeds_qs,
    digits = 2,
    col.names = c("Quantile", "Distance (bp)"),
    caption = "Within-chromosome sparsified seed distance quantiles"
)
```
Then we also reduce the BSseq object to windows around those sparsified seeds before region building,
keeping the subsequent DMR construction focused and memory-conscious, using `CMEnt::reduceBSseqAroundSeeds`.
```{r reduce_bsseq}

if (need_reduced_bsseq_obj) {
    if (!has_hdf5_se(reduced_bsseq_dir)) {
        reduced_bsseq_obj <- CMEnt::reduceBSseqAroundSeeds(filtered_bsseq_obj, sparsified_seeds, expansion_window = 10000)
        reduced_bsseq_obj <- HDF5Array::saveHDF5SummarizedExperiment(reduced_bsseq_obj, reduced_bsseq_dir)
        rm(filtered_bsseq_obj)
        gc(verbose = FALSE)
    } else {
        reduced_bsseq_obj <- HDF5Array::loadHDF5SummarizedExperiment(reduced_bsseq_dir)
    }
}
if (!file.exists(reduced_bsseq_summary_file)) {
    reduced_bsseq_summary <- data.frame(
        Metric = c("Reduced BSseq loci"),
        Value = c(format(nrow(reduced_bsseq_obj), big.mark = ","))
    )
    write.table(reduced_bsseq_summary, reduced_bsseq_summary_file, sep = "\t", row.names = FALSE, quote = FALSE)
} else {
    reduced_bsseq_summary <- read.table(
        reduced_bsseq_summary_file, sep = "\t", header = TRUE, stringsAsFactors = FALSE,
        check.names = FALSE, quote = ""
    )
}

cat("BSseq object loci after reduction to seeds regions: ", reduced_bsseq_summary$Value[reduced_bsseq_summary$Metric == "Reduced BSseq loci"], "\n")
```
This leaves an input BSseq
matrix with `r format(reduced_bsseq_summary$Value[reduced_bsseq_summary$Metric == "Reduced BSseq loci"], big.mark = ",")` loci.

## Running CMEnt

```{r cment}
if (need_beta_handler) {
    beta_handler <- CMEnt::getBetaHandler(beta = reduced_bsseq_obj, genome = genome)
    rm(reduced_bsseq_obj)
}
if (!file.exists(cment_file)) {
    start_time <- Sys.time()
    dmrs_cment <- CMEnt::buildDMRs(
        beta_handler,
        sparsified_seeds,
        pheno,
        genome = genome,
        sample_group_col = sample_group_col,
        casecontrol_col = casecontrol_col,
        entanglement = "weak",
        testing_mode = "parametric",
        aggfun = aggfun,
        expansion_window = 10000,
        min_seeds = 2,
        min_sites = 50,
        max_pval = 0.05,
        ext_site_delta_beta = 0.1,
        max_lookup_dist = 1000,
        score_dmrs = TRUE,
        njobs = as.integer(Sys.getenv("NJOBS", unset=as.character(max(1L, min(8L, parallel::detectCores(logical = TRUE) - 1L))))),
        verbose = 2
    )
    time_cment <- Sys.time() - start_time
    saveRDS(list(dmrs = dmrs_cment, time = time_cment), cment_file)
} else {
    cment_results <- readRDS(cment_file)
    dmrs_cment <- cment_results$dmrs
    time_cment <- cment_results$time
}
cat(sprintf("CMEnt: Found %d DMRs in %d seconds\n", length(dmrs_cment), round(as.numeric(time_cment, units = "secs"))))
```

## CMEnt Region Summary

```{r cment_summary}
if (length(dmrs_cment) > 0L &&
    !all(c("in_promoter_of", "in_gene_body_of", "delta_beta_promoter", "delta_beta_gene_body") %in% colnames(mcols(dmrs_cment))) &&
    requireNamespace(paste0("TxDb.Hsapiens.UCSC.", genome, ".knownGene"), quietly = TRUE) &&
    requireNamespace("org.Hs.eg.db", quietly = TRUE)) {
    dmrs_cment <- CMEnt::annotateDMRsWithGenes(dmrs_cment, genome = genome)
}
for (gene_col in c("in_promoter_of", "in_gene_body_of")) {
    if (!gene_col %in% colnames(mcols(dmrs_cment))) {
        mcols(dmrs_cment)[[gene_col]] <- NA_character_
    }
}
for (delta_col in c("delta_beta_promoter", "delta_beta_gene_body")) {
    if (!delta_col %in% colnames(mcols(dmrs_cment))) {
        mcols(dmrs_cment)[[delta_col]] <- NA_real_
    }
}
if (!"qval" %in% colnames(mcols(dmrs_cment))) {
    mcols(dmrs_cment)$qval <- NA_real_
}

dmr_delta_col <- if ("sites_delta_beta" %in% colnames(mcols(dmrs_cment))) "sites_delta_beta" else "delta_beta"
dmr_delta <- as.numeric(mcols(dmrs_cment)[[dmr_delta_col]])
dmr_summary <- data.frame(
    DMRs = length(dmrs_cment),
    Median_width_bp = median(width(dmrs_cment)),
    Mean_width_bp = round(mean(width(dmrs_cment)), 1),
    Total_coverage_bp = sum(width(dmrs_cment)),
    Median_delta_beta = round(median(dmr_delta, na.rm = TRUE), 4),
    Mean_abs_delta_beta = round(mean(abs(dmr_delta), na.rm = TRUE), 4),
    Hypermethylated_DMRs = sum(dmr_delta > 0, na.rm = TRUE),
    Hypomethylated_DMRs = sum(dmr_delta < 0, na.rm = TRUE)
)

knitr::kable(dmr_summary, caption = "CMEnt DMR summary for PanNET versus neuroendocrine islets")
```

```{r cment_effect_plot, fig.width=8, fig.height=4, fig.cap="DMR width versus methylation effect for the top 1,000 CMEnt-scoring DMRs. Positive values indicate higher methylation in PanNET samples; point size shows within-plot CMEnt score percentile."}
if (length(dmrs_cment) > 0L) {
    max_effect_plot_dmrs <- 1000L
    dmr_effect_df <- data.frame(
        Width = width(dmrs_cment),
        Delta_beta = dmr_delta,
        Score = if ("score" %in% colnames(mcols(dmrs_cment))) as.numeric(mcols(dmrs_cment)$score) else NA_real_
    ) |>
        filter(is.finite(Delta_beta), is.finite(Width), Width > 0) |>
        mutate(Direction = factor(ifelse(Delta_beta > 0, "Hyper in PNET", "Hypo in PNET")))

    finite_score <- is.finite(dmr_effect_df$Score)
    if (any(finite_score)) {
        dmr_effect_df <- dmr_effect_df |>
            filter(is.finite(Score)) |>
            arrange(desc(Score)) |>
            slice_head(n = max_effect_plot_dmrs)
        finite_score <- is.finite(dmr_effect_df$Score)
    }
    dmr_effect_df$Score_percentile <- NA_real_
    if (sum(finite_score) > 1L) {
        dmr_effect_df$Score_percentile[finite_score] <- (
            rank(dmr_effect_df$Score[finite_score], ties.method = "average") - 1
        ) / (sum(finite_score) - 1)
        dmr_effect_df <- dmr_effect_df |> arrange(Score_percentile)
    } else if (sum(finite_score) == 1L) {
        dmr_effect_df$Score_percentile[finite_score] <- 1
    }
    effect_stats <- dmr_effect_df |>
        group_by(Direction) |>
        summarise(
            n = n(),
            rho = if (n() > 2L) {
                suppressWarnings(stats::cor(Delta_beta, log10(Width), method = "spearman"))
            } else {
                NA_real_
            },
            .groups = "drop"
        ) |>
        arrange(Direction) |>
        mutate(
            label = paste0(
                Direction,
                ": n=", comma(n),
                ", rho=", ifelse(is.finite(rho), sprintf("%.2f", rho), "NA")
            )
        )
    effect_stats_label <- paste(
        c("Spearman rho(delta beta, log10 width)", effect_stats$label),
        collapse = "\n"
    )

    p <- ggplot(dmr_effect_df, aes(x = Delta_beta, y = Width, color = Direction)) +
        geom_vline(xintercept = 0, linewidth = 0.2, color = "grey70") +
        stat_density_2d(
            aes(group = Direction),
            contour_var = "ndensity",
            bins = 5,
            linewidth = 0.35,
            alpha = 0.45,
            show.legend = FALSE
        ) +
        scale_y_log10(labels = comma) +
        scale_color_manual(values = c("Hyper in PNET" = "#B8405E", "Hypo in PNET" = "#247BA0")) +
        theme_minimal() +
        theme(legend.position = "bottom") +
        labs(
            title = "DMR Effect Size and Width",
            x = paste0("Delta beta (", case_label, " - ", control_label, ")"),
            y = "DMR width (bp, log scale)",
            color = ""
        ) +
        annotate(
            "label",
            x = -Inf,
            y = Inf,
            label = effect_stats_label,
            hjust = -0.02,
            vjust = 1.05,
            size = 3,
            label.r = grid::unit(0.08, "lines"),
            fill = "white",
            alpha = 0.9
        )
    if (any(finite_score)) {
        p +
            geom_point(aes(size = Score_percentile), alpha = 0.6) +
            scale_size_continuous(
                range = c(0.5, 4),
                limits = c(0, 1),
                breaks = c(0, 0.5, 1),
                labels = percent_format(accuracy = 1),
                name = "Score percentile"
            )
    } else {
        p + geom_point(alpha = 0.65, size = 1.5)
    }
}
```

```{r chromosome_arm_delta_plot, fig.width=4, fig.height=8, out.width="35%", fig.align="center", fig.cap="Chromosome-arm presence of hypermethylated and hypomethylated DMRs. Upper triangles show hypermethylated DMR counts per arm; lower triangles show hypomethylated DMR counts per arm."}
if (length(dmrs_cment) > 0L && requireNamespace("ggnewscale", quietly = TRUE)) {
    get_chromosome_arms <- function(genome) {
        if (!requireNamespace("rtracklayer", quietly = TRUE)) {
            return(NULL)
        }
        for (table_name in c("cytoBandIdeo", "cytoBand")) {
            cytoband <- suppressWarnings(tryCatch({
                session <- rtracklayer::browserSession("UCSC")
                genome(session) <- genome
                query <- rtracklayer::ucscTableQuery(
                    session,
                    track = table_name,
                    table = table_name
                )
                rtracklayer::getTable(query)
            }, error = function(e) NULL))
            if (!is.null(cytoband) && all(c("chrom", "chromStart", "chromEnd", "name") %in% colnames(cytoband))) {
                return(cytoband |>
                    transmute(
                        Chr = sub("^chr", "", chrom),
                        Arm = substr(name, 1L, 1L),
                        Start = chromStart + 1L,
                        End = chromEnd
                    ) |>
                    filter(Arm %in% c("p", "q")))
            }
        }
        NULL
    }

    arm_ranges <- get_chromosome_arms(genome[1])
    if (!is.null(arm_ranges) && nrow(arm_ranges) > 0L) {
        chr_levels <- as.character(1:22)
        dmr_position_df <- data.frame(
            Chr = sub("^chr", "", as.character(seqnames(dmrs_cment))),
            Midpoint = (start(dmrs_cment) + end(dmrs_cment)) / 2,
            Delta_beta = dmr_delta
        ) |>
            filter(Chr %in% chr_levels, is.finite(Midpoint), is.finite(Delta_beta))
        dmr_arm_df <- merge(dmr_position_df, arm_ranges, by = "Chr") |>
            filter(Midpoint >= Start, Midpoint <= End) |>
            select(Chr, Arm, Delta_beta)

        arm_summary <- expand.grid(
            Chr = chr_levels,
            Arm = c("p", "q"),
            stringsAsFactors = FALSE
        ) |>
            left_join(
                dmr_arm_df |>
                    group_by(Chr, Arm) |>
                    summarise(
                        Hyper = sum(Delta_beta > 0, na.rm = TRUE),
                        Hypo = sum(Delta_beta < 0, na.rm = TRUE),
                        .groups = "drop"
                    ),
                by = c("Chr", "Arm")
            ) |>
            mutate(
                Hyper = ifelse(is.na(Hyper), 0L, Hyper),
                Hypo = ifelse(is.na(Hypo), 0L, Hypo),
                x = match(Arm, c("p", "q")),
                y = match(Chr, rev(chr_levels)),
                Tile = paste(Chr, Arm, sep = "_")
            )

        make_arm_triangles <- function(df, value_col, triangle) {
            do.call(rbind, lapply(seq_len(nrow(df)), function(i) {
                x <- df$x[i]
                y <- df$y[i]
                coords <- if (identical(triangle, "upper")) {
                    data.frame(
                        x = c(x - 0.5, x - 0.5, x + 0.5),
                        y = c(y - 0.5, y + 0.5, y + 0.5)
                    )
                } else {
                    data.frame(
                        x = c(x - 0.5, x + 0.5, x + 0.5),
                        y = c(y - 0.5, y - 0.5, y + 0.5)
                    )
                }
                coords$Tile <- df$Tile[i]
                coords$Value <- df[[value_col]][i]
                coords
            }))
        }

        hyper_triangles <- make_arm_triangles(arm_summary, "Hyper", "upper")
        hypo_triangles <- make_arm_triangles(arm_summary, "Hypo", "lower")
        max_hyper <- max(arm_summary$Hyper, na.rm = TRUE)
        max_hypo <- max(arm_summary$Hypo, na.rm = TRUE)
        if (!is.finite(max_hyper) || max_hyper <= 0) max_hyper <- 1
        if (!is.finite(max_hypo) || max_hypo <= 0) max_hypo <- 1

        ggplot() +
            geom_polygon(
                data = hyper_triangles,
                aes(x = x, y = y, group = Tile, fill = Value),
                color = "grey85",
                linewidth = 0.2
            ) +
            scale_fill_gradient(
                low = "white",
                high = "#8B0000",
                limits = c(0, max_hyper),
                name = "Hyper",
                guide = guide_colorbar(
                    label = FALSE,
                    ticks = FALSE,
                    barwidth = 1,
                    barheight = 3,
                    position = "right",
                    direction = "vertical"
                )
            ) +
            ggnewscale::new_scale_fill() +
            geom_polygon(
                data = hypo_triangles,
                aes(x = x, y = y, group = Tile, fill = Value),
                color = "grey85",
                linewidth = 0.2
            ) +
            scale_fill_gradient(
                low = "white",
                high = "#08306B",
                limits = c(0, max_hypo),
                name = "Hypo",
                guide = guide_colorbar(
                    label = FALSE,
                    ticks = FALSE,
                    barwidth = 1,
                    barheight = 3,
                    position = "right",
                    direction = "vertical"
                )
            ) +
            scale_x_continuous(
                breaks = c(1, 2),
                labels = c("p", "q"),
                expand = expansion(add = 0.02)
            ) +
            scale_y_continuous(
                breaks = seq_along(rev(chr_levels)),
                labels = rev(chr_levels),
                expand = expansion(add = 0.02)
            ) +
            coord_fixed(clip = "off") +
            theme_minimal(base_size = 11) +
            theme(
                panel.grid = element_blank(),
                axis.ticks = element_blank(),
                legend.position = "bottom",
                legend.box = "vertical"
            ) +
            labs(
                x = "Chromosome arm",
                y = "Chromosome"
            )
    }
}
```

## Gene Context

Gene annotation is used as a prioritization layer, not as proof of mechanism. Promoter
overlaps are useful for hypotheses about transcriptional regulation, while gene-body
overlaps can reflect broader cell-state or regulatory differences.

```{r gene_context}
gene_context_df <- data.frame(
    Context = c("Promoter", "Gene body", "Promoter and gene body", "No annotated gene overlap"),
    DMRs = c(
        sum(has_text(mcols(dmrs_cment)$in_promoter_of)),
        sum(has_text(mcols(dmrs_cment)$in_gene_body_of)),
        sum(has_text(mcols(dmrs_cment)$in_promoter_of) & has_text(mcols(dmrs_cment)$in_gene_body_of)),
        sum(!has_text(mcols(dmrs_cment)$in_promoter_of) & !has_text(mcols(dmrs_cment)$in_gene_body_of))
    )
) |>
    mutate(Percentage = round(100 * DMRs / max(1L, length(dmrs_cment)), 1))

knitr::kable(gene_context_df, caption = "Gene context of CMEnt DMRs")
```

```{r gene_context_plot, fig.width=6, fig.height=3.5}
ggplot(gene_context_df, aes(x = reorder(Context, DMRs), y = DMRs, fill = Context)) +
    geom_col(width = 0.7) +
    coord_flip() +
    scale_y_continuous(labels = comma) +
    theme_minimal() +
    theme(legend.position = "none") +
    labs(title = "DMR Gene Context", x = "", y = "DMRs")
```

## Candidate CMEnt Gene-Overlapping Loci

The next table moves from global summaries to loci that are easier to review manually.
DMRs are ranked by CMEnt score, adjusted significance, and absolute methylation effect.

```{r candidate_dmrs}
dmr_table <- as.data.frame(dmrs_cment)
dmr_table$dmr_index <- seq_along(dmrs_cment)
dmr_table$DMR <- paste0(dmr_table$seqnames, ":", dmr_table$start, "-", dmr_table$end)
dmr_table$region_delta_beta <- dmr_delta
dmr_table$abs_region_delta_beta <- abs(dmr_delta)
dmr_table$score_for_ranking <- if ("score" %in% colnames(dmr_table)) dmr_table$score else 0
if (!"score" %in% colnames(dmr_table)) dmr_table$score <- NA_real_
if (!"pval" %in% colnames(dmr_table)) dmr_table$pval <- NA_real_
if (!"qval" %in% colnames(dmr_table)) dmr_table$qval <- NA_real_

top_dmr_index <- dmr_table |>
    arrange(desc(score_for_ranking), qval, desc(abs_region_delta_beta)) |>
    slice_head(n = 1) |>
    pull(dmr_index)

top_dmrs <- dmr_table |>
    filter(has_text(in_promoter_of) | has_text(in_gene_body_of)) |>
    arrange(desc(score_for_ranking), qval, desc(abs_region_delta_beta)) |>
    dplyr::select(
        dmr_index,
        DMR,
        width,
        score,
        qval,
        pval,
        region_delta_beta,
        any_of(c("delta_beta_promoter", "delta_beta_gene_body", "seeds_num", "sites_num", "in_promoter_of", "in_gene_body_of"))
    ) |>
    head(20)

knitr::kable(top_dmrs, digits = 4, caption = "Top 20 promoter or gene-body overlapping CMEnt DMRs ranked by score")
```

```{r top_dmr_genes_plot, fig.width=9, fig.height=6}
build_gene_delta_df <- function(context, gene_col, delta_col) {
    if (nrow(dmr_table) == 0L) return(NULL)
    do.call(rbind, lapply(seq_len(nrow(dmr_table)), function(i) {
        delta_beta <- as.numeric(dmr_table[[delta_col]][i])
        genes <- split_ids(dmr_table[[gene_col]][i])
        if (length(genes) == 0L || !is.finite(delta_beta)) return(NULL)
        data.frame(
            Context = context,
            Gene = genes,
            Delta_beta = delta_beta,
            Score = dmr_table$score_for_ranking[i],
            stringsAsFactors = FALSE
        )
    }))
}

gene_dmr_df <- bind_rows(
    build_gene_delta_df("Promoter", "in_promoter_of", "delta_beta_promoter"),
    build_gene_delta_df("Gene body", "in_gene_body_of", "delta_beta_gene_body")
) |>
    mutate(Context = factor(Context, levels = c("Promoter", "Gene body"))) |>
    filter(!grepl("^(LOC|LINC)", Gene))

if (nrow(gene_dmr_df) > 0L) {
    top_gene_df <- gene_dmr_df |>
        group_by(Context, Gene) |>
        summarise(
            DMRs = n(),
            median_delta_beta = median(Delta_beta, na.rm = TRUE),
            score = if (all(is.na(Score))) NA_real_ else max(Score, na.rm = TRUE),
            .groups = "drop"
        ) |>
        group_by(Context) |>
        arrange(desc(score), desc(abs(median_delta_beta)), .by_group = TRUE) |>
        slice_head(n = 20) |>
        ungroup() |>
        arrange(Context, desc(score), desc(abs(median_delta_beta))) |>
        mutate(Gene_label = paste(Context, Gene, sep = ": "))
    top_gene_df$Gene_label <- factor(top_gene_df$Gene_label, levels = rev(top_gene_df$Gene_label))

    ggplot(top_gene_df, aes(x = median_delta_beta, y = Gene_label, fill = score)) +
        geom_vline(xintercept = 0, linewidth = 0.2, color = "grey70") +
        geom_col(width = 0.7) +
        facet_wrap(~Context, scales = "free_y", nrow = 1) +
        scale_y_discrete(labels = function(x) sub("^[^:]+: ", "", x)) +
        scale_fill_continuous(type = "viridis", option = "C", na.value = "grey70") +
        guides(fill = guide_colorbar(barwidth = 8, barheight = 0.5, title = "Score", position = "bottom")) +
        theme_minimal(base_size = 12) +
        theme(
            legend.position = "bottom",
            panel.grid.major.y = element_blank()
        ) +
        labs(
            x = "Median delta beta",
            y = "",
            fill = ""
        )
}
```

## DMR-Level Methylation Separation

The heatmap averages beta values over the highest-ranked DMRs for each sample. It is a
sample-level sanity check rather than a classifier: grey cells indicate missing regional
beta estimates, which can occur in ONT data.

```{r top_dmr_heatmap, fig.width=9, fig.height=6}


if (!file.exists(heatmap_file) && length(dmrs_cment) > 0L) {
    if (!exists("beta_handler")) {
        stop(
            "Missing cached heatmap file: ", heatmap_file,
            ". Regenerate it locally with the reduced BSseq cache available before rendering without large data."
        )
    }
    png(heatmap_file, width = 9, height = 6, units = "in", res = 300)
    tryCatch({
        heatmap_dmrs <- dmr_table |>
            arrange(desc(score_for_ranking), qval, desc(abs_region_delta_beta)) |>
            slice_head(n = 15) |>
            dplyr::select(dmr_index, DMR, in_promoter_of, in_gene_body_of)
        if (nrow(heatmap_dmrs) > 0L) {
            region_beta <- lapply(seq_len(nrow(heatmap_dmrs)), function(i) {
                site_ids <- split_ids(mcols(dmrs_cment)$sites[heatmap_dmrs$dmr_index[i]])
                beta <- beta_handler$getBeta(row_names = site_ids, col_names = rownames(pheno), allow_missing = TRUE)
                colMeans(beta, na.rm = TRUE)
            })
            region_beta <- do.call(rbind, region_beta)
            genes <- vapply(seq_len(nrow(heatmap_dmrs)), function(i) {
                paste(split_ids(c(heatmap_dmrs$in_promoter_of[i], heatmap_dmrs$in_gene_body_of[i])), collapse = ",")
            }, character(1))
            rownames(region_beta) <- short_label(ifelse(has_text(genes), paste0(genes, " | ", heatmap_dmrs$DMR), heatmap_dmrs$DMR), 55)
            sample_order <- order(pheno[[sample_group_col]])
            region_beta <- region_beta[, rownames(pheno)[sample_order], drop = FALSE]
            heatmap_df <- as.data.frame(as.table(region_beta), stringsAsFactors = FALSE)
            colnames(heatmap_df) <- c("DMR", "Sample", "Beta")
            heatmap_df$Sample_Group <- pheno[heatmap_df$Sample, sample_group_col]
            heatmap_df$Sample <- factor(heatmap_df$Sample, levels = colnames(region_beta))
            heatmap_df$DMR <- factor(heatmap_df$DMR, levels = rev(rownames(region_beta)))

            print(ggplot(heatmap_df, aes(x = Sample, y = DMR, fill = Beta)) +
                geom_tile() +
                facet_grid(. ~ Sample_Group, scales = "free_x", space = "free_x") +
                scale_fill_gradientn(colors = c("#247BA0", "white", "#B8405E"), limits = c(0, 1), na.value = "grey90") +
                theme_minimal() +
                theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), axis.text.y = element_text(size = 7)) +
                labs(title = "Top CMEnt DMR Methylation Across Samples", x = "Samples", y = "", fill = "Beta"))
        }
    }, finally = dev.off())
}
if (file.exists(heatmap_file)) {
    knitr::include_graphics(heatmap_file)
}
```

## Top CMEnt DMR Visualization

```{r top_dmr, fig.width=10, fig.height=8}
if (!file.exists(dmr_vis_file) && length(dmrs_cment) > 0L) {
    if (!exists("beta_handler")) {
        stop(
            "Missing cached top-DMR visualization file: ", dmr_vis_file,
            ". Regenerate it locally with the reduced BSseq cache available before rendering without large data."
        )
    }
    top_dmr_index <- dmr_table |>
        arrange(desc(cv_accuracy), desc(seeds_num), desc(abs_region_delta_beta)) |>
        slice_head(n = 1) |>
        pull(dmr_index)
    CMEnt::plotDMR(
        dmrs_cment,
        top_dmr_index,
        beta = beta_handler,
        pheno = pheno,
        genome = genome,
        sample_group_col = sample_group_col,
        max_samples_per_group = Inf,
        output_file = dmr_vis_file,
        width = 10,
        height = 8
    )
}
# show the top DMR plot in the notebook
if (file.exists(dmr_vis_file)) {
    knitr::include_graphics(dmr_vis_file)
}
```

## CMEnt DMR Manhattan Plot

```{r manhattan, fig.width=10, fig.height=4}
if (length(dmrs_cment) > 0L && "score" %in% colnames(mcols(dmrs_cment))) {
    CMEnt::plotDMRsManhattan(dmrs_cment, genome = genome)
}
```


## Session Info

```{r}
sessionInfo()
```
