---
title: "CMEnt Conditional Null Validation on GSE244696"
subtitle: "Regional assembly of published DMPs against chromosome-stratified methylation-profile permutations"
author: "CMEnt Package"
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
        code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    warning = FALSE,
    message = FALSE,
    fig.width = 8,
    fig.height = 5,
    dpi = 300
)

required_packages <- c(
    "CMEnt", "data.table", "GenomicRanges", "ggplot2", "readxl"
)
missing_packages <- required_packages[
    !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages) > 0L) {
    stop("Install required packages before rendering: ", paste(missing_packages, collapse = ", "))
}

if (file.exists("DESCRIPTION") && requireNamespace("devtools", quietly = TRUE)) {
    devtools::load_all(quiet = TRUE)
} else if (file.exists("../DESCRIPTION") && requireNamespace("devtools", quietly = TRUE)) {
    devtools::load_all("..", quiet = TRUE)
}

suppressPackageStartupMessages({
    library(CMEnt)
    library(GenomicRanges)
    library(ggplot2)
})

get_project_file <- function(...) {
    parts <- c(...)
    if (file.exists("DESCRIPTION")) {
        do.call(file.path, as.list(parts))
    } else if (file.exists("../DESCRIPTION")) {
        do.call(file.path, as.list(c("..", parts)))
    } else {
        stop("Render this notebook from the CMEnt repository root or its notebooks directory.")
    }
}

input_dir <- get_project_file("notebooks", "cached_results", "metastudyEpic")
result_dir <- Sys.getenv(
    "CMENT_NULL_RESULT_DIR",
    get_project_file("notebooks", "cached_results", "null_model")
)
dir.create(result_dir, recursive = TRUE, showWarnings = FALSE)

genome <- "hg19"
array_type <- "EPIC"
n_permutations <- suppressWarnings(as.integer(Sys.getenv("CMENT_NULL_NPERM", "999")))
if (length(n_permutations) != 1L || is.na(n_permutations) || n_permutations < 1L) {
    stop("CMENT_NULL_NPERM must be a positive integer.")
}
master_seed <- 20260907L
njobs <- suppressWarnings(as.integer(Sys.getenv(
    "NJOBS",
    as.character(max(1L, min(8L, parallel::detectCores(logical = TRUE) - 1L)))
)))
if (length(njobs) != 1L || is.na(njobs) || njobs < 1L) {
    stop("NJOBS must be a positive integer.")
}

cment_parameters <- list(
    sample_group_col = "tissue.ch1",
    casecontrol_col = "casecontrol",
    testing_mode = "parametric",
    max_bridge_seeds_gaps = 1L,
    max_bridge_extension_gaps = 3L,
    min_seeds = 2L,
    min_sites = 3L,
    ext_site_delta_beta = 0.2,
    max_lookup_dist = 1000,
    expansion_window = -1,
    max_pval = 0.05,
    entanglement = "weak",
    aggfun = "median",
    score_dmrs = FALSE,
    annotate_with_genes = FALSE,
    extract_motifs = FALSE,
    njobs = njobs,
    verbose = 0
)
```

## Objective

This analysis tests whether CMEnt organizes an externally supplied seed set into
regional methylation structures more strongly than expected when methylation
profiles have no privileged genomic placement. The published GSE244696 DMPs are
held fixed as the seeds. No CMEnt seed-discovery helper is called.

The null transformation independently permutes intact locus-level methylation
profiles among measured loci within each chromosome. It therefore preserves the
phenotype, sample structure, locus-profile distribution, seed count, and exact
seed geometry while disrupting the alignment between genomic proximity and
methylation-profile similarity.

## Inputs and fixed seeds

```{r inputs}
beta_file <- file.path(input_dir, "GSE244696_beta_values.tsv.gz")
samplesheet_file <- file.path(input_dir, "GSE244696_samplesheet.tsv")
published_dmp_file <- file.path(input_dir, "GSE244696_DMPs.xlsx")

zenodo_base_url <- "https://zenodo.org/records/20592944/files/"
for (path in c(beta_file, samplesheet_file, published_dmp_file)) {
    if (!file.exists(path)) {
        curl::curl_download(
            paste0(zenodo_base_url, basename(path), "?download=1"),
            path
        )
    }
}

pheno <- utils::read.table(
    samplesheet_file,
    header = TRUE,
    stringsAsFactors = FALSE,
    sep = "\t",
    check.names = FALSE
)
if (all(rownames(pheno) == as.character(seq_len(nrow(pheno))))) {
    rownames(pheno) <- pheno[[1L]]
}
pheno$casecontrol <- pheno$tissue.ch1 == "Tumor"

beta_handler <- CMEnt::getBetaHandler(
    beta_file,
    array = array_type,
    genome = genome,
    njobs = njobs
)
beta_mat <- as.matrix(beta_handler$getBeta())
locs <- as.data.frame(beta_handler$getBetaLocs())

common_samples <- intersect(colnames(beta_mat), rownames(pheno))
if (length(common_samples) < 6L) {
    stop("Fewer than six beta samples match the phenotype row names.")
}
beta_mat <- beta_mat[, common_samples, drop = FALSE]
pheno <- pheno[common_samples, , drop = FALSE]

common_sites <- intersect(rownames(locs), rownames(beta_mat))
locs <- locs[common_sites, , drop = FALSE]
beta_mat <- beta_mat[common_sites, , drop = FALSE]

published_dmps_raw <- suppressWarnings(suppressMessages(readxl::read_xlsx(
    published_dmp_file,
    sheet = "Suppl_Table 1",
    range = readxl::cell_cols(1:21),
    .name_repair = "unique"
)))
published_seed_ids <- unique(as.character(published_dmps_raw[[1L]]))
published_seed_ids <- published_seed_ids[
    !is.na(published_seed_ids) & nzchar(published_seed_ids) & published_seed_ids %in% common_sites
]
if (length(published_seed_ids) < 2L) {
    stop("Fewer than two published DMPs are present in the beta matrix.")
}

# Deliberately exclude published p-values: this validation conditions only on
# the externally supplied seed identities and tests CMEnt regional assembly.
seeds <- data.frame(site_id = published_seed_ids, stringsAsFactors = FALSE)

input_summary <- data.frame(
    Metric = c("Samples", "Measured loci", "Fixed published seeds", "Null permutations"),
    Value = c(ncol(beta_mat), nrow(beta_mat), nrow(seeds), n_permutations)
)
knitr::kable(input_summary, row.names = FALSE, caption = "Null-validation inputs")
```

## Analysis functions

```{r functions}
cache_signature <- list(
    analysis_version = 1L,
    master_seed = master_seed,
    seed_ids = published_seed_ids,
    beta_md5 = unname(tools::md5sum(beta_file)),
    samplesheet_md5 = unname(tools::md5sum(samplesheet_file)),
    dmp_md5 = unname(tools::md5sum(published_dmp_file)),
    cment_parameters = cment_parameters
)

read_cache <- function(path, signature) {
    if (!file.exists(path)) {
        return(NULL)
    }
    value <- readRDS(path)
    if (!is.list(value) || !identical(value$signature, signature)) {
        return(NULL)
    }
    value$result
}

write_cache <- function(path, result, signature) {
    saveRDS(list(result = result, signature = signature), path)
}

permute_profiles_within_chromosome <- function(beta, locations, seed) {
    set.seed(seed)
    source_index <- seq_len(nrow(beta))
    chromosome_rows <- split(seq_len(nrow(locations)), as.character(locations$chr))
    for (idx in chromosome_rows) {
        if (length(idx) > 1L) {
            source_index[idx] <- sample(idx, length(idx), replace = FALSE)
        }
    }
    permuted <- beta[source_index, , drop = FALSE]
    rownames(permuted) <- rownames(beta)
    permuted
}

run_cment <- function(beta) {
    handler <- CMEnt::getBetaHandler(
        beta = beta,
        genome = genome,
        sorted_locs = locs,
        njobs = njobs
    )
    do.call(
        CMEnt::buildDMRs,
        c(
            list(
                beta = handler,
                seeds = seeds,
                seeds_id_col = "site_id",
                pheno = pheno
            ),
            cment_parameters
        )
    )
}

split_ids <- function(x) {
    ids <- unlist(strsplit(as.character(x), ",", fixed = TRUE), use.names = FALSE)
    unique(ids[!is.na(ids) & nzchar(ids)])
}

summarise_dmrs <- function(dmrs, replicate_id, is_observed) {
    if (is.null(dmrs) || length(dmrs) == 0L) {
        return(data.frame(
            replicate_id = replicate_id,
            observed = is_observed,
            incorporated_seeds = 0L,
            dmrs = 0L,
            supporting_sites = 0L,
            total_coverage_bp = 0,
            median_connection_pval = NA_real_
        ))
    }

    metadata <- S4Vectors::mcols(dmrs)
    incorporated_seeds <- if ("seeds" %in% colnames(metadata)) {
        length(split_ids(metadata$seeds))
    } else {
        0L
    }
    supporting_sites <- if ("sites" %in% colnames(metadata)) {
        length(split_ids(metadata$sites))
    } else {
        0L
    }
    connection_pvals <- if ("connection_corr_pval" %in% colnames(metadata)) {
        suppressWarnings(as.numeric(metadata$connection_corr_pval))
    } else {
        numeric(0)
    }
    connection_pvals <- connection_pvals[is.finite(connection_pvals)]

    data.frame(
        replicate_id = replicate_id,
        observed = is_observed,
        incorporated_seeds = incorporated_seeds,
        dmrs = length(dmrs),
        supporting_sites = supporting_sites,
        total_coverage_bp = sum(GenomicRanges::width(GenomicRanges::reduce(dmrs))),
        median_connection_pval = if (length(connection_pvals) == 0L) {
            NA_real_
        } else {
            stats::median(connection_pvals)
        }
    )
}
```

## Observed regional assembly

```{r observed}
observed_cache <- file.path(result_dir, "observed.rds")
observed_summary <- read_cache(observed_cache, cache_signature)
if (is.null(observed_summary)) {
    observed_dmrs <- run_cment(beta_mat)
    observed_summary <- summarise_dmrs(observed_dmrs, 0L, TRUE)
    write_cache(observed_cache, observed_summary, cache_signature)
}
knitr::kable(observed_summary, row.names = FALSE, caption = "Observed CMEnt regional assembly")
```

## Chromosome-stratified null permutations

```{r null-permutations}
null_rows <- vector("list", n_permutations)
for (permutation_id in seq_len(n_permutations)) {
    permutation_cache <- file.path(result_dir, sprintf("permutation_%04d.rds", permutation_id))
    permutation_signature <- c(cache_signature, list(permutation_id = permutation_id))
    cached_result <- read_cache(permutation_cache, permutation_signature)
    if (is.null(cached_result)) {
        null_beta <- permute_profiles_within_chromosome(
            beta_mat,
            locs,
            seed = master_seed + permutation_id
        )
        null_dmrs <- run_cment(null_beta)
        null_rows[[permutation_id]] <- summarise_dmrs(
            null_dmrs,
            permutation_id,
            FALSE
        )
        write_cache(permutation_cache, null_rows[[permutation_id]], permutation_signature)
        rm(null_beta, null_dmrs)
        invisible(gc(verbose = FALSE))
    } else {
        null_rows[[permutation_id]] <- cached_result
    }
}
null_summary <- do.call(rbind, null_rows)
all_results <- rbind(observed_summary, null_summary)
data.table::fwrite(
    all_results,
    file.path(result_dir, "null_validation_summary.tsv"),
    sep = "\t"
)
```

## Global conditional-null test

The primary statistic is the number of unique supplied seeds incorporated into
the final multi-seed DMRs. The Monte Carlo p-value includes the observed result
in the numerator and denominator.

```{r inference}
observed_statistic <- observed_summary$incorporated_seeds[[1L]]
null_statistics <- null_summary$incorporated_seeds
exceedances <- sum(null_statistics >= observed_statistic)
empirical_p <- (exceedances + 1) / (n_permutations + 1)
exceedance_interval <- stats::binom.test(exceedances, n_permutations)$conf.int

inference_summary <- data.frame(
    Statistic = c(
        "Observed incorporated seeds",
        "Null median incorporated seeds",
        "Null 95th percentile",
        "Null exceedances",
        "Monte Carlo p-value",
        "Exceedance probability 95% CI lower",
        "Exceedance probability 95% CI upper"
    ),
    Value = c(
        observed_statistic,
        stats::median(null_statistics),
        unname(stats::quantile(null_statistics, 0.95)),
        exceedances,
        empirical_p,
        exceedance_interval[[1L]],
        exceedance_interval[[2L]]
    )
)
knitr::kable(inference_summary, digits = 4, row.names = FALSE, caption = "Primary conditional-null test")

cat(sprintf(
    paste0(
        "Across %d chromosome-stratified methylation-profile permutations, ",
        "CMEnt incorporated %s published seeds into multi-seed DMRs in the observed data, ",
        "compared with a null median of %s (95th percentile %s; Monte Carlo p = %.4g)."
    ),
    n_permutations,
    format(observed_statistic, big.mark = ","),
    format(stats::median(null_statistics), big.mark = ","),
    format(unname(stats::quantile(null_statistics, 0.95)), big.mark = ","),
    empirical_p
))
```

```{r primary-plot}
ggplot(null_summary, aes(x = incorporated_seeds)) +
    geom_histogram(bins = 40, fill = "#0072B2", color = "white") +
    geom_vline(xintercept = observed_statistic, color = "#D55E00", linewidth = 1) +
    theme_minimal(base_size = 11) +
    labs(
        title = "CMEnt regional assembly under the conditional null",
        subtitle = sprintf("Observed value shown in orange; %d chromosome-stratified permutations", n_permutations),
        x = "Unique supplied seeds incorporated into DMRs",
        y = "Null permutations"
    )
```

Secondary summaries describe the resulting regional structures but are not
treated as additional confirmatory tests.

```{r secondary}
secondary_summary <- data.frame(
    Measure = c("DMRs", "Supporting sites", "Total coverage (bp)"),
    Observed = c(
        observed_summary$dmrs,
        observed_summary$supporting_sites,
        observed_summary$total_coverage_bp
    ),
    Null_median = c(
        stats::median(null_summary$dmrs),
        stats::median(null_summary$supporting_sites),
        stats::median(null_summary$total_coverage_bp)
    ),
    Null_95th_percentile = c(
        stats::quantile(null_summary$dmrs, 0.95),
        stats::quantile(null_summary$supporting_sites, 0.95),
        stats::quantile(null_summary$total_coverage_bp, 0.95)
    )
)
knitr::kable(secondary_summary, digits = 1, row.names = FALSE, caption = "Descriptive regional-assembly summaries")
```

## Interpretation

This is a conditional global test of CMEnt regional assembly given the published
seed coordinates. It does not evaluate how the seeds were selected, does not use
CMEnt's optional seed-discovery helpers, and does not claim per-DMR false-discovery
rate control. A small empirical p-value indicates that the fixed published seeds
are incorporated into coherent multi-seed regions more extensively than expected
when intact methylation profiles are randomly reassigned among loci on the same
chromosome.

```{r session-info}
sessionInfo()
```
